### 2015-10-20 p00-ch00-intro/intro.tex
Initial Commit

% -*- root: ../thesis.tex -*-

\chapter[Introduction]{Introduction}
\label{ch:intro}

\section{From actor \emph{modelling} to actor \emph{programming}}
\label{sec:intro:abs}

Object-oriented programming~\cite{booch1982object,meyer1988object} has been of one the dominant paradigms for software engineering in the past three decades.
One object interacts with another object using the notion of a method.
Method invocations are \emph{blocking}; i.e. the caller object waits until it receives the result of the method call from the callee object.
Surprisingly, such model of interaction was \emph{never} the intention of the creator of the paradigm.
Originally, object interactions were meant to be \emph{messaging} among objects and objects behaved as autonomous entities possibly on remote locations on a network; Alan Kay clarified later~\cite{alank1,alank2}. 
On the contrary, almost all the object-oriented languages at hand have followed the blocking and synchronous model of messaging based on a common misunderstanding. 
The original definition of object-oriented is close to another model of computation: actor model.

One of the fundamental elements of actor model~\cite{actors:agha,agha97} is \emph{asynchronous} message passing.
In this approach, interactions between objects are modelled as \emph{non-blocking} messages.
One object, the sender, communicates a message to the other object, the receiver.
At the receiver side, a message eventually leads to a method invocation (in an object-oriented paradigm) or a functional call (in a functional paradigm).
In actor model, the locality of the objects is transparent from the messages.
A system is composed of objects with communicating messages.

Considerable amount of research have been carried out to mix object-oriented programming languages with actor model. 
Language extensions, libraries, and even new programming languages are the naturual outcomes of such research.
The more mainstream the language, the more research and development are available in the intersection of object orientation and actor model.
For example, \cite{actor_frameworks_jvm:agha} presents a comparative analysis of such research on Java and JVM languages.

With the rise of computation power, a new aspect added to create a triangle: object-oriented, actor model, and \emph{concurrency}.
Concurrency makes it harder to verify programs in terms of correctness of runtime behavior. 
Scheduling of object messages in threads of executions is one of main challenges.
In addition, in the presence of scheduling and thread interleaving, it is important that objects are able to protect their state from concurrent modifications for correctness properties.
In this research, we take advantage of a modelling language that aims to mix the three aspects in addition to the capabilities of formal semantics and verification.

ABS~\cite{johnsen2012abs,hahnlehjlssw11} is a modelling language for concurrent objects and distributed systems.
ABS provides the semantics of actor model~\cite{actors:agha} as a modelling language.
ABS language provides a set of core features including algebraic data types (ADT), functional layer, and concurrency layer.
ABS concurrency layer lays its basis on co-operative scheduling of active objects~\cite{creol:broch:owe}.
ABS semantics is completely expressed in structural operational semantics~\cite{plotkin:sos}.
This allows ABS models to take advantage of various verification methods and static analysis techniques.
% 
Co-operative scheduling in ABS has been additionally extended for real-time scheduling considering priorities and time constraints~\cite{bjork2013:rtabs,johnsen2012modeling}.
Thus, in general, ABS concurrency layer is a perfect fit if it can be integrated with industry-level real-world programming languages.
% With distributed computing as one of the recent de-facto requirements in every business use case, an enormous effort has gone to improve the current mainstream programming languages to fit for the new requirements of distributed and concurrent sett1ings.

Considering the mainstream programming languages
\footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}}, Java~\cite{gosling2000java} is one of the most commonly-used.
Additionally, since Java 1.5 and the rise of concurrency API in Java~\cite{jsr166}, a remarkable effort by the community focus on how concurrency and distribution can be improved in Java as in provided by ABS semantics and actor model.
However, not all the development as Java libraries and frameworks form their basis on formal semantics. 
Therefore, this gives rise to issues and challenges in terms of correctness, semantic preservation, and reasoning.

We target to deliver ABS concurrency layer in form of a Java library API.
However, Java language specification (JLS) clearly concludes that Java language is not comptabile and ready to be extended as a functional language with the support of algebraic data types.
This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
Scala language~\cite{odersky2004scala} is a result of the research that is a dynamic, functional, and object-oriented language on JVM.
Although such extensions over Java towards a functional language exist, they lack formal semantics in the context of concurrency and actor model.

We primarily focus on ABS concurrency layer as a Java library API.
We argue that the advantage of a correct semantic translation from ABS concurrency layer to its Java equivalent constructs overcomes the additional support of functional layer but with \emph{partial} formal semantics.
Our approach enables programmers to model their Java systems in terms of ABS concurrency semantics and then create a correct mapping and implementation from ABS to Java using an API.
In addition, we aim to facilitate ways to actually program in ABS models as if the programmer is able to take advantage of standard Java libraries.

In the next section, we sketch a high-level overview of the architecture design goals and how this research plans its objectives.

\section{Architecture}
\label{sec:intro:arch:crit}

\paragraph{Polyglot Programming}
With the rise of distributed computing challenges, software engineering practice has turned to methods that combine multiple programming languages and models to complete a task. 
In this approach, different languages with different focus and abstractions contribute to the same problem statement in different layers. 
Polyglot programming essentially enables software practice to apply the \emph{right} language in the appropriate layer of the problem statement. 
In this research, we follow the question in which we try to deliver ABS semantics and features in a polyglot approach. 
The programmer develops models with ABS that \emph{partically} take advantage of the target language features (e.g. from Java).
This approach is also referred to as \emph{Foreign Function Interface} in the context of ABS modelling.
Listing~\ref{lst:abs:java} shows an ABS code that uses \jtt{java.util.ArrayList} as a data strucuture.
With Polyglot programming in ABS, the programmer does \emph{not} need to define the abstractions necessary for the list construct.

\begin{lstlisting}[float=h,language=Java,caption=Using Java in ABS,label=lst:abs:java]
List<String> params = new ArrayList<>();
myObj ! doSomething(params);   
\end{lstlisting}

\paragraph{Scalability} 
The distribution level that asynchronous message passing mechinism deliver
challenges performance efficiency.
In such a context, the number of messsages delivered among actors in the 
environment is not predictable at runtime.
The goal is to ensure the actor system scales in performance with least 
influence from the number of asynchronous messages delivered in the system.
We direct benchmarks on the system to present the performance of the system
using different number of messages delivered in the system.

\paragraph{Modularity} 
The scope of the research spans to a number of layers revolving around ABS
 language:
\begin{itemize}
\item \emph{Compiling ABS to a target programming language}
One first objective it compile an ABS model to a target programmig language.
The compilation can be at \emph{source-to-source} or \emph{source-to-bytecode}
level.
Target languages potentially include mainstream programming languages such as
Java, Erlang, Haskell, and Scala.
Thus, the ABS programmer should be able to use an ABS compiler to compile
their ABS models to the target languages.
\item \emph{Using ABS semantics as an API in a programming language}
ABS language is precisely expressed with structural operational semantics~\cite{johnsen2012abs} in addition to its syntax definition.
Therefore, the semantics of ABS can be delivered in a programming language
as Application Programming Interface (API) as long as the programming language provides sufficient constructs to respect ABS language semantics.
In addition, such ABS programming API should be \emph{verifiable}.
If such ABS mapping to a programming API is provided, a programmer is able to
take advantage of ABS semantics without directory programming in ABS.
Such capability from ABS enables industry users of mainstraim languages to
model their systems in ABS semantics using their programming languages and
platforms.
\item \emph{Modelling in ABS}
ABS language provides solid structure and semantics to model concurrent
and distributed systems.
The user (that can be a programmer, an analysist, or a researcher) rightfully
demands to have access to a tool-set and IDE that allows working with ABS 
models in a user-friendly way.
The ABS IDE and tool-set developers should be able to easily reuse and compose
over existing modules and components.
\end{itemize}

Based on the scopes and layers of usage of this research, we aim to provide 
a highly modular architecture in design and implementation.
The goal is verifiable based on produced artifacts of this research in terms
of their \emph{usability}, \emph{re-usability}, and \emph{composability} in
different contexts.

\paragraph{Extensibility}
Delivering ABS language features as an API brings about a new requirement to be able to configure and extende the API. 
It is a common practice in software engineering that components are developed with a mixture of \emph{configuration}, \emph{extension} and \emph{composition}.
Therefore, the ABS programmer should be able to extend such API to custom requirements. 
The extensibility is provided either through configuration of the API or API extension.

Figure~\ref{fig:arch} presents the general architecture of the implementation.

\begin{figure}[h]
\centering
\includegraphics[scale=0.3]{../figs/Arch.eps}
\caption[General Architecture]{General Architecture of ABS API and Java Language Backend}
\label{fig:arch}
\end{figure}

\section{Literature Overview}
\label{sec:intro:rel}

\subsection{Programming Languages Overview}
\label{sec:intro:proglangs}

In this section, we briefly provide an overview of the programming languages that have targetted the similar problem statements.
It is important to note the timeline and the rapid evolution of programming languages trying to provide an actor-based model of asynchronous message passing.
In addition, we identify different types of research and work related to actor model and concurrency programming:
\begin{description}
\item[First-Class Citizen]
is the type of programming languages that deliver actor model programming support as a first-class citizen of the language.
In such languages, actor model is by-design part of the syntax and semantics of the language.
\item[Implicit By Design]
refers to the type that there is no explicit notion of actor models in the language syntax or semantics.
However, the programming language provides fundamental constructs for concurrenty and asynchronous message passing.
Thus, it becomes a trivial task in such type of programming languages to create an abstraction to support the actor model by coding.
\item[External Library]
refers to the type of programming languages for which actor model support is possible through using an external library developed for the language.
\end{description}

Table~\ref{tbl:actor:pl} presents a summary.

\begin{table}[h]
\centering
\begin{tabular}{lll}
\textsfb{Language} & \textsfb{Abstraction} & \textsfb{Type} 
\\ \toprule
Erlang\cite{erlang:armstrong,erlang:actor} & Process & Implicit By Design 
\\ \midrule
Elixir\cite{elixir,elixir:actor} & Agent & Implicit By Design 
\\ \midrule
Haskell\cite{con_haskell:wiki} & forkIO \& MVars & Implicit By Design 
\\ \midrule
Go\cite{go:actor} & Goroutine & Implicit By Design 
\\ \midrule
Rust\cite{rust:2014,rust:actor} & Send \& Sync & Implicit By Design 
\\ \midrule
Scala\cite{haller09tcs} & Akka Actors
\footnote{Scala 2.11.0 adopts Akka as default actor model implementation: \url{http://docs.scala-lang.org/overviews/core/actors-migration-guide.html}}
& External Library 
\\ \bottomrule
\end{tabular}
\caption{Actor Model Support in Programming Languages}
\label{tbl:actor:pl}
\end{table}

\subsection{Frameworks and Libraries}
\label{sec:intro:libs}

In the previous section, we presented an overview of the programming languages and their categories that aim to provide actor model~\cite{agha86book}.
Since many programming languages faced different challenges to provide the necessary syntax and semantics for this at the level of the language, many libraries and frameworks have been born with a motivation to fill this gap.
We observe that the more the language itself is close to the actor model semantics, the less external libraries and frameworks target this gap. 
In the following, we briefly provide a few frameworks
\footnote{A more comprehensive list can be obtained at \cite{KarmaniSA09} and  \url{https://en.wikipedia.org/wiki/Actor_model\#Programming_with_Actors}}
and libraries for Java as one of the main focus areas of this research is to study this gap for a mainstream language such as Java.

\begin{table}[h]
\centering
\begin{tabular}{lll}
\textsfb{Library} & \textsfb{Technique} & \textsfb{JVM Language} 
\\ \toprule
Killim\cite{srinivasan2008kilim,kilim} & Byte-Code Modification & Java 
\\ \midrule
Quasar\cite{quasar} & Byte-Code Modification, Java~8 & Clojure, Java 
\\ \midrule
Akka\cite{akka,scala:actors:ordersky} & Scala Byte-Code on JVM & Scala, Java 
\\ \bottomrule
\end{tabular}
\caption{Actor programming libraries in Java}
\label{tbl:actor:libs}
\end{table}

One of the main techniques used in libraries to deliver actor programming in Java is byte-code engineering~\cite{dahm1999byte,bruneton2002asm,asm}.
Byte-code engineering modifies the generated byte-code for compiled classes in Java either during compilation or at runtime.
Although, this technique is commonly used and argued to provide better performance optimization~\cite{vallee1999soot}, it introduces deep challenges regarding the verification of the running byte-code~\cite{leroy2001java,leroy2003java}.
Thus, one of the main objectives in this research to provide a \emph{statically} verifiable technique to provide actor programming in Java.

\section{Contributions}
\label{sec:intro:contribs}

Based on what we discuss in objectives and related work, the contributions of this thesis can be summarized as the following table:

\begin{table}[h]
\centering
\begin{tabular}{p{7cm}p{3cm}}
\textsfb{Topic} & \textsfb{Chapter/Section}
\\ \toprule
{Formalization of the mapping from ABS to Java including the operational semantics and ABS co-operative scheduling in Java} & Chapter~\ref{ch:p01:ch01} and \ref{ch:p01:ch02}
\\ \midrule
Design and implementation of ABS concurrency layer in Java & Chapter~\ref{ch:p02:ch01} and \ref{ch:p02:ch02}
\\ \midrule 
Design and implementation of a massive-scale monitoring system based on ABS API in Java & Chapter~\ref{ch:p03:ch01}
\\ \bottomrule 
\end{tabular}
\caption{Thesis Contributions Summary}
\label{tbl:thesis}
\end{table}

### 2015-10-20 p01-ch01-2012/compilation.tex
Initial Commit

******************************************************************

In this section, we briefly discuss how different constructs in Creol and Java
co-relate to one another in the compilation process.
Figure \ref{fig:creol_object} depicts a general concurrent object in Creol that is transformed 
to an equivalent Java object with a single thread of execution on a processor.

\subsubsection{Active Objects to Task Objects}
In the transformation process, one of the most important questions is that: how
Creol's concurrency model maps to concurrent Java?

Based on Creol's notion of active objects, each object conceptually possesses
one processing core in Creol. It means that each object is a thread of execution
and sends messages to other objects. Each Creol object is mapped to one Java
object and one concurrent task. One major advantage of this approach is that
since all the object's belongings stay on the same core, the problems of
synchronization and shared state issues will not occur.

Every object maintains a storage of the messages that is used to receive and
store messages sent from other objects (mailbox). At some time $t_i$, a message
is received in the object's mailbox. However, $t_i$ is not necessarily the time
that the message is processed as the same as what happens in Creol. As some
other $t_e$, the objects looks in the mailbox to select a message to process (it
can be the case that $t_i = t_e$).

In the ``Active Segment'' in Figure \ref{fig:creol_object}, when a method
invocation is received in the object, it carries information on priority
specification possibly from both the client and the server. This information is
used by a ``priority manager'' component in the object to assign the message a
final priority value based on which the message is stored in the mailbox of the object.
On the other hand, each object consists of a ``scheduler'' responsible to select
one message at a time for execution. The scheduler component may use predefined
policies specified by the user programmer how to select one message at a time.
It may also be developed and injected by the user programmer to provide further
extensibility and customization. When a message is selected, the object executes
the requested method.

The Java object has access to an instance of
{\footnotesize\texttt{ExecutorService}} to encapsulate a single thread of
execution on a processor. Each method invocation is transformed to a variation
of  {\footnotesize\texttt{Callable}} and stored in the object's mailbox. The
method invocations return instances of {\footnotesize\texttt{Future}} to hold
the eventual result of the method.

\subsubsection{Asynchronous Method Calls}
With an asynchronous method call in Creol, the object continues with the
next instruction after the call and the response from the invoked method will be
stored in a labeled variable that can be later checked or fetched. Creol can
also model a synchronous method with an asynchronous method call.

Similarly, concurrent Java provides an implementation of
{\footnotesize\texttt{Future}}\footnote{\texttt{java.util.concurrent.Future}}
computation. It encapsulates a value that will be determined sometime later in
the time of execution or cancelled during execution. Having a future variable in
Java, a Creol asynchronous method call can be mapped to an instance of
{\footnotesize\texttt{FutureTask}}\footnote{\texttt{java.util.concurrent.FutureTask}}.

When a method invocation request is received in the object, an instance of
{\footnotesize\texttt{FutureTask}} is created to encapsulate the original method
invocation. As the same time, an instance of {\footnotesize\texttt{Future}} is
returned as the ``future'' result of this method invocation. The actual result
will be eventually available when the message is selected for execution. On the
other hand, the incoming message carries priority scheduling specification. So,
the priority manager of the object uses such information to map different levels
of priorities possibly specified for the method invocation to a final priority
value. This value is assigned to the message being stored in the mailbox and is
used by the scheduler. When the scheduler of the object selects one method
invocation from the mailbox, it uses the mapped final priority values.

\subsubsection{Creol Processor Release Points}
One of the most important primitives in Creol language is the ability to
\textit{manually} release a processor. Creol allows the programmer to decide
through syntax whether to yield the processor or continue the process. On the
other hand, using concurrent Java provides the same functionality using
{\footnotesize\texttt{Condition}}s\footnote{\texttt{java.util.concurrent.locks.Condition}}.
When a program in Java takes advantage of a condition object, it is actually
controlling the enclosing executing thread. The condition mechanism is in
analogy with the older Java's Object {\footnotesize\texttt{wait}} and
{\footnotesize\texttt{notify}} mechanism. The transformation process utilizes
this feature in Java to provide the explicit {\footnotesize\texttt{await}}
syntax in Creol.

### 2015-10-20 p01-ch01-2012/langs.tex
Initial Commit

\documentclass{article}
\usepackage[top=0.01in,bottom=0.01in,left=0.01in,right=0.01in,paperwidth=5.7in,paperheight=2.3in]{geometry}

\usepackage{alltt}
\usepackage{array}
\usepackage{amsmath}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{pifont}
\usepackage{url}
\usepackage[table]{xcolor}

\newcommand{\superscript}[1]{\ensuremath{^{\textrm{#1}}}}
\newcommand{\subscript}[1]{\ensuremath{_{\textrm{#1}}}}
\newcommand{\thickhline}{\noalign{\hrule height 1.0pt}}

\begin{document} 
\pagestyle{empty}
\definecolor{lgray}{rgb}{0.85,0.85,0.85}
\rowcolors{2}{lgray}{white}
\begin{center} 
\begin{tabular}{m{0.5in} m{1.2in} m{1in} m{1in} m{1in}}
  & Java & C++ & Python & C\# \\
\thickhline 
Actors &  Scala, Erlang, SALSA, Axum, Ptolemy, Kilim, ActorFoundry, Jetlang,
JavAct, AJ & Theron, Act++ & Stackless, V3.2+ & Retlang \\

STM & Multiverse, Clojure, DSTM2 (Fortress), Deuce, JVSTM & TL2 (Sun), Intel
C++ STM & Durus & SXM \\

Data Flow & Java 7+ JSR 166y Fork/Join, Hadoop & OpenMP, TBoost.STM, Hadoop Streaming &
Disco (Nokia) & Dryad, Hadoop Streaming \\
\end{tabular}
\end{center}
\end{document}

### 2015-10-20 p01-ch01-2012/paper.tex
Initial Commit

% -*- root: ../thesis.tex -*-

\chapter[Application-Level Scheduling]{Programming and Deployment of Active Objects with Application-Level Scheduling%
\footnote{This work is partially supported by the EU FP7-231620 project: HATS.}
}
% 
\label{ch:p01:ch01}
% 
\chapterauthor{Behrooz Nobakht, Frank S. de Boer, Mohammad Mahdi Jaghoori,  Rudolph Schlatte}

% --- Author Metadata here ---
% \conferenceinfo{SAC'12}{March 25-29, 2012, Riva del Garda, Italy.}
% \CopyrightYear{2011}
% \crdata{978-1-4503-0857-1/12/03}
% --- End of Author Metadata ---

% \title{Programming and Deployment of Active Objects with Application-Level Scheduling%
% \titlenote{This work is partially supported by the EU FP7-231620 project: HATS.}
% }

% \numberofauthors{2}  %set to two to enable longer lines for names and

% \author{
% \alignauthor Behrooz Nobakht\\
% 	\affaddr{Leiden Institute of Advanced Computing}\\
% 	\affaddr{Universiteit Leiden, Leiden, The Netherlands}\\
% 	\email{bnobakht@liacs.nl}
% \alignauthor Frank S. de Boer\\
% 	\affaddr{Centrum Wiskunde and Informatica}\\
% 	\affaddr{Amsterdam, The Netherlands}\\
% 	\email{frb@cwi.nl}
% \and
% \alignauthor Mohammad Mahdi Jaghoori\\
% 	\affaddr{Centrum Wiskunde and Informatica}\\
% 	\affaddr{Amsterdam, The Netherlands}\\
% 	\email{jaghoori@cwi.nl}
% \alignauthor Rudolf Schlatte\\
% 	\affaddr{University of Oslo}\\
% 	\affaddr{Oslo, Norway}\\
% 	\email{rudi@ifi.uio.no}
% }

\section*{Abstract}
% Active objects provide a natural means for programming distributed systems.
We extend and implement a modeling language based on concurrent active objects with application-level
scheduling policies.
The language allows a programmer to assign priorities at the application level, for example,
to method definitions and method invocations, and assign corresponding  policies to the individual active objects
for scheduling the messages.
Thus, we   leverage scheduling and performance related issues,
which are becoming increasingly important in multi-core and cloud  applications, from the underlying operating system to the application level.
We describe a tool-set to transform models of active objects extended with application-level
scheduling policies  into Java.
This tool-set allows a direct  use of Java class libraries; thus,  we
obtain  a full-fledged programming language
based on active objects which allows for high-level control of deployment related issues.

% \category{D.1.3}{Programming Techniques}{Concurrent Programming}
% \category{D.2.3}{Software Engineering}{Coding Tools and Techniques}
% \category{D.3.0}{Programming Languages}{General}
% \terms{Languages, Design}
% \keywords{Application-level scheduling, ~Priority scheduling, ~Creol, Java, Actor model, Concurrent active objects}

\section{Introduction} \label{sec:introduction}
One of the major challenges in the design of programming
languages is to provide high-level support for  multi-core and cloud  applications which are becoming increasingly important.
Both multi-core and cloud  applications require an explicit and precise treatment of non-functional properties, e.g., resource requirements.
On the cloud, services execute in the
context of virtual resources, and the amount of resources actually available to a service is subject to change.
Multi-core applications require techniques 
to help the programmer optimally use potentially many cores. At
the operating system level, resource management is greatly affected by scheduling which
is largely beyond the control of most existing high-level programming languages.
Therefore, for optimal use of the available resources, we cannot avoid leveraging
scheduling and performance related issues from the underlying operating system
to the application level. However, the very nature of high-level languages is to
provide suitable abstractions that hide
implementation details from the programmer. The main challenge in designing programming languages
for  multi-core and cloud  applications
is to find a balance between these two conflicting requirements.

We investigate in this paper how concurrent active objects in a high-level object-oriented
language can be used for high-level scheduling of resources. We use the notion of
concurrent objects in Creol \cite{creol:broch_owe,mpd:andrews}. A
concurrent object in Creol has control over one processor; i.e. it has a
single thread of execution that is controlled by the object itself.
Creol processes never leave the enclosing object; method invocations
result in a new process inside the target object.  Thus, a concurrent
object provides a natural basis for a deployment scheme where each
object virtually possesses one processor. Creol further provides high-level
mechanisms for synchronization of the method invocations in an object;
however, the scheduling of the method invocations are left
unspecified. Therefore, for the deployment of concurrent objects in
Creol, we must, in the very first place, resolve the
basic scheduling issue; i.e. which \textit{method} in which
\textit{object} to select for execution.
We show how to introduce priority-based scheduling of the messages of the individual objects
at the application-level itself.

In this paper we also propose a 
% pluggable and extensible 
tool
architecture to deploy Creol applications. To prototype the tool
architecture, we choose Java as it provides low-level concurrency features, i.e., threads, futures, etc.,
required for multi-core deployment of object-oriented applications. The tool
architecture prototype transforms Creol's constructs for concurrency to their
equivalent Java constructs available in the
{\jtt{java.util.\-concurrent}} package. As such, Creol provides
a high-level structured programming discipline based on active objects on top of Java.
Every active object in Creol is transformed to an object in Java that uses a priority manager
and scheduler to respond to the incoming messages from other objects. Besides,
through this transformation, we allow the programmer to seamlessly use, in the
original Creol program, Java's standard library including all the data types.
Thus, our approach converts Creol from  a modeling language to a full-fledged
 ``programming'' language.

Section \ref{sec:creol} first provides an overview of the Creol language with application-level scheduling. 
In Section \ref{sec:compiler}, we elaborate on the design  of the tool-set and the
prototype. The use of the tool-set is exemplified by a case study in Section \ref{sec:caseStudy}.
Section \ref{sec:relwork} summarizes the related work. 
Finally, we conclude in Section \ref{sec:conclusion}.

%\section{Preliminaries: Creol} \label{sec:creol}
%\input{Creol}

\lstset{language=Creol,escapeinside={(*}{*)}}

\section{Application-Level Scheduling}
\label{sec:creol}

Creol \cite{creol:broch_owe} is a full-fledged object-oriented language with formal semantics for modeling and analysis of systems of concurrent objects. 
Some Creol features include interface and class inheritance and being strongly typed such that safety of dynamic class upgrades can be statically ensured \cite{yu06fmoods}.
In this section, we explain the concurrency model of Creol using a toy example: an exclusive resource, i.e., a
resource that can be exclusively allocated to one object at a time, behaving as a mutual exclusion token. 
Further, we extend Creol with priority-based application-level scheduling.

The state of an object in Creol is initialized using the \lstinline{init} method.
Each object then starts its active behavior by executing its
\lstinline{run} method if defined. % (e.g., in class \lstinline{User}). 
When receiving a method call, a new process is created to execute the method. 
Creol promotes {\em cooperative} non-preemptive scheduling for each active object. 
It means that a method runs to completion unless it explicitly releases the processor. 
As a result, there is no race condition between different processes accessing object variables. 
%Methods can have processor release points which define interleaving points explicitly.
Release points can be conditional, e.g.,   \lstinline{await }$\sim$\lstinline{taken}. 
If the guard at a release point evaluates to true, the
process keeps the control, otherwise, it releases the processor and becomes
disabled as long as the guard is not true. Whenever the processor is free, an
enabled process is {\em nondeterministically} selected for execution, i.e.,
scheduling is left unspecified in standard Creol in favor of more abstract modeling.

%\Mahdi{Do we want to talk about with clauses?}

To explain extending Creol with priority specification and scheduling, we take a {\em client/server} perspective.
Each {caller} object is viewed as a client for the {callee} object who behaves as a server. 
We define priorities at the level of language constructs like method invocation or definition rather than low-level concepts like processes.

\begin{lstlisting}[float=t, label=lst:exres, caption=Exclusive Resource in Creol %, multicols=2
]
interface Resource begin
	(* \label{lst:priority_range} *)(*
	%with User  \label{lst:with} 
	*)
		op request()
		op release()
end

class ExclusiveResource implements Resource begin
	var taken := false;
	(* \label{lst:priority_class} *)
	op request () == 
		await (*$\sim$*)taken; 			(* \label{lst:await_taken} *)
		taken := true;
	op release () ==
		taken := false
end
\end{lstlisting}

On the server side, an interface may introduce a \textit{priority
range} that is available to all clients. For instance, in Line
\ref{lst:priority_range} of {\footnotesize\texttt{Resource}} in Listing
\ref{lst:exres}, we can define a priority range:

\begin{lstlisting}[frame=none, numbers=none]
priority range 0..9
\end{lstlisting}

On the client side, method calls may be given priorities within the range specified in the server interface. For example, calling the \ftt{request} method of the \ftt{mutex} object:

\begin{lstlisting}[frame=none, numbers=none]
mutex ! request() priority(7);
\end{lstlisting}

Scheduling \textit{only} on the basis of client-side priority requirements is too restrictive.
For example, if there are many \ftt{request} messages with high priorities and a low priority \ftt{release}, the server might block as it would fail to schedule the \ftt{release}.
In this particular example, we can solve this problem by allowing the servers to
prioritize their methods.
This involves a declaration of a priority range generally depending on the structure of the class. 
In our example, assuming a range 0..2 added in Line \ref{lst:priority_class}, this requires changing the method signatures in the {\ftt{Exclusive\-Resource}} class:

\begin{lstlisting}[frame=none, numbers=none]
op request() priority(2) == ...
op release() priority(0) == ...
\end{lstlisting}

This gives \ftt{release}  a higher priority over \ftt{request}, because by default, the smaller values indicate higher priorities.
Furthermore, the server may also introduce priorities on certain 
characteristics of a method invocation such as the kind of ``release statement''
being executed. 
For example, a process waiting at Line \ref{lst:await_taken} could be given a higher priority over new requests by assigning a smaller value:

\begin{lstlisting}[frame=none, numbers=none]
await (*$\sim$*)taken priority(1);
\end{lstlisting}

The priority can be specified in general as an expression; we used here
only constant values.
Evaluation of this expression at runtime should be within the given
range.
If no priority is specified for a method invocation or definition, a
default priority value will be assigned.

% 
% This is an example of {\em method level} priority specification, which is specified inside a class together with the definition of a method. 
% Method definitions without an explicit priority are given a default priority value. 
% %As discussed in Section \ref{sec:conclusion}, 
% Other possible levels of priority include method invocation level, specified by the caller, or interface level priorities. 
% %

We discussed different levels of application-level priorities. 
Note that now each method invocation in
the queue of the object involves a tuple of priorities.
We define a general function $\delta$ as an ``abstract priority manager'':
%
\[ \delta : P_1 \times P_2 \times P_3 \times \ldots \times P_n \longrightarrow  P \]

The function $\delta$ maps the different levels of priority in the object
($\{P_1,\ldots,P_n\}$) to a single priority value in $P$ that is internally used by the
object to order all the messages that are queued for execution. Each
method in the object may have a $\delta$ assigned to it. In an extended
version of $\delta$, it can also involve the internal state of the object,
e.g., the fields of the object. 
In this case, we have dynamic priorities.

For example, in {\ftt{ExclusiveResource}}, we have two different levels of priorities, namely  the client-side and server-side priorities, which range over $P_1 = \{0, \ldots, 9\}$ and $P_2 = \{0, 1, 2\}$, respectively. 
So, we define $\delta: P_1 \times P_2 \rightarrow P$ as:
%
\[
\delta(p_1, p_2) = p_1 + p_2 \times |P_1|
\]

To see how it works, consider a \ftt{release} and a \ftt{request} message, both sent with the client side priority of 5. 
Considering the above method priorities, we have $\delta(5, request) = \delta(5,2) = 25$ and  $\delta(5, release) = \delta(5,0) = 5$.
It is obvious that the range of the final priority value is $P = \{0,\ldots,29\}$.

Note that the abstract priority manager in general does \textit{not} completely
fix the ``choice'' of which method invocation to execute. 
In our tool-set, we include an extensible library of predefined scheduling policies such as strong or weak fairness that further refine the application-specific multi-level priority scheduling. 
The policies provided by the library are available to the user to annotate the classes.
We may declare a scheduling policy for the \ftt{ExclusiveResource} class by adding at Line
\ref{lst:priority_class} in Listing \ref{lst:exres}: 
%
\begin{quote}
\lstinline{scheduling policy StronglyFairScheduler;} 
\end{quote}

% These scheduling policies will also be represented in terms of a corresponding
% priority range in the $\delta$ function.
% In summary, there will be, first, priorities that directly stem from the
% model and are statically specified by the client or the server and, second, 
% there will be priorities that dynamically change through time based on
% the runtime information of the application. In the latter case, 
A scheduling policy may
use runtime information to re-compute the dynamic priorities and ensure
properties such as fairness of the selected messages; for instance, it may
take advantage of the ``aging'' technique to avoid starvation. 
%Thus, $\delta$ will take into consideration all the different concerns when assigning a final priority value to a message.

%\Mahdi{Starvation?}

\section{Tool Architecture} \label{sec:compiler}

\begin{figure}[t]
\begin{center}
  %\vspace{-.1cm}
  \includegraphics[scale=0.36]{figs/manticore-concurrency-v4}
  \caption{\Crisp Architecture: Structural Overview}
  \label{fig:api}
\end{center}
\end{figure}

We have implemented a tool to translate Creol programs into Java programs for execution, called {\em \Crisp} (Creolized Interacting Scheduling Policies).
\Crisp provides a one-to-one mapping from Creol classes and methods to their equivalent Java constructs. 
In order to implement active objects in Creol, we use the \jtt{java.util.concurrent} package (see Figure \ref{fig:api}).
Each active object consists of an instance of a process store  and an execution manager to store and execute the method invocations.

Incoming messages to the active object are modeled as instances of \MethodInvocation, a subclass of \jtt{java.util.\-concurrent.FutureTask} that  wraps around the original method call.
Therefore the caller can later get the result of the call.
Additionally, \MethodInvocation encapsulates information such as priorities assigned to the message. 
% Asynchronous messages are stored in the process store as instances of \MethodInvocation. The result of the asynchronous method call can be fetched through the method invocation that encapsulates a future value.
% To support Creol's processor release statements, method invocations provide the
% means to be stopped and resumed.

The \ftt{ProcessStore} itself uses an implementation of the \ftt{BlockingQueue} interface in \jtt{java.util.concurrent} package. Implementations of \ftt{BlockingQueue} are thread-safe, i.e.,  all methods in this interface operate atomically using internal locks encapsulating the implementation details from the user.

The \ftt{ExecutionManager} component is responsible for selecting and executing a pending method invocation.
It makes sure that only one method runs at a time, and takes care of processor release points.
%It is implemented using \jtt{java.util.concurrent} in order to provide deployment on multi-core  architectures. We now discuss how in overall an active object operates on a multi-core architecture.

In the following, we explain how the active object behaves in different scenarios, from a ``client/server'' perspective.  
% for the sake of explanation; each active object
% essentially can be seen as a server, in which case all the other objects act as its clients that
% request a method to be executed.

% \subsection{The Problem}
% \input{problem}

\subsection{A New Method Invocation}
\label{sec:add}

\begin{figure}[t]
%\begin{center}
\includegraphics[scale=0.36]{figs/seq-diag-1}
~\\[-3.5cm]ADD:\\[2.5cm]
  \caption{Adding the new MethodInvocations are performed on the Client side.}
  \label{fig:add}
%\end{center}
\end{figure}

A method call needs to be sent from the client to the server in an asynchronous way. 
To implement this in Java, the client first constructs an instance of
\MethodInvocation that wraps around the original method call for the server.
Then, 
%the client needs to directly fetch the \ftt{ProcessStore} of the
%server and \ftt{add}s the instance of the method invocation to the server.
%
%Active objects consist of a \texttt{\textbf{ProcessStore}} which stores
% method invocations as described above and a \texttt{\textbf{ExecutionManager}}
% which controls the actual execution of method invocations.
% When an instance of \texttt{\textbf{MethodInvocation}} is created by a client, 
there are two implementation options how to add it to the server's process store:
\begin{enumerate}
  \item The client calls a method on the server to store the instance.
  \item The client directly adds the method invocation into the process
store of the server. 
\end{enumerate}

In option $1$, the server may be busy doing something else. Therefore,
in this case the client must wait until the server is free,  which is
against the asynchronous nature of communication. In Option $2$, the
Java implementation of each active object exposes its process store as
an immutable public property. Thus, the actual code for adding the new
method invocation instance is run in the execution thread of the
client. 
We adopt the second approach as depicted in Figure \ref{fig:add}.
At any time, there can be concurrent clients that are
storing instances of \ftt{MethodInvocation} into the server's process store, but since
the process store implementation encapsulates the mechanisms for concurrency and
data safety, the clients have no concern on data synchronization and concurrency
issues such as mutual exclusion.
The method or policy used to store the method invocation in the process store of
the server is totally up to the server's process store implementation details.
%Storing a method invocation in the process store of the server takes place in
%the client side thread of execution. Listing \ref{lst:jfib_method_fibonacci}
%presents a sample generated Java code for an asynchronous method call in Listing
%\ref{lst:cfib}.

\subsection{Scheduling the Next Method Invocation}

On the server side of the story,  an active object 
repeatedly fetches an instance of method invocation from its process
store for execution (cf. Fig \ref{fig:take}).
The process store uses its instance of
\ftt{SchedulingManager} to choose one of the method invocations. 
\Crisp has some predefined scheduling policies that can be used as scheduling managers;
nevertheless, new scheduling policies can be easily developed
and customized based on the requirements by the user. 
%Listing \ref{lst:aoloop} presents the main loop for the active object.

\begin{figure}[t]
~\\TAKE:\\
%\begin{center}
%\vspace{-1.3cm}
  \includegraphics[scale=0.3]{figs/seq-diag-2}
\vspace{-.7cm}
  \caption{An active object selects a method invocation based on its local scheduling policy. After a method finishes execution, the whole scenario is repeated.}
  \label{fig:take}
%\end{center}
\end{figure}

\ftt{SchedulingManager} is an interface the implementations of which
introduce a function to \emph{select} a method invocation based on different
possible criteria (such as time or data) that is either predefined or customized
by the user. The scheduler manager is a component used by process store when
asked to remove and provide an instance of method invocation to be executed.
Thus, the implementation of the scheduling manager is responsible how to choose
one method invocation out of the ones currently stored in the process store of
the active object. Different flavors of the scheduling manager may include
time-based, data-centric, or a mixture.
%\Mahdi{What are concrete examples of scheduling managers?} 
%\Rudi{There is a paper submitted to FMCO post-proceedings about user-defined
%scheduling policies; unfortunately there is no reference to cite yet}

Every method invocation may carry different levels of priority information,
e.g., a server side priority assigned to the method or a
client side priority. 
The \ftt{PriorityManager} 
%is an interface the implementations of which are supposed to 
provides a function to determine and
resolve a final priority value in case there are different levels of priorities
specified for a method invocation. 
% This feature is used, if specified, by the
% process store to compute the ``resolved priority'' of the method invocation
% based on different levels of priorities specified and then add it to the
% storage. 
%An example of a user-defined priority manager is given in Section \ref{sec:caseStudy}.
Postponing the act of resolving priorities to this point rather than when inserting new processes to the store enables us to handle dynamic priorities.

\subsection{Executing a Method Invocation}

To handle processor release points, Creol processes should preserve
their state through the time of awaiting.
This is solved by assigning an instance of a Java thread to each method
invocation. 
An \ftt{ExecutionManager} instance, therefore, employs a ``thread pool''
for execution of its method invocations. 
To create threads, it makes use of the factory pattern:
\ftt{ThreadFactory} is an interface used by the execution manager to
initiate a new thread when new resources are required. 
We cache and reuse the threads so that we can control and tune the performance of resource allocation.

When a method invocation has to release the processor, its thread must
be suspended and, additionally, its continuation must be added to the
process store. To have the continuation, the thread used for the method
invocation should be preserved to hold the current state; otherwise the
thread may be taken away and the continuation is lost.
The original \ftt{wait} in Java does not provide a way to achieve this
requirement.
Therefore, we introduce \ftt{InterruptibleProcess} as an extension of
\jtt{java.lang.Thread} to preserve the relation.

As shown in Figure \ref{fig:execute}, the thread factory creates threads of type  {\ftt{InterruptibleProcess}}.
The execution manager thread blocks as soon as it starts the interruptible process which executes the associated  method invocation. 
If the method releases the processor before completion, it will be added back to the process store as explained in Section \ref{sec:add}.
When a suspended method invocation is resumed, the execution manager skips the creation of a new thread and  reuses the one that was assigned to the method invocation before.

\begin{figure}[t]
~\\EXECUTE:\\
% \begin{center}
% \vspace{-1.3cm}
 \includegraphics[scale=0.35]{figs/seq-diag-3}
\vspace{-.7cm}
\caption{A method invocation is executed in an interruptible process. The execution manager thread is blocked while the interruptible process is running.}
  \label{fig:execute}
% \end{center}
\end{figure}

% Our \ftt{ExecutionManager} is an implementation of the standard Java ~interface 
% \jtt{java.util.concurrent.Executor\-Service}. 
% Each active object holds a reference to one execution manager instance.
% Each \ftt{ActiveObject} implements \jtt{java.lang.\-Run\-na\-ble}. So, when
% it gets initialized, it adds \ftt{this} (itself) to
% its execution manager component. This procedure \textit{deploys} the active object
% onto the execution manager corresponding to the underlying multi-core platform.

% FRB: the above last  sentence is not very clear and seems more to hide than to clarify.
% Also the remainder of the section needs to be improved.
% I suggest to add some sequence diagrams.
% Question: when a method invocation executes an await what actually happens to its executing thread?
% Why do we need a thread-pool because only one method at a time is executed?

\subsection{Extension Points}

Besides the methods \ftt{add} and \ftt{take} for adding and removing method invocations, \ftt{ProcessStore} provides methods such as \ftt{preAdd} and \ftt{postAdd} along with \ftt{preTake} and \ftt{postTake} respectively to enable  further customization of the behavior before/after adding or taking a method invocation to/from the store. These extension points enable the customization of priority or scheduling management of the method invocations.

\Crisp provides two generic interfaces for priority specification and scheduling management: \ftt{PriorityManager} and \ftt{SchedulingManager} respectively. These two interface can be freely developed by the programmer to replace the generated code for priorities and scheduling of the messages. It will be the task of the programmer to configure the generated code to use the custom developed classes.

\section{Case Study} \label{sec:caseStudy}

In this section, we demonstrate the use of application-level scheduling and \Crisp with a more complicated example: we program the ``Sieve of Eratosthenes'' %\cite{Sieve_Eratosthenes_wiki}
to generate the prime numbers. 
To implement this algorithm, the 
% \ftt{Sieve} class generates the natural numbers and sends them to the \ftt{Prime} class instances to check whether they can divide these numbers: if a number is not divisible by all primes before that, then it is a new prime number.
% %Initialization of an object happens in its \ftt{init} method, while the \ftt{run} method specifies its active behavior.
% The 
\ftt{Sieve} is initialized by creating an instance of the \ftt{Prime} object representing the first prime number, i.e., \ftt{two}. The active behavior of \ftt{Sieve} consists of generating all natural numbers up to the given limit (100000 in our example) and passing them to the object \ftt{two}. 
A \ftt{Prime} object that cannot divide its input number passes it on to the next \ftt{Prime} object; if there is no next object, then the input number is the next prime and therefore a new object is created.

We parallelize this algorithm by creating active objects that run in parallel.
The numbers are passed asynchronously as a parameter to the \ftt{divide} message. 
Correctness of the parallel algorithm essentially depends on the numbers being processed in increasing order.
For example, if object \ftt{two} processes 9 before 3, it will erroneously treat 9 as a prime, because 3 is not there yet to eliminate 9. 
%Such a scenario is possible if the scheduling of the tasks in the \ftt{Prime} objects is not specified. 
%
To avoid erroneous behavior, we use the actual parameter $n$ in \ftt{divide} method to define its priority level, too (see line \ref{line:priority}). 
As a result, every invocation of this method generates a process with a priority equal to its parameter.
The default scheduling policy for objects always selects a process for execution that has the smallest priority value. 
This guarantees that the numbers sent to a \ftt{Prime} object are processed exactly in increasing order.

\lstset{language=Creol,escapeinside={(*}{*)}}  
\begin{lstlisting}[float=t, label=lst:prime-sieve-creol, caption=Prime Sieve in Creol]
interface IPrime begin
	op divide(n:Int)
end

class Sieve begin
	var n: Int, two: IPrime
	op init == two := new Prime(2); n := 3
	op run ==
		!two.divide(n);
		if n < 100000 then n := n + 1; !run() end
end

class Prime(p: Int) implements IPrime begin (*\label{lst:priority}*)
	var next: IPrime
	op divide(n: Int) priority (n) == (*\label{line:priority}*)
		if (n % p) /= 0 then 
			if next /= null then
				!next.divide(n)
			else
				next := new Prime(n)
		end end
end
\end{lstlisting}
% 
% class Test
% begin
% 	var g : Sieve
% 	op run == g := new Sieve
% end

We used two different setups to execute the prime sieve program and compare the results. In one setting, we ran the parallel prime sieve compiled by \Crisp; in the other, we executed a sequential program developed based on the same algorithm that uses a single thread of execution in JVM. We performed the experiments on a hardware with $2$ CPU's of each $2$GHz with a main memory of size $2$GB. We ran both programs for $max \in \{10000, 20000, 30000, 50000, 100000\}$.
% We compared two interesting properties of the executions: the thread stack and the number of threads required.

The first interesting observation was that \Crisp prime sieve utilizes all the CPU's on the underlying hardware as much as possible during  execution. 
This can be seen in Figure \ref{fig:multi-cpu-1} which shows the CPU usage. 
Both CPUs are fairly in use while running this program.
Figure \ref{fig:prime_sieve_cacoj} depicts the results of the monitoring of the parallel prime sieve in \Crisp using Visual VM tool. 
It depicts the number of threads generated for the program.
This shows that \Crisp can handle a massive number of concurrent tasks.

One interesting feature of \Crisp is that the execution of any program under \Crisp can constantly utilize the \emph{minimum} memory that can be allocated for each thread in JVM (thread stack). 
In JVM, the size of thread stack can be configured using \ftt{-Xss} option for every run. 
To demonstrate this feature of \Crisp, we collected the minimum stack size needed for every program run in Table \ref{tbl:cmp-thread-stack}. 
All \Crisp runs use the minimum thread stack size of \ftt{64k} that is possible for the JVM.
On the contrary, the stack size required for the sequential version of the sieve program increases by the number of primes detected. 
This is also expected because of the long chain of method calls in the sequential sieve.

% The results depict that the sequential version, through time, increasingly required memory allocation for the thread stack, however, \Crisp performs on a constant size of thread stack. 
% Table \ref{tbl:cmp-thread-stack} summarizes the thread stack allocation on different runs.

\begin{table}[b]\scriptsize
\begin{center}
\begin{tabular}{c c c c c c}
$max$ & $10000$ & $20000$ & $30000$ & $50000$ & $100000$ \\ \thickhline
Sequential & 64k & 72k & 96k & 160k & 190k \\ 
\Crisp & 64k & 64k & 64k & 64k & 64k \\ \thickhline
\end{tabular}
\caption{Thread stack allocated for different executions}
\label{tbl:cmp-thread-stack}
\end{center}
\end{table}

Having the constant thread stack size feature, \Crisp provides another interesting feature. It can handle huge number of thread generation if required. Table \ref{tbl:cmp-thread-num-gen} summarizes the thread generation data for parallel prime sieve in \Crisp. It shows scalability of \Crisp as $p$ rises for parallel prime sieve.
\begin{table}[b]\scriptsize
\begin{center}
\begin{tabular}{c c c}
$max$ & Live Peak & Total \\ \thickhline
$10000$ & 817 & 540591 \\
$20000$ & 1468 & 1854067 \\
$30000$ & 2204 & 4054814 \\
$50000$ & 3707 & 11852985 \\ \thickhline
\end{tabular}
\caption{Number of live threads and total threads created for different runs of parallel prime sieve}
\label{tbl:cmp-thread-num-gen}
\end{center}
\end{table}

As the results show the use of Java threads is costly; first, \Crisp
does not need much of the memory allocated to each thread and, second,
the context switch cost is higher for larger memory allocation. In line
with this, JVM uses a \emph{one-to-one} mapping from an
application-level Java thread to a native OS-level thread. 
In the current setting, the context switch of the threads are in
the OS level. When the context switch is taken to the application
level, we leverage the performance issue from the OS level to the
application level. We further discuss this in Section
\ref{sec:conclusion}.

% In future work, in order to support high numbers of concurrent tasks,
% \Crisp will map a set of Creol messages to one Java thread for each
% active object; thus, indirectly resolving the thread performance
% issue.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.5]{figs/p50000vvm}
\caption{Increasing parallelism in \Crisp for Prime Sieve}
\label{fig:prime_sieve_cacoj}
% \end{center}
% \end{figure}
% 
% \begin{figure}[t]
% \begin{center}
  \includegraphics[scale=0.44]{figs/p50000sysmon} 
  \caption{Utilizing both CPUs with Prime Sieve in \Crisp}
  \label{fig:multi-cpu-1}
\end{center}
\end{figure}

\section{Related Work} \label{sec:relwork}
\input{related2}

\section{Conclusion} \label{sec:conclusion}
In this paper, we proposed \Crisp as an implementation scheme for
application-level scheduling of active objects. 
\Crisp first introduces asynchronous message passing with fine-grain
priority management and scheduling of messages. 
Additionally, it introduces a Creol to Java compiler that translates
the active objects in Creol into an equivalent Java application. 
\Crisp compiler seamlessly integrates Java class libraries into Creol
including data types that turns Creol from a modeling language to a
fully-fledged one in the hands of the programmer. 
% \Crisp compiler takes advantage of template engines; i.e. the code
% generation templates can easily be customized in the direction of the
% requirements of the problems at hand.

The \jtt{java.util.\-con\-current} package provides useful API for
concurrent programming. Java futures facilitate modeling
asynchronous message passing. However, for processor release
points, we had to preserve threads (using \jtt{Interruptible\-Process})
to allow continuations which leads to their OS-level context switching
that is costly. Moreover, we were tightly directed to use the
out-of-the-box \jtt{Executor\-Service} which is limitedly extensible. We
had no control over the scheduling mechanisms of the internal queue used
in the service implementations. Thus, we needed to re-implement some of
the concepts. Through prototyping \Crisp, we learned that
there are two major
challenges ahead. Firstly, we need to integrate continuations into Java
using a many-messages-to-one-thread mapping model. Secondly, we need
complete control over scheduling of messages and threads in
\jtt{Executor\-Service}'s internal queue. Table \ref{tbl:eval}
summarizes this discussion.

\begin{table}[h]\small
\begin{center}
\begin{tabular}{p{1.5cm} p{2cm} p{2cm} p{1.5cm}}
 & Asynchronous Communication & Processor Release Point & Scheduling \\
\thickhline
Modeling & \ding{51} & \ding{51} & \ding{53} \\
Performance & \ding{51} & \ding{53} & \ding{51} \\ \thickhline
 & Java Futures & Interruptible Process & Executor Service \\
\end{tabular}
\caption{Overview of evaluation of challenges}
\label{tbl:eval}
\end{center}
\end{table}

In future, we will focus on thread performance
for \Crisp such that thread scalability can be achieved to a certain
limit. Additionally, the development of concurrency features on
multi-core in \Crisp is one of the major future concentrations.
% We will extend Creol to provide the necessary syntax to express all
% different levels of priorities.
% including method level, message level, and object level priorities. 
Moreover, another line of future work involves profiling and monitoring
objects  at runtime  to be used for optimization and performance
improvement.
In addition, we intend to extend and integrate into our tool set 
model checking engines such as Modere \cite{JaghooriMS06}.

% \bibliographystyle{plain}
% \bibliography{references}

% Local Variables:
% fill-column: 80
% End:

### 2015-10-20 p01-ch01-2012/problem.tex
Initial Commit

In this section we briefly sketch the main software engineering problem of deploying 
application-level scheduling of Creol objects onto multicore.

A simple implementation  maps each Creol object unto a single Java thread (e.g., {\jtt{java.lang.Thread}}) which
executes messages taken from a message queue according to some
scheduling policy.  The major drawback of this implementation strategy
is scalability.  OS-level threads are costly resources, and a typical
system written in object-oriented style might instantiate tens of
thousands of objects at runtime.  Thus, any real-world program would
quickly overwhelm current operating systems, see for example our case study in Section \ref{sec:caseStudy}.
Another major software engineering problem of this scheme is that it requires an implementation of  asynchronous method calls 
(e.g., polling/waiting for return values and storing/retrieving local variables).

The basic mechanism of asynchronous method calls is in fact provided by the {\jtt{java.util.\-concurrent}} package by
means of thread pools
which leverage hardware level constructs  to allow a reduction of the overhead of the multi-threaded Java programs by using lock-free, and wait-free thread-safety mechanisms.
This package, e.g., the implementation of {\jtt{FutureTask}}, requires another implementation strategy for a Java backend for Creol which uses
one Java thread per Creol process and allows for a one-to-one translation of Creol classes and methods to Java classes and methods.
However,  one of the major problems in leveraging scheduling issues is that this package does not allow high-level control of the scheduling of the threads.
The main contribution of the paper is the introduction of a general architecture on top of the {\jtt{java.util.\-concurrent}} package
to provide such high-level control for concurrent objects.

% Local Variables:
% mode: LaTeX
% TeX-master: "paper.tex"
% End:

### 2015-10-20 p01-ch02-2013/paper.tex
Initial Commit

% -*- root: ../thesis.tex -*-

\chapter[The Future of a Missed Deadline]{The Future of a Missed Deadline}
% 
\label{ch:p01:ch02}
% 
\chapterauthor{Behrooz Nobakht, Frank S. de Boer, Mohammad Mahdi Jaghoori}

\section*{Abstract}
In this paper, we introduce a real-time actor-based programming language
and provide a formal but intuitive  operational semantics for it. 
The language supports a general mechanism for handling exceptions raised by missed deadlines
and the specification of application-level scheduling policies.
We discuss  the implementation of the language and 
illustrate the use of its constructs with an industrial case study from distributed e-commerce and marketing domain.

% \keywords{actors, application-level scheduling, real-time, deadlines, futures, Java.}

\section{Introduction} \label{sec:introduction}

In real-time applications, rigid deadlines necessitate stringent scheduling strategies.
Therefore, the developer must ideally be able to program the scheduling of different tasks inside the application.
Real-Time Specification for Java (RTSJ) \cite{jsr1,jsr282} is a major extension of Java, as a mainstream programming language, aiming at enabling real-time application development. 
Although RTSJ extensively enriches Java with a framework for 
the  specification of real-time applications, 
it yet remains at the level of conventional \textit{multithreading}.
The drawback of multithreading is that it involves the programmer with OS-related concepts like threads, whereas a real-time Java developer should only be concerned about high-level entities, i.e., objects and method invocations, also with respect to 
real-time requirements.

The actor model \cite{fnd:actor:cmpt:agha} and actor-based programming languages, which
have re-emerged in the past few years \cite{kilim:Srinivasan:Mycroft,erlang:armstrong,scala:actors:ordersky,creol:broch:owe,salsa:agha}, 
provide a different and promising paradigm for concurrency and distributed computing, in which threads are transparently encapsulated inside actors.
As we will argue in this paper, this paradigm is much more suitable for real-time programming because it enables the programmer to 
obtain the appropriate   high-level view which allows the management of   complex
real-time requirements.

In this paper, we introduce an actor-based programming language Crisp for real-time applications. 
Basic real-time requirements include deadlines and timeouts.
In Crisp, deadlines are associated with asynchronous messages and timeouts  with futures \cite{BoerCJ07}.
Crisp further supports a general actor-based mechanism for handling exceptions raised by  missed deadlines.
By the  integration  of these basic real-time control mechanisms with the application-level policies supported by Crisp for scheduling of the messages inside an actor,
more complex real-time requirements of the application can be met with more flexibility and finer granularity.

We formalize the design of Crisp by means of structural  operational semantics \cite{plotkin:sos} and 
describe its implementation as a full-fledged programming language.
This implementation uses both the Java and Scala language with extensions of Akka library.
We illustrate the use of the programming language with an industrial case study from SDL Fredhopper that provides enterprise-scale distributed e-commerce solutions on the cloud.

% In previous work \cite{crisp-sac}, we presented a general framework, called Crisp, 
% to port the notion of {\em concurrent objects} from Creol \cite{creol:broch:owe,mpd:andrews} to Java. 
% A concurrent object encapsulates one processor, i.e., it has a
% single thread of execution. % that is controlled by the object itself.
% In other words, existing objects execute concurrently.
% In this setting, the thread of execution never leaves the enclosing object; instead, method invocations
% result in a new process inside the target object, i.e., they are asynchronous.  
% % Thus, a concurrent
% % object provides a natural basis for a deployment scheme where each
% % object virtually possesses one processor.
% Such a context brings the scheduling problem to the level of each active object in order to resolve the basic issue
% of which method in the object to select for execution.
% The Crisp framework introduces a scheduler for each object as a mathematical function on the values of different priority levels associated to each method definition, the current invocation or the caller.
% We concretize this idea in Crisp by allowing the developer to program a scheduler function inside each concurrent object definition.
% % This function can take as input the invocation deadline as well as other static and dynamic priorities associated to the method. 
% Additionally, active objects are able to use their own internal state to determine how the scheduling of 
% messages are done; this brings the problem of scheduling for real-time from low-level threads to the 
% level of objects and methods.
% 
% When message scheduling is elevated to the level of application inside the active object using the
% scheduler function, real-time requirements of the application can be met with more flexibility and finer granularity.
% Such real-time requirements include deadlines and timeouts. Using Crisp, each active object is able to
% take control of the different ways it behaves and reacts to deadlines and timeouts. Crisp allows the
% active object to define local timeouts in case the response is not received in due time. Additionally, 
% an active object may impose a deadline on the callee active object when sending a message to expect a response
% in a due time. Thus, Crisp provides seamless and direct mechanisms to meet real-time requirements
% at application level which relieves the programmer from many lower-level concerns.
% 
% 
% We extend our previous work \cite{crisp-sac} for Crisp. 
% Originally, Scala \cite{scala:actors:ordersky,coord:ordersky}, that runs on JVM, introduced a package
% for actor model programming. As an extension, Akka \cite{akka:homepage} extends Scala's actor library 
% with more facilities. We develop Crisp using both Java and Scala language with extensions of Akka library.
% In Crisp, actor features from Akka are extended and developed further to provide high- and application-level
% capabilities at the level of active objects to allow them take control of scheduling of the messages. 
% In Crisp, each active object may define a scheduler function while it may also use the notions of timeouts
% and deadlines when sending messages to other active objects. It is also possible to mix different scheduling
% mechanisms in Crisp which makes it more flexible for the active object in order to schedule message
% executions. The source code of Crisp along with previous work can be found at 
% \ftt{\url{http://github.com/nobeh/crisp}}.

The paper continues as follows: Section \ref{sec:deadlines} introduces the language constructs
and provides informal semantics of the language with a case study in Section \ref{sec:frh}. 
Section \ref{sec:sos} presents the operational semantics of Crisp.
Section \ref{sec:impl} follows to provide a detailed discussion on the implementation.
The case study continues in this section with further details and code examples.
Section \ref{sec:relwork} discusses related work of research and 
finally Section \ref{sec:conclusion} concludes the paper and proposes future line of research.

\section{Programming with deadlines} \label{sec:deadlines}
In this section, we introduce the basic concepts underlying  the notion of ``deadlines'' for asynchronous messages between actors. 
The main new constructs specify how a message can be sent with a deadline, how the message response can be processed, and what happens when a deadline is missed.
We discuss the informal semantics of these concepts  and illustrate them using a case study in Section \ref{sec:frh}.

Listing \ref{fig:grammar} introduces a minimal version of the  real-time
actor-based language Crisp.
Below we discuss the two main new language constructs presented at lines (7) and (8).
% In Line (6), we introduce an extension on how each concurrent object can send a message to another concurrent object annotated with an extra
% property that represents a deadline ($d$).

\begin{figure}[t]
\begin{align}
C &::= \textbf{\jtt{class}} \; \mathsf{N} \; \textbf{\jtt{begin}} \; V^? \; \{M\}^* \; \textbf{\jtt{end}} \\
M_{sig} &::=
%\textbf{\jtt{op}} \; 
\mathsf{N}(\overline{T \: x}) \\ 
M &::= \{ M_{sig} == \{V \; ;\}^? S \} \\
V &::= \textbf{\jtt{var}} \; \{\{x\},^+ \; : \; \mathsf{T} \; \{ = e\}^? \},^+ \\
S &::=  x := e\; | \\
 &::= x := \textbf{\jtt{new}} \; \mathsf{T}(e^?) \; |\\
&:= f = e \; ! \; m(\overline{e}) \; \textbf{\jtt{deadline}}(e) \: | \\
 &::= x := f.\gett(e^?) \; | \\
  &::= \textbf{\jtt{return}} \; e  \; | \\
  &::= S \; ; \; S \; | \\
  &::= \textbf{\jtt{if}} \; (b) \; \textbf{\jtt{then}} \; S \; \textbf{\jtt{else}} \; S \; \textbf{\jtt{end}}  \; | \\
  &::= \textbf{\jtt{while}} \; (b) \; \{ \; S \; \} \; | \\
  &::= \textbf{\jtt{try}} \: \{ S \} \: \jtt{\textbf{catch}}(\mathsf{T}_{\jtt{Exception}} \; x) \: \{ \: S \: \}  
%E &::= B \; | \; V \\
%B &::= \textbf{\jtt{true}} \; | \; \textbf{\jtt{false}} \; | \; B \wedge B \; | \; B \vee B \; | \; %\textbf{\jtt{not}} \; B
\end{align}
\caption{
A kernel version of the real-time programming language.
The bold scripted \textbf{\jtt{keywords}} denote the reserved words in the language.
The over-lined $\overline{v}$ denotes a sequence  of syntactic entities $v$.
Both local and instance variables are denoted by $x$.
We assume distinguished local variables  $\jtt{this}$, $\jtt{myfuture}$, and  $\jtt{deadline}$ which
denote the actor itself, the  unique future corresponding to the process, and  its deadline,  respectively.
A distinguished instance variable $\jtt{time}$  denotes  the current time.
Any subscripted type $\mathsf{T}_{specialized}$ denotes a \textit{specialized} type of general type $\mathsf{T}$; 
e.g. $\mathsf{T}_{\jtt{Exception}}$ denotes all ``exception'' types.
A variable $f$ is in $\mathsf{T}_{future}$. 
\textsf{N} is a name (identifier) used for classes and method names.
%$e$ is in \textsf{Expression} ($E$);
$C$ denotes a class definition which consists of  a definition of its instance variables
and its methods;
$M_{sig}$ is a method signature; $M$ is a method definition; $S$ denotes a statement.
We abstract from the syntax the side-effect free expressions $e$ and
boolean expressions $b$.
%
%For simplicity, we drop the notation syntax to specify how $V$ can expand to different values along with arithmetic operations.
%\emph{Note} that small letters (e.g. $e$) are of type of their capital constructs (e.g. $E$).
}
\label{fig:grammar}
\end{figure}

\parbf{How to send a message with a deadline?} 
% Let us consider that object $o$ sends message $m$ to object $o'$ with deadline $d$:
The construct
$$
f = e_0 \; ! \; m(\overline{e}) \; \textbf{\jtt{deadline}}(e_1)
$$
describes an asynchronous message with a deadline specified by $e_1$ (of type $\mathsf{T}_{time}$).
Deadlines  can be specified  using a notion of time unit such as millisecond, second, minute or other units of time.
% The value of the deadline $d$ is relative from the perspective of the caller object $o$.
% The value of the deadline $d$ is translated in the side of callee object $o'$ to an absolute value in a transparent way.
The caller expects the  callee (denoted by $e_0$) to process   the message  within the  units of time
specified by $e_1$.
Here processing a message means starting the execution of the process generated by the message.
A deadline  is missed if and only if the callee  does not  start  processing  the message within the specified units of time.
% In other words, deadline $d$ is missed if the response for message $m$ is not ready within $d$ unit of time at the side of object $o'$.
% \textsl{Note} that we do not consider I/O, infrastructure, and network times as part of the effective times of processing message $m$ 
% and sending the correspondent response.

\parbf{What happens when a deadline is missed?}
Messages received by an actor generate processes.
Each actor contains one active process and all its other processes are queued. 
Newly generated processes are \emph{inserted} in the queue according to an application-specific policy.
When a \emph{queued} process misses its deadline it is \emph{removed} from the queue and a
corresponding \emph{exception} is recorded by its future (as described below).
When the currently active process is terminated the process at the head of the queue is activated
(and as such  dequeued).
The active process  cannot be \emph{preempted} and is forced
to  \emph{run to completion}.
In  Section \ref{sec:impl} we discuss the implementation details of this design choice.

\parbf{How to process the response of a message with a deadline?}
In the above example of an asynchronous message, the \emph{future} result of processing the  message   
is denoted by the  variable $f$ which has the type of \textbf{\jtt{Future}}.
Given a future variable $f$, the programmer can query  the  availability of the  result by the construct

$$
v = f.\gett(e)
$$
The execution of the \gett \space operation terminates successfully when  the future variable $f$ contains the result value.
In case the future variable $f$ records an exception, e.g. in case the corresponding process has missed its  deadline,
the \gett\space operation is \emph{aborted} and the exception is \emph{propagated}. 
Exceptions can be caught by \jtt{try-catch} blocks.

\lstset{language=Creol,escapeinside={(*}{*)}} 
\vspace{-20pt}
\begin{lstlisting}[float=h, label=lst:try:catch, caption=Using try-catch for processing future values]
try {
	x = f.get(e)
	S_1
} catch(Exception x) {
	S_2
}
\end{lstlisting}
\vspace{-20pt}
For example, in Listing \ref{lst:try:catch}, if the \gett \space operation raises an exception  control,  is transferred to line (5);
otherwise, the execution continues in line (3). 
In the \jtt{catch} block, the programmer has also access to the occurred exception that can be 
any kind of exception including an exception that is caused by a missed deadline.
In general, any uncaught exception gives rise to abortion of the active process and is recorded by its future.
Exceptions in our actor-based model thus are propagated by futures.

The additional parameter $e$ of the get operation is of type  $\mathsf{T}_{time}$ and specifies a \emph{timeout};
i.e., the \gett \space operation will timeout after the specified units of time.

% Since, we provide a reference implementation in Java (c.f. Section \ref{sec:impl}), we take advantage of exception ``stack traces''; 
% i.e. an instance of exception can produce a chain of ``causes'' that has led to the occurrence of the exception.
% This means that cathing an exception may in turn lead to throwing another exception or just rethrowing it. 
% Therefore, if the programmer decides not to use a \jtt{try-catch} block, the exception is propagated through the call hierarchy.
% It means that the propagation of the exception may also contribute to the stack trace of another exception.
% If any exception is not eventually caught, the execution environment will receive it and the application will stop.

% As discussed in detail in our previous work \cite{crisp-sac}, object $o'$ maintains a queue of processes for the incoming messages such as $m$.
% While a message $m$ is in the queue of $o'$, there is a chance that the deadline $d$ for message $m$ would be missed.
% We say a deadline $d$ for message $m$ on object $o'$ is missed when
% \begin{enumerate}
%  \item The process for message $m$ is \emph{not} the current process of $o'$. 
%  In other words, we do \emph{not} terminate the current process in case it missses its deadline; 
%  i.e. we allow \emph{run to completion} for the current process. And,
%  \item The response for message $m$ is not ready and the \emph{current} time has passed the deadline $d$ expectation.
% \end{enumerate}
% 
% When a deadline-miss is detected, it is required to be reflected in the future response that is being waited on in object $o$. 
% Therefore, for a deadline $d$ that is missed on message $m$:
% 
% \begin{enumerate}
%  \item The process that is related to message $m$ is aborted and removed from $o'$'s  process queue.
%  \item The aborted state is recorded in the corresponding future value $f$.
%  \item Whenever, \jtt{f.get()} is invoked, the \emph{aborted} state causes an exception. 
%  The exception is supposed to be handled by the user of the future; otherwise, it is propagated in the call hierarchy.
% \end{enumerate}
% 
% \emph{Starvation?} \Behrooz{Talk about the responsiveness of the system.}
% 
% \emph{Discussion?} Compare the approaches: event-based with no get() and exceptions as messages vs. messages with exceptions
\subsection{Case Study: Fredhopper Distributed Data Processing} \label{sec:frh}

\begin{wrapfigure}{r}{0.4\textwidth}
\vspace{-40pt}
\begin{center}
\includegraphics[scale=0.4]{figs/DM}
\end{center}
\vspace{-20pt}
\caption{Fredhopper's Controller life cycle for remote data processing}
\label{fig:frh-dm}
\vspace{-10pt}
\end{wrapfigure}

Fredhopper is an SDL company since 2008 and a leading search, merchandising and personalization solution provider, whose
products are uniquely tailored to the needs of online business. Fredhopper operates behind the scenes of more than 100 of the largest online
sellers. The Fredhopper Access Server (FAS) provides access to high quality product catalogs.  Typically deployments have about 10
explicit attribute values associated with a product over thousands of attribute dimensions. This challenging task involves working on
difficult issues, such as the performance of information retrieval algorithms, the scalability of dealing with huge amounts of data and
in satisfying large amounts of user requests per unit of time, the fault tolerance of complex distributed systems, and the executive
monitoring and management of large-scale information retrieval operations. Fredhopper offers its services and facilities to e-Commerce
companies (customers) as services (SaaS) over the cloud computing infrastructure (IaaS); which gives rise to different challenges in
regards with resources management techniques and the customer cost model and service level agreements (SLA).

To orchestrate different services such as FAS or data processing, Fredhopper takes advantage of a service controller (a.k.a. Controller).
Controller is responsible to passively manage different service installations for each customer.
For instance, in one scenario, a customer submits their data along with a processing request to their data hub server.
Controller, then picks up the data and initiates a data processing job (usually an ETL job) in a data processing service.
When the data processing is complete, the result is again published to customer environment and additionally becomes available through FAS services.
Figure \ref{fig:frh-dm} illustrates an example scenario that is described above.

In the current implementation of Controller, at Step 4, a data job instance is submitted to a remote data processing service.
Afterwards, the future response of the data job is determined by a periodic remote check on the data service (Step 4). 
When the job is finished, Controller continues to retrieve the data job results (Step 5) and eventually publishes it to customer environment (Step 6).

In terms of system responsiveness, Step 4 may never complete. 
Step 4 failure can have different causes. 
For instance, at any moment of time, there are different customers' data jobs running on one data service node;
i.e. there is a chance that a data service becomes overloaded with data jobs preventing the periodic data job check to return.
If Step 4 fails, it leads the customer into an \emph{unbounded waiting} situation.
According to SLA agreements, this is \emph{not} acceptable.
It is strongly required that for any data job, the customer should be notified of the result: either a completed job with \textsl{success/failed} status, a job that is not completed, or a job with an unknown state.
In other words, Controller should be able to guarantee that any data job request terminates.

To illustrate the contribution of this paper, we extract a closed-world simplified version of the scenario in Figure \ref{fig:frh-dm} from Controller.
In Section \ref{sec:impl}, we provide an implementation-level usage of our work applied to this case study.

\section{Operational Semantics} \label{sec:sos}
We describe the semantics of the language by means of a two-tiered labeled transition system:
a local transition system describes the behavior of a single actor and a global transition system
describes the overall behavior of a system of interacting actors.
We define an actor state as a pair $\langle p,q \rangle$, where
\begin{itemize}
%  \item $o$ is the object identifier
 \item $p$ denotes the current active process of the actor, and
 \item $q$ denotes a queue of pending processes.
\end{itemize}
Each pending process is a pair  $(S, \tau)$ consisting of the current executing statement $S$ and the assignment $\tau$ of 
values  to the  \emph{local} variables (e.g., formal parameters).
The active process consists of a  pair  $(S,\sigma)$, where $\sigma$ assigns values to the local variables
and additionally assigns values to the instance variables
of the actor.

% We use structural approach in operational semantics \cite{plotkin-sos} to present the semantics of application-level scheduling
% with deadlines in Creol.

% \emph{Object configuration.} We define an object configuration as a pair of $(O, F)$ with $ O = (o, p, q)$ in which:
% \begin{itemize}
%  \item $o$ is the object identifier
%  \item $p$ is the current process of the object that is executing in the single thread of execution for the object. 
%  Each process is a combination $(S, \tau)$ of current executing statement, $S$, and the bindings of both object's internal fields and local variables, $\tau$.
%  \item $q$ is the queue of processes for the object
%  \item $F$ is the bindings of created objects which represents futures; i.e. $F(o)$ denotes a record. 
%  Each record present two propeties; \textsf{val} representing the actual value of a future and \textsf{aborted} if the future has failed.
% \end{itemize}
% 
% \Behrooz{Add time notations for global transition system.}
% \Behrooz{tau can hold the identifier of an object to simplify.}
% \Behrooz{Mention ``run to completion'' for the current process.}

\subsection{Local transition system}
The local transition system defines transitions among actor configurations of the form $\langle p,q,\phi\rangle $,
where $(p,q)$ is an actor state and for any object $o$ identifying a created future,  $\phi$ denotes the shared heap
of the created future objects, i.e., $\phi(o)$, for any future object $o$ existing in $\phi$,  denotes a record with a field
\textsf{val} which represents the return value  and a boolean field \textsf{aborted} which indicates abortion of the process
identified by $o$.

In the local  transition system we make use of the following axiomatization of the occurrence of exceptions.
Here $(S,\sigma,\phi)\uparrow v$ indicates that $S$  raises an exception $v$:
\begin{itemize}\itemsep5pt
 \item $ (x=f.\gett(),\sigma,\phi) \uparrow \sigma(f)$ where $\phi(\sigma(f)).\mathsf{aborted}=\jtt{true}$,
 \item $\frac{\displaystyle (S,\sigma,\phi)\uparrow v}{\trycatch{S}{T}{u}{S'}\uparrow v}$ where $v$ is not of type $\mathsf{T}$, and, 
 \item $\frac{\displaystyle (S,\sigma,\phi)\uparrow v}{\displaystyle (S;S,\sigma,\phi) '\uparrow v}$.
\end{itemize}

We present here the following transitions describing internal computation steps
(we denote by $ \textsf{val}(e)(\sigma)$ the value of the expression $e$ in $\sigma$
and by $f[u\mapsto v]$ the result of assigning the value $v$ to $u$ in the function $f$).

\parbf{Assignment statement} is used to assign a value to a variable:

$$ \langle (x=e; S, \sigma), q, \phi\rangle \rightarrow \langle (S, \sigma [x \mapsto \textsf{val}(e)(\sigma)]), q, \phi\rangle $$

\parbf{Returning a result} consists of setting the field 
\textsf{val} of the future of the process:

$$
\langle (\textbf{\jtt{return}}\; e ;S , \sigma), q, \phi\rangle \rightarrow 
\langle (S, \sigma), q, \phi[\sigma(\jtt{myfuture}).\textsf{val} \mapsto 
\textsf{val}(e)(\sigma)]\rangle
$$

\parbf{Initialization of \jtt{timeout} in \gett  \space operation} 
assigns to a  distinguished (local) variable \jtt{timeout} its initial \emph{absolute} value:
\begin{multline*} \langle  (x=f.\gett(e); S, \sigma), q, \phi\rangle \rightarrow  \\
 \langle (x=f.\gett(e);S, \sigma [\jtt{timeout} \mapsto \textsf{val}(e+\mathsf{time})(\sigma), q, \phi\rangle 
\end{multline*}

\parbf{The \gett \space operation} is used to assign the value of a future to a variable:
$$ \langle  (x=f.\gett(); S, \sigma), q, \phi\rangle \rightarrow \langle (S, \sigma [x \mapsto \phi(\sigma(f)).\mathsf{val}]), q, \phi\rangle $$
where $ \phi(\sigma(f)).\mathsf{val} \neq \bot$ .

\parbf{Timeout} is operationally presented by the following transition:
$$ \langle  (x=f.\gett(); S, \sigma), q, \phi\rangle \rightarrow \langle (S, \sigma), q, \phi\rangle $$
where $\sigma(\jtt{time})< \sigma(\jtt{timeout})$.

\parbf{The \textnormal{\textbf{\jtt{try-catch}}} block} semantics is presented by:

$$ 
\frac
{\langle (S,\sigma),q, \phi\rangle \rightarrow \langle (S', \sigma'), q', \phi'\rangle}
{\langle (\trycatch{S}{T}{x}{S''};S''', \sigma), q, \phi\rangle \rightarrow \langle (\trycatch{S'}{T}{x}{S''};S''', \sigma), q', \phi'\rangle}
$$

\parbf{Exception handling.} 
We provide the operational semantics of exception handling in a general way in the following:
$$
\frac
{(S,\sigma,\phi) \uparrow v}
{\langle (\trycatch{S}{T}{x}{S''};S''', \sigma), q, \phi\rangle \rightarrow \langle  (S'';S''', \sigma[x\mapsto v]), q, \phi\rangle}
$$
where the exception $v$ is  of type \jtt{T}.

\parbf{Abnormal termination} of the active process is generated by an uncaught exception:
$$
\frac
{(S,\sigma,\phi)\uparrow v}
{\langle (S;S', \sigma), q, \phi\rangle \rightarrow \langle (S'',\sigma'), q', \phi'\rangle}
$$
where $q = (S'',\tau) \cdot q' $ and $\sigma'$ is obtained from restoring the values of the local variables
as specified by $\tau$ (formally, $\sigma'(x)=\sigma(x)$, for every instance variable $x$, and $\sigma'(x)=\tau(x)$, for every local variable $x$), and  $\phi'(\sigma(\jtt{myfuture})).\jtt{aborted} = \jtt{true}$
($\phi'(o)=\phi(o)$, for every $o\not= \sigma(\jtt{myfuture})$).

\parbf{Normal termination} is presented by:
$$
\langle (E, \sigma), q, \phi\rangle \rightarrow \langle  (S,\sigma'), q', \phi\rangle
$$
where $q = (S,\tau) \cdot q'$ and $\sigma'$ is obtained from restoring the values of the local variables
as specified by $\tau$ (see above).
We denote by $E$ termination (identifying $S;E$ with $S$).

\parbf{Deadline missed.}
Let  $(S', \tau)$ be some  pending process  in $q$ such that $\tau(\textbf{\jtt{deadline}}) < \sigma(\jtt{time})$.
Then
$$
\langle  (S,\sigma), q, \phi\rangle \rightarrow \langle p, q', \phi'\rangle 
$$
where $q'$ results from $q$ by removing $(S',\tau)$ and $\phi'(\tau(\jtt{myfuture})).\jtt{aborted} = \jtt{true}$
($\phi'(o)=\phi(o)$, for every $o\not= \tau(\jtt{myfuture})$).

\bigskip
% Given a
% method signature $m(\overline{T \: x})$, corresponding incoming messages 
% $o?m(\bar{v})$ and outgoing messages $o!m(\overline{v})$ specify  besides the callee $o$  and the actual values of the specified formal parameters $\bar{x}$
% the values of the variables  $\jtt{myfuture}$ and $\jtt{deadline}$.
% To model locally incoming and outgoing  messages we introduce the following
% labeled transitions.
A message $m(\tau)$ specifies for the method $m$ the  initial assignment $\tau$ of its  local variables
(i.e., the formal parameters and the variables \jtt{this}, \jtt{myfuture}, and \jtt{deadline}).
To model locally incoming and outgoing  messages we introduce the following
labeled transitions.

\parbf{Incoming message.}
Let the active process $p$ belong to the actor $\tau(\jtt{this})$ (i.e., 
 $\sigma(\jtt{this})=\tau(\jtt{this})$ for the assignment $\sigma$ in $p$):
$$
\langle p , q, \phi\rangle \xrightarrow[]{m(\tau)}  \langle p , \mathsf{insert}(q, m(\overline{v}, d)), \phi\rangle
$$
where   $\mathsf{insert}(q, m(\tau))$  defines the result of inserting the process  $(S,\tau)$, where $S$ denotes the body of method $m$,
in $q$, according to some application-specific policy (described below in Section \ref{sec:impl}).

\parbf{Outgoing message.} 
We model an outgoing message by:
$$
\langle (f = e_0 \; ! \; m(\bar{e}) \; \textbf{\jtt{deadline}}(e_1);S,\sigma)  , q, \phi\rangle 
\xrightarrow[]{m(\tau)} 
\langle (S,\sigma[f\mapsto o])  , q, \phi' \rangle 
$$
where 
\begin{itemize}\itemsep5pt
\item $\phi'$ results from $\phi$ by extending its domain  with a new future object $o$
such that $\phi'(o).\textsf{val}=\perp$\footnote{$\perp$ stands for "uninitialized"}and $\phi'(o).\textsf{aborted}=\jtt{false}$,
 \item $\tau(\jtt{this})=\textsf{val}(e_0)(\sigma)$,
 \item $\tau(x)= \textsf{val}(e)(\sigma)$, for every formal parameter $x$ and corresponding actual parameter $e$, 
 \item $\tau(\jtt{deadline})=\sigma(\mathsf{time})+ \textsf{val}(e_1)(\sigma)$,
  \item  $\tau(\jtt{myfuture})=o$.
\end{itemize}

\subsection{Global transition system}
A (global) system configuration $S$ is a pair $(\Sigma, \phi)$ consisting of a set $\Sigma$
of actor states and a global heap $\phi$ which stores the created future objects.
We denote actor states by $s$, $s'$, $s''$, etc.

\parbf{Local computation step.} 
The interleaving of local computation steps of the individual actors is
modeled by the rule:
$$
\frac{(s, \phi) \rightarrow (s', \phi')}
{(\{s\}\cup\Sigma, \phi)\rightarrow (\{s'\}\cup\Sigma, \phi')}
$$

\parbf{Communication.} 
Matching a  message sent by one actor with its  reception by  the specified callee
is described by the rule:
$$
\frac
{(s_1, \phi) \xrightarrow[]{m(\tau)} (s'_1, \phi')\quad (s_2, \phi) \xrightarrow[]{m(\tau)} (s'_2, \phi) }
{(\{s_1,s_2\}\cup\Sigma, \phi) \rightarrow (\{s'_1 , s'_2 \} \cup \Sigma, \phi')}
$$
Note that only an outgoing message affects the shared heap $\phi$ of futures.

\parbf{Progress of Time.}
The following transition uniformly updates the local clocks (represented by the instance variable \jtt{time})
of the actors.
$$
(\Sigma,\phi)\rightarrow (\Sigma',\phi)
$$
where
$$
\Sigma'=\{\langle (S,\sigma'), q, \phi\rangle\mid
\langle (S,\sigma), q, \phi\rangle\in \Sigma,\; \sigma'=\sigma [\jtt{time}\mapsto \sigma(\jtt{time})+\delta]\}
$$
for some positive $\delta$.

% \emph{Time transition.}
% $$
% \langle\mathcal{O}, F, \mathcal{T}\rangle \rightarrow \langle\mathcal{O}, F, \mathcal{T} + \Delta\rangle
% $$
% 
% \emph{Updating a future.}
% 
% $$
% \frac
% {\langle\mathcal{O}, F, \mathcal{T}\rangle \rightarrow \langle\mathcal{O}', F', \mathcal{T} + \Delta\rangle}
% {\langle\{O\} \cup \mathcal{O}, F,\mathcal{T}\rangle \rightarrow \langle\{O'\} \cup \mathcal{O}, F', \mathcal{T} + \Delta\rangle }
% $$

% \subsection{Previously written in case notations would be reused}
% 
% \Behrooz{Previous wrong version. May be removed later.}
% \emph{Receive a new message.} The operational semantics when an object with identifier $o$ receives a message to invoke method $m$ with deadline $d$.
% 
% \begin{align}
% \frac
% {\mathsf{in} = m(\overline{v}, d) + \dot{p} = \mathsf{new()}}
% {(o, (\dot{p}.\jtt{deadline} \mapsto \mathsf{time()} + d,s[\overline{v}], d), p, \mathsf{enq}(q, \dot{p}), \sigma)}
% \end{align}
% 
% in which the operation \textsf{new()} creates a new instance of a process for the concurrent object.
% Additionally, as we discssued in Equation \ref{eq:orderedq}, each concurrent object owns an ordered queue for the processes.
% The process queue is ordered based on the result of $\sigma$ function:
% 
% $$ \sigma(\dot{p}) = \dot{p}.\jtt{deadline} $$
% 
% Thus, the operational semantics of \textsf{enq} ensures to find a proper index place $i$ for the new process $\dot{p}$ such that:
% 
% $$
% \forall k < i \::\: q_k.\jtt{deadline} < \dot{p}.\jtt{deadline} \wedge \forall k > i \::\: q_k.\jtt{deadline} > \dot{p}.\jtt{deadline}
% $$
% 
% And, then, $q_i = \dot{p}$ completes the enqueuing action. Since, $q$ is an \emph{ascending} ordered queue, the required action to deqeue the 
% ``next'' process is simply to remove and return the \emph{head} of the queue; namely $q_0$.
% 
% \emph{Deadline Miss Detection.} A deadline miss is detected by a separate watchdog thread of the active concurrent object.
% The watchdog thread observes all the processes but the current process of the object to allow it complete its execution:
% 
% \begin{align}
% \frac
% {\forall p \neq \overline{p} \::\: (o,s[f_p \mapsto \jtt{null}],\overline{p},q,\sigma) + p.\jtt{deadline} < \mathsf{time()}}
% {(o,s[f_p.\jtt{aborted} \mapsto \jtt{true}], \overline{p}, q - p, \sigma)}
% \end{align}
% 
% Consequently, when the future of a missed-deadline process $p$ is accessed, it causes the application to generate and throw an exception:
% 
% \begin{align}
% \frac
% {(o,(f.\jtt{get()},s[f.\jtt{aborted} = \jtt{true}],d),q,\sigma)}
% {(o,(x \mapsto \jtt{new Exception},s[f \mapsto \jtt{null}],d),q,\sigma)}
% \end{align}

\section{Implementation} \label{sec:impl}
We base our implementation on Java's concurrent package: \jtt{java.util.concurrent}. 
The implementation consists of the following major components:
\begin{enumerate}
 \item An extensible language API that owns the core abstractions, architecture, and implementation. 
For instance, the programmer may extend the concept of a scheduler to take full control of how, i.e., in what order, the processes of the individual actors are queued
(and as such scheduled for execution).
We illustrate the scheduler extensibility with an example in the case study below.
 \item Language Compiler that translates the modeling-level programs into Java source.
 We use ANTLR \cite{antlr} parser generator framework to compile modeling-level programs to actual implementation-level source code of Java.
 \item The language is seamlessly integrated with Java.
 At the time of programming, language abstractions such as data types and third-party libraries from either Crisp or Java are equally usable by the programmer.
\end{enumerate}

We next discuss the underlying deployment of actors and the implementation
of real-time  processes with deadlines.

\parbf{Deploying actors onto JVM threads.}
%As a major improvement to our previous work \cite{crisp-sac}, we changed the model of deployment of actors onto Java threads for simplicity, performance, and better reasoning.
In the implementation, each actor owns a main thread of execution, that is,
the implementation does \emph{not} allocate \emph{one} thread per process because
threads are  \emph{costly} resources and
allocating to each process one thread in general leads to a poor performance:
there can be an arbitrary number of actors in the application and each may receive numerous messages which thus give rise to  a number of threads that  goes beyond the  limits of memory and resources.
Additionally, when processes go into pending mode, their correspondent thread may be reused for other processes.
Thus, for better performance and optimization of resource utilization, the implementation assigns a single thread for all processes inside each actor.

Consequently, at any moment in time, there is only one process that is executed inside each actor.
On the other hand, the actors share a thread which is used
for the execution of a  watchdog for the deadlines of the queued processes (described below) because allocation of such a   thread to each actor  in general slows down the performance.
%Actors take turn in using the shared thread to process and detect possible deadlines on this thread.
Further  this sharing  allows the implementation to decide, based on the underlying resources and hardware, to optimize the allocation of the watchdog thread to actors.
For instance, as long as the resources on the underlying hardware are abundant, the implementation decides to share as less as possible the watchdog thread.
This gives each actor a better opportunity with higher precision to detect missed deadlines.

%\parbf{Implementation of $(S, \tau)$ and $(S, \sigma)$.} 
%As discussed, when an actor receives an incoming message, a process is created for the messages.
%Within the process creation, the process allocates a data structure that behaves as a repository for the process.
%When the process is pending, the process has only access to the local variables and actual parameters of the message.
%When the process is executing, the access is extended to the level of actor's instance variables.
%The access level is managed by Crisp's core architecture when transferring the process from pending state to executing state (and the other way).
%At either state, the process has access to special convenient instances/methods including \jtt{myActor}, \jtt{myFuture}, \jtt{myDeadline};
%respectively, they denote the owning actor for the process, the future that will be the result of the process, and the deadline specified for the process.

\parbf{Implementation of processes with deadlines.}
A process itself is represented in the implementation by a data structure
which encapsulates the values of its local variables and the method to be executed.
Given a relative deadline $d$ as specified by a call
we compute at run-time its absolute deadline (i.e. the expected starting time of the process) by
$$
 \jtt{TimeUnit.toMillis}(d) + \jtt{System.currentTimeMillis()}
$$
which is a \emph{soft} real-time requirement.
%As shown in the above equation, the received deadline is expected to express a \emph{relative} amount of time.
%Each actor maintains a queue of processes to be executed along with the reference to the current executing process.
As in the operational semantics, in the real-time implementation
always the head of the process queue  is scheduled for execution.
This allows the implementation of a \emph{default}
earliest deadline first (EDF) scheduling policy
by  maintaining a queue ordered by  the above absolute time values for the deadlines.

The important consequence of our non-preemptive mode of execution 
for the implementation is the resulting simplicity of thread management
because preemption requires additional thread interrupts that facilitates the abortion of a process in the middle of execution.
% 
As stated above,  a single  thread in the implementation detects if a process has missed its deadline.
This task  runs periodically and to the end of all actors' life span.
%The detection is performed for all processes in the actor's queue (that does not include the current process).
To check for a missed deadline it suffices to simply check  for a process 
that the  above absolute time value of its deadline is \emph{smaller} than
\jtt{System.currentTimeMillis()}.
When a process misses its deadline,  the actions as specified by the corresponding transition of the operational semantics are subsequently performed.
% 
The language API provides  extension points which allow for each actor
the definition of a customized watchdog process and scheduling policy
(i.e., policy for enqueuing processes).
The customized  watchdog processes are still  executed by a single thread.

%In line with our previous work \cite{crisp-sac}, by default, application-level scheduling is realized and implemented using the \jtt{deadline} property of processes.
%In other words, the programmer is able to use asynchronous messages that are extended by deadlines at a higher level to specify application-level scheduling expectations.
%Nonetheless, the programmer still has the flexibility to extend the scheduling API if needed. 
%If the scheduling API is extended by the programmer, they are expected to provide an implementation that is fair in scheduling.
%By fairness, it is expected that no process should enter an unbounded waiting (starvation) state.
%Additionally, the order of processes needs to be determined by the programmer's custom scheduler that can be application specific.

\parbf{Fredhopper case study.}
As introduced in Section \ref{sec:frh}, we extract a closed-world simplified version from Fredhopper Controller.
We apply the approach discussed in this paper to use deadlines for asynchronous messages.
% Controller operates in a high-demand distributed setting as in Figure \ref{fig:ctl:deploy}.

Listing \ref{lst:frh:polling} and \ref{lst:frh:deadlines} present the difference in the previous Controller and the approach in Crisp.
The left code snippet shows the Controller that uses polling to retrieve data processing results.
The right code snippet shows the one that uses messages with deadlines.
\lstset{language=Creol,escapeinside={(*}{*)}} 
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\vspace{-25pt}
\begin{lstlisting}[label=lst:frh:polling, caption=With polling]
class DataProcessor begin
	op process(d: Data) ==
		var p := allocDataProcessor(d)
		p ! process (d)
		do {
			s := p ! getStatus (d)
			if (s <> nil)
				var r := p ! getResults(d)
				publishResult(r)
			wait(TimeUnit.toSecond(1))
		} while (true)
end
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\vspace{-25pt}
\begin{lstlisting}[label=lst:frh:deadlines, caption=With deadlines]
class DataProcessor begin
	op process(d: Data) == 
		var p := allocDataProcessor(d)
		var D := estimateDeadline(d)
		var f := 
			p ! process (d) deadline (D)
		try {
			publishResult(f.get())
		} catch (Exception x) {
			if (f.isAborted)
				notifyFailure(d)
		}
end 
\end{lstlisting}
\end{minipage}
\vspace{-10pt}
\end{center}

% We simplify details in the listing by higher level abstractions using straightforward naming.
% \jtt{Data} is an abstraction for customer data that carries all information required to initiate a data job request on the data processing service.
% Since Controller manages a distributed setting, we need to have \jtt{allocDataProcessor} to find and allocate the proper data processing service node to the current received data.
% Each data processing service understand a service message \jtt{process} that receives and processes the data job.

% In the left snippet, \jtt{getStatus()} and \jtt{getResults()} are operations to fetch the status of a job and publish its results.
% In the right snippet, we need to determine the deadline based on each \emph{customer} job history and size of data; 
% so, we introduce an operation to calculate an estimate of deadline by \jtt{estimateDeadline()}.

When the approach in Crisp in the right snippet is applied to Controller, it is guaranteed that all data job requests are terminated in a \emph{finite} amount of time.
Therefore, there cannot be complains about never receiving a response for a specific data job request.
Many of Fredhopper's customers rely on data jobs to eventually deliver an e-commerce service to their end users.
Thus, to provide a guarantee to them that their job result is always published to their environment is critical to them.
As shown in the code snippet, if the data job request is failed or aborted based on a deadline miss, 
the customer is still eventually informed about the situation and may further decide about it.
However, in the previous version, the customer may never be able to react to a data job request because its results are never published.
% \begin{wrapfigure}{r}{0.35\textwidth}
% \vspace{-35pt}
% \begin{center}
% \includegraphics[scale=0.35]{frh-ctl}
% \end{center}
% \vspace{-20pt}
% \caption{Fredhopper's Controller distribution model}
% \label{fig:ctl:deploy}
% \vspace{-15pt}
% \end{wrapfigure}

In comparison to the Controller using polling, there is a way to express timeouts for future values.
However, it does not provide language constructs to specify a deadline for a message that is sent to data processing service.
A deadline may be simulated using a combination of timeout and periodic polling approaches (Listing \ref{lst:frh:polling}).
Though, this approach cannot guarantee eventual termination in all cases; as discussed before that Step 4 in Figure \ref{fig:frh-dm} may never complete.
Controller is required to meet certain customer expectations based on an SLA.
Thus, Controller needs to take advantage of a language/library solution that can provide a higher level of abstraction for real-time scheduling of concurrent messages.
When messages in Crisp carry a deadline specification, Controller is able to guarantee that it can provide a response to the customer.
This termination guarantee is crucial to the business of the customer.

Additionally, on the data processing service node, the new implementation takes advantage of the extensibility of schedulers in Crisp.
As discussed above, the default scheduling policy used for each actor is EDF based on the deadlines carried by incoming messages to the actor.
However, this behavior may be extended and replaced by a custom implementation from the programmer.
In this case study, the priority of processes may differ if they the job request comes from specific customer; i.e.
apart from deadlines, some customers have priority over others because they require a more real-time action on their job requests while others run a more relaxed business model.
To model and implement this custom behavior, a custom scheduler is developed for the data processing node.
\lstset{language=Creol,escapeinside={(*}{*)}} 
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\vspace{-30pt}
\begin{lstlisting}[label=lst:frh:dm, caption=Data Processor class]
class DataProcessor begin
	var scheduler := new DataScheduler()
	op process(d: Data) ==
		// do process
end
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\vspace{-30pt}
\begin{lstlisting}[label=lst:frh:sched, caption=Custom scheduler]
class DataScheduler extends DefaultSchedulingManager {
	boolean isPrior(Process p1, Process p2) {
		if (p1.getCustomer().equals("A")) {
			return true;
		}
		return super.isPrior(p1, p2);
	}
}
\end{lstlisting}
\end{minipage}
\vspace{-15pt}
\end{center}
In the above listings, Listing \ref{lst:frh:sched} defines a custom scheduler that determines the priority of two processes with custom logic for specific customer.
To use the custom scheduler, the only requirement is that the class \jtt{DataProcessor} defines a specific class variable called \jtt{scheduler} in Listing \ref{lst:frh:dm}.
The \textsl{custom scheduler} is picked up by Crisp core architecture and is used to schedule the queued processes.
% In the listing, \jtt{DefaultSchedulingManager} denotes the default EDF scheduler that is implemented in Crisp.
Thus, all processes from customer \jtt{A} have priority over processes from other customers no matter what their deadlines are.

We use Controller's logs for the period of February and March 2013 to examine the evaluation of Crisp approach.
We define \emph{customer satisfaction} as a property that represents the effectiveness of futures with deadline.
\begin{wraptable}{l}{0.25\textwidth}
\vspace{-15pt}
\begin{tabular}{c c}\thickhline
$s_1$ & $s_2$ \\ \thickhline
$88.71\%$ & $94.57\%$ \\ \hline
\end{tabular}
\caption{Evaluation Results}
\label{tbl:eval}
\vspace{-15pt}
\end{wraptable}
For a customer $c$, the satisfaction can be denoted by $s = \frac{r^F_c}{r_c}$; in which $r^F_c$ is the number of finished data processing jobs and $r_c$ is the total number of requested data processing jobs from customer $c$. 
We extracted statistics for completed and never-ended data processing jobs from Controller logs ($s_1$).
We replayed the logs with Crisp approach and measured the same property ($s_2$).
We measured the same property for 180 customers that Fredhopper manages on the cloud.
In this evaluation, a total number of about 25000 data processing requests were included.
The results show $6\%$ improvement in Table \ref{tbl:eval} (that amounts to around 1600 better data processing requests). 
Because of data issues or wrong parameters in the data processing requests, there are requests that still fail or never end and should be handled by a human resource.

You may find more information including documentation and source code of Crisp at \surl{http://nobeh.github.com/crisp}.

\section{Related Work} \label{sec:relwork}
The programming language presented in this paper is a real-time  extension of the  language introduced in  \cite{crisp-sac}. This new extension features
\begin{itemize}
 \item integration of  asynchronous messages with  deadlines and futures with timeouts;
 \item  a general mechanism for handling exceptions raised by missed deadlines;
  \item high-level specification of application-level scheduling policies; and
  \item a formal operational semantics.
\end{itemize}
To the best of our knowledge the resulting language is the first implemented 
real-time actor-based programming language which formally integrates the above features.

In several works, e.g,  \cite{ACIHHS11} and \cite{Nielsen96r}, asynchronous messages in actor-based languages are extended with deadlines.
However these languages do not feature futures with timeouts, a general mechanism for handling exceptions raised by missed deadlines or
support the specification of application-level scheduling policies.
Futures and fault handling are considered in the ABS language \cite{abs:2012}.
This work describes recovery mechanisms  for  failed get operations on a future.
However,  the language does not support the specification of real-time requirements,
i.e., no  deadlines for asynchronous messages are  considered and no timeouts on futures.
Further, when a  get operation on a future fails,  \cite{abs:2012} does  not provide any context or information about the exception or the cause for the failure.
Alternatively, \cite{abs:2012} describes  a way to ``compensate'' for a failed get operation on future.
In  \cite{einar:rt:sched:co}, a real-time extension of  ABS with
scheduling policies to model distributed systems is introduced. 
In contrast to Crisp,  Real-Time ABS is an executable  \emph{modeling} language which supports the explicit specification
of the progress of time by means of duration statements for the \emph{analysis} of real-time requirements.
The language does not support however asynchronous messages with  deadlines and futures with timeouts.

%The concurrency model used in this paper is derived from the Actor model enriched by synchronization mechanisms and coupled with strong typing. 
%The Actor model \cite{actors:agha} provides  a suitable formal basis  for multi-core and distributed programming, 
%as objects (actors) are inherently concurrent and autonomous
%entities with a single thread of execution which makes them a natural fit for
%distributed deployment \cite{actor_frameworks_jvm:agha}. 
Two successful examples of actor-based programming languages are Scala and Erlang. 
Scala \cite{scala:actors:ordersky,coord:ordersky} is a hybrid object-oriented and functional programming language inspired by Java. 
% The most important concept introduced in  is that Scala Actors unify
% \textit{thread-based} and \textit{event-based} programming model to fill the gap
% for concurrency programming. 
Through the event-based model, Scala also provides the notion of continuations. Scala further provides  mechanisms for
scheduling of tasks similar to those provided by concurrent Java:
it does not provide a direct and
customizable platform to manage and schedule  messages
received by an individual actor.
Additionally, Akka \cite{akka:homepage} 
extends Scala's actor programming model and as such provides a direct integration with
both Java and Scala. 
% Akka also applies separation of concerns in different 
% aspects of actor programming; it tries to decouple actor model in separate
% including message storage, message dispatching mechanism and message execution.
% It also provides a basic level of message priorities using a simple priority
% generator service.
%To the best of our knowledge neither Scala nor Akka provide a formal operational semantics of the programming language constructs 
%they propose to address the challenges of 
%concurrency, application-level scheduling and real-time requirements.
Erlang \cite{erlang:armstrong} is a dynamically typed functional language that
was developed at Ericsson Computer Science Laboratory with telecommunication
purposes \cite{actors_highly:Correa}. Recent developments in the deployment of
Erlang support the assignment of a scheduler to each processor
\cite{erlang_scheduling} (instead of one global scheduler for the entire
application) but it does not, for example, support application-level scheduling policies.
In general,  none these  languages 
provide  a formally defined real-time extension which integrates the above features.
% This is a crucial improvement in Erlang, because the massive
% number of light-weight processes in the asynchronous setting of Erlang turns
% scheduling into a serious bottleneck. However, the scheduling policies are not
% yet controllable by the application.

There are well-known efforts in Java to bring in the
functionality of asynchronous message passing onto multicore including Killim
\cite{kilim:Srinivasan:Mycroft}, Jetlang \cite{jetlang}, ActorFoundry
\cite{actor_frameworks_jvm:agha}, and SALSA \cite{salsa:agha}. In
\cite{actor_frameworks_jvm:agha}, the authors present a comparative analysis of
actor-based frameworks for JVM platform. 
Most of these frameworks support futures with timeouts but do not provide asynchronous messages with deadlines, or
a general mechanism for handling exceptions raised by missed deadlines.
Further, pertaining to the domain of
priority scheduling of asynchronous messages, these efforts in general  provide a predetermined
approach or a limited control over message priority scheduling.
As another example, in  \cite{maia_rtsj_11} the use of Java Fork/Join is described  to optimize  mulicore applications.  This work is also based on a \textit{fixed priority} model. 
Additionally, from embedded hardware-software research domain, Ptolemy \cite{ptolemy:lee,actor:lee} is an actor-oriented open architecture and platform that is used to design, model and simulate embedded software. Their approach is hardware software co-design. It provides a platform framework along with a set of tools.

In general, existing high-level programming languages provide the programmer with little real-time control over
scheduling. The state of the art allows specifying priorities for threads or
processes that are  used by the operating system, e.g.,
Real-Time Specification for Java (RTSJ \cite{jsr1,jsr282}) and Erlang. 
Specifically in RTSJ, \cite{zerzel_rtsj} extensively introduces and discusses a framework
for application-level scheduling in RTSJ. It presents a flexible framework to allow scheduling
policies to be used in RTSJ. 
However, \cite{zerzel_rtsj}  addresses the problem mainly in the context of  the standard  multithreading approach to concurrency which in general does not provide
the most suitable approach to distributed applications.
%Thus, all the interesting discussion and extensive features
%relate to the level of \textit{threads} which is not the center of focus in the actor programming model.
In contrast,  in this paper
we have shown that an actor-based  programming
language provides a suitable formal basis for a fully integrated real-time control
in distributed applications.

% The current work contrasts however in the sense that it allows a \textit{fine-grain} and 
% \textit{application-level} scheduling feature in the level of active object with better \textit{flexibility}
% as in the scheduler function.

% Finally, in our previous work, 
% \cite{jaghoori:dating} deals with a real-time extension of a \emph{modeling} language
% based on active objects. In this paper we describe the actual deployment of active objects
% in terms of an implementation of an extension of Java with active objects, using Scala and Akka.
% \cite{crisp-sac} only deals with application-level scheduling policies using priorities.
% In this paper we consider real-time information based on an extension of active objects
% with deadlines and time-outs.

\section{Conclusion and future work} \label{sec:conclusion}
In this paper, we presented both a formal semantics and  an implementation of a real-time actor-based programming language.
We presented how asynchronous messages with deadline can be used to control application-level scheduling with higher abstractions.
We illustrated the language usage with a real-world case study from SDL Fredhopper along the discussion for the implementation.
Currently we are investigating  further optimization of  the implementation of  Crisp and   the formal verification of real-time properties of  Crisp applications using schedulability analysis \cite{sched:elena}.
% Additionally, Crisp targets to support integration with languages that support higher-order abstractions (e.g. Scala).

% \bibliographystyle{plain}
% \bibliography{references}

### 2015-10-20 p02-ch01-2014/j8lambda.tex
Initial Commit

\begin{figure}[t]
\begin{align}
L ::=& \; L_P \; \to \; \{ \; L_B \; \} \\
L_P ::=& \; I \; | \; ( \: P_{\mathsf{list}}^? \: ) \; | \; ( \: P_{\mathsf{list}}^i \: ) \\
P_{\mathsf{list}}^i ::=& \; I \; | \; P_{\mathsf{list}}^i \; , \; I \\
L_B ::=& \; E \; | \; K \\
P_{\mathsf{list}} ::=& \; P_{\mathsf{last}} \; | \; P \; , \; P_{\mathsf{last}} \\
P ::=& \; P_{\mathsf{single}} \; | \; P \; , \; P_{\mathsf{single}} \\
P_{\mathsf{single}} ::=& \; F^? \; T \; I \\
P_{\mathsf{last}} ::=& \; F^? \; T... \; I \; | \; P_{\mathsf{single}} \\
M ::=& \; [I.]^? \; I \; ( \; A_{\mathsf{list}}^? \; )  \\
A_{\mathsf{list}} ::=& \; E \; | \; A_{\mathsf{list}} \; , \; E \\
M_{\mathsf{ref}} ::=& \; (C \; | \; I) \; :: \; I \; | \\
                 ::=& \; C \; :: \; \bfjtt{new}
\end{align} 
\caption{
In the above grammar, $L$ is Lambda Expression block in Java; 
$L_P$ is Lambda's parameters and $L_B$ is Lambda's body; 
$I$ is a Java identifier;
$P_{\mathsf{list}}$ is Java's formal parameter \emph{list};
$P_{\mathsf{list}}^i$ is inferred formal parameter \emph{list};
$E$ is a Java expression and $L$ is \emph{also} an expression;
$K$ is Java code block wrapped in \jtt{\{\}};
$P_{\mathsf{last}}$ is the ``last'' formal parameter in which $T...$ refers to Java's \emph{varargs} feature;
$P_{\mathsf{single}}$ is a Java formal parameter;
$F$ is standard Java variable modifiers such as \jtt{final} or \jtt{static};
$T$ is a Java type;
$M$ denotes a method invocation in Java;
$A_{\mathsf{list}}$ is the optional argument list;
$M_{\mathsf{ref}}$ denotes a method reference in Java 8;
$C$ is (possibly) a fully qualified Java class name that introduces the method to be referred to.
}
\label{fig:java8:syntax}
\end{figure}

### 2015-10-20 p02-ch01-2014/paper.tex
Initial Commit

% -*- root: ../thesis.tex -*-

\chapter[Programming with actors in Java 8]{Programming with actors in Java~8%
\footnote{This paper is funded by the EU project FP7-610582 ENVISAGE: Engineering Virtualized Services, \surl{http://www.envisage-project.eu}.}
}
% 
\label{ch:p02:ch01}
% 
\chapterauthor{Behrooz Nobakht, Frank S. de Boer}

\section*{Abstract}
There exist numerous languages and frameworks that support an implementation of a variety of actor-based programming models in Java using concurrency utilities and threads.
% 
Java~8 is released with fundamental new features: lambda expressions and further dynamic invocation support.
%
We show in this paper that such features in Java~8 allow for a high-level actor-based methodology for programming distributed systems which supports the programming to interfaces discipline.
%
The embedding of our actor-based Java API is shallow in the sense that it abstracts from the actual thread-based deployment models. 
%
We further discuss different concurrent execution and thread-based deployment models and an extension of the API for its actual parallel 
and distributed implementation.
% 
We present briefly the results of a set of experiments which provide evidence of  the potential impact of lambda expressions in Java~8  regarding the adoption of the actor concurrency model in large-scale distributed applications.

% \keywords{Actor model, Concurrency, Asynchronous Message, Java, Lambda Expression}

\section{Introduction}
\label{sec:introduction}
Java is beyond doubt one of the mainstream object oriented programming languages 
that supports a \emph{programming to interfaces} discipline \cite{oop:patterns,oop:survey}.
Through the years, Java has evolved from a mere programming language to a huge platform to drive and envision standards for mission-critical business applications.
Moreover, the Java language itself has evolved in these years to support its community with new language features and standards.
One of the noticeable domains of focus in the past decade has been distribution and concurrency in research and application.
This has led to valuable research results and numerous libraries and frameworks with an attempt to provide distribution and concurrency at the level of Java language.
However, it is widely recognized that the thread-based model of concurrency in Java that is a well-known approach is not appropriate for realizing distributed systems because of its inherent synchronous communication model.
On the other hand, the event-driven actor model of concurrency introduced by Hewitt \cite{Hewitt69} is 
a powerful concept for modeling distributed and concurrent
systems \cite{Agha90,Agha97}. 
Different extensions of actors are proposed in several domains and are claimed to be the most suitable model of computation for many  applications \cite{Hewitt07-commitment}. 
Examples of these domains include designing embedded systems \cite{LeeACtorEmbedded03,LeeLN09}, 
wireless sensor networks \cite{CheongSensor05}, multi-core programming \cite{KarmaniSA09} 
and delivering cloud services through SaaS or PaaS \cite{Chang:DAIS07}.
This model of concurrent computation forms the basis of the
programming languages Erlang \cite{Armstrong10Erlang} and Scala
\cite{haller09tcs} that have recently gained in popularity, in part
due to their support for scalable concurrency.  Moreover, based on the
Java language itself, there are numerous libraries  that
provide  an implementation of an actor-based programming model.

% The main problem addressed in this paper is that in general existing actor-based programming techniques use extension and inheritance over composition of interfaces; 
% i.e. they require explicit inheritance of objects to allow for message passing and handling through specific methods designed in the framework or technique.
% They allow for generic and efficient invocation of special methods in objects that handle messages by inheritance.
% However, they \emph{bypass} the the basis of programming to interfaces and design-by-contract discipline~\cite{meyer:design}.
% This style of combination of object orientation and event-driven (or asynchronous messaging) in such frameworks and techniques makes actor-based programs developed difficult to reason about and formalize.
% This clearly hampers the promotion of actor-based programming in mainstream industry which is heavily based on object-oriented software engineering practice.
% 
% The main result of this paper is a Java API for programming distributed systems
% using asynchronous message passing and a corresponding actor-based programming methodology 
% which abstracts invocation from execution (e.g. thread-based deployment).
% The Java API fundamentally follows programming to interfaces discipline such that implementations only utilize a natural object-oriented design approach to program with actors.
% The API is desgined and implemented using Java~8 features.
% We further discuss the API architecture, its properties, and different concurrent execution models for the actual implementation.
% 
% Our main approach consists of the explicit realization of an actor in terms of its \emph{interface}.
% To this goal, a set of features including lambda expressions introduced in Java~8 are used in the  implmentation of asynchronous message passing and actor model.
% The corresponding actor-based methodology is formalized in terms of an executable modelling language
% which lends itself to formal analysis and verification, ABS~\cite{johnsen2012abs}.
% We provide a semantic translation of the concurrent modelling language ABS into Java~8.
% We present the operational semantics of such translation and how model extraction is a direct benefit of an implementation based on such API.

The main problem addressed in this paper is that in general existing actor-based programming techniques
are based on an explicit encoding of mechanisms at the application level for message passing and handling, 
and as such overwrite the general object-oriented approach of method look-ups that 
forms the basis of programming to interfaces and the design-by-contract discipline~\cite{meyer:design}.
The entanglement of event-driven (or asynchronous messaging) and object-oriented method look-up 
makes  actor-based programs developed using such techniques extremely difficult to  reason about and formalize.
This clearly hampers the promotion of actor-based programming in mainstream industry that heavily practices object-oriented software engineering.

The main result of this paper is a Java~8 API for programming distributed systems
using asynchronous message passing and a corresponding actor programming methodology 
which abstracts invocation from execution (e.g. thread-based deployment) and fully supports
 programming to interfaces discipline.
We  discuss the API architecture, its properties, and different concurrent execution models for the actual implementation.

Our main approach consists of the explicit description of an actor in terms of its \emph{interface}, 
the use of the recently introduced lambda expressions in Java~8 in the  implementation of asynchronous message passing,  
and the formalization of a corresponding high-level actor programming methodology in terms of an executable modeling language
which lends itself to formal analysis, ABS ~\cite{johnsen2012abs}.
% We provide a semantic translation of the concurrent modelling language ABS~\cite{johnsen2012abs} into Java~8 supporting actor programming.
% We present the operational semantics of such translation and how model extraction be a direct benefit of an implementation based on such API.

The paper continues as follows: 
in Section \ref{sec:relatedwork}, we briefly discuss a set of related works on actors and concurrent models especially on JVM platform. 
Section \ref{sec:example} presents an example that we use throughout the paper, we start to model the example using a library. 
Section \ref{sec:actor:programming} briefly introduces a concurrent modeling language and implements the example.
Section \ref{sec:j8:features} briefly discusses Java 8 features that this works uses for implementation.
Section \ref{sec:j8:actors:realization} presents how an actor model maps into programming in Java~8.
Section \ref{sec:impl:arch} discusses in detail the implementation architecture of the actor API.
Section \ref{sec:experiments} discusses how a number of benchmarks were performed for the implementation of the API and how they compare with current related works.
Section \ref{sec:conclusion} concludes the paper and discusses the future work.

\section{Related Work}
\label{sec:relatedwork}

There are numerous works of research and development in the domain of actor modeling and implementation in different languages.
We discuss a subset of the related work in the level of modeling and implementation with more focus on Java and JVM-based efforts in this section.

Erlang \cite{Armstrong10Erlang} is a programming language used to build massively scalable soft real-time systems with requirements on high availability. 
Some of its uses are in telecoms, banking, e-commerce, computer telephony and instant messaging. Erlang's runtime system has built-in support for concurrency, distribution and fault tolerance.
While threads require external library support in most languages, Erlang provides language-level features for creating and managing processes with the aim of simplifying concurrent programming. 
Though all concurrency is explicit in Erlang, processes communicate using message passing instead of shared variables, which removes the need for locks.
Elixir \cite{elixir} is a functional meta-programming aware language built on top of the Erlang VM. It is a dynamic language with flexible syntax with macros support that leverages Erlang's abilities to build concurrent, distributed, fault-tolerant applications with hot code upgrades.

Scala is a hybrid object-oriented and functional programming language inspired by Java. 
The most important concept introduced in \cite{haller09tcs} is that Scala actors unify
\textit{thread-based} and \textit{event-based} programming model to fill the gap for concurrency programming. 
Through the event-based model, Scala also provides the notion of continuations. 
Scala provides quite the same features of scheduling of tasks as in concurrent Java; 
i.e., it does not provide a direct and customizable platform to manage and schedule priorities on messages sent to other actors.
Akka \cite{haller2012integration} is a toolkit and runtime for building highly concurrent, distributed, and fault tolerant event-driven applications on the JVM based on actor model.
% Akka implements a hybrid of the actor model, fault tolerance through a
% ``let it crash'' mind set, and location transparency to support distributed environments.

Kilim \cite{srinivasan2008kilim} is a message-passing framework for Java that provides ultra-lightweight threads and facilities for fast, safe, zero-copy messaging between these threads.
It consists of a bytecode postprocessor (a ``weaver''), a run time library with buffered mailboxes (multi-producer, single consumer queues) and a user-level scheduler and a type system that puts certain constraints on pointer aliasing within messages to ensure interference-freedom between threads.
% 
The SALSA \cite{varela2007salsa,KarmaniSA09} programming language (Simple Actor Language System and Architecture) is an active object-oriented programming language that uses concurrency primitives beyond asynchronous message passing, including token-passing, join, and first-class continuations.
% 
% Clojure \cite{hickey2008clojure} is a dynamic programming language that targets the Java Virtual Machine. 
% It is designed to be a general-purpose language, combining the approachability and interactive development of a scripting language with an efficient and robust infrastructure for multithreaded programming. 
% Clojure is a compiled language - it compiles directly to JVM bytecode, yet remains completely dynamic.
% Clojure is influenced by Scheme \cite{scheme} which is a Lisp dialect.
% Scheme was introduced through ``Lambda Papers''.
 % \footnote{\url{http://en.wikipedia.org/wiki/Lambda\_Papers\#The\_Lambda_Papers}}.
% Neither Clojure nor scheme do directly implement the actor model, however, through functional programming of lambda calculus that they provide, they are suitable for concurrent applications built with lambda statements.

RxJava \cite{rxjava} by Netflix is an implementation of reactive extensions \cite{rx} from Microsoft.  
Reactive extensions try to provide a solution for composing asynchronous and event-based software using observable pattern and scheduling.  
An interesting direction of this library is that it uses reactive programming to avoid a phenomenon known as ``callback hell''; a situation that is a natural consequence of composing \jtt{Future} abstractions in Java specifically when they wait for one another.  
However, RxJava advocates the use of asynchronous functions that are triggered in response to the other functions.
% 
In the same direction, LMAX Disruptor \cite{lmax,lmax:mf} is a highly concurrent event processing framework that takes the approach of event-driven programming towards provision of concurrency and asynchronous event handling.
The system is built on the JVM platform and centers on a Business Logic Processor that can handle 6 million events per second on a single thread. 
The Business Logic Processor runs entirely in-memory using event sourcing. 
The Business Logic Processor is surrounded by Disruptors - a concurrency component that implements a network of queues that operate without needing locks.

\section{State of the Art: An example}
\label{sec:example}
In the following, we illustrate the state of the art in actor programming by means of a simple example using
the Akka~\cite{akka} library which features asynchronous messaging and which is used to program actors in both Scala and Java.
We want to model in Akka an ``asynchronous ping-pong match'' between two actors represented by the 
two interfaces \jtt{IPing} and \jtt{IPong} which are depicted in Listings \ref{lst:ping:interface} and \ref{lst:pong:interface}.
An asynchronous call by the actor implementing the \jtt{IPong} interface of the \jtt{ping} method of the actor
implementing the \jtt{IPing} interface should generate an asynchronous call of the \jtt{pong} method of the callee,
and vice versa.
We intentionally design \jtt{ping} and \jtt{pong} methods to take  arguments in order to demonstrate 
how method arguments may affect the use of an actor model in an object-oriented style.

\lstset{language=Java}
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Ping as an interface,label=lst:ping:interface]
public interface IPing {
  void ping(String msg);
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Pong as an interface,label=lst:pong:interface]
public interface IPong {
  void pong(String msg);
}
\end{lstlisting}
\end{minipage}
\end{center}

To model an actor in Akka by a class, say  \jtt{Ping},  with interface \jtt{IPing}, this class is required  \emph{both} to \emph{extend}
a given pre-defined  class \jtt{UntypedActor}  and \emph{implement} the interface \jtt{IPing}, 
as depicted in Listings \ref{lst:ping:akka} and \ref{lst:pong:akka}.
The class \jtt{UntypedActor} provides two Akka framework methods \jtt{tell} and \jtt{onReceive} 
which are used to enqueue and dequeue asynchronous messages.
An asynchronous call to, for example,  the method \jtt{ping} then can be modeled by passing a user-defined  encoding of this call,
in this case by prefixing the string argument with the string ``pinged'', to a (synchronous) call
of the \jtt{tell} method which results in enqueuing the message.
In case this message is dequeued
the implementation of the \jtt{onReceive} method as provided by the \jtt{Ping} class then calls the \jtt{ping} method.

\lstset{language=Java}
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Ping actor in Akka,label=lst:ping:akka]
public class Ping(ActorRef pong)
  extends UntypedActor 
  implements IPing {

  public void ping(String msg) {
    pong.tell("ponged," + msg)
  }

  public void onReceive(Object m) {
    if (!(m instanceof String)) {
      // Message not understood.
    } else 
    if (((String) m).startsWith("pinged") {
      // Explicit cast needed.
      ping((String) m);
    } 
   }
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Pong class in Akka,label=lst:pong:akka]
public class Pong 
  extends UntypedActor 
  implements IPong {

  public void pong(String msg) {
    sender().tell(
      "pinged," + msg); 
  }

  public void onReceive(Object m) {
    if (!(m instanceof String)) {
      // Message not understood.
    } else 
    if (m.startsWith("ponged") {
      // Explicit cast needed.
      ping((String) m);
    } 
   }
}

\end{lstlisting}
\end{minipage}
\end{center}

Access to the sender of the message in Akka is provided by \jtt{sender()}.
In the main method as described in Listing \ref{lst:main:akka} we show how the initialize and start the ping/pong match.
Note that a reference to the ``pong'' actor is passed to the ``ping'' actor.

\begin{wrapfigure}{r}{0.48\textwidth}
\vspace{-20pt}
\begin{lstlisting}[caption=main in Akka,label=lst:main:akka]
ActorSystem s = ActorSystem.create();
ActorRef pong = s.actorOf(Props.create(Pong.class));
ActorRef ping = s.actorOf(Props.create(Ping.class, pong));
ping.tell(""); // To get a Future
\end{lstlisting}
\vspace{-20pt}
\end{wrapfigure}

Further, both the \jtt{onReceive} methods are invoked by Akka \jtt{ActorSystem} itself. 
In general, Akka actors are of type  \jtt{ActorRef} which is an abstraction provided by Akka to allow actors send asynchronous messages to one another.
% 
An immediate consequence  of the above use of inheritance is that the class \jtt{Ping} is now exposing a public behavior 
that is \emph{not} specified by its \emph{interface}. 
Furthermore, a ``ping'' object refers to a ``pong'' object by the type \jtt{ActorRef} .
This means that the interface \jtt{IPong} is not directly visible to the ``ping'' actor.
Additionally, the implementation details of receiving a message should be 
``hand coded'' by the programmer into the  special method \jtt{onReceive} to define the  responses to the received messages.
In our case, this implementation consists of a decoding of the message  (using type-checking)
in order to \emph{look up} the method that subsequently should be invoked.
This fundamentally interferes with the general object-oriented mechanism for method look-up which forms the basis of the programming to interfaces discipline.
In the next section, we continue the same example and discuss an actor API for directly calling asynchronously methods
using the general object-oriented mechanism for method look-up.
Akka has recently released a new version that supports Java 8 features
\footnote{Documentation available at \surl{http://doc.akka.io/docs/akka/2.3.2/java/lambda-index-actors.html}}.
However, the new features can be categorized as syntax sugar on how incoming messages are filtered through object/class matchers to find the proper type.

\section{Actor Programming in Java}
\label{sec:actor:programming}

We first describe informally the actor programming model assumed in this paper. This model is based on the
Abstract Behavioral Specification language (ABS)  
introduced in \cite{johnsen2012abs}.
ABS uses asynchronous method calls, futures, interfaces for encapsulation, 
and cooperative scheduling of method invocations inside concurrent (active) objects. 
This feature combination results in a concurrent object-oriented model which is inherently compositional.
More specifically, actors in ABS  have an identity and behave as active
objects with encapsulated data and methods which represent their state
and behavior, respectively.  Actors are the units of concurrency: conceptually an actor has a dedicated processor. Actors can only send
asynchronous messages and have queues for receiving messages.  An
actor progresses by taking a message out of its queue and processing
it by executing its corresponding method.  A method is a piece of
sequential code that may send messages.
% 
\lstset{language=Java}
\begin{wrapfigure}{r}{0.48\textwidth}
\vspace{-20pt}
\begin{lstlisting}[caption=main in ABS,label=lst:main:ABS]
ABSIPong pong;
pong = new ABSPong;
ping = new ABSPing(pong);
ping ! ping("");
\end{lstlisting}
\vspace{-20pt}
\end{wrapfigure}
% 
Asynchronous method calls use futures as dynamically generated references  to return values.
The execution of a method can be (temporarily) suspended by release statements which give rise to a form
of  cooperative scheduling of method invocations inside concurrent (active) objects.
Release statements can be conditional (e.g., checking a future for the return value).
Listings \ref{lst:ping:abs}, \ref{lst:pong:abs} and \ref{lst:main:ABS} present an implementation of ping-pong example in ABS.
By means of the statement on line 6 of Listing \ref{lst:ping:abs}
a ``ping'' object  directly calls asynchronously the \jtt{pong} method  of its ``pong'' object, and vice versa.
Such a call is stored in the message queue and the called method is executed when the message is dequeued.
Note that variables in ABS are declared by interfaces.
In ABS, \jtt{Unit} is similar to \jtt{void} in Java.

\lstset{language=abs}
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Ping in ABS,label=lst:ping:abs]
interface ABSIPing {
  Unit ping(String msg);
}
class ABSPing(ABSIPong pong) implements ABSIPing {
  Unit ping(String msg) {
  pong ! pong("ponged," + msg);
  }
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Pong in ABS,label=lst:pong:abs]
interface ABSIPong {
  Unit pong(String msg);
}
class ABSPong implements ABSIPong {
  Unit pong(String msg) {
    sender ! ping( "pinged," + msg);
  }
}
\end{lstlisting}
\end{minipage}
\end{center}
\lstset{language=Java}

\section{Java 8 Features}
\label{sec:j8:features}

In the next section, we describe how ABS actors are implemented in Java~8 as API.
In this section we provide an overview of the features in Java~8 that facilitate an efficient, expressive, and precise implementation of an actor model
in ABS.

\textsl{Java Defender Methods}
Java \emph{defender} methods (JSR 335~\cite{jsr335}) use the new keyword \bfjtt{default}. 
Defender methods are declared for \bfjtt{interface}s in Java. 
In contrast to the other methods of an interface, a default method is not an abstract method but  must have an implementation.
From the perspective of a client of the interface, defender methods are no different from ordinary interface methods.
From the perspective of a hierarchy descendant, an implementing class can optionally \textit{override} a default method and change the behavior.
It is left as a decision to any class implementing the interface whether or not to override the default implementation.
For instance, in Java 8 \jtt{java.util.Comparator} provides a default method \jtt{reversed()} that creates a reversed-order comparator of the original one.
Such default method eliminates the need for any implementing class to provide such behavior by inheritance.

\textsl{Java Functional Interfaces}
Functional interfaces and lambda expressions (JSR 335~\cite{jsr335}) are fundamental changes in Java 8. 
A \bfjtt{@Functional\-Interface} is an annotation that can be used for interfaces in Java.
Conceptually, any class or interface is a functional interface if it consists of exactly one \emph{abstract} method.
A lambda expression in Java 8, is a runtime translation~\cite{jsr335:lambda:translation} of any type that is replaceable by a functional interface.
Many of Java's classic interfaces are functional interfaces from the perspective of Java 8 and can be turned into lambda expressions; 
e.g. \jtt{java.lang.Runnable} or \jtt{java.util.Comparator}.
For instance,

$$
\jtt{(s1, s2)} \to \jtt{return s1.compareTo(s2);}
$$

is a lambda expression that can be statically cast to an instance of a \jtt{Comparator<String>};
because it can be replaced with a functional interface that has a method with two strings and returning one integer.
Lambda expressions in Java 8 \emph{do not} have an intrinsic type.
Their type is bound to the context that they are used in but their type is always a functional interface.
For instance, the above definition of a lambda expression can be used as:

$$
\jtt{Comparator<String> cmp1 = (s1, s2)} \to \jtt{return s1.compareTo(s2);}
$$

in one context while in the other:

$$
\jtt{Function<String> cmp2 = (s1, s2)} \to \jtt{return s1.compareTo(s2);}
$$

given that \jtt{Function<T>} is defined as:

$$
\bfjtt{interface} \; \jtt{Function<T> \{\;
  int apply(T t1, T t2); \;\}}
$$

In the above examples, the same lambda expression is statically cast to a different matching functional interface based on the context.
This is a fundamental new feature in Java 8 that facilitates application of functional programming paradigm in an object-oriented language.

This work of research extensively uses this feature of Java~8.
Java~8 marks many of its own APIs as functional interfaces most important of which in this context are \jtt{java.lang.Runnable} and \jtt{java.util.concurrent.Callable}.
This means that a lambda expression can replace an instance of \jtt{Runnable} or \jtt{Callable} at runtime by JVM.
We will discuss later how we utilize this feature to allow us model an asynchronous message into an instance of a \jtt{Runnable} or \jtt{Callable} as a form of a lambda expression.
A lambda expression equivalent of a \jtt{Runnable} or a \jtt{Callable} can be treated as a queued message of an actor and executed.

\textsl{Java Dynamic Invocation}
Dynamic invocation and execution with method handles (JSR 292~\cite{jsr292:invokedyn}) enables JVM to support efficient and flexible execution of method invocations in the absence of static type information. 
JSR 292 introduces a new byte code instruction \jtt{invokedynamic} for JVM that is available as an API through \jtt{java.lang.invoke.MethodHandles}.
This API allows translation of lambda expression in Java 8 at runtime to be executed by JVM.
In Java 8, use of lambda expression are favored over anonymous inner classes mainly because of their performance issues~\cite{lambda:perf}.
The abstractions introduced in JSR 292 perform better that Java Reflection API using the new byte code instruction. 
Thus, lambda expressions are compiled and translated into method handle invocations rather reflective code or anonymous inner classes.
This feature of Java~8 is indirectly use in ABS API through the extensive use of lambda expressions.
Moreover, in terms of performance, it has been revealed that invoke dynamic is much better than using anonymous inner classes~\cite{lambda:perf}.

\section{Modeling actors in Java~8}
\label{sec:j8:actors:realization}

In this section, we discuss how we model ABS actors  using Java~8 features.
% We discuss how the message queue of actor is realized and how co-operative scheduling is achieved.
In this mapping, we demonstrate how new features of Java~8 are used.

\paragraph{The \jtt{Actor} Interface}
We introduce an interface to model actors  using Java~8 features discussed in Section~\ref{sec:j8:features}.
Implementing an interface in Java means that the object exposes public APIs specified by the interface that is considered the behavior of the object.
Interface implementation is opposed to inheritance extension in which the object is 
possibly forced to expose behavior that may not be part of its intended interface.  
Using an interface for an actor  allows an object to preserve its own interfaces, and second, 
it allows for multiple interfaces to be implemented and composed.

A Java API for the implementation of ABS models should have the following main three  features.
First, an object should be able to send asynchronously an arbitrary message in terms of a method invocation to a receiver actor object.
Second, sending a message can optionally generate a so-called future which is used to refer to the return value.
Third, an object during the processing of a message should be able to access the ``sender'' of a message such that it can reply to the message by another message.
All the above must co-exist with the 
fundamental requirement that for an object to act like an actor (in an object-oriented context)
should \emph{not} require a modification of its intended interface.

The \jtt{Actor} interface (Listings \ref{lst:j8:run} and \ref{lst:j8:enq})
provides a set of \bfjtt{default} methods, namely the \jtt{run} and \jtt{send} methods,  which the implementing classes do not need to re-implement.
This interface further  encapsulates a queue of messages that supports concurrent features of Java API
\footnote{Such API includes usage of different interfaces and classes in \jtt{java.util.concurrent} package \cite{jsr166}.
The concurrent Java API supports blocking and synchronization features in a high-level that is abstracted from the user.}.
We distinguish two types of messages:
messages that are  not expected to generate  any result and messages that are  expected to generate a result captured by a future value; i.e. 
an instance of \jtt{Future} in Java~ 8.
The first kind of messages are modeled as instances of \jtt{Runnable} and the second kind are modeled instances of \jtt{Callable}.
The default \jtt{run} method then takes a message from the queue, checks its type and executes the message correspondingly.
On the other hand, the default (overloaded) \jtt{send} method stores the sent message and creates a future
which is returned to the caller, in case of an instance of \jtt{Callable}.

\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Actor interface (1) ,label=lst:j8:run]
public interface Actor {
  public void run() {
    Object m = queue.take();

    if (m instanceof Runnable) {
      ((Runnable) m).run();
    } else

    if (m instanceof Callable) {
      ((Callable) m).call();
    } 
  }

  // continue to the right
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Actor interface (2),label=lst:j8:enq]

  public void send(Runnable m) {
    queue.offer(m);
  }

  public <T> Future<T> 
    send(Callable<T> m) {
      Future<T> f = 
        new FutureTask(m);
      queue.offer(f);
      return f;
  }
}
\end{lstlisting}
\end{minipage}
\end{center}

\paragraph{Modeling Asynchronous Messages} 
\label{par:asynchronous_messages}
We model an asynchronous call 
$$
\jtt{Future<V>} \; f = e_0\; !\; m(e_1,\ldots,e_n)
$$
to a method in ABS by the Java~8 code snippet of Listing \ref{lst:j8:call0}.
The final local variables $\jtt{u}_1$, $\ldots$, $\jtt{u}_n$ (of the caller) are used to store the values of
the Java~8 expressions $\jtt{e}_1$, $\ldots$, $\jtt{e}_n$ corresponding to the actual parameters  $e_1,\ldots,e_n$.
The types \jtt{T}$_i$, $i=1,\ldots,n$,  are the corresponding Java~8 types of  $e_i$, $i=1,\ldots,n$.
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape,caption=Async messages with futures,label=lst:j8:call0]
final T$_1$ u$_1$ = e$_1$;
. . . 
final T$_n$ u$_n$ = e$_n$;
Future<V> v = e$_0$.send(
  () $\to$ { return m(u$_1$,$\ldots$,u$_n$); }
);
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape,caption=Async messages w/o futures,label=lst:j8:call1]
final T$_1$ u$_1$ = e$_1$;
. . . 
final T$_n$ u$_n$ = e$_n$;
e$_0$.send(
  { () $\to$ m(u$_1$,$\ldots$,u$_n$); }
);
\end{lstlisting}
\end{minipage}
\end{center}

The lambda expression  which encloses the above method invocation  is an instance of the  functional interface; e.g. \jtt{Callable}.
Note that the generated object which  represents the  lambda expression will contain the local context of the caller of the method ``$m$'' (including the local variables storing the values of the expressions $e_1,\ldots,e_n$), which will be restored upon execution of the lambda expression. 
Listing \ref{lst:j8:call1} models an asynchronous call to a method without a return value.

As an example, Listings \ref{lst:ping} and \ref{lst:pong} present the running ping/pong example, using the above API.
The main program to use ping and pong implementation is presented in Listing~\ref{lst:main:api}.

\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape, caption=Ping as an Actor,label=lst:ping]
public class Ping(IPong pong) implements IPing, Actor {
 public void ping(String msg) {
  pong.send( () $\to$ { pong.("ponged," + msg) } );
  }
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape, caption=Pong as an Actor,label=lst:pong]
public class Pong implements IPong, Actor {
  public void pong(String msg) {
    sender().send(( ) $\to$ { ping.("pinged," + msg) } );
  }
}
\end{lstlisting}
\end{minipage}
\end{center}

% In Listing \ref{lst:ping}, the implementation uses a reference to another actor; namely \jtt{pong}.
% It sends a message to the actor, captures the result into a future value and waits for the result.
% The processing of the message in \jtt{pong} leads to the invocation of the method ``pong'' 
% using the provided argument and returns the result (Listing \ref{lst:pong}).
% 
As demonstrated in the above examples, the ``ping'' and ``pong'' objects \emph{preserve} 
their own \emph{interfaces}  contrary to the example depicted in Section \ref{sec:example} 
in which the objects \emph{extend} a specific ``universal actor abstraction'' to inherit methods and behaviors to become an actor.
Further,  messages are processed \emph{generically} by the \jtt{run} method described in Listing \ref{lst:j8:run}.
Although, in the first place, sending an asynchronous may look like to be able to change the recipient actor's state, this is not correct.
The variables that can be used in a lambda expression are \emph{effectively} final. 
In other words, in the context of a lambda expression, the recipient actor only provides a snapshot view of its state that cannot be changed.
This prevents abuse of lambda expressions to change the receiver's state.

% 
% To the third feature discussed above, when an actor is processing a message sent to it, can refer to the actor that has sent the message.
% In the above example, the pong actor sends the result of the method as a message to the ``sender'' of the message.
% In this scenario, the pong actor assumes that the sender of the message is always a ping actor; i.e. a ``ping'' message can be sent to the sender.

\paragraph{Modeling Cooperative Scheduling}
The ABS statement $\bfjtt{await}\; g$, where $g$ is a boolean guard, allows an active object to preempt the current method and schedule another one.
% We describe the  translation of await statements in the context of a general translation of statements in the ABS language.
We model cooperative scheduling by means of a call to the \jtt{await} method in Listing~\ref{lst:j8:await}.
Note that the preempted process is thus passed as an additional parameter and as such queued in case the guard is false,
otherwise it is executed.
Moreover, the generation of the continuation of the process is an optimization task for the code generation process to prevent code duplication.
\begin{center}
\begin{minipage}[t]{0.48\textwidth}
% \begin{lstlisting}[mathescape, caption=Java~8 conceptual await,label=lst:j8:await]
%  this.await($b$, $()\rightarrow \{S\}$)
%  // S is the continuation
% \end{lstlisting}
\begin{lstlisting}[mathescape,caption=main in ABS API,label=lst:main:api]
IPong pong = new Pong();
IPing ping = new Ping(pong);
ping.send(
  () -> ping.ping("")
);
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[mathescape,caption=Java~8 await implementation,label=lst:j8:await]
void await(final Boolean guard, 
           final Runnable cont) {
  if (!guard) { 
    this.send(() $\to$ 
      { this.await(guard, cont) })
  } else { cont.run() }
}
\end{lstlisting}
\end{minipage}
\end{center}

% where $\theta(g)$ denotes  the Java~8 realization of the boolean guard of the await statement
% and $\theta(S')$ denotes the Java~8 realization of the statement $S'$, following inductively the above scheme.

% \textsl{Actor API in Java~8} 
% \label{par:actor_api_in_java_8}

\section{Implementation Architecture}
\label{sec:impl:arch}

Figure \ref{fig:actors:arch} presents the general layered architecture of the actor API in Java 8.
It consists of  three layers: the  routing layer  which
forms the foundation for the support of distribution and location transparency~\cite{KarmaniSA09} of actors,
the queuing layer  which allows for different implementations of the message queues, and finally, the processing layer
which implements the actual execution of the messages.
Each layer  allows for further customization by means of plugins.
The implementation is available at \surl{https://github.com/CrispOSS/abs-api}.

% 
% The routing layer  forms the foundation for the support of distribution and location transparency~\cite{KarmaniSA09} of actors.
% From the perspective of the sender of a message, the location of the receiver is hidden and only referred to by a name or an address.
% It is the responsibility of the routing layer to find the appropriate receiver actor for the message.
% As a result of routing a message, the queue in which the message should be placed into is identified.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.6]{figs/abs-api-java8-2}  
\end{center}
\caption{Architecture of Actor API in Java~8}
\label{fig:actors:arch}
\end{figure}

We discuss the architecture from bottom layer to top.
The implementation of actor API preserves a faithful mapping of message processing in ABS modeling language.
An actor is an active object in the sense that it controls how the next message is executed and may release any resources to allow for co-operative scheduling.
Thus, the implementation is required to optimally utilize JVM threads.
Clearly, allocating a dedicated thread to each message or actor is not scalable.
Therefore, actors need to \emph{share} threads for message execution and yet be in full control of resources when required.
The implementation fundamentally separates \emph{invocation} from \emph{execution}.
An asynchronous message is a reference to a method invocation until it starts its execution.
This allows to minimize the allocation of threads to the messages and facilitates sharing threads for executing messages.
Java concurrent API~\cite{jsr166} provides different ways to deploy this separation of invocation from execution.
We take advantage of Java Method Handles~\cite{jsr292:invokedyn} to encapsulate invocations.
Further we utilize different forms of \jtt{ExecutorService} and \jtt{ForkJoinPool} to deploy concurrent invocations of messages in different actors.

In the next layer, the actor API allows for different implementations of a queue for an actor.
A dedicated queue for each actor simplifies the process of queuing messages for execution but consumes more resources.
However, a  shared queue for a set of actors allows for memory and storage optimization.
This latter approach of deployment, first, provides a way to utilize the computing power of multi-core; 
for instance, it allows to use work-stealing to maximize the usage of thread pools.
Second, it enables  application-level scheduling of messages.
The different implementations cater for a variety of plugins, like
one that releases computation as long as there is no item
in the queue and becomes active as soon as an item is placed into the queue; e.g. \jtt{java.util.concurrent.BlockingQueue}.
Further, different plugins can be injected to allow for scheduling of messages extended with deadlines and priorities
~\cite{Nobakht:sched}.

%Each layer can be \emph{distributed} independently of another layer in a transparent way.
%  For instance, not only the routing layer can provide distribution, the queue layer of the architecture may also be remote to take advantage of cluster storages for actor messages. 
%
%
%
%
%
%
%The process of sending an asynchronous message leads to the following steps according to the actor model in ABS at the level of implementation.
%
%
%A message has a receiver actor that needs to be identified. 
%The receiver in essence has no location limitation and as such can be local or remote.
%Thus, the actor API should manage a way to \emph{route} the message to the receiver.
%The routing layer of actor API forms the foundation for the support of distribution and location transparency~\cite{KarmaniSA09} of actors.
%From the perspective of the sender of a message, the location of the receiver is hidden and only referred by a name or an address.
%It is the responsibility of the routing layer to find the appropriate receiver actor for the message.
%As a result of routing a message, the queue in which the message should be placed into is identified.

%An actor owns a queue of messages.

% When a message is sent to an actor, an \emph{envelope} is created.
% The envelope encapsulates the sender, the receiver and the method invocation of the receiver with its actual parameters.
% An envelope is \emph{routed} to its receiver; there may exist different implementations of routers such as local or remote.
% Routing an envelopes consists of discovering the appropriate \emph{inbox} to which the envelope is posted.
% When an envelope is received in an inbox, it is delegated to an \emph{opener}.
% Note that each actor is ``related'' to an inbox;
% such relation can be \emph{own}ing individually or \emph{shar}ing an inbox with other actors.
% Openning an envelope is left as an implementation detail.
% For instance, a direct simple implementation is to \emph{synchronously} open and process the envelope.
% In an asynchronous implementation, openning an envelope actually \emph{queues} it to be executed. 
% A queue of envelopes may take advantage of different JVM execution services; e.g. direct usage of \jtt{Thread}s or utilizing different types of concurrent executor services.

%The API architecture described in the above has certain properties:
%\begin{itemize}
%  \item From the perspective of the user of the API, an actor is abstraction that facilitates sending asynchronous messages and working with future values according to the actor model. Any extension or custom implementation should guarantee the same semantics of asynchronous message passing.
%  \item Each layer of the actor API is \emph{pluggable}; i.e. specific implementations can be injected into the actor API context using standard Java service loading mechanisms.
%  \item Each layer can be \emph{distributed} independently of another layer in a transparent way.
%  For instance, not only the routing layer can provide distribution, the queue layer of the architecture may also be remote to take advantage of cluster storages for actor messages. 
%  \item The \emph{invocation} is separated from \emph{execution}. 
%  An asynchronous message encapsulates and carries a method invocation; however, its execution and the properties of its execution such as priority in terms of time or order is left to a different layer. 
%  Such separation of concerns maximizes the concurrency level and utilization of computing resources in a lock-free style.
%\end{itemize}

% We discuss the distribution of actors in this architecture in the next section.
% The implementation is available at \surl{https://github.com/CrispOSS/abs-api}.
% 
% \section{Distribution of Actors}
% \label{sec:distributtion}

We discuss next the distribution of actors in this architecture.
In the architecture presented  in Figure~\ref{fig:actors:arch}, each layer can be \emph{distributed} independently of another layer in a transparent way.
Not only the routing layer can provide distribution, the queue layer of the architecture may also be remote to take advantage of cluster storage for actor messages. 
A remote routing layer can provide access to actors transparently through standard naming or addresses.
% The implementation of remote actors is left as an the extension or plugin that is injected into the system.
% For instance, an extension may choose to provide remote actor access through HTTP layer; however, another may opt for a network layer implementation.
We exploit the main properties of actor model~\cite{Agha97,Agha90} to distribute actors based on our implementation.
From a distributed perspective, the following are the main requirements for  distributing actors:
\begin{description}
\item[Reference Location Transparency] 
Actors communicate to one another using references. 
In an actor model, there is no in-memory object reference; however, every actor reference denotes a location by means of which the actor is accessible.
The reference location may be local to the calling actor or remote.
The reference location is \emph{physically} transparent for the calling actor.
\item[Communication Transparency]
A message $m$ from actor $A$ to actor $B$ may possibly lead to transferring $m$ over a network such that $B$ can process the message.
Thus, an actor model that supports distribution must provide a layer of remote communication among its actors that is transparent, i.e., 
when actor $A$ sends message $m$, the message is transparently transferred over the network to reach actor $B$.
For instance, actors existing in an HTTP container that transparently allows such communication.
Further, the API implementation is required to provide a mechanism for serialization of messages.
By default, every object in JVM cannot be assumed to be an instance of \jtt{java.io.Serializable}.
However, the API may enforce  that any remote actor should have the required actor classes in its JVM during runtime which
allows the use of the JVM's general object serialization
\footnote{Java Object Serialization Specification: \surl{http://docs.oracle.com/javase/8/docs/platform/serialization/spec/serialTOC.html}}
to send messages to remote actors and receive their responses.
Additionally, we model asynchronous messages with lambda expressions for which Java~8 supports serialization by specification
\footnote{Serialized Lambdas: \surl{http://docs.oracle.com/javase/8/docs/api/java/lang/invoke/SerializedLambda.html}}.
\item[Actor Provisioning]
During a life time of an actor, it may need to create new actors.
Creating actors in a local memory setting is straightforward. 
However, the local setting \emph{does} have a capacity of number of actors it can hold.
When an actor creates a new one, the new actor may actually be initialized in a remote resource.
When the resource is not available, it should be first provisioned.
However, this resource provisioning should be transparent to the actor and only the eventual result (the newly created actor) is visible.
\end{description}

We extend the ABS API to ABS Remote API\footnote{The implementation is available at \surl{https://github.com/CrispOSS/abs-api-remote}.} 
that provides the above properties for  actors in a seamless way.
A complete example of using the remote API has been developed\footnote{An example of ABS Remote API is available at \surl{https://github.com/CrispOSS/abs-api-remote-sample}.}. 
Expanding our ping-pong example in this paper, Listing~\ref{lst:ping:remote} and \ref{lst:pong:remote} present how a remote server of actors is created for the ping and pong actors.
In the following listings, \jtt{java.util.Properties} is used provide input parameters of the actor server;
namely, the address and the port that the actor server responds to.

\begin{center}
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Remote ping actor main,label=lst:ping:remote]
Properties p =  new Properties();
p.put("host", "localhost");
p.put("port", "7777");
ActorServer s = new ActorServer(p);
IPong pong = 
 s.newRemote("abs://pong@http://localhost:8888",
  IPong.class);
Ping ping = new Ping(pong);
ping.send(
  () -> ping.ping("")
);
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\begin{lstlisting}[caption=Remote pong actor main,label=lst:pong:remote]
Properties p = new Properties();
p.put("host", "localhost");
p.put("port", "8888");
ActorServer s = new ActorServer(p);
Pong pong = new Pong();
\end{lstlisting}
\end{minipage}
\end{center}

In Listing~\ref{lst:ping:remote}, a remote reference to a pong actor is created that exposes the \jtt{IPong} interface.
This interface is proxied
\footnote{Java Proxy: \surl{http://docs.oracle.com/javase/8/docs/api/java/lang/reflect/Proxy.html}} by the implementation 
to handle the remote communication with the actual pong actor in the other actor server.
This mechanism hides the communication details from the ping actor and as such allows the ping actor to use 
the same API to send a message to the pong actor (without even  knowing that the pong actor is actually remote).
When an actor is initialized in a distributed setting it transparently identifies its actor server and registers with it.
The above two listings are aligned with the similar main program presented in Listing~\ref{lst:main:api} that presents the same in a local setting.
The above two listings run in separate JVM instances and therefore do not share any objects.
In each JVM instance, it is required that both interfaces \jtt{IPing} and \jtt{IPong} are visible to the classpath; however, 
the ping actor server only needs to see \jtt{Ping} class in its classpath and similarly the pong actor server only needs 
to see \jtt{Pong} class in its classpath.

\section{Experiments}
\label{sec:experiments}

In this section, we explain how a series of benchmarks were directed to evaluate the performance and functionality of actor API in Java 8.
For this benchmark, we use a simple Java application that uses the ``Ping-Pong'' actor example discussed previously. 
An application consists of one instance of \jtt{Ping} actor and one instance of \jtt{Pong} actor. 
The application sends a \jtt{ping} message to the ping actor and waits for the result.
The ping message depends on a \jtt{pong} message to the pong actor.
When the result from the pong actor is ready, the ping actor completes the message;
this completes a round trip of a message in the application.
To be able to make comparison of how actor API in Java 8 performs, the example is also implemented using Akka~\cite{akka} library. 
The same set of benchmarks are performed in isolation for both of the applications.
To perform the benchmarks, we use JMH~\cite{jmh} that is a Java microbenchmarking harness developed by OpenJDK community and used to perform benchmarks for the Java language itself.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.45]{figs/benchmarks}  
\end{center}
\caption{
Benchmark results of comparing sampling time of message round trips in ABS API and Akka.
An example reading of above results is that the time shows for $p(90.0000)$ reads as ``message round trips were completed under $10{\mu}s$ for $90\%$ of the sent messages''.
The first two columns show the ``minimum'' and ``mean'' message round trip times in both implementations.
}
\label{fig:benchmarks}
\end{figure}

The benchmark is performed on the round trip of a message in the application. 
The benchmark starts with a warm-up phase followed by the running phase.
The benchmark composes of a number of iterations in each phase and specific time period for each iteration specified for each phase.
Every iteration of the benchmark triggers a new message in the application and waits for the result.
The measurement used is \emph{sampling time} of the round trip of a message.
A specific number of samples are collected. 
Based on the samples in different phases, different \emph{percentile} measurements are summarized.
An example percentile measurement $p(99.9900) = 10 \; {\mu}s$ is read as $99.9900\%$ of messages in the benchmark took 10 micro-seconds to complete.

Each benchmark starts with 500 iterations of warm-up with each iteration for 1 micro-second.
Each benchmark runs for 5000 iterations with each iteration for 50 micro-seconds.
In each iteration, a maximum number of 50K samples are collected.
Each benchmark is executed in an isolated JVM environment with Java 8 version b127.
Each benchmark is executed on a hardware with 8 cores of CPU and a maximum memory of 8GB for JVM.

The results are presented in Figure \ref{fig:benchmarks}.
The performance difference observed in the measurements can be explained as follows.
An actor in Akka is expected to expose a certain behavior as discussed in Section \ref{sec:example} (i.e. \jtt{onReceive}).
This means that every message leads to an eventual invocation of this method inside actor.
However, in case of an actor in Java~8, there is a need to make a look-up for the actual method to be executed with expected arguments. 
This means that for every method, although in the presence of caching, there is a need to find the proper method that is expected to be invoked. 
A constant overhead for the method look-up  in order to adhere to the object-oriented principles is naturally to be expected.
Thus, this is the minimal performance cost that the actor API in Java~8 pays to support  programming to interfaces.

\section{Conclusion}
\label{sec:conclusion}

In this paper, we discussed an implementation of the actor-based ABS modeling language
in Java~8 which supports  the basic object-oriented mechanisms
and principles of method look-up and programming to interfaces.
In the full version of this paper we have developed an operational semantics of Java~8 features including lambda expressions 
and have proved formally the correctness of the embedding in terms of a bisimulation relation.

The underlying modeling language has an executable semantics
and supports a variety of formal analysis techniques, including
deadlock and schedulability analysis \cite{GiachinoGLLW13,JaghooriBCS09}.
Further it supports a formal 
behavioral specification of interfaces \cite{HahnleHJLSSW11}, to be used as contracts.

We intend to expand this work in different ways.
We aim to automatically \emph{generate} ABS models from Java code which follows the ABS
design methodology.
Model extraction allows industry level applications be abstracted into models and 
analyzed for different goals such as deadlock analysis and concurrency optimization.
This approach of model extraction we believe will greatly enhance industrial uptake of formal methods.
We aim to further extend the implementation of API to support different features especially regarding distribution of actors especially in the queue layer, and scheduling of messages using application-level policies or real-time properties of a concurrent system.
Furthermore, the current implementation of ABS API in a distributed setting allows for instantiation of remote actors. 
We intend to use the implementation to model ABS deployment components~\cite{johnsen2012modeling} and simulate a distributed environment.

% Bibliography
% \bibliographystyle{plain}
% \bibliography{refs}

### 2015-10-20 p02-ch02-design-pattern/manual.tex
Initial Commit

% -*- root: ../thesis.tex -*-

\chapter{Design Pattern}
\label{ch:p02:ch02}

We design and implement a
\href{http://en.wikipedia.org/wiki/Sieve_of_Eratosthenes}{prime sieve
generator} using \href{https://github.com/CrispOSS/abs-api-parent}{ABS
API}. In this example, we strictly follow
\href{http://en.wikipedia.org/wiki/Object-oriented_programming}{object-oriented
design principles} and eventually mix in the actor model with the
implementation.

The following summarizes the prime sieve generation:

\begin{enumerate}
\itemsep1pt\parskip0pt\parsep0pt
\item
  The implementation works with a fixed input number \lstinline!N!.\\
\item
  The result is a sequence of prime numbers less than or equal to
  \lstinline!N!.\\
\item
  The implementation is required to be \emph{concurrent} and
  \emph{asynchronous}.
\end{enumerate}

The following diagram depicts a general structure of the prime sieve
generator in this example:

\includegraphics{figs/PrimeSieves.jpg}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \lstinline!SieveMaster! creates a number of buckets; i.e.
  \lstinline!SieveWorker!, each of which are responsible for a specific
  range.\\
\item
  The bucket size is chosen as \lstinline!5000! as a fixed number is
  this example.\\
\item
  \lstinline!SieveMaster! sends a message \lstinline!sieve! to all
  \lstinline!SieveWorker!s.\\
\item
  When a \lstinline!SieveWorker! finishes its computation it responds to
  the master by sending a message \lstinline!done!.\\
\item
  When the last worker is done, the prime generation completes.
\end{itemize}

Next, we design a Java interface for the prime sieves. Both
\lstinline!SieveMaster! and \lstinline!SieveWorker! are different
implementations of
\href{https://github.com/CrispOSS/prime-sieves/blob/master/src/main/java/com/github/crisposs/sieves/Sieve.java}{\lstinline!Sieve!}:

\lstset{language=Java}
\begin{lstlisting}[caption=Sieve Actor]
interface Sieve {
  void sieve();
  void done(List<Long> primes);
  Long last();
}
\end{lstlisting}

The above interface allows \lstinline!SieveMaster! to:

\begin{enumerate}
\itemsep1pt\parskip0pt\parsep0pt
\item
  break the input \lstinline!N! into a number of buckets\\
\item
  create a \lstinline!SieveWorker! for every bucket\\
\item
  initiate the algorithm with the first worker and moving on one by one
  when each worker is done\\
\item
  return \lstinline!last()! when the last worker finishes the
  computation
\end{enumerate}

The following presents the above steps in \lstinline!SieveWorker!'s
\lstinline!sieve()! method:

\begin{lstlisting}
SieveWorker w = getNextWorker() or finish
(M) send a message to `w` to `sieve`
\end{lstlisting}

and equivalently in \lstinline!SieveWorker!'s
\lstinline!done(List<Long> primes)! method:

\begin{lstlisting}
Collect generated `primes`
`sieve` again
\end{lstlisting}

When \lstinline!sieve()! in the master is done, the collected primes are
the generated prime numbers up to input \lstinline!N!.

The interesting part of the problem above in master's
\lstinline!sieve()! method is:

\begin{lstlisting}
(M) send a message to `w` to `sieve`
\end{lstlisting}

We model \emph{a message} by using

\begin{enumerate}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \lstinline!java.lang.Runnable! to present a message with no interest
  in its return value\\
\item
  \lstinline!java.util.concurrent.Callable<V>! to present a message with
  a specific type of return value \lstinline!<V>!
\end{enumerate}

Through this model, we \textbf{separate} \emph{message delivery} from
\emph{message execution}. Essentially, a message execution is eventually
running the method invocation encapsulated in \lstinline!Runnable! or
\lstinline!Callable!.

For example:

\begin{enumerate}
\itemsep1pt\parskip0pt\parsep0pt
\item
  a message can be delivered in the same thread but executed in a
  different thread\\
\item
  a message can be delivered in a separate thread but executed in the
  same delivery thread\\
\item
  a message can be delivered in a separate thread and executed in a
  thread again different from the delivery thread
\end{enumerate}

The above models of message delivery and execution build the foundation
for \emph{synchronous} and \emph{asynchronous} message passing in ABS
API. By default, ABS API uses the 3rd approach in which all objects in
the runtime of the system share the same pool of threads.

We present \lstinline!(M)! in the following as:

\begin{lstlisting}[language=Java]
Runnable msg = new Runnable() {
  void run() {
    w.sieve();
  }
}
\end{lstlisting}

In Java 8, we can simply use
\href{https://docs.oracle.com/javase/tutorial/java/javaOO/lambdaexpressions.html}{Lambda
expressions}, so:

\begin{lstlisting}[language=Java]
Runnable msg = () -> w.sieve();
\end{lstlisting}

If we want to rewrite \lstinline!(M)! from the above:

\begin{lstlisting}[language=Java]
Runnable msg = () -> w.sieve();
send(w, msg);
\end{lstlisting}

in which \lstinline!send(Object, Object)! is expected to be a method
that semantically provides Approach 2 explained above to deliver the
message and execute it. Now the question is where
\lstinline!send(Object, Object)! can come from?

Next, let us zoom at \lstinline!SieveWorker!. A \lstinline!SieveWorker!
is also a \lstinline!Sieve! implementation a sketch of which can be
presented as:

\begin{lstlisting}
Sieve my range for prime numbers
(W) Reply to the "sender" with generated prime numbers
\end{lstlisting}

When a sieve worker has generated the prime numbers in its own range, it
is expected to reply back to the master. We designed
\lstinline!void done(List<Long>)! for this purpose in \lstinline!Sieve!.
We can rewrite \lstinline!(W)! from above as:

\begin{lstlisting}[language=Java]
List<Long> primes = // compute primes in my range
(W) Reply to master (sender) with message with invoking `done(primes)`
\end{lstlisting}

which means as before, we need to \emph{send a message} but the
difference here is that this is a \emph{reply message}. We can rewrite
then as:

\begin{lstlisting}[language=Java]
List<Long> primes = // compute primes in my range
Runnable replyMsg = () -> {
  (S) Sieve sender = sender();
  s.done(primes);
};
(R) reply(replyMsg);
\end{lstlisting}

We can \emph{functionally} define \lstinline!reply! in \lstinline!(R)!
as:

\begin{lstlisting}
reply = send(sender(), Object)
\end{lstlisting}

in which \lstinline!sender()! in \lstinline!(S)! refers to the
\emph{sender} object of the current message being processed by
\lstinline!this! object.

Naturally, we need to sketch the implementation of
\lstinline!done(List<Long>)! in \lstinline!SieveMaster! as it will be
the main receive of done messages:

\begin{lstlisting}[language=Java]
void done(List<Long> primes) {
  Store the primes
  sieve();
}
\end{lstlisting}

Simply, after each receipt of \lstinline!done!, the master stores the
newly generated primes and sieve again to see if there are still sieve
workers to do their computation.

We summarize the requirements that we have to be able to design this
example following both object-oriented design and asynchronous
computation:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \lstinline!send(Object o, Object msg)! sends a message \lstinline!msg!
  to an object \lstinline!o! in an asynchronous way\\
\item
  \lstinline!sender()! refers to the sender object of the currently
  being processed message in \lstinline!this! object\\
\item
  \lstinline!reply(Object msg)! send a message \lstinline!msg! as a
  reply to the \lstinline!sender()! of the processed message in
  \lstinline!this! object
\end{itemize}

This is where we mix object-oriented principles with Actor model brought
by ABS API. To add the above capabilities to our example, we simple
change \lstinline!Sieve! interface as:

\begin{lstlisting}[language=Java]
interface Sieve extends Actor {
}
\end{lstlisting}

to extend \lstinline!abs.api.Actor! interface. Now, all implementations
of \lstinline!Sieve! interface have access to the methods we draw as
expectations for this example. Applying the actor model API to this
sieve example as explained above can be depicted conceptually as:

\includegraphics{figs/PrimeSievesControlFlow.jpg}

TODO: Generalize how we mixed the original object-oriented principles
and Actor model principles implemented in ABS API through Java 8.

### 2015-10-20 p03-ch01-2015-monitoring/paper-esocc.tex
Initial Commit

%!TEX root = ../thesis.tex

\chapter[Formal verification of service level agreements through distributed monitoring]{Formal verification of service level agreements through distributed monitoring%
\footnote{This paper is funded by the EU project FP7-610582 ENVISAGE: Engineering Virtualized
Services, \url{http://www.envisage-project.eu}.}}
% 
\label{ch:p03:ch01}
% 
\chapterauthor{Behrooz Nobakht, Stijn de Gouw, Frank S. de Boer}

\section*{Abstract}
\label{abstract}
In this paper, we introduce a  formal model of the  availability, budget compliance
and  sustainability of distributed services, where service   sustainability is
a new concept which arises as 
% an extension of 
the composition of 
service  availability and budget compliance.
The model formalizes a distributed  platform for monitoring the above
service characteristics in terms of a parallel composition of task automata,
where dynamically generated tasks model asynchronous events with deadlines.
The main result of this paper is a formal model to optimize and reason about service characteristics
through monitoring.
In particular, we use schedulability analysis of the
underlying timed automata to optimize and guarantee service sustainability.

%to capture characteristics of a distributed deployment environment.
%In the distributed deployment, a service runs across several resources in a virtualized environment.
%We formalize characteristics of a distributed service by service availability and service budget compliance.
%We introduce service sustainability as an extension and composition of the two.
%We design timed automata to ensure the given service level agreement for each characteristic.
%The formal model is realized by devising an actor-based distributed monitoring platform using the Abstract Behavioral Specification language ABS.
%We discuss how the characteristics of the model can be utilized in the application and business in different aspects.
%In addition, we discuss how monitoring model parameters; e.g. monitoring window; influence maintaining service characteristics in a distributed environment.
% We present algorithms to generate the monitoring platform components given input parameters in order to guarantee the defined service level agreements.

% \keywords{runtime monitoring, service availability, budget compliance, service sustainability, distributed architecture, cloud computing, service level agreement}

\input{sec01-intro}
\input{sec02-related}
\input{sec03-frh-example}
\input{sec045-defs}
\input{sec05-monitoring-model}
\input{sec06-impl}
\input{sec07-conclusion}

% \bibliographystyle{abbrv}
% \bibliography{refs}

% \end{document}

### 2015-10-20 p03-ch01-2015-monitoring/sec01-intro.tex
Initial Commit

% -*- root: paper-podc.tex -*-

\section{Introduction}
\label{sec:introduction}

Cloud computing provides the elastic technologies for virtualization. 
Through virtualization, software itself can be offered as a service (Software as a Service,
SaaS). 
One of the aims of SaaS is to allow service providers to offer reliable software services while scaling up and down allocated resources based on their availability, budget, service throughput and the Service Level Agreements (SLA).
Thus, it becomes essential that virtualization technologies facilitate elasticity in a way that enables business owners to \emph{rapidly} evolve their systems to meet their customer requirements and expectations.

The fundamental technical challenge to a SaaS offering is maintaining the quality
of service (QoS) promised by its SLA. In SaaS, providers must
ensure a consistent QoS in a dynamic virtualized environment with variable
usage patterns. Specifically, virtualized environments such as the cloud
provide elasticity in resource allocation, but they often do not offer an SLA that can
guarantee constant resource availability. As a result, SaaS providers are
required to react to resource availability at runtime. Furthermore, by offering a
$24/7$ software service, SaaS providers must be able to react to certain service
usage patterns, such as an increase in throughput to ensure the SLA is
maintained.

Runtime monitoring~\cite{Logean_monitoring,BratanisDS10} is a dynamic analysis approach
based on
%observing the runtime of a software system and
extracting relevant information about the execution.
Runtime monitoring may be employed to collect statistics about the service usage over time, and to detect and react to service behavior. 
This latter ability is fundamental in the SaaS approach to guarantee the SLA of a service
% to the provision of virtualized services
%This ability
and is the focus of this paper.

The monitoring model that is presented in this paper is designed to \emph{observe} 
% the evolution 
in real-time certain  service characteristics and \emph{react} to them to ensure the evolution of the system towards its SLA.
% In this paper we develop an actor-based monitoring model to study the dynamic
% provisioning of resources to services for maintaining service availability given a
% certain level of budget compliance. Our monitoring model formalizes the notion of service
% sustainability, that is, whether a reliable service can be maintained that
% meets the SLA.
%Specifically, our actor-based model is defined in terms of the Abstract
%Behavioral Specification (ABS) language~\cite{johnsen2012abs} and Real-Time ABS~\cite{johnsen2012modeling}, %that is
%an extension of ABS that allows modeling resources and measuring resource usage.
%%providing scheduling of time concerns for ABS. 
%ABS is an object oriented modeling
%language with a formal executable semantics that is designed for modeling
%highly adaptable distributed concurrent systems. 
%%We provide an operational semantics to our model, and 
%% We formalize the notion of service availability, budget compliance, and service sustainability.
%%We apply the formalization results to present
%% 
%We choose actor model~\cite{agha86book} to base the monitoring model on.
%The monitoring model is required to be reactive to changes; 
%i.e. synchronization risks must be avoided as much as possible.
Asynchronous communication is an essential feature of a monitoring model 
in a distributed context.
Asynchronous communication accomplishes non-intrusive observations of the service runtime. 
Further, the monitoring model is expected to operate according to certain real-time constraints specified by the SLA of the service.
Satisfying the real-time constraints is the main challenge in a distributed monitoring model.

%  based on a  scheduling policy;
% i.e. it should be able to use application-level scheduling mechanisms~\cite{nobakht2013future} to meet its requirements.
% A basic assumption underlying our monitoring model is that gathering service statistics and information at runtime should not affect the runtime behavior of the services themselves.

%In addition, the monitoring model should operate in a distributed environment of services.
%Distributed remote monitors operate
%%against their configured
%services to
%%guard them
%ensure their SLAs.

In this paper, we formalize service availability and budget compliance in a distributed deployment environment.
This formalization is based on  
high-level task automata models~\cite{alur:1994:timedautomata,fersman2007task,jaghoori2010time}.
The automata capture the real-time evolution of the resources provided by a distributed deployment platform and 
the above two main service characteristics.
These  task  automata represent the  real-time generation of the asynchronous events 
extended with deadlines~\cite{bjork2013:rtabs,nobakht2013future}
by the monitoring platform for managing resources (i.e. allocation or  deallocation).
The main result of this paper is a formal model to optimize and reason about the above service  characteristics through monitoring.
In particular, the \emph{schedulability} of the underlying timed automata implies service availability and budget compliance.
%present how the semantics of the timed automaton evolves the system towards its service level agreements.
%We also present the semantics of the distributed environment using a separate timed automaton.
%We prove theorems on the above timed automata and how their semantics guarantee achievement of the service level agreements for service availability and service budget compliance.
%In the semantics of the timed automata, we utilize asynchronous communication along with deadlines for messages.
Furthermore, we introduce a  composition of  service availability and  budget compliance which captures service sustainability.
We show that service sustainability presents a multi-objective optimization problem.
% and to be able to ensure this property, elements of the monitoring platform needs to changes during the runtime of the system.
% We show how they can be used in the application in business levels and how to design and generate corresponding monitors.

%The monitoring model that is presented in this research is able to \emph{observe} a system based on the above service characteristics and \emph{react} to them to ensure the evolution of the system towards its SLA.
%% In this paper we develop an actor-based monitoring model to study the dynamic
%% provisioning of resources to services for maintaining service availability given a
%% certain level of budget compliance. Our monitoring model formalizes the notion of service
%% sustainability, that is, whether a reliable service can be maintained that
%% meets the SLA.
%Specifically, our actor-based model is defined in terms of the Abstract
%Behavioral Specification (ABS) language~\cite{johnsen2012abs} and Real-Time ABS~\cite{johnsen2012modeling}, %that is
%an extension of ABS that allows modeling resources and measuring resource usage.
%%providing scheduling of time concerns for ABS. 
%ABS is an object oriented modeling
%language with a formal executable semantics that is designed for modeling
%highly adaptable distributed concurrent systems. 
%%We provide an operational semantics to our model, and 
%% We formalize the notion of service availability, budget compliance, and service sustainability.
%%We apply the formalization results to present
%% 
%We choose actor model~\cite{agha86book} to base the monitoring model on.
%The monitoring model is required to be reactive to changes; 
%i.e. synchronization risks must be avoided as much as possible.
%Asynchronous communication allows the monitoring model to be concurrent and reactive.
%The monitoring model is expected to expose certain properties based on a time schedule;
%i.e. it should be able to use application-level scheduling mechanisms~\cite{nobakht2013future} to meet its requirements.
%The monitoring model should be non-intrusive and distributed. 
%Gathering service statistics and information at runtime should not affect the runtime behavior of the services themselves.
%In addition, the monitoring model should operate in a distributed environment of services.
%Distributed remote monitors operate
%%against their configured
%services to
%%guard them
%ensure their SLAs.

%The rest of the paper is structured as follows: 
%In Section \ref{sec:relatedwork} we discuss related work in relation with runtime monitoring and service level agreements.
%We provide a real-life example from SDL Fredhopper in Section \ref{sec:fredhopper_example} that introduces an industrial case from which the model is extracted and on which it is evaluated. 
%We present the elements of a distributed environment in Section \ref{sec:modelling} and then formalize necessary assumptions and definitions.
%In Section~\ref{sec:timed:fsm}, we present the timed automata for service characteristics and theorems.
%In Section~\ref{sec:implementation}, we discuss the evaluation of the monitoring model.
%Section~\ref{sec:conclusion} presents the future work and concludes the paper.

### 2015-10-20 p03-ch01-2015-monitoring/sec02-related.tex
Initial Commit

% -*- root: paper-esocc.tex -*-

% Section: Related Work

\section{Related Work}
\label{sec:relatedwork}

Vast research work present different aspects of runtime monitoring.
We focus on those that present a line of research for distributed deployment of services.

%MONINA~\cite{inzinger:monitoring,inzinger2013specification} is a DSL language based on which a generic monitoring architecture is designed.
%The architecture and the language use a rule engine based on facts.
%To use MONINA, it is required to model system parts as ``component'' offered by the DSL language.
%We build our model using timed automata~\cite{alur:1994:timedautomata} with a mapping to ABS~\cite{johnsen2012abs}, a concurrent object-oriented modelling language which is essential for schedulability analysis~\cite{fersman2007task} of real-time and distributed systems.
%MONINA offers a solution based on facts and ``derived'' facts used by a rule engine to emit corrective events.
%The DSL language in MONINA introduces \emph{cost} and \emph{capacity}.
%In contrast, we introduce a generic service metric function to allow for generic definition of a metric from the service's standpoint. 
%In our model, we also study similar metrics such as cost.
%To deploy MONINA on the cloud, the authors recommend using the notion of \emph{adaptation in cycles}.
%It is similar to \emph{monitoring window} that we define further in the paper.
%However, we formally prove how the selection of the monitoring window is key to the evolution of the service.
MONINA~\cite{inzinger:monitoring} is a DSL with a monitoring architecture
which supports certain mathematical optimization techniques.
A prototype implementation is available.
%To use MONINA requires developing a MONINA component.
Accurately capturing the
behavior of an in-production legacy system coded in a
conventional language seems challenging: it requires
developing MONINA components, which
generate events at a specified fixed rate,  there are no control structures (if-else, loops), the data types that can be used in events are pre-defined, and there are no OO-features. 
We use ABS~\cite{johnsen2012abs}, an executable modeling language
that supports all of these features and offers a wide range of tool-supported analyses~\cite{BubelMH14,WongBBGGHMS15}.  The mapping from ABS to timed automata~\cite{alur:1994:timedautomata} allows to
exploit the state-of-the-art tools for timed automata, in
particular for reasoning about real-time properties (and, as we show, SLAs using
schedulability analysis~\cite{fersman2007task}).
MONINA offers \emph{two pre-defined} parameters that can be used in monitoring to adapt
the system: cost and capacity.  Our service metric function generalizes this to
\emph{arbitrary user-defined} parameters, including cost and capacity.

% Furthermore, the architecture applies a set of adaptation rules.
% The architecture uses an event-driven approach and message passing mechanisms.
% Our approach leverages the well-established formalism of timed automata~\cite{alur:1994:timedautomata} which allows for static analysis of real-time properties of a service.
% We do not impose a new language; however, we use the established formalism of timed automata~\cite{alur:1994:timedautomata}.
% We propose a generic model to be able to characterize arbitrary service properties that may come from different sources.
% The model in the paper allows for a language-agnostic implementation. 
%However, MONINA provides an architecture implementation based on Rule Engine technologies and its domain-specific language design.

Hogben and Pannetrat examine in \cite{hogben2013defavail} the challenges of defining and measuring availability to support real-world service comparison and dispute resolution through SLAs.
They show how two examples of real-world SLAs would lead one service provider to report 0\% availability while another would report 100\% for the same system state history but using a different period of time.
The transparency that the authors attempt to reach is addressed in our work by the concept of monitoring window and expectation tolerance in Section \ref{sec:modelling}.
Additionally, the authors take a continuous time approach contrasted with ours that uses discrete time advancements.
Similarly, they model the property of availability using a two-state model.

%There are related work in the domain of service level agreements (SLA).
The following research works provide a language or a framework that allows to formalize service level agreements (SLA).
However, they do not study how such SLAs can be used to monitor the service and evolve it as necessary.
WSLA~\cite{keller2003wsla} introduces a framework to define and break down customer agreements into a technical description of SLAs and terms to be monitored.
In ~\cite{mahbub2011translationsla}, a method is proposed to translate the specification of SLA into a technical domain directed in SLA@SOI EU project.
In the same project, \cite{comuzzi2009defavail} defines terms such as availability, accessibility and throughput as notions of SLA, however, the formal semantics and properties of the notions are not investigated.
In \cite{chen2007sladecompose}, authors describe how they introduce a function how to decompose SLA terms into measurable factors and how to profile them.
Timed automata is used in~\cite{raimondi2008fsmsla} to detect violations of SLA and formalize them.

Johnsen~\cite{johnsen2012modeling} introduce ``deployment components'' using Real-Time ABS~\cite{bjork2013:rtabs}.
A deployment component enables an application to acquire and release resources on-demand based on a QoS specification of the application.
A deployment component is a high level abstraction of a resource that promotes an application to a resource-aware level of programming.
Our work is distinguished by the fact that we separate the monitors from the application (service) themselves.
We argue that we aim to design the monitoring model to be as \emph{non-intrusive} as possible to the service runtime. 
Thus, we do not deploy the monitors inside the service runtime.

In Quanticol EU project\footnote{Quanticol EU project with no. 600708: \url{http://quanticol.eu/}}, authors in \cite{coles2011cost} and \cite{gilmore2011non} use statistical approaches to observe and guarantee service level agreements for public transportation.
We also present that service characteristics can be composed together. 
This means that evolving a system based on SLAs turns into a multi-object optimization problem.
In addition, in COMPASS EU project\footnote{COMPASS EU project with no. 287829: \url{http://www.compass-research.eu/}}, CML~\cite{woodcock2014contracts} defines a formal language to model systems of systems and the contracts between them.
CML studies certain properties of the model and their applications.
CML is used in the context of a Robotics technology to model and ensure how emergency sensors should react and behave according to the SLAs defined for them.
Our approach is similar to provide a generic model for service characteristics definition, however, we utilize timed and task automata.

### 2015-10-20 p03-ch01-2015-monitoring/sec03-frh-example.tex
Initial Commit

% -*- root: paper-podc.tex -*-

\section{SDL Fredhopper Cloud Services} % (fold)
\label{sec:fredhopper_example}

In this section, we introduce a running example in the context of SDL Fredhopper. 
We use the example in different parts of the paper and also in the experiments.

SDL Fredhopper develops the \frhcsp to offer search and targeting facilities on a large product database to e-Commerce companies as services (SaaS) over the cloud computing infrastructure (IaaS).  
\frhcsp provides several SaaS offerings on the cloud. 
These services are exposed at endpoints. 
In practice these endpoints typically are implemented to accept connections over HTTP.
For example, one of the services offered by these endpoints is the Fredhopper Query API, which allows users to query over their product catalog via full text search\footnote{\url{http://en.wikipedia.org/wiki/Full_text_search}} and faceted navigation\footnote{\url{http://en.wikipedia.org/wiki/Faceted_navigation}}.

A customer of SDL Fredhopper using Query API owns a \emph{single} HTTP endpoint to use for search and other operations.
However, internally, a number of resources (virtual machines) are used to deliver Query API for the customer.
The %set of
resources used for a customer are managed by a load balancer.
% to provide a single HTTP endpoint.
In this model of deployment, each resource is launched to serve \emph{one} instance of Query API; i.e. resources are \emph{not} shared among customers.

When a customer signs a contract with SDL Fredhopper, there is a clause in the contract that describes the minimal QoS levels of the
%customer regarding the
Query API.
For example, we have a notion of query per second (\qps) that defines the number of completed queries per second for a customer. 
An agreement is a bound on the expected \qps and forms
%which becomes
the basis of many decisions (technical or legal) thereafter. 
The agreement is used by the operations team to set up an environment for the customer which includes the necessary resources described above.
The agreement is additionally used by the support team to manage communications with the customer during the lifetime of the service for the customer.
%

Maintaining the services for more than 250 customers on more than 1000 servers is not an easy operation task
\footnote{Figures are indication of complexity and scale. Detailed confidential information may be shared upon official request.}.
Thus, to ensure the agreements in a customer's contract:
\begin{itemize}
\item The operation team maintains a monitoring platform to get notifications on the current metrics.
\item The operation team performs \emph{manual} intervention to ensure that sufficient resources are available for a customer (launching or terminating).
\item The monitoring platform depends on \emph{human} reaction.
\item The cost that is spent for a customer on the basis of safety can be \emph{optimized}.
\end{itemize}

In this paper, we use the notion of \qps as an example in the concepts that are presented in this research. 
We use the example here to demonstrate how the model that is proposed in this research can address the issues above and alleviate the \emph{manual} work with \emph{automation}. 
The manual life cycle depends on the domain-specific and contextual knowledge of the operations team for every customer service that is maintained in the deployment environment.
This is labor-intensive as the operations team stands by $24\times7$.
In such a manual approach, the business is forced to over-spend to ensure service level agreements for customers.

### 2015-10-20 p03-ch01-2015-monitoring/sec045-defs.tex
Initial Commit

% -*- root: paper-esocc.tex -*-

\section{Distributed Monitoring Model}
\label{sec:modelling}

We introduce a distributed monitoring platform and its components and
discuss some underlying assumptions and definitions.
Further, we define the notion of service availability and service budget compliance.
%\begin{wrapfigure}{r}{0.3\textwidth}
%\vspace{-20pt}
%\begin{center}
%\includegraphics[scale=0.7]{figs/deployment}
%\end{center}
%\vspace{-20pt}
%\caption{{\footnotesize High-level deployment architecture}}
%\label{fig:deployment}
%\vspace{-10pt}
%\end{wrapfigure}
%Figure \ref{fig:deployment} presents a high-level overview of services deployed in a distributed environment (e.g. ``the cloud'').
In the deployment environment (e.g., ``the cloud''), every  server from the IaaS provider is  used for a \emph{single} service of a customer, such as
the Query Service API for a customer of SDL Fredhopper (c.f. Section \ref{sec:fredhopper_example}).
Typically, multiple servers are allocated to a single customer.
The number of servers allocated for a customer is not visible to the customer.
The customer uses a single endpoint - in the load balancer layer - to access all their services.

The ultimate goal is to maintain the environment in such a way that customers and their end users experience the delivered services up to their expectations while minimizing the cost of the system.
%environment does not have a negative impact on the business of SDL Fredhopper.
The first objective can be addressed by adding resources; however, this conflicts with the second goal since it increases the cost of the environment for the customer.
In this section, we formalize the above intuitive notions as \emph{service availability} and \emph{service budget compliance}.

We then develop a distributed monitoring platform that aims to optimize these service
characteristics in a deployment environment.
 % given in Figure \ref{fig:deployment}.
The monitoring platform works in two cyclic phases: \emph{observation} and \emph{reaction}.
The observation phase takes measurements on services in the deployment environment.
Subsequently, the corresponding levels of the service characteristics are calculated.
In the reaction phase, if needed, a platform API is utilized to make the necessary changes to the deployment environment (e.g. adjust the number of allocated resources) to optimize the service characteristics.
% In the monitoring platform, not all components reside in the same resource in the deployment environment; all communications and interactions are performed in an \emph{asynchronous} way.
The monitoring platform builds on top of a real-time extension of the 
actor-based language ABS~\cite{johnsen2012abs}.
To ensure non-intrusiveness of the monitor with the running service,
each monitor is an active object (actor) running on a separate resource
from that which runs the service itself, and the components of the monitoring platform communicate through \emph{asynchronous messages} with deadlines~\cite{johnsen2012modeling}.
% The following elements are the major high-level components of the monitoring platform:
%for a distributed deployment environment:

Below, we discuss assumptions and basic oncepts that will be used in the
analysis of the formal properties of the monitoring platform and corresponding theorems.
We assume that the external infrastructure provider has an  \emph{unlimited} number of resources.
Further, we assume that all resources are of the \emph{same type}; i.e. they have the same computing power, memory, and IO capacity.
Finally, we assume that every resource   is initialized within  at most $t_i$ amount of time.
%We illustrate and motivate each assumption and definition and discuss how they are generalized from real use-cases from the large deployment environment of the
%SDL Fredhopper cloud.

%\begin{assm}[Unlimited Capacity]
%\label{def:platform:cap}
%We assume that the external infrastructure provider is capable to provision an \emph{unlimited} number of resources.
%\end{assm}

%In reality, every service provider such as SDL Fredhopper has a separate contract with an IaaS provider such as AWS.
%In these contracts, there cannot be a guarantee for \emph{unlimited} capacity.
%We simplify this limitation by assuming that the platform API (IaaS layer abstraction) is capable to provision as many resources as requested.

%\begin{assm}[Single Resource Type]
%\label{def:single:resource:type}
%To simplify reasoning, we assume that all resources are of the \emph{same type}; i.e. they have the same computing power, memory, and IO capacity.
%For example, if we are using AWS,
%we could choose the instance type \jtt{m1.large} for all the services.
%\end{assm}

%If a business delivers different types of services, it cannot avoid the fact that different services may require different \emph{capabilities} from their underlying resources.
%% i.e. one service might require high I/O throughput for its process whereas another might demand high parallelism support from the hardware.
%% Such capability profiles are provisioned through different types of resources from a IaaS provider.
%For example, AWS offers\footnote{\url{http://aws.amazon.com/ec2/instance-types/}} families of EC2 instance types each of which expose different sets of capabilities in terms of computation power, I/O, and network.
%We simplify our analysis by assuming that there is a \emph{single resource type} that is able to provide the necessary capabilities for all services.% in this research scope.

%\begin{assm}[Resource Initialization Time $t_i$]
%\label{def:resource:init:time}
%We assume that every resource $r$ that is initialized is ready for use in at most $t_i$ amount of time.
%% The time $t_i$ is bounded by a finite constant value.
%\end{assm}

%Initialization time can vary between different resource types. The above Assumption \ref{def:single:resource:type} simplifies this: we can just choose the initialization time (a fixed constant) of the single resource type that was chosen.
%The initialization time is also part of the contract with the IaaS provider.
%As discussed above in Assumption \ref{def:single:resource:type}, when a resource is launched and initialized, it might take a different amount of time to be in a state that is \emph{operational} and ready to be used.
%Thus,
%We simplify the property of initialization time of a resource to be a fixed constant for the single resource type that is used in the context of this research.

In our framework
time $T$ is a universally shared clock based on the NTP
\footnote{\url{https://tools.ietf.org/html/rfc1305}} that is used by all elements of the system in the same way. 
$T$ is discrete.
We fix that the unit of time is \emph{milliseconds}.
This level of granularity of time unit means that between
two consecutive milliseconds, the system is not observable.
For example, we use the UTC time standard for all services, monitors and platform API.
We refer to the current time by $t_c$.

We denote by $r$ a resource which provides computational power and storage
and by $s$ a general abstraction of a service in the deployment environment. 
A service exposes an API that is accessible through a delivery layer, such as HTTP.
In our example, a service is the Query API (c.f. Section~\ref{sec:fredhopper_example}) that is accessible through a single HTTP endpoint.

In our framework, 
\emph{monitoring platform}  $P$ is responsible for (de-)allocation of resources for computation or storage.
We abstract from a specific implementation of the monitoring platform $P$ through an API in Listing~\ref{lst:platform}. 
% 
\lstset{language=java,aboveskip=-20pt,belowskip=-20pt}
\begin{wrapfigure}{r}{0.5\textwidth}
\vspace{-15pt}
\begin{center}
\begin{lstlisting}[mathescape,caption=Platform API,label=lst:platform]
interface Platform {
  void    allocate(Service s);
  void    deallocate(Service s);
  Number  getState(Service s);
  boolean verify$_\alpha$(Service s);
  boolean verify$_\beta$(Service s);
}
\end{lstlisting}
\end{center}   
\vspace{-15pt}
\end{wrapfigure}
% 
There is only \emph{one} instance of $P$ available.
In this paper, $P$ internally uses an external infrastructure provisioning API to provide resources (e.g. AWS EC2).
The term ``platform'' is interchangeably used for monitoring in this paper.
The platform provides a method \jtt{getState(Service s)} which returns
the number of resources allocated to the given service $s$ at time $t_c$.

%Monitoring is a runtime analysis approach~\cite{Logean_monitoring}.
We use monitoring to observe the external behavior of a service.
We formalize the external behavior of a service with its service-level agreement (SLA).
An SLA is a contract between the customer (service consumer) and the service provider
which defines (among other things) the minimal quality of the offered service,
and the compensation if this minimal level is not reached.
To formally analyze an SLA, we introduce the notion of a service metric function.
We make basic measurements of the service externally in a given monitoring window (a duration).
The service metric function aggregates the basic measurements into a single number
that indicates the quality of a certain service characteristic (higher numbers are better).
%
%monitoring window measurement.
%The quantified measurements projected to a monitoring window lead to an understanding of the quality of the service based on its SLA.

\emph{Basic measurement} $\mu(s,r,t)$ is a function that produces a real number of a \emph{single} monitoring check on a resource $r$ allocated to service $s$ at some time $t$.  
For example, for SDL Fredhopper cloud services, a basic measurement is the number of completed queries at the current time. 
%

\emph{Service Metric} $f_s$ is a function that aggregates a sequence of basic non-negative measurements to a single non-negative real value:
$f_s : \bigcup_n \mathbb{R}^n \rightarrow \mathbb{R}$.
For example, for SDL Fredhopper cloud services, the service metric function
$f_s$ calculates the average number of queries per second (\qps) given a list of basic measurements.

\emph{Monitoring Window} is a duration of time $\tau$ throughout which basic measurements for a service are taken.
%

\emph{Monitoring Measurement} is a function that aggregates the basic measurements for a service over its resources in the last monitoring window.
The last monitoring window is defined as $[t_c-\tau, t_c]$.
To produce the monitoring measurement, $f_s$ is applied.
Formally:
\[
\mu(s,r,\tau) = 
f_s \big(\langle \mu_i(s,r,t) \rangle^{\infty}_{i=0}\big) \;\; \textnormal{where} \;\; t \in [t_c-\tau, t_c] 
\]
in which $\mu_i(s,r,t)$ is the $i$-th basic measurement of services $s$ on resource $r$ at time $t$ where $t \in [t_c-\tau,t_c]$.

\begin{defn}[Service Availability $\alpha(s,\tau,t_c)$]
\label{def:service:availability}
% 
First, we need a few auxiliary definitions before we can define service availability.
%

\emph{Service Capacity} $\kappa_\sigma(s,\tau) = \sum_{r \in \sigma(s)} \mu(s,r,\tau)$ denotes the  capability of service $s$ that is the aggregated monitoring measurements of its resources over the monitoring window $\tau$ and $\sigma(s)$ is the number of allocated resources to service $s$.
%

\emph{Agreement Expectation} $E(s,\tau, t_c)$ is the minimum number of requests that a customer expects to complete in a monitoring window $\tau$.
The agreement expectation depends on the current time $t_c$ because the expectation may change over time.
For example, SDL Fredhopper customers expect a different \qps during Christmas.
%

We define the availability of a service $\alpha(s,\tau,t_c)$ in every monitoring window $\tau$ as:
\vspace{-10pt}
\[
\alpha(s,\tau,t_c) = \frac{\kappa_\sigma(s,\tau)}{E(s,\tau,t_c)}
\]
% in every monitoring window $\tau$.
%

\emph{Capacity Tolerance} $\varepsilon_\alpha(s,\tau)) \in [0,1]$ defines how much $\kappa_\sigma(s,\tau)$ can deviate from $E(s,\tau,t_c)$ in every time span
of duration $\tau$.
\end{defn}
%

\emph{Service Guarantee Time $t_G$} is the duration within which a customer expects service availability reaches an acceptable value after a violation. 
Typically, $t_G$ is an input parameter from the customer's contract.

\begin{exmp}
Intuitively, $\alpha(s,\tau,t_c)$ presents the actual capability of a service $s$ over a time period $\tau$ compared to the expectation on the service $E(s,\tau)$. 
For values $\alpha(s,\tau,t_c) \ll 1 - \varepsilon_\alpha(s,\tau))$, the resource for service $s$ are at ``under-capacity'' while for values $\alpha(s,\tau,t_c) \gg 1 + \varepsilon_\alpha(s,\tau))$, there is ``over-capacity''. The goal is optimize $\alpha(s,\tau,t_c)$ towards a value of 1.
%

For example, we expect a query service to be able to complete 10 queries per second. 
We define the monitoring window $\tau = $ 5 minutes; thus, $E(s,\tau,t_c) = 10 \times 60 \times 5 = 3000$.
Suppose we allocate only one resource to the service,
measure the service during a single monitoring window $\tau$
and find $\mu(s,r,\tau) = 2900$. Then $\alpha(s,\tau,t_c) = \frac{2900}{3000} = 0.966$.
If we have $\varepsilon_\alpha(s,\tau)) = 0.03$, this means that service $s$ is under-capacity because $\alpha(s,\tau,t_c) < 1 - \varepsilon_\alpha$.
\end{exmp}

\begin{defn}[Budget Compliance $\beta(s,\tau)$]
\label{def:budget:compliance}
% 
We first provide a few auxiliary definitions.
%

\emph{Resource Cost} $\euro(r,\tau) \in \mathbb{R}^+$ is the cost of resource $r$ in a monitoring window $\tau$ which is determined by a fixed resource cost per time unit.

\emph{Service Cost} $\euro_\sigma(s,\tau) \in \mathbb{R}^+$ is the cost of a service $s$ in a monitoring window $\tau$ and defined as $\euro_\sigma(s,\tau) = \sum_{r \in \sigma(s)} \euro(r,\tau)$.
%

\emph{Service Budget} $B(s,\tau)$ specifies an upper bound of the expected cost of a service in the time span $\tau$.
Intuitively $B(s,\tau)$ is the allowed budget that can be spent for service $s$ over the time span $\tau$.
%Note that
The service budget is typically chosen to be fixed over any time span $\tau$.
%

We are now ready to define service budget compliance $\beta(s,\tau)$ that,
intuitively, represents how a service complies with its allocated budget:
\[
\beta(s,\tau) = \frac{\euro_\sigma(s,\tau)}{B(s,\tau)}
\]

\emph{Budget Tolerance} $\varepsilon_\beta(s,\tau) \in [0,1]$ specifies how much the service cost $\euro(s,\tau)$ can deviate from $B(s,\tau)$ in every time span of duration $\tau$.

\emph{Service Guarantee Time} $t_G$ is similar to that defined for service availability.
% 
\end{defn} % end - budget compliance

\begin{exmp}
% during its runtime. 
Assume every resource on the environment costs $1$ (e.g. $\euro$) per hour.
Suppose we set a budget of $1.5$ per hour for every service,
allocate \emph{one} resource to the service and define a monitoring window of $\tau = 5$ minutes.
Every hour has $12$ monitoring windows.
This means that each resource costs $\euro(r,\tau) = \frac{1}{12} \approx 0.08$ per monitoring window.
Since there is only one resource, the service cost is $\euro(s,\tau) = \sum_{r\in\sigma(s)}\euro(s,\tau) \approx 0.08$ per monitoring window.
On the other hand, if we calculate the budget for one monitoring window, we have $B(s,\tau) = \frac{1.5}{12} = 0.125$ per monitoring window.
This yields budget compliance as $\beta(s,\tau) = \frac{0.08}{0.125} = 0.64$.
\end{exmp}
%

% The formal definitions above provide a basis
% to react appropriately to changes or failures in an environment.
% We enumerate a few as follows.
% 
% A service $s$ is ``under-capacity'' as long as $\alpha(s,\tau,t_c) < 1 - \varepsilon_\alpha(s,\tau))$; 
% i.e. intuitively this means that the service cannot provide the expected quality to
% end-users.
% To fix an under-capacity service, \emph{new} resources through the monitoring platform $P$ can be requested (scale up) to increase the capacity of $s$.
% Similarly, service $s$ can be ``over-capacity'';
% i.e. $\alpha(s,\tau,t_c) \gg 1 + \varepsilon_\alpha(s,\tau))$.
% An important disadvantage of an over-capacity service is that it \emph{costs} more than necessary to deliver what is expected from it.
% To improve an over-capacity service, at least one of the current resources allocated to service $s$ can be terminated (scale down) as long as this does not result in an ``under-capacity'' service.
%Similarly, a service can be ``over-budget'' and ``under-budget'', and corresponding actions can be taken to resolve that.

The formal definitions of service availability and budget compliance provide a rigorous 
basis for automatic deployment of resource-aware services with an appropriate quality of service, taking costs into account.
%Properties $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$ of a service $s$ build a foundation facilitate \emph{automation of operation} of the service according to its properties. 
This in particular includes automated scaling up or down of the service with the help of monitoring checks that are installed for the service. 
%However,
The fundamental challenge in ensuring service availability and budget compliance is that they have \emph{conflicting} objectives:
\[
\alpha(s,\tau,t_c) \uparrow \iff \beta(s,\tau) \downarrow
\]
Intuitively, if more resources are used to ensure the availability of a service; then $\alpha(s,\tau,t_c)$ increases.
However, at the same time, the service costs more; i.e. budget compliance $\beta(s,\tau)$ decreases.

% In the following sections, we first present two separate timed automata~\cite{alur:1994:timedautomata} to model the behavior of $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$.
% Then, we formalize how the two can be composed to form a new property called service sustainability $\gamma(s,\tau)$.

% In the next section, we design a timed automata for each service characteristics.
% We show how the timed automata can be used to evolve the systems towards SLA optimizations.
% We present a generalization of the characteristics and how they can be composed together.

### 2015-10-20 p03-ch01-2015-monitoring/sec05-monitoring-model.tex
Initial Commit

% -*- root: paper-esocc.tex -*-

\section{Service Characteristics Verification}
\label{sec:timed:fsm}

In this section, we use timed automata and task automata to model the behavior of a monitoring platform $P$, the deployment environment $E$, and the monitoring components for service availability $\alpha(s,\tau,t_c)$ and budget compliance $\beta(s,\tau)$. 
\cite{jaghoori2010time} defines a task automata as an extension of timed automata in which each task is a piece of executable program with $(b,w,d)$: best/worst time and deadline of the task. A task automata uses a scheduler for the tasks to schedule each task with a location on a queue.
%

Modeling the elements of the monitoring platform is necessary to be able to study certain properties of the system.
The most important goal of a monitoring platform is to enable the autonomous operation of a set of services according to their SLA.
Thus, it is essential how to analyze that the monitoring platform can provide certain guarantees about the service and its SLA.
In addition, it is important be able to verify the monitoring platform through model checking and schedulability analysis.
Using timed automata and task automata facilitates model checking and verification through formal method tools such as UPPAAL~\cite{uppaal2004} supporting advanced methods such as state-space reduction~\cite{larsen1997efficient}.

We use task automata as defined in \cite{fersman2007task,jaghoori2012composing,jaghoori2010time}.
Task automata are an extension of timed automata~\cite{alur:1994:timedautomata}.
In addition, we design the automata for the monitoring platform using the real-time extension of task automata presented in \cite{jaghoori2010time} p. 92 in which the author presents a mapping from Real-Time ABS~\cite{johnsen2012modeling} to the equivalent task automata.

A task type is a piece of executable program/code represented by a tuple $(b,w,d)$, where $b$ and $w$ respectively are the best-case and worst-case execution times and $d$ is the deadline.
In a task automata, there are two types of transitions: \emph{delay} and \emph{discrete}.
A delay transition models the execution of a running task by idling for other tasks.
A discrete transition corresponds to the arrival of a new task.
When a new task is triggered, it is placed into a certain position in the queue based on a scheduling policy~\cite{Nobakht:sched,nobakht2013future}.
Examples of a scheduling policy are \textsf{FIFO} or \textsf{EDF} (earliest deadline first).
The scheduling policy is modeled as a timed automaton \textsf{Sch}.
Every task has its own stop watch.
The scheduler also maintains a separate stop watch for each task to determine if a task misses its deadline.
All stop watches work at the same clock speed specified by $T$.

We design separate automata for each service $s$ characteristic: service availability $\alpha(s,\tau,t_c)$ by an automata $M_{\alpha_s}$ and service budget compliance $\beta(s,\tau)$, by an automata $M_{\beta_s}$.
Each automaton is responsible for one goal: to optimize the service characteristic.
$M_{\alpha_s}$ aims to improve $\alpha(s,\tau,t_c)$ whereas $M_{\beta_s}$ aims to improve $\beta(s,\tau)$.
$M_{\alpha_s}$ uses \jtt{allocate} to launch a new resource in the environment and improve the service $s$.
In contrast, $M_{\beta_s}$ uses \jtt{deallocate} to terminate a resource to decrease the cost of the service.

We use task automata to design $M_{\alpha_s}$.
Periodically, $M_{\alpha_s}$ checks whether the service availability
is within the thresholds, taking tolerance into account
(Definition~\ref{def:service:availability}).
If the condition fails, $M_{\alpha_s}$ generates a task for monitoring platform $P$ to allocate a new resource to service $s$ with a deadline of $\tau$.
We define the period to be $\tau$.
We use the semantics of a task automata in \cite{jaghoori2010time} p. 92 in the transitions of the task automata.
Figure~\ref{fig:fsm:task:alpha} and \ref{fig:fsm:task:beta} present $M_{\alpha_s}$ and $M_{\beta_s}$.
Both $M_{\alpha_s}$ and $M_{\beta_s}$ share state with the monitoring platform $P$.
The state keeps the current number of resources for a service $s$ that is denoted by $\sigma(s)$.
All timed automata and task automata in the monitoring platform have shared access to $\sigma(s)$.
In the automata, we use a conditional statement to check the service characteristics $\alpha(s,\tau,t_c)$ or $\beta(s,\tau)$.
If the condition fails, $M_{\alpha_s}$ requests $P$ to \jtt{allocate} a new resource to $s$ and $M_{\beta_s}$ requests $P$ to \jtt{deallocate} a resource.
In addition, $M_{\alpha_s}$ triggers a new task $\jtt{verify}_\alpha$ with deadline $t_G$.
Intuitively, this means the service characteristic $\alpha(s,\tau,t_c)$ is verified to be within the expected thresholds after at most $t_G$ time.
\begin{figure}[h]
\vspace{-15pt}
\captionsetup[subfigure]{font=footnotesize}
\centering
\subcaptionbox{\scriptsize{$M_{\alpha_s}$ task automata for $\alpha(s,\tau,t_c)$}\label{fig:fsm:task:alpha}}[0.45\textwidth]{
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=7cm,thick,scale=0.6,every node/.style={scale=0.6},level distance=20mm]
\node[state,initial]
  (a)   {};
\node[state]
  (b) [right of=a]  {};
\path[]
(a)
  edge
  node {$\bfjtt{duration}(\tau,\tau)$}
  (b)
(b)
  edge [bend right,above] 
  node [align=center] 
    {$\bfjtt{if}\big((1-\varepsilon_\alpha(s,\tau,t_c)) > \alpha(s,\tau,t_c)\big) $ \\ \{ %\\
     $P \; ! \; \jtt{allocate}(s,\tau)$ ; %\\
     $P \; ! \; \jtt{verify}_\alpha(s,t_G)$ \}
    }
  (a)
;
\end{tikzpicture}
}%
\subcaptionbox{\scriptsize{$M_{\beta_s}$ task automata for $\beta(s,\tau)$}\label{fig:fsm:task:beta}}[0.45\textwidth]{
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=7cm,thick,scale=0.6,every node/.style={scale=0.6},level distance=20mm]
\node[state,initial]
  (a)   {};
\node[state]
  (b) [right of=a]  {};
\path[]
(a)
  edge 
  node {$\bfjtt{duration}(\tau,\tau)$}
  (b)
(b)
  edge [bend right, above] 
  node [align=center] 
       {$\bfjtt{if}\big((1-\varepsilon_\beta(s,\tau)) > \beta(s,\tau)\big) $ 
       \\ \{ %\\
        $P \; ! \; \jtt{deallocate}(s,\tau)$ ; %\\
        $P \; ! \; \jtt{verify}_\beta(s,t_G)$ \}
       }
  (a)
;
\end{tikzpicture}
}
\vspace{-10pt}
% \caption{{\scriptsize Task Automata for $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$}}
\vspace{-10pt}
\end{figure}

We use a separate task automaton for each service characteristic to verify the SLA of the service after $t_G$ time.
Respectively, $M^{\alpha}_{\jtt{V}}$ and $M^{\beta}_{\jtt{V}}$ execute tasks $\jtt{verify}_\alpha$ and $\jtt{verify}_\beta$ (Figures~\ref{fig:fsm:task:verify:alpha} and \ref{fig:fsm:task:verify:beta}).
$M^{\alpha}_{\jtt{V}}$ uses \bfjtt{await} to ensure the condition of the SLA.
In addition, the task is controlled by the scheduler using a deadline that is specified as $t_G$ in the generated task $\jtt{verify}_\alpha(s,t_G)$ in $M_{\alpha_s}$. 
If $t_G$ passes before the guard statement in \bfjtt{await} statement holds, it leads to a \emph{missed deadline}.
\begin{figure}[h]
\vspace{-20pt}
\captionsetup[subfigure]{font=scriptsize}
\centering
\subcaptionbox{\scriptsize{$M^{\alpha}_{\jtt{V}}$ to execute $\jtt{verify}_\alpha$}\label{fig:fsm:task:verify:alpha}}[0.45\textwidth]{
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=6.5cm,thick,scale=0.6,every node/.style={scale=0.6},level distance=20mm]
\node[state,initial]
  (a)   {};
\node[state,accepting]
  (b) [right of=a]  {};
\path[]
(a)
  edge
  node [align=center] 
    {$\bfjtt{await} \; \alpha(s,\tau,t_c) \geq 1-\varepsilon_\alpha(s,\tau,t_c)$}
  (b)
;
\end{tikzpicture}
}%
\subcaptionbox{\scriptsize{$M^{\beta}_{\jtt{V}}$ to execute $\jtt{verify}_\beta$}\label{fig:fsm:task:verify:beta}}[0.45\textwidth]{
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=6cm,thick,scale=0.6,every node/.style={scale=0.6},level distance=20mm]
\node[state,initial]
  (a)   {};
\node[state,accepting]
  (b) [right of=a]  {};
\path[]
(a)
  edge
  node [align=center] 
    {$\bfjtt{await} \; \beta(s,\tau) \geq 1-\varepsilon_\beta(s,\tau)$}
  (b)
;
\end{tikzpicture}%
}
\vspace{-10pt}
% \caption{{\scriptsize Task automata to execute $\jtt{verify}_\alpha$ and $\jtt{verify}_\beta$}}
\vspace{-10pt}
\end{figure}

Both $M_{\alpha_s}$ and $M_{\beta_s}$ are specific to one particular service $s$.
A generalized automaton for all services is obtained as their parallel composition:$M_\alpha = (\parallel_s M_{\alpha_s})$ and $M_\beta = (\parallel_s M_{\beta_s})$.
The tasks generated by $M_\alpha$ and $M_\beta$ (triggered by the calls to
\jtt{allocate} and \jtt{deallocate}) are executed by the task automata for platform $M_P$.

We model monitoring platform $P$ by a task automata $M_P$.
The task types are $\{\jtt{A}(\jtt{allocate}),\allowbreak \jtt{D}(\jtt{deallocate})\}$.
For task type \jtt{A} in $M_P$, we use $(b,w,d) = (t_i,\tau,\tau)$; i.e.
the best-case execution time of a task is the resource initialization time, 
the worst-case is the length of the monitoring window, and
the deadline is the length of the monitoring window.
For task type \jtt{D} in $M_P$, we use $(b,w,d) =(0, \tau,\tau)$. 
We do not fix the scheduling policy \textsf{Sch}.
The error state $q_{err}$ in $M_P$ is defined when either a deadline is missed or when the platform fails to provision a resource.
Thus the monitoring platform $P$ contains the following ingredients:
\[
M_P = \langle M_{\jtt{A}} \parallel M_{\jtt{D}} \parallel 
M^{\alpha}_{\jtt{V}} \parallel M^{\beta}_{\jtt{V}}
, \textsf{Sch}, \tau \rangle
\]
We define $M_{\jtt{A}_s}$ as the timed automata to execute the tasks of type \jtt{allocate} in $M_P$.
We use the model  semantics presented in \cite{jaghoori2010time} p. 92 to design $M_{\jtt{A}_s}$. The resulting automata is presented in Figure~\ref{fig:fsm:alloc_s}.

\begin{figure}[h]
\vspace{-20pt}
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm,thick,scale=0.7,every node/.style={scale=0.7}]
\node[state,initial]    
  (a)   {};
\node[state]
  (b) [right of=a]   {};
\node[state,accepting]
  (c) [right of=b]  {};
\path[]
(a) 
    edge    
    node {$\bfjtt{duration}(t_i,\tau)$}          
    (b)
(b) 
    edge    
    node {$\sigma(s) \gets \sigma(s) + 1$} 
    (c)
;
\end{tikzpicture}
\vspace{-15pt}
\caption{{\scriptsize{$M_{\jtt{A}_s}$: Timed Automaton to execute task type \jtt{allocate} in $M_P$}}
\vspace{-20pt}
\label{fig:fsm:alloc_s}}
\end{figure}

Then, we define $M_{\jtt{A}}$ in $M_P$ as:
$M_{\jtt{A}} = \; \parallel_s M_{\jtt{A}_s}$; i.e.
the composition of all timed automata to execute a task \jtt{allocate} for some service $s$.
% 
Similarly, we design $M_{\jtt{D}_s}$ to execute task type \jtt{deallocate} in Figure~\ref{fig:fsm:dealloc_s}.
Therefore, we also have $M_{\jtt{D}}$ in $M_P$ as: $M_{\jtt{D}} = \; \parallel_s M_{\jtt{D}_s}$.
% 
\begin{figure}[h!]
\vspace{-10pt}
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm,thick,scale=0.7,every node/.style={scale=0.7}]
\node[state,initial]    
  (a)   {};
\node[state]
  (b) [right of=a]   {};
\node[state,accepting]
  (c) [right of=b]  {};
\path[]
(a) 
    edge    
    node {$\bfjtt{duration}(t_i,\tau)$}          
    (b)
(b) 
    edge    
    node {$\sigma(s) \gets \sigma(s) - 1$} 
    (c)
;
\end{tikzpicture}
\vspace{-15pt}
\caption{{\scriptsize{$M_{\jtt{D}_s}$: Timed Automaton to execute task type \jtt{deallocate} in $M_P$}}
\vspace{-20pt}
\label{fig:fsm:dealloc_s}}
\end{figure}
%

For a particular service $s$, its automaton $M_{\alpha_s}$ regularly measures the service
characteristics and calculates $\alpha(s,\tau,t_c)$. When $s$ is under-capacity, $M_{\alpha_s}$ requests to \jtt{allocate} a new resource for $s$ through monitoring platform $P$.
This generates a new task in $M_P$ that is executed by $M_{\jtt{A}_s}$.
When the task completes, the state of the service $\sigma(s)$ is updated; strictly increased.
Thus, in isolation, the combination of $M_{\alpha_s}$ and $M_{\jtt{A}_s}$ increase the value of service availability $\alpha(s,\tau,t_c)$ for service $s$ over time.
Similarly, in isolation, the combination of $M_{\beta_s}$ and $M_{\jtt{D}_s}$ increase the value of service budget compliance $\beta(s,\tau)$ for service $s$ over time. 
Because in the latter, \jtt{deallocate} is used to decrease the cost of the service and as such increases $\beta(s,\tau)$.

In reality, resources might fail in the environment.
The failure of a resource is not and cannot be controlled by the monitoring platform $P$.
However, the failure of a resource affects the state of a service and its characteristics.
Thus, we model the environment, including failures, as an additional timed automata, $M_E$.
In $M_E$, in every monitoring window, there is a probability that some resources fail.
For example, we present a particular instance of $M_E$ in Figure~\ref{fsm:env}.
In this environment, in every monitoring, an unspecified constant ($c$) number of resources fail.

\begin{figure}[h]
\vspace{-20pt}
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=5cm,thick,scale=0.7,every node/.style={scale=0.7}]
\node[state,initial]
  (a) {};
\node[state]
  (b) [right of=a] {};
\node[state]
  (c) [right of=b] {};
\path[]
(a)
  edge
  node {$\bfjtt{duration}(0,\tau)$}
  (b)
(b)
  edge
  node {$\sigma(s)\gets\sigma(s)-c$}
  (c)
;
\end{tikzpicture}
\vspace{-13pt}
\caption{\scriptsize{An example behavior for $M_E$}}
\label{fsm:env}
\vspace{-20pt}
\end{figure}

We define system automata~\cite{jaghoori2010time} (p. 33, Definition 3.2.7) for each service characteristic;
$\mathcal{S}_\alpha$ for $\alpha(s,\tau,t_c)$ and $\mathcal{S}_\beta$ for $\beta(s,\tau)$:
\[
\mathcal{S}_\alpha = M_\alpha \parallel M_E \parallel M_P
\quad\quad\textit{ and }\quad\quad
\mathcal{S}_\beta = M_\beta \parallel M_E \parallel M_P
\]

With the above automata that we designed for $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$, we are now ready to present the main results.
%

\begin{thm}
\label{thm:alpha}
If the SLA for service $s$ on $\alpha(s,\tau,t_c)$ is violated, either:
\begin{itemize}
\item $\mathcal{S}_\alpha$ re-establishes the condition $\alpha(s,\tau,t_c) \geq 1 - \varepsilon_\alpha(s,\tau)$ (thereby satisfying the SLA) within $t_G$ time, or, 
\item there exists at least one task $\jtt{verify}_\alpha$ in $M^{\alpha}_{\jtt{V}}$ with a missed deadline.
\end{itemize}
\end{thm}

\begin{proof}
At any given time in $T$:
\begin{itemize}
\item If $\alpha(s,\tau,t_c) \geq 1-\varepsilon_\alpha(s,\tau)$, then the SLA for service availability $\alpha$ is satisfied. 
\item If the above condition does not hold, on every monitoring window $\tau$, $M_\alpha$ generates a new task \jtt{allocate} in $M_{\jtt{A}}$. 
In addition, a new task $\jtt{verify}_{\alpha}$ is generated with a deadline $t_G$. 
After a duration of $t_G$, the \bfjtt{await} statement allows $M^{\alpha}_{\jtt{V}}$ to complete the task $\jtt{verify}_\alpha$ only if the condition $\alpha(s,\tau,t_c) \geq 1-\varepsilon_\alpha(s,\tau)$ holds.
If this is not the case, since $t_G$ has passed, the scheduler generates a missed deadline (moving to its error state). 
\end{itemize}
\end{proof}

\begin{thm}
\label{thm:beta}
If the SLA for service $s$ on $\beta(s,\tau)$ is violated, either:
\begin{itemize}
\item $\mathcal{S}_\beta$ re-establishes the condition $\beta(s,\tau) \geq 1-\varepsilon_\beta(s,\tau)$ (thereby satisfying the SLA) within $t_G$ time, or,
\item there exists at least one task $\jtt{verify}_\beta$ in $M^{\beta}_{\jtt{V}}$ with a missed deadline. 
\end{itemize}
\end{thm}

\begin{proof}
Similar to the proof of Theorem~\ref{thm:alpha}.
\end{proof}

In practice, the guarantee of $\mathcal{S}_\alpha$ and $\mathcal{S}_\beta$
in isolation to eventually evolve the system to satisfy the SLA is not enough.
In reality, a service provider tries ensure both simultaneously to reduce their cost of service delivery while ensuring the delivered service is of the expectations agreed upon with the customer.
However, these goals conflict.
When $\alpha(s,\tau,t_c)$ increases because of adding a new resource,
it means that service $s$ costs more, hence $\beta(s,\tau)$ decreases.
%Under the same $B(s,\tau)$, this means that $\beta(s,\tau)$ decreases.
The same applies in the other direction: increasing $\beta(s,\tau)$
negatively affects $\alpha(s,\tau,t_c)$.

To capture the combined behavior of service availability and budget compliance,
we compose them. 
We define \emph{service sustainability} $\gamma(s,\tau)$ as the composition of $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$.
We present the composition by system automata $\mathcal{S}_\gamma$ as:
\[
\mathcal{S}_\gamma = \mathcal{S}_\alpha \; \parallel \; \mathcal{S}_\beta
\]

Authors in~\cite{fersman2007task} define that a task automata is \emph{schedulable} if there exists no task on the queue that misses its deadline. 
The next theorem presents the relationship between schedulability analysis of service sustainability and satisfying its SLA.

\begin{thm}
\label{thm:gamma}
If $\mathcal{S}_\gamma$ is \emph{schedulable} given input parameters $(\tau, t_i, t_G)$, then the SLA for both service characteristics $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$ is satisfied within $t_G$ time after a violation.
\end{thm}

\begin{proof}
When a violation of the SLA occurs in $\mathcal{S}_\gamma$, either $\mathcal{S}_\alpha$ or $\mathcal{S}_\beta$ (or both) start to evolve the service based on Theorems~\ref{thm:alpha} and \ref{thm:beta}.
Therefore, there exists at least one task of $\jtt{verify}_\alpha$ or $\jtt{verify}_\beta$ with a deadline $t_G$.
% 
Hence, if $\mathcal{S}_\gamma$ is schedulable, then neither $\jtt{verify}_\alpha$ nor $\jtt{verify}_\beta$ miss their deadline.
Thus, both $\mathcal{S}_\alpha$ and $\mathcal{S}_\beta$ are schedulable.
This means that both $\jtt{verify}_\alpha$ and $\jtt{verify}_\beta$ complete successfully.
Therefore, the SLA of the service is guaranteed within $t_G$ after a violation in $\mathcal{S}_\gamma$.
\end{proof}

Using the algorithm presented in Chapter 6~\cite{jaghoori2010time}, we translate the above task automata into traditional timed automata.
This allows to leverage well-established model checking techniques such as UPPAAL~\cite{uppaal2004} to determine the schedulability of $\mathcal{S}_\gamma$.
Moreover, the results of the schedulability analysis serves as a method to optimize the input parameters of the monitoring model including $\tau$ and $t_G$.

### 2015-10-20 p03-ch01-2015-monitoring/sec06-impl.tex
Initial Commit

% -*- root: paper-esocc.tex -*-

\section{Evaluation of the monitoring model} % (fold)
\label{sec:implementation}

In this section, we evaluate the implementation of the monitoring model.
% that we presented in the previous sections.

We set up an environment to evaluate how the monitoring evolves a service according to its SLA. 
In the environment, a single instance of monitoring platform is present to provide new resources as necessary.
Every resource hosts only one service.
We define two customers in the environment.
For both customers, we deploy the same service, Fredhopper Query API.
For every resource that hosts a service, we set up a monitor that measures \qps and reports it to the platform.
Both customers run with the same SLA: the \qps expectation is $E(s,\tau,t_c) = 10$ and $\varepsilon_\alpha(s,\tau,t_c) = 0.1$.
We launch every customer service with only one resource.
Monitors observe the customer service and calculate the service availability of every customer service $\alpha(s,\tau,t_c)$.

We run the environment setup for different monitoring windows $\tau \in \{1,5,10\}$ (seconds).
We fix the initialization time of a resource to $t_i = 2.5$ seconds.
We set $t_G = 300$ seconds; i.e. we verify the service after this time and evaluate if the service is guaranteed based on its SLA.

Figure~\ref{fig:alpha} plots the service availability $\alpha(s,\tau,t_c)$ over time with the different monitoring windows.
The following summarizes the behavior:
\begin{itemize}
\item As the monitoring window $\tau$ increases, the system converges with a slower pace towards the expected $\alpha(s,\tau,t_c)$.
\item When the monitoring window is chosen such that $\tau < t_i$, the evolution of the system becomes \emph{non-deterministic}.
\item The setting $\tau < t_i$ causes a missed deadline in $\jtt{verify}_\alpha$ because after a duration of $t_G$ the service availability has not yet reached the expected value.  
\end{itemize}

% 
\begin{wrapfigure}{r}{0.6\textwidth}
\vspace{-40pt}
\begin{center}
\includegraphics[scale=0.6]{figs/alpha.eps}  
\end{center}
\vspace{-20pt}
\caption{{\scriptsize Evolving $\alpha(s,\tau,t_c)$ with different $\tau$}}
\label{fig:alpha}
\vspace{-20pt}
\end{wrapfigure}
%
Every monitoring measurement is performed in a monitoring window $\tau$.
Monitoring measurements are aggregated and calculated in every window and form the basis of reactions necessary to evolve the service to meet their SLA.
Thus, selection of an appropriate monitoring window length $\tau$ is crucial, as we also discussed how schedulability analysis can be used to optimize it.
The authors in \cite{hogben2013defavail} present that for the same setup and deployment of services, measurements using different monitoring windows yield to very different understanding of service properties such as service availability.
Therefore, it is essential to choose the value of $\tau$ such that monitoring measurements do not lead to \emph{unrealistic} understanding and inappropriate reactions.

If $\tau < t_i$, Theorem \ref{thm:alpha} does not hold because every task \jtt{allocate} in $M_{\jtt{A}}$ misses its deadline.
Thus, it is essential that $\tau \geq t_i$.
Analogously, choosing monitoring window as $\tau \gg 2 \times t_i$ also has a counter-productive effect on the service deployments.
% 
% Assumption \ref{def:single:resource:type} says all the resources are of the same type and as such $t_i$ is the same for all resources.
In a real setting, different services may use different types of resources.
In such a setting, the monitoring window should be chosen as the largest $t_i$ of any resource type that is available in the platform: $\tau \geq \mathsf{max}(t_i) \; \forall r \in P$.

% section implementation (end)

### 2015-10-20 p03-ch01-2015-monitoring/sec07-conclusion.tex
Initial Commit

% -*- root: paper-esocc.tex -*-

\section{Future work}
\label{sec:conclusion}

%In this paper, we introduced a distributed monitoring model in order to evolve a service towards its SLA. 
%We formalized characteristics of a distributed service: service availability $\alpha(s,\tau,t_c)$ and service budget compliance $\beta(s,\tau)$.
%We designed task and timed automata for each characteristic to react to the changes of the service in the environment to re-establish its SLA.
%% We presented theorems how SLA for service availability and budget compliance can be met.
%We composed the two characteristics as service sustainability.
%% We showed how the schedulability analsysis of service sustainability builds a model.
%The monitoring model can be used to analyze, model check, and optimize the input parameters to the system (such as $\tau$ and $t_G$), leveraging well-established model checking and schedulability analysis techniques.
%We evaluated the monitoring model on the SDL Fredhopper cloud services.
%We discussed how the results of the evaluation confirm the theorems presented in Section~\ref{sec:timed:fsm} from a practical standpoint.

% Using the formal model of the service characteristics, we design a monitoring model using actors. 
% Each monitor in the system is responsible for one service.
% Monitors use asynchronous communication through Real-Time ABS to interact with the platform API.
% The monitors observe the system and react to the changes or expectations of the properties. 
% They make changes to the system (\jtt{allocate} or \jtt{deallocate}) through Platform API to ensure the system satisfies its SLA.
% We presented an evaluation of the implementation of the  monitoring platform.
% We showed how choosing certain parameters in the system influences the evolution of the services at runtime.

% There is a wealth of future work for this research.
We continue to generalize the notion of the distributed service characteristics and investigate how the composition of an arbitrary number of such properties can be formalized and reasoned about.
In the context of the ENVISAGE project, industry partners define their service characteristics in this framework and monitor the service evolution.
Moreover, the work will be extended to generate parts of the monitoring platform based on an input of different SLA formalizations such as SLA$\star$~\cite{kearney2010sla}.
Currently, we are integrating our automated monitoring infrastructure
into the in-production SDL Fredhopper cloud services (cf. Section~\ref{sec:fredhopper_example}).

### 2015-10-20 p03-ch01-2015-monitoring/sec55-app.tex
Initial Commit

% -*- root: paper.tex -*-

\section{Applications and Future Work} % (fold)
\label{sec:application}

\subsection{Code Generation for Monitors}

The parameters $E(s,\tau)$ and $B(s,\tau)$ are significantly driven from a business perspective.
Such parameters enforce business-level constraints on the services that are also bound from a legal perspective.
In the past decade, there is considerable research on how to specify and model such business constraints generally referred to as service level agreements (SLA).
A number of languages, frameworks and models have been developed to be able to model and specify SLAs for the deployment of services.
Among others, we refer to SLA$\star$~\cite{kearney2010sla}, WSLA~\cite{keller2003wsla}, and SLAng~\cite{lamanna2003slang}.
The focus of this research is not to enter this domain and yet build on top of it a model to be able to reason about ``distributed service properties'' such as service availability or budget compliance. 
In the following, we present a set of pseudo algorithms for the model in Section \ref{sec:modelling} that present how the monitoring can be generated based on the agreements that are specified
\footnote{This is a work in progress in ENVISAGE project WP2}.

In Algorithm \ref{alg:gen:monitor}, the monitoring actor method \jtt{check} is presented that collects the measurements and publishes them to the platform.

% 
\begin{algorithm}
\caption{Monitor for service $s$}
\label{alg:gen:monitor}
\begin{algorithmic}
\Function{check}{}  
  \For{$r \in \sigma(s)$}
    \State $p \; ! \; \jtt{publish}(\mu(s,r,\tau))$
    \State $\bfjtt{this} \; ! \; \jtt{check}() \; \bfjtt{duration}(\tau)$
  \EndFor
\EndFunction
\end{algorithmic}
\end{algorithm}
%

When platform receives a measurement, it updates its state, computes the service properties and initiates an asynchronous message to process and react to them.
This is presented in Algorithm \ref{alg:publish}.

% 
\begin{algorithm}
\caption{Process measurements}
\label{alg:publish}
\begin{algorithmic}
\Function{publish}{$\mu$}
  \State $\alpha(s,\tau,t_c) \gets$ \Call{compute$_{\alpha}$}{}
  \State $\beta(s,\tau) \gets$ \Call{compute$_{\beta}$}{}
  \State Update $\sigma(s)$ and $h(s)$
  \State $\bfjtt{this} \; ! \;$ \Call{process}{$\mu,\alpha(s,\tau,t_c),\beta(s,\tau)$}
\EndFunction
\end{algorithmic}
\end{algorithm}
%

As a consequence of publishing measurements, the platform react to them in case they do not match the expectations.
The behavior is discussed in Theorem \ref{thm:sustain:conv}.
Algorithm \ref{alg:process} presents the pseudo code for the behavior that makes changes to $M_\alpha$ and $M_\beta$.

% 
\begin{algorithm}
\caption{$M_\alpha$ and $M_\beta$}
\label{alg:process}
\begin{algorithmic}
\Function{process}{$\mu,\alpha(s,\tau,t_c),\beta(s,\tau)$}
  \If{$\alpha(s,\tau,t_c) < 1 - \varepsilon_\alpha$}
    \State $n_r \gets$ \Call{compute$_{n_r}$}{}
    \State $r \gets p \; ! \; \jtt{allocate}(s, n_r)$
  \EndIf
  \If{$\beta(s,\tau) - B(s, \tau) < 1 - \varepsilon_\beta$}
    \State $p \; ! \; \jtt{deallocate}(s)$
  \EndIf
  \State Update $\sigma(s)$ and $h(s)$
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Monitoring Window $\tau$}

Every monitoring measurement is performed in a monitoring window $\tau$.
Monitoring measurements are aggregated and calculated in every window and form the basis of reactions and corrections that are necessary to balance the running services to meet their expectations.
The selection of an appropriate monitoring window length $\tau$ is crucial to manage the deployments of the services according to their agreements. 
The authors in \cite{hogben2013defavail} present that for the same setup and deployment of services, measurements using different monitoring windows yield to very different understanding of service properties such as service availability.
Therefore, it is essential to choose the value of $\tau$ such that monitoring measurements do not lead to unrealistic understanding and inappropriate reactions.
To this end, the following should be considered when selecting a monitoring window $\tau$:
\begin{itemize}
\item We defined $\hat{t}$ to be the time a resources needs to initialized in the platform.
Under Assumption \ref{def:resource:init:time}, all resources require same amount of time to initialize.
If $\tau < \hat{t}$, Theorem \ref{def:service:availability} invalidates because the measurements are done in a time window that is not enough to be able to take the newly allocated resource into account for a service.
Thus, it is essential that $\tau \geq \hat{t}$.
It can be similarly argued that choosing monitoring window as $\tau \gg 2 \times \hat{t}$ also has a counter-productive effect on the service deployments.
Widening the monitoring window causes to lose fine granularity of monitoring measurements to be able to react to them in proper time.
\item For simplicity of reasoning, we considered Assumption \ref{def:single:resource:type};
i.e. all the resources provisioned by the platform are of the same type and as such $\hat{t}$ is the same.
However, different services may fit into different types or resources and using different types of resource may fit better for budget compliance as they cost differently.
In such setting, the monitoring window should be at least the maximum $\hat{t}$ of any resource type that is available in the platform: $\tau \geq \mathsf{max}(\hat{t}) \; \forall r \in p$; otherwise, Theorem \ref{thm:service:budget:conv} violates.
\end{itemize}

% section applications_and_discussion (end)

### 2015-10-20 p03-ch01-2015-monitoring/sec99-appendix.tex
Initial Commit

\appendix
\label{appendix}

\newpage

\section{Appendix: Previous write-ups}

\Behrooz{What follows may be obsolete. The content will be used/removed gradually. Do not remove.}

\subsection{Operational Semantics} % (fold)
\label{sub:operational_semantics}
We use operational semantics to present the semantics of the runtime monitoring. 
The global configuration is
$$
\mathcal{C} = \langle \mathcal{S}, \mathcal{A}, \mathcal{M}, P \rangle
$$
which respectively composes of the services, agreements, monitors, and platform.
Note that there is only one platform in this configuration.
For any actor in the system, we use a local configuration as $C = \langle Q, \sigma \rangle$ in which $Q$ is the queue of the messages of the actor and $\sigma$ is a local state that the actor may use for its responsibilities.

\paragraph{Instantiate a service} 
When a new service $s$ is instantiated, the global configuration makes a transition:
\[
\langle \mathcal{S}, \mathcal{A}, \mathcal{M}, P \rangle
\xrightarrow[m = \bfjtt{new} \jtt{ Monitor}(s, a)]{s=p \; ! \; \jtt{instantiate}(\jtt{name}, a)}
\langle \mathcal{S'}, \mathcal{A'}, \mathcal{M'}, P \rangle
 \\
\]
where $\mathcal{S'} = \mathcal{S} \cup \{s\}$, $\mathcal{A'} = \mathcal{A} \cup \{a\}$, and $\mathcal{M'} = \mathcal{M} \cup \{m\}$.
And, \jtt{name} is the name of the new service $s$ and $m$ are the monitor(s) established for $s$.

\paragraph{Monitor a service}
A monitor actors performs checks based on the agreements for a specific service.
The completion of a check on a service can yield an instance of measurement.
\[
\langle \mathcal{S}, \mathcal{A}, \mathcal{M}, P \rangle
\xrightarrow[p \; ! \; \jtt{publish}(\omega)]{\omega = m \; ! \; \jtt{check}()}
\langle \mathcal{S}, \mathcal{A}, \mathcal{M}, P' \rangle
\]
where monitoring platform $P$ state $\sigma$ is updated using the measurement: $\sigma_{P'} = \sigma_{P} \pm \omega$.

\paragraph{Allocate a new resource}
The platform instance $p$ is able to allocate a new resource $r$ to service $s$:
\[
\langle \mathcal{S}, \mathcal{A}, \mathcal{M}, P \rangle
\xrightarrow[]{r = p \; ! \; \jtt{allocate}(s) \; \bfjtt{duration}(t,t)}
\langle \mathcal{S}, \mathcal{A}, \mathcal{M'}, P' \rangle
\]
where $r$ is either the resource that is allocated to service $s$ or carries a failure for the allocation.
The allocation of a resource can be attributed with a deadline $t$ using ABS semantics where $t$ is the amount of time to expect $r$ is ready or failed.
If the operation succeeds, $\sigma_{P'} = \sigma_{P} \pm R(s) \cup \{r\}$ where $R(s)$ is the set of resources allocated to $s$.
% subsection operational_semantics (end)

% section modelling (end)

\subsection{A Timed Automata for Budget Compliance}
\label{sec:prop:budget}

Similar to Section \ref{sec:service:avail:fsm}, we use a timed automata (see Definition \ref{def:timed:automata}) to model the behavior of service budget compliance $\beta(s,\tau)$ (see Definition \ref{def:budget:compliance}) represented in Figure \ref{fig:fsm:budgetcomp}.
% 
\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm,thick]
\node[state,accepting]        (A)   {$q_B$};
\node[state]  [right of=A]    (N)   {$q_{\overline{B}}$};
\path
(A) edge    [loop left, left]  node    {$\delta_M^1=(\bot,g_B,\tau)$}     (A)
(A) edge    [bend right, below]       node    {$\delta_M^2=(\dot{a},g_{\overline{B}},\tau)$}     (N)
(N) edge    [loop right, right]  node    {$\delta_M^3=(\dot{a},g_{\overline{B}},\tau)$}     (N)
(N) edge    [bend right, above]       node    {$\delta_M^4=(\bot,g_B,\tau)$}     (A)
;
\end{tikzpicture}
\caption{A timed automata for Budget Compliance $\beta(s,\tau)$}
\label{fig:fsm:budgetcomp}
\end{figure}
%

$M_\beta$ has two states, $Q = \{q_B, q_{\overline{B}} \}$.
In comparison, $M_\beta$ is simpler than $M_\alpha$.
Computation of the cost of a service is independent of whether the service is faulty or not responsive.
Thus, $M_\beta$ does not have a state of $q_e$.
The invariants that guard transitions in $M_\beta$ are as follows:
\begin{align}
g_B \; &: \; 0 \geq \beta(s,\tau) < \varepsilon_{\euro} \\
g_{\overline{B}} \; &: \; |\beta(s,\tau)| > \varepsilon_{\euro}
\end{align}
The transitions of $M_\beta$ can be described similar to that of $M_\alpha$.
The different is in the action $\dot{a}$ produced.
An action in $M_\beta$ is a message to monitoring platform $p$ that deallocates a resource from services $s$:
\[
p \; ! \; \jtt{deallocate(s)}
\]
that terminates one resource allocated to $s$ from platform $p$.
This action is immediate and never fails.

\section{Distributed Service Sustainability} % (fold)
\label{sec:distributed_service_sustainability}
\begin{defn}[Service Sustainability $\gamma(s,\tau)$]
\label{def:service:sustain}
We define the service sustainability as:
\[
\gamma(s,\tau) = \alpha(s,\tau,t_c) \times \beta(s,\tau)
\]
that is the composition of service availability $\alpha(s,\tau,t_c)$ and budget compliance $\beta(s,\tau)$ over the time span $\tau$.
Note that under a fixed agreement $a$ for $s$, $\alpha(s,t)$ and $\beta(s,t)$ have an inverse relation to one another:
\[
\alpha(s,t) \uparrow \iff \beta(s,t) \downarrow
\]
because consuming resources for service $s$ for $\alpha$ mean more cost for $s$ under the same agreement $a$ that leads to lower $\beta$ (and vice versa).
\end{defn}

% section distributed_service_sustainability (end)

\begin{theorem}[Budget Compliance Convergence]
\label{thm:budget:comp:conv}
The automata $M_\beta$ on budget compliance $\beta(s,\tau)$ makes a transition to state $q_B$ in a finite amount of time under Assumptions \ref{def:single:resource:type} and \ref{def:resource:cost:constant} given that $\tau \geq \hat{t}$ and agreement $a$ for service $s$ is fixed and not changing through time.
\end{theorem}

\begin{proof}[Budget Compliance Convergence]
Proof can be completed using same technique in Theorem \ref{thm:service:avail:conv}.
\newline$\Box$
\end{proof}

\subsection{A Timed Automata for Service Sustainability}
\label{sec:sustain}

The timed automata $M_\gamma$ for $\gamma(s,\tau)$ is constructed by $M_\alpha \times M_\beta$.
$M_\gamma$ has two states, $Q = \{ q_S, q_{\overline{S}}\}$.
$M_\gamma$ enters its final state $q_S$ when $M_\alpha$ is in $q_A$ and $M_\beta$ is in $q_B$.
$M_\gamma$ enters its non-final state $q_{\overline{S}}$ when either $M_\alpha$ is not in $q_A$ or $M_\beta$ is not in $q_B$.

\begin{theorem}[Sustainable Service Convergence]
\label{thm:sustain:conv}
The sustainability of service $s$ converges to:
$$
\lim_{T \to \infty} \; \gamma(s, \tau) = (E(s,\tau) - \varepsilon) \times (E_B(s,\tau) - \varepsilon_{\euro})
$$
if and only if, agreement $a$ for service $s$ is \emph{modifiable} from a monitoring window $\tau$ to the next given that either $\alpha(s,\tau,t_c)$ or $\beta(s,\tau)$ fails in one $\tau$.
\end{theorem}

\begin{proof}[Sustainable Service Convergence]
In other words, $M_\gamma$ remains in state $q_S$ as long as both $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$ are compliant.
$M_\gamma$ does not remain in state $q_S$ if $\alpha(s,\tau,t_c)$ or $\beta(s,\tau)$ change under a fixed agreement $a$.
The proof is by contradiction.
We study different cases each of which has a \emph{fixed} element and we conclude a contradiction.

Let $M_\alpha$ be in $q_A$ and $M_\beta$ in $q_B$ at a given time $T$.
At time $T$, if $g_{\overline{A}}$ holds, $M_\alpha$ makes a transition to $q_{\overline{A}}$ and starts to allocate a new resource $r$ for $s$.
At a later time $T' \geq T + \tau$, $M_\alpha$ makes a transition to $q_A$ based Theorem \ref{thm:service:avail:conv}.
However, if $g_{\overline{B}}$ holds at time $T'$, $M_\beta$ makes a transition to $q_{\overline{B}}$ because the newly allocated resource $r$ makes budget compliance $\beta(s,\tau)$ violate.
Based on Theorem \ref{thm:budget:comp:conv}, $M_\beta$ makes a transition to $q_B$ after deallocating at least one resource from $s$ at a later time $T'' \geq T' + \tau$.
At time $T''$, the system recurs the behavior at time $T$ and thus $\gamma(s,\tau)$ for every subsequent $\tau$ cannot converge to $(1-\varepsilon)\times(1-\varepsilon_{\euro})$.
Similarly, it can be reasoned how changes in $M_\beta$ lead to changes in $M_\alpha$ and eventually destabilize $M_\gamma$.

The proof is complete if the modification to be applied to $a$ of $s$ is calculated at a given monitoring window $\tau$.
When $M_\alpha$ makes a transition to $q_{\overline{A}}$, the action is to allocate a new resource $r$ to $s$.
We complement the action by another one.
At time $T$, if pre-calculation of $\euro(s,\tau)$ reveals violation of $\beta(s,\tau)$, we update $B(s,\tau)$ by $B(s,\tau) = B(s,\tau) + \overline{\euro}_r \times \tau$; i.e. the cost difference of a single resource over monitoring window is reflected into service budget.
At time $T'$, $M_\beta$ does not violate because of the newly added resource because the service budget already includes the necessary changes to pass.
Similarly, when $M_\beta$ makes a transition to $q_{\overline{B}}$, the action is to deallocate a resource from $s$.
We complement the action by another one.
At time $T$, if pre-calculation of $\alpha(s,\tau,t_c)$ reveals a violation, we update $E(s,\tau)$ by $E(s,\tau) = E(s,\tau) + \Delta(s,\tau)$; i.e. the availability distance caused by deallocation single resource over monitoring window is reflected into service expectation.
At time $T'$, $M_\alpha$ does not violate because of the adjustment. 
Thus, in either case, when $M_\alpha$ or $M_\beta$ make a transition to an unacceptable state, we are able to make an adjustment to the agreement $a$ of service $s$ such that the convergence of $\gamma(s,\tau)$ during the next monitoring window is preserved. 
\newline
$\Box$
\end{proof}

\section{Application of Actor-based Monitoring Model}
\label{sec:app}

If we expand $\gamma(s,\tau)$, we have:
\begin{align*}
\gamma(s,\tau) &= \alpha(s,\tau,t_c) \times \beta(s,\tau) \\
 &= \frac{|\, \{\; \omega \; | \; 0 \leq v(\omega) - e(s(\omega),a(\omega)) < \epsilon \; \}\,|}{|\{\omega\}|}
 \times \frac{\euro(s,\tau) - B(s,\tau)}{B(s,\tau)} \\
 &=_{T \rightarrow \infty} (E(s,\tau) - \varepsilon) \times (E_B(s,\tau) - \varepsilon_{\euro})
\end{align*}
in which the parameters that can possibly change for service $s$ are:
\begin{description}
\item[Agreement Expectation $e(s, a)$] defines the values that are expected for each agreement $a$ of service $s$.
\item[Service Budget $B(s,\tau)$] defines the limit of the budget that can be spent for service $s$ in a period of time $\tau$.
\item[Service Availability Tolerance $\varepsilon$] defines the accepted deviation of service availability calculated for $s$ over a period of $\tau$.
\item[Budget Compliance Tolerance $\varepsilon_{\euro}$] defines the accepted deviation of budget compliance calculated for $s$ over a period of $\tau$.
\item[Monitoring Window $\tau$] define the time span $\tau$ with which all monitoring checks are performed and based on which measurements, the above properties are calculated and reacted upon.
\end{description}
In the following, we discuss how changes of the above parameters affect service sustainability and how they can be used to synthesize or drive the generation of monitoring model for a service based on its agreements.

\subsection{Tolerance Flexibility $\varepsilon$ and $\varepsilon_{\euro}$}

Theorem \ref{thm:sustain:conv} proved that $\gamma(s,\tau)$ converges to $(E(s,\tau)-\varepsilon)\times(E_B(s,\tau)-\varepsilon_{\euro})$ given certain conditions.
The values of $E(s,\tau)$ and $E_B(s,\tau)$ are bound to $1$.
To simplify, we define $f(\varepsilon,\varepsilon_{\euro}) = (1 - \varepsilon)\times(1-\varepsilon_{\euro})$.
On its own, this function is interesting.
Figure \ref{fig:tolerance:func} presents the landscape of this function with $\varepsilon$ and $\varepsilon_{\euro}$ as its variables.
%

The following summarizes the characteristics of the function and how it affects $\gamma(s,\tau)$:
\begin{itemize}
\item The following property holds:
\[
(1-\varepsilon) \times (1-\varepsilon_{\euro}) = 1 \iff \varepsilon = 0 \;\lor\; \varepsilon_{\euro} = 0
\]
This property shows that choosing $0$ value for $\varepsilon$ or $\varepsilon_{\euro}$ makes $\gamma(s,\tau)$ more strict and less tolerant. 
Unless it is not a hard business requirement, the tolerance parameters must allow for room of error for either $\alpha(s,\tau,t_c)$ or $\beta(s,\tau)$; otherwise, the sustainability $\gamma(s,\tau)$ exposes more fragility.
\item Similarly, choosing $\varepsilon = 1$ or $\varepsilon_{\euro} = 1$ means that respectively $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$ is \emph{not} a business driver.
In this setting, the service sustainability $\gamma(s,\tau)$ depends on a single property (either $\alpha(s,\tau,t_c)$ or $\beta(s,\tau)$) which makes its maintenance easier.
Additionally, choosing such model allows to design a business in which exclusively either service availability or budget compliance is crucial but not both.
\item Apart from the corner values, the value of the function is \emph{confined} by the arcs that are projected on the ceiling of the box in Figure \ref{fig:tolerance:func}.
This means that the function values changes co-dependently based on the changes of $\varepsilon$ and $\varepsilon_{\euro}$; i.e. optimization of $\gamma(s,\tau)$ is freely achievable by changing each variable.
\item It should be also noted that the values for $\varepsilon$ and $\varepsilon_{\euro}$ should \emph{not} be so relaxed; 
i.e. values higher than $0.1$ would create a large range of error which may not be desired.
\end{itemize}

\subsection{Business Constraints $E_B(s,\tau)$ and $E(s,\tau)$ for Monitor Generation}

The parameters $E(s,\tau)$ and $E_B(s,\tau)$ are significantly driven from a business perspective.
Such parameters enforce business-level constraints on the services that are also bound from a legal perspective.
In the past decade, there is a considerable research on how to specify and model such business constraints generally referred to as service level agreements (SLA).
A number of languages, frameworks and models have been developed to be able to model and specify SLAs for the deployment of services.
Among others, we refer to SLA$\star$~\cite{kearney2010sla}, WSLA~\cite{keller2003wsla}, and SLAng~\cite{lamanna2003slang}.
The focus of this research is not enter this domain and yet build on top it a model to be able to reason about ``distributed service properties'' such as service availability or budget compliance. 
In the following, we present a set of pseudo algorithms for the model in Section \ref{sec:modelling} that present how the monitoring can be generated based on the agreements that are specified. 
In Algorithm \ref{alg:gen:monitor}, the monitoring actor method \jtt{check} is presented that collects the measurements and publishes them to the platform.

% 
% section service_synthesis_based_monitoring_model_parameters (end)

\subsection*{A nice picture}

The top surface (including the arcs) is the 2-D projected values for $\Gamma$.
% 
\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
  view={45}{45},
  colorbar,
  3d box=complete,
  enlargelimits=false,
  domain=0:1,
  xlabel=$\varepsilon_\alpha$,
  ylabel=$\varepsilon_\beta$,
  zlabel=$\Gamma$
]
\addplot3[surf]
  {(1-x) * (1-y)};
\addplot3[contour gnuplot={contour dir=z, labels=false, draw color=black}, thick, z filter/.code=\def\pgfmathresult{1}]
  {(1-x) * (1-y)};
\end{axis}
\end{tikzpicture}
\caption{
The landscape for $\Gamma$.
} 
\label{fig:sustain:factor}
\end{figure}
%

\subsection{Generalization of $\alpha$, $\beta$ and $\gamma$}

For every characteristics of the service, define a predicate function:

\[
F_C(s,\tau, C) \in \{0, 1\}
\]

in which $0$ is \jtt{false}, and $1$ is \jtt{true}.
$C$ denotes the set of properties that is defined for the service $s$ for the predicate function.

For example, in the context of $\alpha$, the predicte function is:

\[
F^\alpha_{C_\alpha}(s, \tau, \{E(s,\tau), \varepsilon_\alpha\}) = \begin{cases}
0  &\;  \alpha(s,\tau,t_c) < 1 - \varepsilon_\alpha \\
1 &\; \textnormal{otherwise}
\end{cases}
\]

We gather the set of all characteristics for a service $s$ by
\[
\mathcal{C} = \{C_1, \ldots, C_n\}
 \]
then for each $C_i$, we define the predicate function $F_{C_i}$.
Similar to $M_\alpha$ and $M_\beta$, we build a timed automata for each $C_i$ using $F_{C_i}$. 
Thus, we build a set of timed automata
\[
\mathcal{M} = \{M_{c_i}, \ldots, M_{c_n}, M_P\}
\]

$\mathcal{M}$ can be partitioned in two \emph{disjoint} subsets, $\mathcal{M}^+$ and $\mathcal{M}^-$, such that $\mathcal{M}^+$ only containts those that use platform $P$ API to \jtt{allocate} a new resource to service $s$ whereas $\mathcal{M}^-$ only contains those that use platform $P$ API to \jtt{deallocate} a resource from service $s$:
\begin{itemize}
  \item $\forall \mathcal{M}_k \in \mathcal{M}^+$ Thoerem \ref{thm:service:avail:conv} holds for $\mathcal{M}_k$.
  \item $\forall \mathcal{M}_k \in \mathcal{M}^-$ Theorem \ref{thm:service:budget:conv} holds for $\mathcal{M}_k$.
  \item $\forall \mathcal{M}_k \in \mathcal{M}^+ \; , \mathcal{M}_l \in \mathcal{M}^- \;, \mathcal{M}_m = \mathcal{M}_k \times \mathcal{M}_l$ Theorem \ref{thm:service:sust} holds for $\mathcal{M}_m$. 
\end{itemize}

\subsection{A Timed Automata for Platform} 
\label{sec:automata:platform}

To model the behavior of the platform $P$ with a timed automata, we use the notion of ``deadline'' from Real-Time ABS~\cite{JaghooriBCS09,bjork2013:rtabs}.
A message is expected to complete (either successfully or with failure) within a specific amount of time, the deadline.
We choose monitoring window $\tau$ as the deadline for any message that is internally processed by platform $P$.
In practice, a deadline for a message is part of the implementation details of any platform API for an IaaS provider.
Figure \ref{fig:fsm:platform} presents a timed automata for platform $P$.

In the timed automata that we design for different elements of the system, the following apply.
Every timed automaton has only one state which is the start and end state.
We use $\dot{c}$ to denote the clock of the timed automaton.
The timed automaton makes a transition at a random time in every monitoring window $\tau$.
A transition may have two forms: \emph{periodic} by the clock or \emph{reactive} through an external event or API call.
For example, $M_P$ has a trasition for \jtt{allocate} a new resource which is reactive to an external platform API call.
And, it has a transition for updating service state $\sigma(s)$ which is periodic and regulated by the clock. 
Every reactive transition has a fixed deadline equal to $\tau$.
A reactive transition is presented with a ``dotted'' line in Figure~\ref{fig:fsm} whereas a periodic transition is presented with a ``solid'' line.
Every transition is presented as $\frac{\textsf{condition}}{\textsf{action}}$.
The \textsf{condition} is a boolean expression which evaluates to either \jtt{true} or \jtt{false}.
The \textsf{action} is an atomic operation or an asynchronous message using Real-Time ABS with a fixed deadline equal to $\tau$.
For every timed automaton, there is an implicit transition as: $\frac{\dot{c} \geq \tau}{\dot{c} \gets 0}$.

Every transition in $M_P$ has a deadline of $\tau$;
it is expected that the result of the action of the transition is completed before the end of the next monitoring window.
In particular, this implies that the state of a service ($\sigma(s)$), and the history of a service ($h(s)$) are updated every monitoring window.

Capturing a failure in the platform is achieved by identifying a resource $r$ for some service $s$ such that the resource is recorded in the history of the service, but is not any more in the current state $\sigma$ of the service.

The platform also receives requests to allocate and deallocate specified by its API. 
All such requests use the same deadline $\tau$.

% 
\begin{figure}[t]
\centering

% M_P
\subcaptionbox{{\small{$M_P$: Timed Automaton for Platform $P$}}\label{fig:fsm:platform}}
{
\begin{tikzpicture}[->,>=stealth',auto,node distance=6cm,thick,scale=0.7,every node/.style={scale=0.7}]
\node[state,accepting]   
    (s)   {{\small$\dot{c} < \tau$}};
% \node[state]   [right of=s]      (e)   {$\dot{c} = t$};
\path[every loop/.style={looseness=7}]
(s) 
    edge    
    [loop left,style=densely dotted]  
    node {$\jtt{allocate}(s,n) \; \bfjtt{deadline}(\tau)$}          
    (s)
(s) 
    edge    
    [loop right,style=densely dotted]   
    node {$\jtt{deallocate}(s) \; \bfjtt{deadline}(\tau)$} 
    (s)
(s)
    edge
    [loop,out=150,in=120,looseness=10]
    node {$\sigma(s) \gets \jtt{getState}(s) \; \bfjtt{deadline}(\tau)$}
    (s)
(s)
    edge
    [loop,in=30,out=60,looseness=10]
    node  {$h(s) \gets \jtt{getHistory}(s) \; \bfjtt{deadline}(\tau)$}
    (s)
(s)
    edge
    [loop above]
    node {$\frac{\forall s\; \exists r: \; r \in h(s) \wedge r \not\in \sigma(s)}{\bfjtt{skip}}$}
    (s)
;
\end{tikzpicture}
}

% M_alpha
\subcaptionbox{{\small{$M_\alpha$: Timed Automaton $\alpha(s,\tau,t_c)$}}\label{fig:fsm:serviceavail}}
{
\begin{tikzpicture}[->,>=stealth',auto,node distance=6cm,thick,scale=0.7,every node/.style={scale=0.7}]
\node[state,accepting]
(s)   
    {{\small$\dot{c} < \tau$}};
\path[every loop/.style={looseness=7}]
(s) 
    edge    
    [loop left]  
    node {$\frac{\alpha_\sigma(s,\tau) < 1 - \varepsilon_\alpha}{P \; ! \; \jtt{allocate}(s,n_r)}$}          
    (s)
(s) 
    edge    
    [loop right]   
    node {$\frac{1 - \varepsilon_\alpha \leq \alpha_\sigma(s,\tau) \leq 1 + \varepsilon_\alpha}{h(s) \gets \sigma(s)}$} 
    (s)
;
\end{tikzpicture}
}

% M_beta
\subcaptionbox{{\small{$M_\beta$: Timed Automaton for $\beta(s,\tau)$}}\label{fig:fsm:budgetcompliance}}
{
\begin{tikzpicture}[->,>=stealth',auto,node distance=6cm,thick,scale=0.7,every node/.style={scale=0.7}]
\node[state,accepting]   (s)   {$\dot{c} < \tau$};
\path[every loop/.style={looseness=10}]
(s) 
    edge    
    [loop left]  
    node {$\frac{\beta(s,\tau) < 1 - \varepsilon_\beta}{P \; ! \; \jtt{deallocate}(s)}$}          
    (s)
(s) 
    edge    
    [loop right]   
    node {$\frac{1 - \varepsilon_\beta \leq \beta(s,\tau) \leq 1 + \varepsilon_\beta}{\sigma(s) \gets h(s)}$} 
    (s)
;
\end{tikzpicture}
}

% M_gamma
\subcaptionbox{{\small{$M_\gamma$: Timed Automaton for $\gamma(s,\tau)$}}\label{fig:fsm:sustain}}
{
\begin{tikzpicture}[->,>=stealth',auto,node distance=6cm,thick,scale=0.7,every node/.style={scale=0.7}]
\node[state,accepting]   (s)   {$\dot{c} < \tau$};
\path[every loop/.style={looseness=10}]
(s) 
    edge    
    [loop left]  
    node {$\frac{\alpha_\sigma(s,\tau) < 1 - \varepsilon_\alpha}{P \; ! \; \jtt{allocate}(s,n_r)}$}          
    (s)
(s) 
    edge    
    [loop right]   
    node {$\frac{1 - \varepsilon_\alpha \leq \alpha_\sigma(s,\tau) \leq 1 + \varepsilon_\alpha}{h(s) \gets \sigma(s)}$} 
    (s)
(s) 
    edge    
    [loop,in=30,out=60,looseness=10]  
    node {$\frac{\beta(s,\tau) < 1 - \varepsilon_\beta}{P \; ! \; \jtt{deallocate}(s)}$}          
    (s)
(s) 
    edge    
    [loop,in=120,out=150,looseness=10]   
    node {$\frac{1 - \varepsilon_\beta \leq \beta(s,\tau) \leq 1 + \varepsilon_\beta}{\sigma(s) \gets h(s)}$} 
    (s)
;
\end{tikzpicture}
}

\caption{Timed automata for elements of the monitoring platform}
\label{fig:fsm}
\end{figure}

\subsection{A Timed Automata for Service Availability}
\label{sec:service:avail:fsm}

We use timed automata~\cite{alur:1994:timedautomata} $M_\alpha$ to manage service availability (see Definition \ref{def:service:availability}).
$M_\alpha$ is designed to specifically address an \emph{under-capacity} service:

The timed automata $M_\alpha$ has two \emph{possible} transitions.
%A transition is chosen according to the condition check. 
If the condition evaluates to \jtt{true}, the corresponding action is taken and the clock is reset.
An action in $M_\alpha$ is an \emph{asynchronous message} in the monitoring model in ABS or an update to $\sigma(s)$ or $h(s)$.
Below we explain the possible conditions and associated actions:

\textbf{Available Service}
  When service $s$ is available, the action is to update the history of the service to record the number of resources allocated to $s$. 
  Specifically, $h(s)$ is the last observed number of resources that ensured $s$ was available.
  This is captured in the right transition in $M_\alpha$.

\textbf{Under-capacity Service}
  When service $s$ is not available, the action is to allocate \emph{at least} one resource.
  This strategy can be improved since we captured the number of resources that
  most recently ensured availability ($h(s)$). Hence, the number of resources to allocate to service $s$, can be calculated as:
  \[
  n_r = 
  \begin{cases}
    \mathsf{max}(1, h(s)-\sigma(s)) &\; h(s) > 1 \\
    1 &\; \text{otherwise}
  \end{cases}
  \]
  Intuitively, when there is no history for $s$, \emph{one} resource is allocated.
  As soon as $h(s)$ is recorded, the required number of resources is the difference between the current number of resources and the last number that ensured $s$ is available.
  This is captured in the left transition in $M_\alpha$.

The automaton $M_\alpha$ guarantees the following important property:
% present an interesting property of $M_\alpha$ in the following theorem.
%

\begin{defn}[Service Availability Convergence]
\label{thm:service:avail:conv}
Given that: 
\begin{itemize}
  \item platform $P$ provides resources with no limits (Assumption~\ref{def:platform:cap}), and,
  \item monitoring window $\tau$ is chosen such that it's larger that any resource initialization time ($\tau > \hat{t}$) (Assumption~\ref{def:resource:init:time}), 
  \item the number of failed resources
  %in every monitoring window $\tau$
  is bounded by a constant, 
\end{itemize}
then, the timed automata $M_\alpha$ for service availability $\alpha(s,\tau,t_c)$ ensures that service $s$ is available in a finite amount of time.
\end{defn}

\begin{proof}%[Service Availability Convergence]
By induction on $n_r$.
% $\Box$
\end{proof} % end thm:service:avail:conv

\subsection{A Timed Automata for Service Budget Compliance}
\label{sec:automata:budget}

Analogously, we define a timed automata $M_\beta$ for service budget compliance (Definition~\ref{def:budget:compliance}). The automaton is shown in Figure~\ref{fig:fsm:budgetcompliance}.  $M_\beta$ is specifically designed to address a service that is \emph{over-budget}.
%

$M_\beta$ has three possible condition with corresponding actions:

\textbf{Budget Compliant Service}
  When service $s$ is budget-compliant, the action is to update the history of the service to record the current number of resources allocated to $s$. 
  Formally, $h(s)$ denotes the last observed number of resources that ensured $s$ was budget-compliant.

\textbf{Over-budget Service}
  When service $s$ is over-budget and not compliant, the action is to deallocate \emph{one} resource.
  We do not maximize deallocation to prevent sudden service unavailability.
  % \item[Failure in Platform]
  % We capture failures in the platform $P$ by deallocating a resource that is known to be allocated to service $s$ but is not recognized as operational by the platform itself.
  % There is no reaction to platform failures for budget compliance, because when a resource is deallocated from a service, it increases the budget compliance per cut-down on cost.
  % Platform failures \emph{can} lead to a reaction in the previously discussed automaton $M_\alpha$ to ensure service availability.

The above definition of $M_\beta$ guarantees the following important property:
%exposes an interesting property that is presented in the following theorem.

\begin{defn}[Service Budget Compliance Convergence]
\label{thm:service:budget:conv}
The timed automata $M_\beta$ for service budget compliance $\beta(s,\tau)$ ensures that service $s$ is budget compliant in a finite amount of time.
\end{defn}

\begin{proof}%[Service Budget Compliance Convergence]
Similar to that of Theorem~\ref{thm:service:avail:conv}.
\end{proof}

\subsection{Service Sustainability}
\label{sec:service:sustain}

In the previous two sections, we modelled service availability $\alpha(s,\tau,t_c)$ and service budget compliance $\beta(s,\tau)$ using two timed automata.
In isolation, the automata guarantee two important theorems: Theorem \ref{thm:service:avail:conv} and \ref{thm:service:budget:conv}.
%  The automata two theorems on the two behave in the absence of the other limitation.
%In reality,
In practice, the main challenge is to be able to ensure both $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$ at the same time.
However, recall that these two properties of a service $s$ have conflicting goals: $\beta(s,\tau)$ decreases as $\alpha(s,\tau,t_c)$ increases (and vice versa).
%as an \emph{inverse} relation to one another.

### 2015-11-02 p00-ch00-intro/intro.tex
Introduction polishing

@@
 One object, the sender, communicates a message to the other object, the receiver.
-At the receiver side, a message eventually leads to a method invocation (in an object-oriented paradigm) or a functional call (in a functional paradigm).
+At the receiver side, a message eventually leads to a method invocation (in  object-oriented paradigm) or a function call (in functional paradigm).
 In actor model, the locality of the objects is transparent from the messages.
@@
 
-With the rise of computation power, a new aspect added to create a triangle: object-oriented, actor model, and \emph{concurrency}.
+With rapid increase of computation power, a new aspect added to create a triangle: object-oriented, actor model, and \emph{concurrency}.
 Concurrency makes it harder to verify programs in terms of correctness of runtime behavior. 
@@
 
-We target to deliver ABS concurrency layer in form of a Java library API.
-However, Java language specification (JLS) clearly concludes that Java language is not comptabile and ready to be extended as a functional language with the support of algebraic data types.
+% We deliver ABS concurrency layer in form of a Java library API.
+% However, 
+Java language specification (JLS)~\cite{gosling2000java} clearly concludes that Java language is not comptabile and ready to be extended as a functional language with the support of algebraic data types.
 This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
@@
 
-We primarily focus on ABS concurrency layer as a Java library API.
-We argue that the advantage of a correct semantic translation from ABS concurrency layer to its Java equivalent constructs overcomes the additional support of functional layer but with \emph{partial} formal semantics.
+We primarily focus on ABS concurrency layer and deliver its semantics as a Java API.
+We argue that the advantage of a correct semantic translation from ABS concurrency layer to its Java equivalent constructs overcomes the additional support of functional layer with \emph{partial} formal semantics.
 Our approach enables programmers to model their Java systems in terms of ABS concurrency semantics and then create a correct mapping and implementation from ABS to Java using an API.
@@
 
-In the next section, we sketch a high-level overview of the architecture design goals and how this research plans its objectives.
-
 \section{Architecture}
 \label{sec:intro:arch:crit}
+
+In this section, we present a high-level overview of design goals that we pursue in this research. 
+Equivalently, the effectivness and success of this research is measured upon the achievement of the discussed goals and objectives.
 
@@
 \paragraph{Scalability} 
-The distribution level that asynchronous message passing mechinism deliver
-challenges performance efficiency.
-In such a context, the number of messsages delivered among actors in the 
+Asynchronous message passing in ABS is a fit for distributed systems.
+In such systems, the number of messsages delivered among actors in the 
 environment is not predictable at runtime.
+Therefore, distribution challenges performance efficiency or scalability.
 The goal is to ensure the actor system scales in performance with least 
 influence from the number of asynchronous messages delivered in the system.
-We direct benchmarks on the system to present the performance of the system
-using different number of messages delivered in the system.
 
@@
 \item \emph{Compiling ABS to a target programming language}
-One first objective it compile an ABS model to a target programmig language.
+One first objective is to compile an ABS model to a target programmig language.
 The compilation can be at \emph{source-to-source} or \emph{source-to-bytecode}
@@
 If such ABS mapping to a programming API is provided, a programmer is able to
-take advantage of ABS semantics without directory programming in ABS.
+take advantage of ABS semantics without directly programming in ABS.
 Such capability from ABS enables industry users of mainstraim languages to
@@
 
-Based on the scopes and layers of usage of this research, we aim to provide 
-a highly modular architecture in design and implementation.
-The goal is verifiable based on produced artifacts of this research in terms
-of their \emph{usability}, \emph{re-usability}, and \emph{composability} in
-different contexts.
+% Based on the scopes and layers of usage of this research, we aim to provide 
+% a highly modular architecture in design and implementation.
+% The goal is verifiable based on produced artifacts of this research in terms
+% of their \emph{usability}, \emph{re-usability}, and \emph{composability} in
+% different contexts.
 
 \paragraph{Extensibility}
-Delivering ABS language features as an API brings about a new requirement to be able to configure and extende the API. 
+Delivering ABS language features as an API brings about a new requirement: to configure or extend the API. 
 It is a common practice in software engineering that components are developed with a mixture of \emph{configuration}, \emph{extension} and \emph{composition}.
@@
 
+\Behrooz{It's worth to include the figure but needs more work to justify its presence and how it relates to this thesis.}
 Figure~\ref{fig:arch} presents the general architecture of the implementation.
@@
 
-\subsection{Programming Languages Overview}
+We discuss a brief overview of related work in the context of programming languages, actor model, and concurrency.
+We devide the overview in two levels; one is at the level of the programming languages and the other is for the external (third-party) libraries developed for programming languages.
+
+\subsection{Programming Languages}
 \label{sec:intro:proglangs}
@@
 
-In the previous section, we presented an overview of the programming languages and their categories that aim to provide actor model~\cite{agha86book}.
-Since many programming languages faced different challenges to provide the necessary syntax and semantics for this at the level of the language, many libraries and frameworks have been born with a motivation to fill this gap.
+% In the previous section, we presented an overview of the programming languages and their categories that aim to provide actor model~\cite{agha86book}.
+Since many programming languages faced different challenges to provide the necessary syntax and semantics for actor model and concurrency at the level of the language, many libraries and frameworks have been born with a motivation to fill this gap.
 We observe that the more the language itself is close to the actor model semantics, the less external libraries and frameworks target this gap. 
@@
 Although, this technique is commonly used and argued to provide better performance optimization~\cite{vallee1999soot}, it introduces deep challenges regarding the verification of the running byte-code~\cite{leroy2001java,leroy2003java}.
-Thus, one of the main objectives in this research to provide a \emph{statically} verifiable technique to provide actor programming in Java.
 
-\section{Contributions}
+\section{Oraganization and Contributions}
 \label{sec:intro:contribs}
 
-Based on what we discuss in objectives and related work, the contributions of this thesis can be summarized as the following table:
+In this thesis, we focus on the concurrency layer of ABS.
+We formalize a subset of Java language in regards with ABS.
+We provide the operaional semantics of the concurrency layer for scheduling of actors with an extension for deadlines and priorities. 
+We deliver ABS concurrency as a Java API.
+We utilize Java~8 and Java Concurrency API to preserve the semantics of ABS concurrency layer.
+We sketch a general architecture on how ABS language can be modularly translated into different programming languages.
+We present how ABS API library can be used in the translation of ABS models into Java.
+
+Table~\ref{tbl:thesis} summarizes the structure of the research and contributions in the chapters of this text:
 
@@
 \centering
-\begin{tabular}{p{7cm}p{3cm}}
-\textsfb{Topic} & \textsfb{Chapter/Section}
+\begin{tabular}{p{7cm}p{3cm}p{3cm}}
+\textsfb{Topic} & \textsfb{Part} & \textsfb{Chapter/Section}
 \\ \toprule
-{Formalization of the mapping from ABS to Java including the operational semantics and ABS co-operative scheduling in Java} & Chapter~\ref{ch:p01:ch01} and \ref{ch:p01:ch02}
+{Formalization of the mapping from ABS to Java including the operational semantics and ABS co-operative scheduling in Java} & Programming Model (Part~\ref{p:model}) & Chapter~\ref{ch:p01:ch01} and \ref{ch:p01:ch02}
 \\ \midrule
-Design and implementation of ABS concurrency layer in Java & Chapter~\ref{ch:p02:ch01} and \ref{ch:p02:ch02}
+Design and implementation of ABS concurrency layer in Java & Implementation (Part~\ref{p:impl}) & Chapter~\ref{ch:p02:ch01} and \ref{ch:p02:ch02}
 \\ \midrule 
-Design and implementation of a massive-scale monitoring system based on ABS API in Java & Chapter~\ref{ch:p03:ch01}
+Monitoring method call sequences using annotations & Application (Part~\ref{p:app}) & Chapter X
+\\ \midrule
+Design and implementation of a massive-scale monitoring system based on ABS API in Java & Application (Part~\ref{p:app}) & Chapter~\ref{ch:p03:ch01}
 \\ \bottomrule 

### 2015-11-24 p03-ch02-2014-jmseq/grammar-sample-usage-1.tex
Add JMSeq Journal Paper from SCP 2014

--- 2015-10-20 p01-ch01-2012/langs.tex
@@
 \documentclass{article}
-\usepackage[top=0.01in,bottom=0.01in,left=0.01in,right=0.01in,paperwidth=5.7in,paperheight=2.3in]{geometry}
+\usepackage[top=0.01in,bottom=0.01in,left=0.01in,right=0.01in,paperwidth=5.1in,paperheight=0.4in]{geometry}
 
 \usepackage{alltt}
-\usepackage{array}
 \usepackage{amsmath}
+\usepackage[usenames,dvipsnames]{color}
+%\usepackage{courier}
 \usepackage{graphics}
 \usepackage{graphicx}
-\usepackage{pifont}
 \usepackage{url}
-\usepackage[table]{xcolor}
 
@@
 \newcommand{\subscript}[1]{\ensuremath{_{\textrm{#1}}}}
-\newcommand{\thickhline}{\noalign{\hrule height 1.0pt}}
 
@@
 \pagestyle{empty}
-\definecolor{lgray}{rgb}{0.85,0.85,0.85}
-\rowcolors{2}{lgray}{white}
-\begin{center} 
-\begin{tabular}{m{0.5in} m{1.2in} m{1in} m{1in} m{1in}}
-  & Java & C++ & Python & C\# \\
-\thickhline 
-Actors &  Scala, Erlang, SALSA, Axum, Ptolemy, Kilim, ActorFoundry, Jetlang,
-JavAct, AJ & Theron, Act++ & Stackless, V3.2+ & Retlang \\
-
-STM & Multiverse, Clojure, DSTM2 (Fortress), Deuce, JVSTM & TL2 (Sun), Intel
-C++ STM & Durus & SXM \\
-
-Data Flow & Java 7+ JSR 166y Fork/Join, Hadoop & OpenMP, TBoost.STM, Hadoop Streaming &
-Disco (Nokia) & Dryad, Hadoop Streaming \\
-\end{tabular}
-\end{center}
+\begin{alltt} 
+\centering\small
+{\color{red}\{}call(* *.A.m_a(..)){\color{red}[}{\color{blue}\{}call(* *.B.m_b(..)){\color{blue}[}\{call(* *.C.m_c(..))\}{\color{blue}]}{\color{blue}\}}{\color{red}]}{\color{red}\}}
+\end{alltt}
 \end{document}

### 2015-11-24 p03-ch02-2014-jmseq/grammar-sample-usage-2.tex
Add JMSeq Journal Paper from SCP 2014

--- 2015-11-24 p03-ch02-2014-jmseq/grammar-sample-usage-1.tex
@@
 \centering\small
-{\color{red}\{}call(* *.A.m_a(..)){\color{red}[}{\color{blue}\{}call(* *.B.m_b(..)){\color{blue}[}\{call(* *.C.m_c(..))\}{\color{blue}]}{\color{blue}\}}{\color{red}]}{\color{red}\}}
+{\color{red}\{}call(* *.A.m_a(..)){\color{red}[}\{call(* *.B.m_b(..))\}{\color{red}]}{\color{red}[}\{call(* *.C.m_c(..))\}{\color{red}]}{\color{red}\}}
 \end{alltt}

### 2015-11-24 p03-ch02-2014-jmseq/jmseq-elsa.tex
Add JMSeq Journal Paper from SCP 2014

%!TEX root = ../thesis.tex

\chapter[Monitoring Method Call Sequences using Annotations]{Monitoring Method Call Sequences using Annotations%
\footnote{This research is partly funded by the EU project FP7-231620 HATS: Highly Adaptable and Trustworthy Software using Formal Models (\url{http://www.hats-project.eu/})}
}
% 
\label{ch:p03:ch02}
% 
\chapterauthor{Behrooz Nobakht, Frank S. de Boer, Marcello M. Bonsangue, Stijn de Gouw, Mohammad M. Jaghoori}
%

\section*{Abstract}
In this paper we introduce JMSeq, a Java-based tool for 
% specification and runtime verification by
monitoring  sequences of
method calls. JMSeq provides a simple but expressive language
to specify the observables of  a Java program in terms of sequences of possibly nested method calls. 
% JMSeq additionally provides a simple way to verify the probable exceptions
% during the execution of a Java program. 
Similar to many
monitoring-oriented environments, verification in JMSeq is done
at runtime; unlike all other approaches based on
aspect-oriented programming, JMSeq uses code
annotation rather than instrumentation,
and therefore is suitable for component-based software verification.

\subparagraph*{Journal Publication}
\emph{Science of Computer Programming, 2014, Volume 94, Part 3: Selected Best Papers from 7th International Workshop on Formal Aspects of Component Software -- FACS 2010, Pages 362--378, DOI \smalltt{10.1016/j.scico.2013.11.030}}

% \begin{keyword}
% Object Monitoring \sep Runtime Verification \sep Method Call Sequence Specification \sep Code Annotation \sep Component-based Testing
% \end{keyword}

 \section{Introduction} \label{sec:intro}
Monitoring refers to the activity of tracking observable information from an
execution of a system with the goal of checking it against the constraints imposed
by the system specification. The observable information of the monitored system
typically includes behavior, input and output, but may also contain quantitative
information. A monitoring framework consists of a monitor that extracts the
observable behavior from a program execution, and a controller that checks the
behavior against a set of specifications. Should an execution
violate the constraints imposed by the specification, corrective
actions can be taken.

What can be monitored depends, of course, on what can be observed in a system.
When the application source code is available, its code can be instrumented 
to produce informative messages about the execution of an application at run
time. New code is inserted  for the management of
monitoring and verification mechanisms, while preserving the original logic
of the application. 
Bytecode instrumentation is another technique widely used 
for monitoring and tracing execution of Java programs.
Bytecode instrumentation consists of inserting bytecode into classes either at compile-time or at runtime. 
It is typically preferred to source code instrumentation because it can be used to
enhance a program temporarily for debugging or
analysis purposes, and no source code is required.
Main disadvantages however of  bytecode
instrumentation are the following:

\begin{itemize}
\item
Although there exist tools for facilitating bytecode
instrumentation, building instrumentation in bytecode
requires an in-depth knowledge of bytecode; thus, it is \emph{error-prone}.
%Moreover, JMSeq does \emph{not} use instrumentation; neither at source-level nor at bytecode-level.
%   Source-level or bytecode-level instrumentation is used in approaches such as aspect orientation in AspectJ.
\item
Instrumentation by  its very nature is an intrusive approach; i.e., the target program is modified either at source level or bytecode level.
This makes it unsuitable, for example, for monitoring classes in the Java standard library, which are supposed to remain unchanged.
Moreover, if two different bytecode agents simultaneously instrument the same program,
they must be compatible; otherwise, the resulting bytecode may contain \emph{conflicts}.
\item
      From the perspective of business security, a verification approach that modifies a program potentially increases its \emph{vulnerability } and as such is of restricted applicability.
\end{itemize}

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.5]{images/arch-diagram-summary-exec}
\caption{JMSeq Execution Architecture}
\label{fig:arch-overview}
\end{center}
\end{figure}

In this paper, we introduce JMSeq, a tool for monitoring  sequences of method calls in Java programs.
JMSeq is based on Java annotations rather than (source code or bytecode) instrumentation. 
%, at source or bytecode level, we use another approach, called JMSeq.
%JMSeq monitors  sequences of method calls using.
Therefore, it does \emph{not} require  access to the source code of the target application.
JMSeq detects the annotated methods from the target application at runtime. 
Furthermore,  Java annotations themselves have no direct effect on program execution or logic of the application;
they carry metadata that can optionally be used at runtime to perform operations orthogonal to the execution of the main application.
   %Annotations are based on a declarative language to specifiy method call sequences using regular expressions for method signatures.
%The program is verified against the specifications during runtime and fails as soon as a violation is detected. 
%The verification is performed in a separate virtual machine to protect the original application from any modification (Figure \ref{fig:arch-overview}). 
%In the following, we provide a  discussion of what distinguishes JMSeq from other related work (c.f. Section \ref{sec:relworks}) in verification:
The main distinguishing overall characteristic of JMSeq is
the integration of the following main features.

\begin{description}
%   \item[Specification language.] JMSeq introduces a simple declarative specification language.
%   The programmer can use the language to express verification requirements using simple expressions from the language.
%   The language is based on regular expressions of method call signatures.
%   The language is simple and intuitive to help the programmer identify different types of sequences of method calls.
   \item[Java annotations.] 
%JMSeq uses Java annotations to specify and verify the sequence of method calls. 
JMSeq annotations specify  method call sequences  in terms of regular expressions of method call signatures.
 This pure declarative language is simple and intuitive in  specification  of  different types of sequences of method calls.
   \item[JVM decoupling.]
To ensure that the target program is in \emph{no} way affected or changed during monitoring, JMSeq runs in a separate JVM from the target application's JVM.
   JMSeq uses Java Platform Debugger API (JPDA) \cite{JPDA_Home} to initiate a separate JVM and verify the target application using events from the target JVM.
  Figure \ref{fig:arch-overview} displays the high-level execution architecture of JMSeq.
   \item[Fail-fast approach.] JMSeq terminates execution as soon as a violation of a specification occurs.  This approach prevents unsafe behavior,
   enables the localization of an error, and facilitates a reaction
   to a failure immediately, thus making the software more robust~\cite{Shore04a}.
   %In mission-critical software, this approach is favored over others because it facilitates a reaction to the situation as soon as it occurs.
   %It helps the software/service provider to be able to provide better software to its end users.
   %It additionally contributes to better disaster recovery of software in failure situation.
\end{description}

An important consequence of these features
is that neither  the source code (even if available) nor bytecode is modified by JMSeq. 
Furthermore,  JMSeq does not require source code to be present.  
As such, JMSeq is suited for
the specification and verification of both
 newly built software and legacy systems, including third-party libraries
(and even the Java standard library).
For example, given  the API of a legacy system, its expected behavior  can be
specified by annotating the methods of a ``driver'' program which triggers
the legacy code.
JMSeq then verifies the triggered behavior of the legacy code using these
annotations.
%
%   In the context of legacy systems, it can be assumed that the sequence of API and method calls are known.
%   It is also known that JMSeq annotations are not placed on the legacy system.
%   This means that a helper program is needed to use the annotations from JMSeq.
%   Therefore, one can write a helper program that uses the legacy system and its API.
%   The helper program is annotated with JMSeq annotations.
%   Then, the helper program is executed through JMSeq.
%   JMSeq verifies the legacy system through the helper program that carries the annotations and as such the specification of method call sequences from the legacy system.
%   JMSeq verification of legacy system is distinguished from approaches that use instrumentation (especially source level) because access to source code cannot be presumed.
 In the context of an ongoing software development process, JMSeq can also be used as mentioned above
by means of a ``driver'' program but also by directly
annotating the ``interfaces'' of the system API.
%   Since Java annotations have no effect on the execution of the application, they behave as metadata at runtime.
%   JMSeq can detect them and verify the system without the need for a helper program.
By annotating the  interfaces of a system any service object that implements the interface behavior inherits the annotations and can be verified by JMSeq.
In this  respect, JMSeq also introduces integration with frameworks such as JUnit (cf. Section \ref{sec:junit}).

The rest of the paper is organized as follows.
The first three sections describe the usage of JMSeq.
Section \ref{sec:spec} presents the language for specifying sequences of method calls, 
and Section \ref{sec:impl_annot} discusses the types of annotations available in JMSeq.
These two sections are connected via examples in Section \ref{sec:samples}.
Afterwards, implementation of the JMSeq framework is explained in detail in Section \ref{sec:implementation}.
The practical applicability of JMSeq is demonstrated by means of a case study from industry in Section \ref{sec:frh} and performance evaluation in Section \ref{sec:results}. 
Finally, related work is discussed in Section \ref{sec:relworks}.
Section \ref{sec:conclusion} concludes the paper. %discuss possible future work.

\section{Method Call Sequence Specification} \label{sec:spec}

We consider a component to be a collection of compiled Java classes. The
relevant dynamic behavior of a component can be expressed in terms of specific
sequences of messages among a small, fixed number of objects. In
Figure \ref{fig:seq-spec}, we see two UML message sequence diagrams each
describing how, and in which order, the four objects interact with each other. In this
section, we develop a specification language for describing these kinds of
interactions.

%Many approaches use finite state machines or regular
%expressions to specify sequences of method calls, and as such it cannot be
%applied to specify recursive call-backs.

%A vertical line is associated
%to each object, representing its lifetime (from top to bottom). The solid
%horizontal arrows are labeled by a method name and represent calls, whereas the
%dashed arrow are return messages.

In addition to specifying the order, a specification language for sequences of method calls
needs to distinguish between the following two cases (depicted in
Figure \ref{fig:seq-spec}).
\begin{description}
  \item[Case $1$] 
  shows a scenario in which (the call to) \texttt{m\_c} is nested in 
  \texttt{m\_b} since \texttt{m\_c} is called during the activation of 
  \texttt{m\_b} (i.e., after \texttt{m\_b} is called and before it returns). 
  Similarly, both \texttt{m\_b} and \texttt{m\_c} are nested in \texttt{m\_a}.

  %exemplifies a nested method call in which methods from
  %different/same objects are called in a nested way creating new instances of
  %stack trace in the program execution. More specifically,
  %method call to \texttt{m\_c} is nested in method call to \texttt{m\_b} that in
  %turn is nested in method call to \texttt{m\_a}.

  \item[Case $2$] represents a method call in which methods from
  different/same objects are called in a sequential rather than nested
  manner. In this example, both \texttt{m\_b} and \texttt{m\_c} are called by
  \texttt{m\_a}.
\end{description}
Typically, a program will need a combination of both cases to specify its dynamic behavior.
Specifying only the order of the method calls is not enough, as both cases above
have the same order of method calls. It is thus required to have a specification
technique that distinguishes between the \textsl{method calls} and
\textsl{method returns}. \begin{figure}[h]
\begin{center}
  \includegraphics[scale=0.4]{images/seq-spec}
  \caption{Examples of Method Call Sequence Specification}
  \label{fig:seq-spec}
\end{center}
\end{figure}

%It is clear from the above examples that regular expressions over method calls
%are not enough to specify the typical nested call structure of a sequence of
%messages in the presence of recursion. In general, such sequences form a
%context free language. However they have a special structure: there is a
%deterministic pushdown automaton accepting them such that it pushes or pops at
%most one symbol only, depending if a method call or return is read,
%respectively. Such an automaton is called a visibly pushdown
%automaton \cite{Alur_abstractvisibly}.

% TODO: mention that events are always accepted for the same object for a single
% method activation

In JMSeq, a specification is written as a post-condition associated with a method. It
specifies the set of possible sequences of method calls, or protocol, of an
object in the context of the relevant part of its environment. Formally,
sequences of method calls can be specified by means
of the grammar in  Figure \ref{fig:spec_grammar}.

\begin{figure}[h]
\begin{center}
\begin{framed}
\includegraphics[scale=1]{grammar}
\end{framed}
\caption{Method Sequence Specification Grammar}
\label{fig:spec_grammar}
\end{center}
\end{figure}

A \textsl{specification} consists of a sequence of calls that can be repeated an
a priori fixed number of times 
("${\langle}$\textsl{Specification}${\rangle}^m$"), with $m \geq 0$, one or
more times ("${\langle}$\textsl{Specification}${\rangle}\$ $"), or zero or more
times ("${\langle}$\textsl{Specification}${\rangle}\#$"). Although JMSeq does
not use aspect-oriented programming in its implementation, we have used the
generic method call join points syntax of AspectJ \cite{kiczales_aspectj},
using for example $\#$ to denote the more standard Kleene star operation.
Additionally, $(\langle$\textsl{Specification}$\rangle)$ is a way to group
specifications to avoid ambiguity. To improve readability of the
specifications, grouping is only used when necessary.

A \textsl{call} is a call signature followed by a (possibly empty) sequence of
inner calls. Each call can be repeated either one or more times, zero or more
times, or exactly $m$-times, for some $m \in \{1, 2, 3, \ldots \}$;
i.e., $m$ is a constant. 
Having the repetition by $m$ does not make the language go beyond the expressiveness of regular expressions.
Additionally, the wild card 
{\small$\{$\textbf{\texttt{call}}$(*)\}$} denotes a call to an arbitrary method.
To support branching, JMSeq also provides
$<\langle$\textsl{Call}$\rangle ? \langle$\textsl{Call}$\rangle>$ to allow the
specification of a choice in a sequence of method executions.

\textsl{Inner calls} are calls that are executed before the outer
call returns, i.e., they are nested. 
We do not have explicit return messages, but rather we specify the scope of a call  using brackets. 
Information about the message call, like the type of the caller and the callee, method
name, and the types of the parameters are expressed using a syntax similar to that of AspectJ. 
Simply put, a call signature is of the form
\begin{alltt}
\small
\texttt{call}([ModifiersPattern] TypePattern
        [TypePattern . ] IdPattern (TypePattern | ".." , ... )
        [ throws ThrowsPattern ]
)
\end{alltt}
reflecting the method declarations in Java.
The two leftmost patterns specify the method modifiers (like   ``static'' or ``public'') and return types.
% that include method names, method
%parameters, return types, modifiers, and throws
%clauses. 
Here \texttt{IdPattern} is a pattern specifying the method name. It can possibly
be prefixed at the left by a specification of the type of the object called.
The method name is followed by a specification of the types of the parameters
passed during the call, and possibly by a specification of the exceptions that may be thrown.
It implies that JMSeq specification grammar can distinguish the overloaded
methods in a class. Patterns may contain wild card symbols ``\texttt{*}'' and
``\texttt{..}'', used for accepting any value and any number of values,
respectively. For example, the call
\begin{alltt}
\centering\small
\texttt{call(* *.A.m\_a(..))}
\end{alltt}
is denoting a call to method \texttt{m\_a} of any object of type \texttt{A}
that is placed in any package, and returning a value of any type. If there is a
need to be more specific, a possible restatement of the same specification
could be:
\begin{alltt}
\centering\small
\texttt{call(int nl.liacs.jmseq.*.A.m_a(Object, double))}
\end{alltt}

Next we give a few examples of correct method sequence specifications. 
For instance, the specification of the sequence in case $1$ of
Figure \ref{fig:seq-spec} is given by:
%\begin{flushleft}
%\includegraphics[scale=1]
\input{grammar-sample-usage-1-1}
%\end{flushleft}
%\begin{alltt}
%\centering\small
%{\color{red}\{}call(* *.A.m_a(..)){\color{red}[}{\color{blue}\{}call(*
%*.B.m_b(..)){\color{blue}[}\{call(*
%*.C.m_c(..))\}{\color{blue}]}{\color{blue}\}}{\color{red}]}{\color{red}\}}
%\end{alltt}
Here it is important to notice that the call to method  \texttt{m\_b} is internal
to \texttt{m\_a}, and the call to \texttt{m\_c} is internal to \texttt{m\_b}.
The sequence in case $2$ of the same figure would be:
%\begin{flushleft}
%\includegraphics[scale=1]
\input{grammar-sample-usage-2-1}
%\end{flushleft}
%\begin{alltt}
%\centering\small
%{\color{red}\{}call(* *.A.m_a(..)){\color{red}[}\{call(*
%*.B.m_b(..))\}{\color{red}]}{\color{red}[}\{call(*
%*.C.m_c(..))\}{\color{red}]}{\color{red}\}}
%\end{alltt}
where both calls to methods  \texttt{m\_b} and \texttt{m\_c} are internal to
\texttt{m\_a}.

These two cases depict a fixed sequence of method calls. More interesting are the cases when,
for instance, the method \texttt{m\_a} should be called at least once before any possible
method call to \texttt{m\_b} or \texttt{m\_c}:
%\begin{flushleft}
%\includegraphics[scale=1]
\input{grammar-sample-usage-3-1}
%\end{flushleft}
%\begin{alltt}
%\centering\small
%{\color{red}\{}call(* *.A.m_a(..)){\color{red}\}}\${\color{red}\{}call(*
%*.B.m_b(..)){\color{red}\}}\#
%\end{alltt}

Such a specification is used in circumstances where the execution of  \texttt{m\_a} is a prerequisite to the execution of either of \texttt{m\_b} or \texttt{m\_c}.
% 
% It is \emph{notable} that the meta-grammar provided in Figure
% \ref{fig:spec_grammar} is a context-free grammar; 
It is notable that the actual specifications are regular expressions;
for example, they may contain  unbounded repetitions and a choice of calls.

\section{Annotations for method call sequences} \label{sec:impl_annot}
Java enables developers to attach extra metadata to their programs using annotations.
%Java provides a means to annotate parts of a program.
These metadata do not directly affect the program semantics, but they do affect the
way programs are treated by tools and libraries, which can in turn affect the
behavior of the running program \cite{Java_5_Annotations}. Annotations can be
read from source files, class files, or reflectively at runtime. 
One can define custom annotation types and use them to provide the desired types of metadata for the declarations in his program. 
In Java, an annotation is a special kind of modifier, and can be used anywhere other
modifiers (such as {\small\texttt{public}}, {\small\texttt{static}}, or
{\small\texttt{final}}) can be used.

We use annotations as an enabling feature to test components without the need to have their source code.
We only use the metadata loaded from the annotations during runtime.
These metadata include the specification of the sequences of method calls
based on the grammar discussed in Section \ref{sec:spec}. The metadata
is placed on classes or methods in a Java program. 
Then, they are loaded into JMSeq when it monitors the actual observable
behavior of the program at runtime.

JMSeq defines two type of annotations: sequenced object annotations and
sequenced method annotations.

\subsection{Sequenced Object Annotations}

To verify the sequence of executions, methods in classes need to be
annotated. Thus, JMSeq needs to detect all  classes that contain
some method carrying an annotation for verification. 
To improve readability and performance, JMSeq introduces
\smalltt{SequencedObject} annotation. This annotation is a marker
for such classes and notifies the annotation metadata loader (cf.
Section \ref{sec:annotation_repo}) that the objects instantiated from the annotated
class contain methods that specify a sequence of method executions. 
The code is given in Listing \ref{lst:seq_obj}.

\lstset{language=Java}
\begin{lstlisting}[label=lst:seq_obj, caption=SequencedObject Annotation Declaration]
@Retention(RetentionPolicy.RUNTIME)
@Target(ElementType.TYPE)
public @interface SequencedObject {
   // we need no properties for this annotation as this is only a marker.
}
\end{lstlisting}

In Listing~\ref{lst:seq_obj}, {\small\texttt{@Retention(RetentionPolicy.RUNTIME)}} declares that this
annotation is only applicable during runtime and may  not be used in other scenarios.
Moreover, {\small\texttt{@Target(ElementType.TYPE)}} declares that this
annotation can only be used on types including classes, interfaces and
enumerated types; in other words, this annotation may not be used to mark a method
in a Java class.

\subsection{Sequenced Method Annotations}
A sequenced method annotation is used to specify the sequences of method
calls beginning from a given method. The annotation requires a string
property declaring the sequence specification discussed in Section
\ref{sec:spec}. Listing \ref{lst:seq_method} presents the declaration of
this annotation type.

\lstset{language=Java}
\begin{lstlisting}[label=lst:seq_method, caption=SequencedMethod Annotation Declaration]
@Retention(RetentionPolicy.RUNTIME)
@Target(ElementType.METHOD)
public @interface SequencedMethod {

   String value();

   boolean allowExceptions() default true;

   Class<?>[] expect() default {};

   Class<VerificationFailureHandler> verificationFailureHandler();
}
\end{lstlisting}

{\small\texttt{@Target(ElementType.METHOD)}} declares that this annotation is
only applicable to \textit{methods}. The string value from
{\small\texttt{value()}} holds the sequence specification (cf.
Section \ref{sec:spec}). 
Additionally, a sequenced method annotation allows verification of
exceptions during runtime. The tester can configure whether exceptions
are allowed or not using the combination of \smalltt{allowExceptions()}
and \smalltt{expect()}:
\begin{description}
  \item[\smalltt{allowExceptions()=true}] implies that exceptions are allowed during execution; 
  this is the default case.
If there is no exception configured using \smalltt{expect()},
\emph{any} exception is acceptable; otherwise,  an exception is acceptable 
only if it is specified in the array
of classes in \smalltt{expect()} or if it is a subclass of one of them.
If an exception occurs and it is ``expected'', JMSeq allows the program to continue execution;
the program should nevertheless be able to properly deal with the exception.

  \item[\smalltt{allowExceptions()=false}]  means that no exceptions are
expected during the execution. In this case, the value specified by
\smalltt{expect()} is ignored.
\end{description}

JMSeq additionally provides a general mechanism to specify how to handle situations in
which the sequenced execution fails to complete. This can happen
because of a method call which was not included in the specification, or
an unexpected exception.
To this end, JMSeq introduces \smalltt{Verification\-FailureHandler} interface as an
extension point where the tester or developer can provide 
custom behavior to gracefully deal with the verification failure.
This interface
introduces a single method: \smalltt{handle\-FailedExecution(Execution
e)}. 
This method
%\smalltt{handle\-FailedExecution(Execution e)} 
receives information about the execution through a parameter of
type \smalltt{Execution} (cf. Section \ref{sec:implementation}).
\smalltt{Execution} is the central data model that carries information
about the method execution sequences; the trail of the
executions may be useful to the developer to examine the failure causes.
%
If no such handler is provided, verification failure results in termination 
of program execution with a failure status.

Section \ref{sec:samples} provides more discussion along with different
examples on using JMSeq.

\section{JMSeq by example} \label{sec:samples}
In this section, we go through different usage scenarios of JMSeq 
by means of illustrative examples.

\subsection{Sequenced Execution Specification}
In  Listing \ref{lst:annotated_spec}, we again go through the example
introduced in Section \ref{sec:spec}.
% give two examples of
% annotations of the class {\small\texttt{Main}} in Figure
% \ref{fig:seq-spec}. 
In this example, we annotate the method \texttt{main()} with the sequences of method calls describing the two scenarios
given in Figure \ref{fig:seq-spec}. 
The JMSeq specifications used here are discussed in Section \ref{sec:spec}.
The listing shows how to use them in a Java program.

\lstset{language=Java}
\begin{lstlisting}[label=lst:annotated_spec, caption=Sample annotated specification]

// Case 1
@SequencedObject
public class Main {

   @SequencedMethod(
     "{call(* *.A.m_a(..))[{call(* *.B.m_b(..))" + 
     "[{call(* *.C.m_c(..))}]}]}") 
   public void main() { 
      // ...
   }

   public void init() {
      // ...
   }
}

// Case 2
@SequencedObject
public class Main {

   @SequencedMethod("{call(* *.A.m_a(..))" + 
     "[{call(* *.B.m_b(..))}][{call(* *.C.m_c(..))}]}") 
   public void main() {
      // ...
   }

   public void init() {
      // ...
   }
}
\end{lstlisting}

\subsection{Exception Verification}
We discuss how JMSeq can be used to verify  occurrence of exceptions during
execution of an application using the example given in Figure
\ref{fig:seq-spec} Case 1.

Let's assume that \smalltt{m\_c} may encounter
\smalltt{NullPointerException} and \smalltt{m\_b} may face
\smalltt{IOException}. But, we do not hold any assumption on the
location in the code where the exceptions may be handled or even
totally ignored. We use the following annotation in Listing
\ref{lst:case-1-exceptions} to express this situation.

\lstset{language=Java}
\begin{lstlisting}[label=lst:case-1-exceptions, caption=Exception
Specification for Case 1]
// Case 1
@SequencedObject
public class Main {

   @SequencedMethod("{call(* *.A.m_a(..))" + 
     "[{call(* *.B.m_b(..))[{call(* *.C.m_c(..))}]}]}",
     allowExceptions = true,
     expect = {NullPointerException.class, 
               IOException.class}) 
   public void main() { 
      // ...
   }

}
\end{lstlisting}

During the execution of the application, if any exception occurs,
JMSeq receives the event of the exception occurrence. Since exceptions
are allowed, there will be no verification failure by JMSeq as long as
the exceptions received and verified are either instances of the
specified exception classes or any of their subclasses.
However, if the underlying application does not handle the exceptions
(i.e., proper use of \smalltt{try/catch}), they are
propagated up in the call stack and this may result in JVM crash.
%because of the application ignorance of the exceptions.

For instance, if an instance of \smalltt{FileNotFoundException} occurs
in \smalltt{m\_c}, JMSeq accepts the exception (because it is a subclass of
\smalltt{IOException}) and allows the application to continue. If the
application at some point deals with the exception, the execution
continues; otherwise JVM stops, complaining that the exception is not handled.
On the other hand, if an \smalltt{IllegalStateException}
happens during the execution of the program, JMSeq does not allow the
application to receive the exception since it is not one of the expected
exceptions in the specification. 
Instead, it runs the failure code provided by the tester (if any).

In another situation in Case 1, assume that \emph{no} exceptions are
acceptable during the execution. For this to happen, we simply use
\smalltt{allowExceptions = false} in the specification annotation. In
this case, no exception should occur during the execution of the
application; otherwise JMSeq issues a verification failure. Besides,
when no exceptions are allowed, using \smalltt{expect()} has no effect
on JMSeq verification behavior.

JMSeq additionally supports integration with testing frameworks such as JUnit. 
An example of such integration is discussed in Section \ref{sec:junit}.

\section{The JMSeq framework} \label{sec:implementation}

In this section, we present the JMSeq monitoring and verification framework and discuss its implementation, which uses Java $5$ Annotations \cite{Java_5_Annotations} and Java Platform Debugger API \cite{JPDA_Home}. 
More implementation details, including examples and documentation, can be found at \url{http://github.com/nobeh/jmseq}.

As discussed above, to use JMSeq, it is assumed that Java components are supplied with sequenced object and sequenced method annotations. 
Multiple annotations are possible within the same class, but not for the same method. 
This implies that a component specification is in fact scattered across its constituent classes, each of which with a local and partial view of the component behavior.
Only the annotated methods will be monitored by JMSeq.
The overall verification process in JMSeq is depicted in Figure \ref{fig:alg-verify}.

\begin{figure*}[t]
\begin{center}
  \includegraphics[scale=0.47]{images/alg-diagram}
\end{center}
\caption{Overview of JMSeq's process to verify a program}
\label{fig:alg-verify}
\end{figure*}

As a Java program, JMSeq executes naturally inside a Java virtual machine (JVM); JMSeq additionally initiates another JVM (called the \textsl{Target JVM}) to be able to control the sequenced execution of the target program.
Through initial parameters, the target JVM is configured to report back to JMSeq some events such as method entry and method exit.
Furthermore, JMSeq does not verify events from all objects but only those specified in these parameters. 
The target JVM takes advantage of the Java Platform Debugger Architecture (JPDA) to access the needed details on the execution of the application while it is running.

Before verification starts, JMSeq inspects the classes in the execution class path searching for methods that are annotated for sequenced verification; during this process, it collects all of the available method call sequence metadata. 
This step (denoted as Load Annotated Metadata in Figure \ref{fig:alg-verify}) does not need the source code, but rather uses the annotations metadata available in the compiled code.

The target program subsequently starts its execution and JMSeq reacts to the events it receives.
JMSeq uses state machines to verify the method calls occurring in an application during runtime.
The state machines are indexed by the pairs of method signature and the callee object; thus JMSeq can uniquely identify the associated state machine.
When JMSeq receives a method execution event, it checks if there is a need to initialize a new state machine for that method execution.
A state machine is initialized if there is no state machine for the current method execution and the method is annotated  \smalltt{@SequencedMethod}.
Then, JMSeq tries to verify the current method execution in the context of all active state machines for other methods.
For every state machine, JMSeq tries to make a valid transition in the state machine. 
Verification will continue if such a transition can be made for all state machines; 
if not, JMSeq stops the execution reporting a violation of the specification. 
% of the outermost specification for the current method is detected.
% When a verification failure is detected, JMSeq stops the execution; if a user defined failure handler is present, it will be executed before termination.

The process of matching method execution events with state machine transitions is done in JMSeq by following these steps:
\begin{enumerate}
  \item %A ``call expression'' is constructed based on the information associated to the received event. 
  First, every event received from the target JVM, together with the information associated to the event, is transformed to an instance of \smalltt{Execution} to capture various details about the execution of a method. 
  Then, at verification time, the instance of \smalltt{Execution} is transformed to an instance of \smalltt{CallExpression} to hold method signature details used by the state machine to make a transition. (Section \ref{sec:event_handling})
  \item Using the metadata available from the annotations of the methods, then, the   next ``possible call expressions'' of the current state, corresponding to the possible next transitions of the state machine,  are built. (Section \ref{sec:annotation_repo}, \ref{sec:exec_verify})
  \item A match making is done between the possible call expressions (state machine transitions) and the   current call expression as the candidate.  
  If a match is found,  the transition is made and the method is  accepted; otherwise, it fails. 
  When a failure occurs, JMSeq will execute the custom verification failure handler that is implemented either as part of
  the component code by the programmer, or externally by the system tester. (Section \ref{sec:exec_verify})
\end{enumerate}

JMSeq computes the possible call expressions on-the-fly instead of constructing the full state machine in the beginning of the verification process. 
This way it avoids the construction of states in the state machine that are never used in the execution. 
To avoid repeated computations of the same states, JMSeq utilizes dynamic programming techniques such as in-memory object caches.

\subsection{JMSeq Architecture}

The overall architecture of JMSeq is given in Figure \ref{fig:arch-diag}. It basically
consists of three main modules: one for handling the events raised by the JVM executing the target program, another module for storing the annotation information, and a third
one to perform the runtime verification.

\begin{figure*}[t]
\begin{center}
  \includegraphics[scale=0.35]{images/arch-diagram}
  \caption{Software Architecture for Method Sequence Specification}
  \label{fig:arch-diag}
\end{center}
\end{figure*}

% The current design of JMSeq is completely \textsl{general} and \textsl{modular};  as it
% allows for replacing the grammar in Figure \ref{fig:spec_grammar} with other
% specification modules, based, for example, on temporal logics or extended
% regular expressions.

\subsubsection{Event Handling Module} \label{sec:event_handling}
Java Debugger Interface (JDI) is the interface
of Java Platform Debugger Architecture (JPDA) that gives access to different details during the execution of a program.
Following the JDI event model, JMSeq takes control over some of the
execution events that JVM publishes during a program execution. Therefore, a
component was designed to model and hold the execution trace events required
for event handling and execution verification.

\begin{enumerate}
  \item \smalltt{Execution} is the central data structure that holds all the
  information about execution that is made available through the JDI event mechanisms.
  Relevant events include ``method entry'', ``method  exit'', and
``exception'' occurrences.
   It provides access to information such as:
   \begin{itemize}
     \item The object that is currently executing (\emph{the callee object}) and
     its unique identifier in JVM.
     \item Event details through subclasses such as method return values or
     \emph{the caller object} reference in case of a method exit event.
     \item \textit{Parent Execution}: every execution can hold a reference to its
     parent execution object forming a directed tree of executions. This helps
     traversing the executions at validation time.
    \end{itemize}
     \item \smalltt{EventHandler} is the event handling interface that
is
     injected into JVM with access to JDI information. The event handler
     receives the events it is subscribed to and possibly takes an associated
     action. In particular it creates an instance of {\small\texttt{Execution}}
     for each sequence annotation and it stores it in the
     {\small\texttt{ExecutionTraceOracle}} registry. In general, all events received and processed by
     {\small\texttt{EventHandler}} are stored in this registry for further use
     by other components, for example, by the
     verification module.
\end{enumerate}

\subsubsection{Annotation Metadata Module} \label{sec:annotation_repo}

As the execution traces are stored in a repository, they are supposed to be
checked and verified against the formal specifications as described on the
{\small\texttt{@SequencedObject}} and {\small\texttt{@SequencedMethod}} annotations
of the compiled classes in the program. It is the task of the
annotation metadata module to store this information. A small utility
component in this module is responsible to read and load the metadata of all
classes that are annotated. This information is used when there is a
need to verify the conformance of an execution event.

\subsubsection{Execution Verification Module} \label{sec:exec_verify}

% \note{Most of this paragraph is repetition. If we have time, this repetition can be reduced ideally.}
During execution of a program, those events that need to be monitored are received and
verified against a message sequence specification. 
For every specification in the metadata repository, a state machine is created on the fly.
At each state, JMSeq computes the Brzozowski derivative of the regular expression associated with the current state and memorizes this computation for further use. 
This avoids the full construction of the state machine at each step of verification.

When a method execution event is received, either the corresponding state machine exists or it should be created.
A state machine is initialized if there is no state machine for the current method execution and the method is annotated  \smalltt{@SequencedMethod}.
JMSeq then tries to verify the event against all active state machines.
For every state machine, an attempt is made to make a valid transition.
If the transition is successfully made, JMSeq continues.
Otherwise, an ``invalid'' transition is detected; verification stops.
At this point, JMSeq checks if an implementation of the
\smalltt{Verification\-Failure\-Handler} interface is present (cf. Section \ref{sec:impl_annot}).
If present, JMSeq tries to instantiate an instance of the verification failure handler and hands over the execution metadata to this handler. 
Otherwise, JMSeq terminates and reports how the specification is violated.

An event may also be sent due to occurrence of an exception. 
In this case, the execution is verified based on the exception specifications expressed using the mechanism described in Section \ref{sec:impl_annot}. 
If the exception is allowed and verified the execution continues; otherwise, it fails and control is delegated to verification failure handler as explained above.

The execution verification component is composed of the following elements:
\begin{enumerate}
  \item \textsl{Call Expression} is a simple component for interacting with each
  state machine object. Basically, it transforms execution events coming from
  the JVM into call expressions used by the state machine components.
  \item \textsl{Call Expression State Machine}: Sequences of
  executions are translated to ``call expressions'' that are to be accepted by
  a state machine. The state machine needs to distinguish the
  method's context on different calls to the same method. At the
  end of a successful sequential execution, the state machine associated to that
  sequenced execution specification should be in a success state. 
  \item \textsl{Call Expression Builder} constructs a call expression out of:
  \begin{enumerate}
    \item A string which is in the format of the JMSeq specification grammar as
    in Figure \ref{fig:spec_grammar}. This service is used the first time a
    sequential execution is detected to build the root of the future possible
    (candidate) call expressions.
    \item An execution event; every time an execution is handed over to the
    verification module, an equivalent  call expression is built for it so
    that it can be compared and matched against the call expression for the
    previous event.
  \end{enumerate}
  \item \textsl{Call Expression Proposer} is a service proposing \textit{possible
  next call expressions} for a given call expression based on the sequences specified
  in the annotation. As described by the grammar in Figure
  \ref{fig:spec_grammar}, for each current call expression there can
  be several possible next call expressions that may appear as the next event, but for
  each of them the state machine associated with the grammar can only make one
  transition (that is, the specification is deterministic). Note that since more specification sequences are possible involving
  the same method, only those call expressions that are valid in all
  specifications will be proposed.
  \item \textsl{Call Expression Matcher} is another service that
  tries to match two call expressions. It is used, for example, to validate
  the current call expression against all those proposed by the previous service.
  If a match is found, the execution continues; otherwise the verification is regarded as
  failed. In this case, if a verification failure handler is provided, the failure data is
  transferred to it for further processing.
\end{enumerate}

\subsection{JUnit Support} \label{sec:junit}
To ease the process of verification, JMSeq provides a seamless integration to JUnit \cite{JUnit}. 
JUnit is a widely accepted unit testing framework for Java programs.

To use JUnit, the programmer is expected to write Java class files that introduce ``test'' methods.
A test method is annotated with \smalltt{@Test} annotation in JUnit.
Each test method is run in isolation and independently of the other test methods.
The programmer may provide special methods \smalltt{setUp} and \smalltt{tearDown} to, respectively, prepare resources or release them for the unit test.
JUnit provides a runner framework that runs all Java classes containing test methods.
Different tools such as Eclipse \cite{eclipse_junit} or Maven \cite{maven_junit} provide easy ways to run unit tests with JUnit.

JMSeq extends the JUnit framework and provides a simple to use mechanism for programmers to write unit tests in order to verify their programs. 
JMSeq introduces the \smalltt{JUnit4Support} class that is to be extended by the programmer to direct the verification  process.
In this unit test class, the programmer uses the \smalltt{@Test} annotation to identify the methods implementing a test, in the same way as in JUnit.
Furthermore, the programmer should override the following methods such that they provide the  information that JMSeq requires to start the verification process:
\begin{description}
  \item[\smalltt{getClassName()}] \emph{should} be overridden to provide
the simple name of the class that contains the entry point to the
application. 
  \item[\smalltt{getPackageBase()}] \emph{should} be overridden to
provide the qualified package address that contains the class file
provided by \smalltt{getClassName()}.
  \item[\smalltt{getExcludedPatterns()}] is a method that
may be overridden to provide the patterns that are excluded in matching
the events received from the underlying virtual machine started by
JMSeq. By default, a set of events
received from the standard packages in Java are excluded. The developer
may add, remove or modify the patterns based on the expected events.

% \note{I am confused, how does it relate to allowExceptions that we had
% before?}
%   \item[\smalltt{getTraceExceptions()}]  is a convenient method that can
% be overridden to command whether or not to verify exceptions in the
% program. By default, the value is \smalltt{false}; i.e. exceptions are
% allowed but ignored in JMSeq verification.

  \item[\smalltt{setUp()}] is a standard base method from JUnit that is
executed before each test method execution. It is supposed to prepare
the pre-conditions before the actual test. If this method is overridden,
it either has to call the parent method first or guarantee that
hierarchy objects are ready for execution.
  \item[\smalltt{tearDown()}] is another standard base method that is
executed after each test method.
  \item[\smalltt{@Test} methods] Each test method should carry
\smalltt{@Test} JUnit annotation to be considered a unit test method.
Each test method should contain the call to \smalltt{startJMSeq()}.
Other than that, it is the choice of the tester what to add in the test
method.
\end{description}

We describe how to use the JUnit support through an example. 
A \smalltt{Basket} is a resource holder for apples. Generally, each
\smalltt{Basket} supports two basic features: \smalltt{put()} and
\smalltt{get()}. The first one inserts an apple (i.e., an instance of a
resource) to the basket while the second one retrieves and removes one
apple from the basket. There are different resource producers and
consumers: producers populate the resource holder (basket) and
consumers use the resources available in the resource holder. As an
example, \smalltt{Pat} is a consumer and for now we forget about the
producer that co-relates to the mutual \smalltt{Basket} and
\smalltt{Pat}. \smalltt{Pat} consumes from the \smalltt{Basket} and gets
apples to \smalltt{eat()}.

In terms of JMSeq sequenced specifications, each invocation of
\smalltt{eat()} in \smalltt{Pat} should be followed by an invocation of
\smalltt{get()} in \smalltt{Basket}. We specify this scenario in Listing
\ref{lst:apples}.

\lstset{language=Java}
\begin{lstlisting}[label=lst:apples, caption=Apples JMSeq Specification]
@SequencedObject
public class Apples {

   private Pat pat;

   public Apples(Basket basket) {
      this.pat = new Pat(basket);
   }

   @SequencedMethod("{call(public void nl..Pat.eat())" +
         "[{call(public int nl..Basket.get())}$]}$")
   public void startLunch() {
     // generally a sequence of "eat" invocations
     pat.eat();
     // still hungry?
     pat.eat();
   }

   public static void main(String[] args) {
     Basket b = new Basket(10);
     Apples apples = new Apples(b);
     apples.startLunch();
   }
}
\end{lstlisting}

We need now to run the tests for the \smalltt{Apples} application.
This is made very easy by JMSeq's support for JUnit. 
We develop a JUnit
test class for this purpose and provide a set of predefined information
to the base test class. The JMSeq test classes should \textsl{extend}
the \smalltt{JUnit4Support} that is provided in the library. The code is
presented in Listing \ref{lst:apples-junit}.

\lstset{language=Java}
\begin{lstlisting}[label=lst:apples-junit, caption=Apples JMSeq JUnit
Test Unit]
public class ApplesDriver extends JUnit4Support {

   @Test
   public void startLunch() {
     startJMSeq();
   }

   @Override
   protected String getClassName() {
     return Apples.class.getSimpleName();
   }

   @Override
   protected String getPackageBase() {
     return Apples.class.getPackage().getName();
   }
}
\end{lstlisting}

\section{The Fredhopper Access Server: A Case Study} \label{sec:frh}

Fredhopper\footnote{\url{http://www.sdl.com/products/fredhopper/}} is a
%is an SDL company since 2008 and a leading
search, merchandising and personalization solution provider, whose
products are uniquely tailored to the needs of online businesses. Fredhopper operates behind the scenes of more than 100 of the largest online
shops
\footnote{\url{http://www.sdl.com/campaign/wcm/gartner-maqic-quadrant-wcm-2013.html?campaignid=70160000000fSXu}}.
%Typically, deployments have about 10
%explicit attribute values associated with a product over thousands of attribute dimensions. This challenging task involves working on
%difficult issues, such as the performance of information retrieval algorithms, the scalability of dealing with huge amounts of data and
%in satisfying large amounts of user requests per unit of time, the fault tolerance of complex distributed systems, and the executive
%monitoring and management of large-scale information retrieval operations.
Fredhopper offers its products and facilities to e-Commerce
companies (referred to as customers) as services over the cloud computing infrastructure.
In this respect, Fredhopper should tackle various challenges in
resource management techniques, the customer cost model, and service level agreements.
One of the major components of their software is the Fredhopper Access Server (FAS), which provides access to high quality product catalogs.

% \begin{wrapfigure}{r}{0.5\textwidth}
\begin{figure}[t]
\vspace{-30pt}
\begin{center}
\includegraphics[scale=0.7]{images/DM}
\end{center}
\vspace{-20pt}
\caption{Fredhopper's Controller life cycle for data processing requests from customers}
\label{fig:frh-dm}
\vspace{-5pt}
\end{figure}
% \end{wrapfigure}

To orchestrate different services such as FAS or data processing, Fredhopper takes advantage of a service controller (referred to as Controller).
The Controller manages different service installations for each customer.
For instance, a customer can submit their data along with a processing request to their data hub server.
The Controller then picks up this data and initiates a data processing job, usually in the cloud.
When the data processing is complete, the result is published to the customer environment and becomes also available through FAS services.
Figure \ref{fig:frh-dm} illustrates this example scenario.

The Controller provides a simple client API for customers to develop applications that use Fredhopper services.
Essentially a data job request consists of three major steps:
\begin{enumerate}
 \item Prepare the data , upload it to the customer data hub and receive a data identifier.
 \item Specify the data processing: reindexing, updating or loading data. Such actions are called \emph{trigger}s.
 \item Collect the results of a trigger and decide how to use them.
\end{enumerate}
%For modularity and loosely coupling, a data identifier is not \emph{strictly} required to create a new trigger.
%There are triggers that actually require no data to be initiated while others do.
The customer is required to have the appropriate data identifier before executing a new trigger. In fact, the customer application
is supposed to use the proper sequence of operations from the API.

In this case study, we examined and measured how JMSeq can help customers improve their usage of client API for data processing; 
i.e. verify data processing triggers and notify customers as soon as they occur.
We used ``rebuild'' and ``update'' triggers in this experiment based on customer environment data from the period of February 2013
\footnote{Fredhopper data in this research is confidential due to privacy protection of customers.}.
In this experiment, we measured the following:
\begin{enumerate}
\item Using the log files, we replayed the triggers twice with different settings.
\item In one setting, we replayed the original customer data processing trigger.
\item In the other setting, we replayed the original customer data processing that was wrapped with JMSeq verification.
\item In each setting, we measured the number of \emph{successful} and \emph{failed} triggers. 
A successful trigger is one for which data processing completes and the result is reported back to customer data environment.
\end{enumerate}

Figure \ref{fig:frh:cs} presents the measurements for each setting without using JMSeq and with using JMSeq for both rebuild and update triggers.
The measurements were based on an estimated number of 30300 triggers in the same period.
Each figure shows the influence of using JMSeq for the data processing task.
On the left, measurements shows that JMSeq verification for rebuilding the catalog helps to decrease failures by 15\% if using JMSeq verification.
On the right, measurements shows that JMSeq verification has less influence to improve update process as it natually involves less erroneous data because of smaller data size in update triggers.

\begin{figure}
\begin{center}
\includegraphics[scale=0.72]{images/frh-case-study-summary-rev2}
\caption{Result of experiment measurements without and with using JMSeq on Fredhopper Client API directed with live environment data from customers.}
\label{fig:frh:cs}
\end{center}
\end{figure}

%In this case study, we apply JMSeq to a sample customer application to verify that the methods of the FAS API are used in the correct order.
%We \emph{cannot} expect to have the source code for the application.
%However, since the customer uses Fredhopper Controller API, we can expect what API is used from Fredhopper services.
Next we present the case study in code listings in which JMSeq monitors calls to methods of the Controller API.
The runtime environment stops as soon as an incorrect sequence of calls is detected.
%Thus, the customer is prevented from further steps.
%Such prevention can have significant impact on the customer's business.
%A customer may create a considerable number of triggers on a daily basis.
%The result of each trigger has a direct influence on the end users of the customer.
%Being able to prevent risky or erroneous product catalog updates substantially helps a customer to plan the services they deliver to the end users.

%Fredhopper Controller exposes an interface to prepare data, submit the data and then receive/collect the data job results.
\lstset{language=java}
\begin{lstlisting}[label=lst:frh:datadispatcher, caption=Controller Data Dispatcher Interface API]
interface DataDispatcher {
   Data create(Object object);
   Future<?> dispatch(String dataId);
   Future<?> dispatch(Data oid, String dataId);
}
\end{lstlisting}

The relevant part of the API of the Fredhopper Controller is presented in Listing \ref{lst:frh:datadispatcher}.
Customer applications use the interface to create a data object reference (using the \smalltt{create(Object)} method) and then use the created data reference to start a data job (using the overloaded \smalltt{dispatch} method, which has an optional \smalltt{Data} parameter).
Note that the dispatch method allows for two different uses by the customer, for instance in Listing~\ref{lst:frh:c1} the data job to be executed performs only a rebuild of the customer's product catalog and as such does not need to have a data reference.
\lstset{language=java}
\begin{lstlisting}[label=lst:frh:c1, caption=A use case with no data requirement]
class CustomerApp {
   public void rebuildCatalog() {
      Future<?> result = 
         dispatcher.dispatch("reindex");
      // further processing of result
   }
}
\end{lstlisting}
If the customer needs to update their catalog with new products, then first a data reference is created. This reference is used to initiate a data job, as described in Listing \ref{lst:frh:c2}.
\lstset{language=java}
\begin{lstlisting}[label=lst:frh:c2, caption=A use case with data requirement]
class CustomerApp {
   public void updateCatalog(Catalog catalog) {
      Data data = dispatcher.create(catalog);
      Future<?> result = 
         dispatcher.dispatch(data.getId(), "update");
      // further processing of result
   }
}
\end{lstlisting}

Note that we cannot expect to have the source code of the above listings.  To verify the correct use of the Fredhopper Controller API
with JMSeq we can create a \smalltt{CustomerAppDriver} class which calls the \smalltt{CustomerApp} methods and monitors the subsequent
method calls.  The expected usage of the API of the two scenarios above is described by the SequencedMethod annotations in Listing~\ref{lst:frh:c3}.
If the \smalltt{CustomerApp} does not behave as expected, the verifier detects an error.

\lstset{language=java}
\begin{lstlisting}[float=t,label=lst:frh:c3, caption=How to verify Fredhopper API usage]
@SequencedObject
class CustomerAppDriver {

   @SequencedMethod(
      "{call(* nl..CustomerApp.rebuildCatalog()" +
      "[{call(* nl..DataDispatcher.dispatch(String))}]}")
   public void verifyRebuildCatalog() {
      // initiate the customer application
      app.rebuildCatalog();
      // finalize verification
   }

   @SequencedMethod(
      "{call(* nl..CustomerApp.updateCatalog(Catalog)" +
      "[{call(* nl..DataDispatcher.create(Object)}" + 
      "{call(* nl..DataDispatcher.dispatch(Data, String))}]}")
   public void verifyUpdateCatalog(Catalog catalog) {
      // initiate the customer application
      app.updateCatalog(catalog);
      // finalize verification 
   }
}
\end{lstlisting}

%Figure \ref{fig:frh:cs} shows how using JMSeq in an evaluation environment for the customers detects violations and
%improves the usage of Controller API and eventually affects the success rate of triggers for customers.

% \newpage
\subsection{Discussion} \label{sec:discuss-jmop}

We use JavaMOP \cite{chen_rosu_jmop,chen_rosu_mop,MOP} for the same case study from Fredhopper. 
Listing \ref{lst:frh:jmop} shows a JavaMOP script to describe the same verification goal.

\lstset{language=[AspectJ]Java, morekeywords={suffix, event,ere, CFG, @fail, @match, __RESET}} 
\begin{lstlisting}[label=lst:frh:jmop, caption=Fredhopper case study with JavaMOP]
UpdateCatalog(Catalog catalog) {
   event create_data before(DataDispatcher d):
      call(* DataDispatcher.create(..)) && target(d) {}

   event dispatch_data before(DataDispatcher d):
      call(* DataDispatcher.dispatch(..)) && target(d) &&
      cflow(dispatch_data) {}

   ere: (create_data dispatch_data)

   @fail { __RESET; }
}
\end{lstlisting}
JavaMOP uses aspect-orientation and Java bytecode instrumentation.
Using aspects (e.g., as in AspectJ \cite{kiczales_aspectj}) means that the eventual customer application bytecode is instrumented.
The question is \emph{how can it be proved that JavaMOP does not have any influence on the application?}
%Moreover, JavaMOP requires source code of the application to generate the required AspectJ sources.
If the customer is not willing to cooperate in such a process, the system cannot be monitored with JavaMOP.
%because they believe the risk of having an impact on the eventual software cannot be controlled if it happens.
JMSeq on the other hand does not change the source- or bytecode, because JMSeq runs on a different JVM as the one running the customer application.
%uses Java annotations to express runtime verification goals.
%By definition, a Java annotation is a syntactic metadata that is carried either at source level or runtime level.
%Java annotations have no effect on the logic of the executing Java application.
%However, if available at runtime, the Java application has the opportunity to use such metadata for orthogonal purposes aside main execution of the application.
%Moreover, JMSeq runs a in a separate JVM from the one that is actually running the Java application.
%Therefore, JMSeq introduces a highly safe approach for runtime verification.
%Since the customer application is affected in no way by JMSeq, they have more inclination to use such approach to verify their applications.

\section{Performance Results} \label{sec:results}

In this section, we present performance measurements of JMSeq and other techniques to verify a program.
We choose a problem that has been studied in the context of JavaMOP \cite{chen_rosu_jmop} to provide a reference for comparison.

The example verifies the correct use of \smalltt{java.util.Iterator}.
Originally, the interface introduces two major methods: \smalltt{hasNext()} and \smalltt{next()}. 
The first one checks whether there are still elements in the sequence that can be visited and the second one gives out the next element in the sequence and moves forward.
An invocation of \smalltt{next()} presumes that it succeeds an invocation of \smalltt{hasNext()}.
Otherwise, an invocation of \smalltt{next()} may lead to an exception.

We use JMSeq to verify an example code for this scenario. 
We use the code snippet displayed in Listing \ref{lst:jmseq-iter} that uses JMSeq annotations to express the correct usage. 
In the code, we use a variable \smalltt{count} to control the number of times \smalltt{hasNext()} is called before there is a call to \smalltt{next()}. 
This example always passes the JMSeq verification with success. 
For performance measurements, we study the ``correct'' use of the example while verification of wrong usages are also possible. 
\lstset{language=Java}
\begin{lstlisting}[label=lst:jmseq-iter, caption=Iterator usage with
JMSeq Annotations]
   @SequencedMethod(
     value = "{call(* java..Vector.iterator())}$"
         + "{call(* java..Iterator.hasNext())}$"
         + "{call(* java..Iterator.next())}", 
        allowExceptions = false)
   public void visitAllElements() {
      Iterator<Integer> i = v.iterator();
      for (int k = 0; k < count && i.hasNext(); ++k) {
         i.hasNext();
         i.next();
      }
   }
\end{lstlisting}

In the following, we provide certain measurements from executions of JMSeq for the example above for different values of \smalltt{count} in \smalltt{\{10000, 15000, 20000, 30000, 50000, 100000, 200000\}}. 
We use VisualVM tool that is a standard profiling and performance measurement tool packaged with standard Java development packages.

We measure the execution time of the following cases:
\begin{description}
\item[original.] We execute the original program and measure the execution time of the program with the above \smalltt{count} values.
\item[instrumented.] We use a technique that uses instrumentation such as JavaMOP \cite{chen_rosu_jmop} and apply it to the program and measure the execution time the same.
\item[original{\textbackslash}JMSeq.] We execute the original program and monitor it by JMSeq. We measure the execution time of the original program while being verified by JMSeq.
\end{description}

Figure \ref{fig:jmseq-performance} displays how different cases compare to each other regarding the execution time. 
As expected, the original program execution has the lowest execution time.
Since techniques based on instrumentation modify the bytecode of the program, this adds to the execution time of the program.
The figure shows that compared to instrumentation techniques, JMSeq adds a lower overhead.

\begin{figure}
\begin{center}
\includegraphics[scale=0.55]{images/perf}
\caption{Comparison of different techniques applied to the same program. 
In this figure, ``original'' denotes the original program that is measured for execution time;
``instrumented'' denotes using a technique such as JavaMOP that uses instrumentation for verification;
 and ``original{\textbackslash}JMSeq'' denotes the measurement of the execution time for the program apart from times for JMSeq.}
\label{fig:jmseq-performance}
\end{center}
\end{figure}

We use the performance results here to argue that the technique used by JMSeq is more suitable to verify applications that are mission-critical or operate on a high load.
Experience shows that, under high load, issues start to emerge that do not exist under normal operational situations.
In such circumstances, the application provider is very likely to look for methods to verify and detect problems.
And yet, they need to make sure that the application remains at least at the same level of execution and performance.
Moreover, the fact that JMSeq runs in a separate JVM increases the safety of our approach, since
JMSeq can be used from even a remote machine to verify a program, without the slightest influence on the resources that run the target program.

% JMSeq process of verification consists of two major activities in the context of performance and profiling:
% \begin{description}
% \item [Execution data model] that first receives the event
% from JPDA and the target JVM and prepares an execution data model for
% JMSeq verification.
% \item [Verification] that uses the execution data model to verify the
% current event based on the annotations, history of executions and the
% next possible execution.
% \end{description}

% \begin{table}[t]
% \begin{center}
% \begin{tabular}{|c|c|c|c|}\thickhline
% \texttt{count} & Total Time (msecs) & Invocations & Single Invocation
% Time \\ \thickhline
% 10,000 & 798 & 28751 & 0.0278 \\
% 15,000 & 1240 & 48204 & 0.0257 \\
% 20,000 & 1830 & 73470 & 0.0249 \\
% 30,000 & 2651 & 109909 & 0.0241 \\
% 50,000 & 4605 & 200010 & 0.0230 \\
% 100,000 & 9873 & 400010 & 0.0247 \\
% 200,000 & 19607 & 800010 & 0.0245 \\ \thickhline
% \end{tabular}
% \end{center}
% \caption{Execution data model performance results with JMSeq}
% \label{tab:results-handleevent}
% \end{table}

% Using VisualVM, we observed different measurements for both
% execution data model and verification. Please also
% note that the verification time is by nature different from
% execution time; i.e. we do not measure the execution time of the
% underlying application but we measure the time JMSeq takes to verify the
% executions in the application and then either allow or disallow the
% execution. In addition, we measure three metrics for the performance:
% total time for each JMSeq activity, number of invocations of each JMSeq
% activity, and a computed time for a single JMSeq activity. Table
% \ref{tab:results-handleevent} summarizes the performance results for
% execution data model and Table \ref{tab:results-verify} summarizes the
% same facts for verification.

% \begin{table}[t]
% \begin{center}
% \begin{tabular}{|c|c|c|c|}\thickhline
% \texttt{count} & Total Time (msecs) & Invocations & Single Invocation
% Time \\ \thickhline
% 10,000 & 1120 & 28751 & 0.0390 \\
% 15,000 & 1698 & 48204 & 0.0352 \\
% 20,000 & 2558 & 73470 & 0.0348 \\
% 30,000 & 3504 & 109909 & 0.0319 \\
% 50,000 & 6388 & 200005 & 0.0319 \\ 
% 100,000 & 12753 & 400005 & 0.0319 \\
% 200,000 & 25805 & 800005 & 0.0323 \\ \thickhline
% \end{tabular}
% \end{center}
% \caption{Verification performance results with JMSeq}
% \label{tab:results-verify}
% \end{table}

% The results show that JMSeq has a \textit{linear} co-relation to the
% number of invocations that it needs to verify (Figure
% \ref{fig:jmseq-performance}). In this example, we set
% up \smalltt{count} to increase in an ``exponential'' way, thus, JMSeq
% performance also is affected in an exponential manner pertaining to the
% linear co-relation to \smalltt{count}. In addition, it is worthwhile
% to note that generating the events from JPDA is another concern. In
% this regards, we tested different \emph{standard} implementations of
% JPDA including the standard reference implementation \cite{JPDA_Home}
% and Eclipse JDT implementation \cite{eclipse_jdt} in which both
% converging to the same results.

% Since JMSeq and the target Java program run in different JVMs, a future
% work for improving the performance of runtime verification is to run
% these JVMs in parallel, for example using a multi-core processor.
% Such a decoupling is possible because JMSeq is based on annotations
% that leave the execution of the original Java program intact.
% We expect a good speed-up because the communication between these two
% JVMs is limited to a one-way (and possibly asynchronous) sending of
% events.

\section{Related Work} \label{sec:relworks}

JML \cite{leavens_baker_ruby_jml_design} provides a robust and rigorous grounds for specification of behavioral checks on methods. 
Although JML covers a wide range of concerns in assertion checking, it does \textit{not} directly address the problem of method call sequence specification  as it is more directed towards the reasoning about the state of an object. 
JML is rather a comprehensive modeling language that provides many improvements to extensions of Design by Contract (DBC) \cite{dbc} including jContractor \cite{jcontractor} and Jass \cite{jass}.

In \cite{yoonsik_ash_mscs}, taking advantage of the concept of
\textit{protocols}, an extension to JML is proposed that provides a syntax for
method call sequences (protocols) along with JML's functional specification
features. Through this extension, the developer can specify the methods' call
sequence through a \textsl{call sequence clause} in JML-style meta-code. In the
proposed method, the state of a program is modeled as a ``history'' of method
calls and return calls using the expressiveness of regular expressions; thus, a
program execution is a set of ``transitions'' on method call histories. The
verification takes place when the execution history is simulated using a finite
state machine and checked upon the specified method call sequence clause.

JML features have been equivalently implemented with AspectJ constructs
\cite{jml_aspects}, using aspect-oriented programming \cite{kiczales_aspectj}.
The authors of this work propose AJMLC (AspectJ JMLC), which integrates AJC and JMLC into a single compiler
so that instead of JML-style meta-code specifications, the developer writes 
\textsl{aspects} to specify the requirements.

In \cite{stijn} an elegant extension of JML with histories is presented.
Attribute Grammars \cite{DBLP:journals/mst/Knuth68} are used as a formal
modeling language, where the context-free grammar describes properties of the
control-flow and the attributes describe properties of the data-flow, providing
a powerful separation of concerns. A runtime assertion checker is
implemented. Our approach differs in several respects. The
implementation of their runtime checker is based on code
instrumentation. Additionally, they use local histories of objects,
so call-backs can not be modeled. However, the behavior of a stack
can be modeled in their approach (and not in ours), since their
specifications are not regular expressions but context-free languages.

In the domain of runtime verification, Tracematches \cite{tracematches} enables the
programmer to specify events in the execution trace of a program that could be specified 
with ``the expressiveness'' of a regular pattern. The specification is done with 
AspectJ pointcuts and upon a match the advised code is run for the pointcut.
Along the same line, J-LO \cite{bodden_jlo} is a tool that provides temporal assertions in
runtime-checking. J-LO shares similar principles as Tracematches with differences 
in specifications using linear time temporal logic syntax.

Additionally, Martin et al. propose PQL \cite{martin_livshits_lam_pql} as a 
program execution trace query language. It enables the programmer to express queries 
on the execution events of objects, methods and their parameters. PQL then takes 
advantage of two ``static'' and ``dynamic'' checkers to analyze the application. The 
dynamic checker instruments the original code to install points of ``recovery'' and
``verification'' actions. The dynamic checker also translates the queries into
state machine for matching criteria. The set of events PQL can deal with
includes method calls and returns, object creations and end of program among
others. Accordingly, generic logic-based runtime verification frameworks are
proposed as in MaC \cite{java_mac}, Eagle \cite{eagle}, and PaX (PathExplorer)
\cite{pax} in which monitors are instrumented using the specification based on
the language specific implementations.

Using runtime verification concepts, Chen and Rosu propose MOP
\cite{chen_rosu_mop,MOP} as a generic runtime framework to verify programs
based on monitoring-oriented programming. As an
implementation of MOP,
JavaMOP \cite{chen_rosu_jmop} provides a platform supporting a large part of
runtime JML features. Safety properties of a program are specified and 
inserted into the program with monitors for runtime verification. Basically,
the runtime monitoring process in MOP is divided into two orthogonal
mechanisms: ``observation'' and ``verification''. The former stores the desired
events specified in the program and the latter handles the actions that are
registered for the extracted events. Our approach follows the same idea as MOP,
but it does not use AOP to implement it. Another major
difference is in the specifications part. MOP specifications are  generic in
four orthogonal segments: logic, scope, running mode and event handlers.  Very
briefly, the \textit{scope} section is the fundamental one that defines and
specifies the parts of the program under test. It also enables the user to
define desired events of the program that need to be verified. The
\textit{logic} section helps the user specify the behavioral specification of
the events using different notations such as regular expressions or
context-free grammars. The \textit{running mode} part lets the user specify
what is the running context of the program under test; for instance, if the
test needs to be run per thread or in a synchronized way. And, \textit{the
event handlers} section is the one to inject customized code of verification or
logic when there is a match or fail based on the event expression logic.
In Section \ref{sec:discuss-jmop}, we discussed how JMSeq contrasts with JavaMOP and similar approaches.

% In Listing \ref{lst:sample_mop} we show an example of JavaMOP code with ERE
% logic for the two specifications given in Figure \ref{fig:seq-spec}.

% \lstset{language=[AspectJ]Java, morekeywords={suffix, event,ere, CFG, @fail, @match, __RESET}} \begin{lstlisting}[label=lst:sample_mop, caption=Sample JavaMOP Specification using CFG logic] // Case 1
% SampleCase1(Main m) {
%  event method_m_a before(A a):
%     call(* A.m_a(..)) && target(a) {}

%  event method_m_b before(B b):
%     call(* B.m_b(..)) && target(b) && 
%     cflow(SampleCase1_method_m_a) {}

%  event method_m_c before(C c):
%     call(* C.m_c(..)) && target(c) && 
%     cflow(SampleCase1_method_m_a) && 
%     cflow(SampleCase1_method_m_b) {}

%  ere: (method_m_a method_m_b method_m_c)*

%  @fail {
%     System.err.println("Invalid Execution");
%     __RESET;
%  }     
% }

% // Case 2
% SampleCase2(Main m) {
%  event method_m_a before(A a):
%     call(* A.m_a(..)) && target(a) {}

%  event method_m_b before(B b):
%     call(* B.m_b(..)) && target(b) && 
%     cflow(SampleCase1_method_m_a) {}

%  event method_m_c before(C c):
%     call(* C.m_c(..)) && target(c) && 
%     cflow(SampleCase1_method_m_a) && 
%     !cflow(SampleCase1_method_m_b) {}

%  ere: (method_m_a method_m_b method_m_c)*

%  @fail {
%     System.err.println("Invalid Execution");
%     __RESET;
%  }     
% }
% \end{lstlisting}

% The code snippet presented in Listing \ref{lst:sample_mop} is one of
% several ways to express the problem described in Figure
% \ref{fig:seq-spec}. JavaMOP allows the use of AspectJ language to
% define the scope of the program points to be verified. This may look
% advantageous since the programmer can use different syntax in
% AspectJ. However, it leads to low-level specification of control flow
% using AspectJ in addition to JavaMOP logic section syntax. In
% particular, the \smalltt{cflow} keyword in the event definitions
% (scope section), already specifies the logic. In contrast, JMSeq allows
% for purely declarative specification language which completely captures
% the operational behavior using regular expressions. For instance,
% in Section \ref{sec:spec}, we used
% %\begin{flushleft}
% %\includegraphics[scale=1]{grammar-sample-usage-1}
% \input{grammar-sample-usage-1-1}
% %\end{flushleft}
% and
% %\begin{flushleft}
% %\includegraphics[scale=1]
% \input{grammar-sample-usage-2-1}
% %\end{flushleft}
% to specify the same program in Figure \ref{fig:seq-spec}.
% Another major difference between JMSeq and JavaMOP is the use of
% annotations versus instrumentation as already discussed in Section
% \ref{sec:intro}.

% It is interesting to note that both MOP specifications have \textit{the same
% ERE expression}. This is because JavaMOP has separated logic and scope
% specification from event handling. The difference in monitoring is
% obtained by the different usage of the command {\small\texttt{cflow}} in the
% scope section. This command is an AspectJ construct to control the context of
% the execution when running some code inside a method
% \cite{kiczales_aspectj}. Seemingly, even though JavaMOP claims to
% support a separation of concerns, the event definitions are mixed with
% behavioral specification in aspects declarations such as
% \smalltt{cflow}. In contrast, JMSeq provides all the specification
% through the annotations that are used by the programmer user.

\section{Conclusion and future work} \label{sec:conclusion}

We proposed JMSeq, a framework for specifying sequences of possibly nested method calls using Java annotations. 
The sequences do not only consist of method names, but may contain information such as object caller and callee, and distinguish overloaded methods. 
JMSeq uses Java Platform Debugger to monitor the execution of a component-based system in Java. 
Monitoring is divided into two phases: observing the events in the program and verifying them at runtime against the local specifications provided through annotated methods. 
JMSeq does not use \emph{any} form of instrumentation; neither source level nor bytecode level.
Moreover, JMSeq provides a simple way to express exceptions and verify them in the execution of the target application. JMSeq is integrated with JUnit.
We presented a case study from Fredhopper distributed cloud services in the domain of e-commerce and marketing. 
Moreover, we provided a set of performance results for JMSeq in different settings for program execution and runtime verification.

JMSeq is a novel approach to runtime verification of software using code annotations. 
The approach is especially suitable for runtime component-based verification. 
We improved the execution data model of JMSeq using optimization techniques such as in-memory object caches with indices.
In the context of event processing, we examined different implementations of JPDA and JDI standards (e.g. using Eclipse platform \cite{eclipse_debug_platform}) and it showed part of the optimization is out of our reach. 
We used VisualVM, a standard Java profiling tool, to measure and discuss the performance of JMSeq.

A future line of work consists of porting JMSeq to multi-core platforms. 
Concurrent and distributed software introduces new challenges for runtime verification of applications. 
In a concurrent setting, objects interact with each other from different threads of execution (which may optionally involve synchronization and locking events).
JMSeq should be able to verify such objects in different threads and consider how synchronization and locking may affect the runtime or flow of an application.
In a distributed context, application objects are located in (physically) remote JVM instances.
In such scenarios, JMSeq needs to load or observe a topology of the involved JVMs and their objects to be able to verify the application for correct sequences of method calls.

We plan to extend JMSeq by providing native features such as mock \cite{frank:mock} implementations in case parts of the system are unavailable.
JMSeq may provide better integration for mocking parts of the system that are not implemented or not available.
Such features will be especially useful in the context of integration testing and verification.
At the time of integration, not all components of the system may be available and complete; and yet, the application should be verified in the parts that are complete.
In comparison with runtime checking, another line of future work can be to extend JMSeq to support static verification of protocols (using approaches as in Copilot \cite{copilot}).

% Moreover, in another direction, the testing framework can even take control of the
% underlying program execution for further checks or verification. 
% In other words, method call sequence specifications may also be used in the pre-conditions of a method such that a method is executed only if it is allowed in the high-level specification.

Finally, we plan to  extend the specification language to a context free language. 
This is in fact simple as JMSeq already implements a simple form of pushdown automata to recognize the call-return structure of Java.
In fact, Java is expressible by a restricted form of deterministic pushdown automata, called  visibly pushdown automata \cite{Alur_abstractvisibly}, where push and pop correspond to call and return, respectively.

% Another potential area of improvement is the data model used
% by JMSeq. Currently
% it overlaps with the event model that JPDA provides when publishing the
% registered events in JVM. The resulting overhead is rather expensive.
% Optimization will allow, for example, to store only the minimal necessary
% information providing a faster indexing for the retrieval of the events.
% Further performance improvement can be obtained by better exploiting the
% connections between JPDA and JVM. For example, there are tools such as Eclipse
% Debug Platform \cite{eclipse_debug_platform} that provide extensive
% facilities through JPDA with high performance. Currently, in JMSeq, it is the
% standard JVM that runs the program and publishes the events that are
% interesting to JMSeq for further verification for which, in turn, JMSeq uses a
% simple state machine to verify the executing events.

% \paragraph{Acknowledgments}
% We thank the referees for their comments and Michiel Helvensteijn for 
% the helpful discussion.

% \bibliographystyle{abbrv}
% \bibliography{references}

% \end{document}

### 2015-11-24 p01-ch01-2012/paper.tex
Add publication information to chapter abstracts

@@
 \chapter[Application-Level Scheduling]{Programming and Deployment of Active Objects with Application-Level Scheduling%
-\footnote{This work is partially supported by the EU FP7-231620 project: HATS.}
-}
+\footnote{This work is partially supported by the EU FP7-231620 project: HATS.}}
 % 
@@
 \chapterauthor{Behrooz Nobakht, Frank S. de Boer, Mohammad Mahdi Jaghoori,  Rudolph Schlatte}
-
-% --- Author Metadata here ---
-% \conferenceinfo{SAC'12}{March 25-29, 2012, Riva del Garda, Italy.}
-% \CopyrightYear{2011}
-% \crdata{978-1-4503-0857-1/12/03}
-% --- End of Author Metadata ---
-
-% \title{Programming and Deployment of Active Objects with Application-Level Scheduling%
-% \titlenote{This work is partially supported by the EU FP7-231620 project: HATS.}
-% }
-
-% \numberofauthors{2}  %set to two to enable longer lines for names and
-
-% \author{
-% \alignauthor Behrooz Nobakht\\
-% 	\affaddr{Leiden Institute of Advanced Computing}\\
-% 	\affaddr{Universiteit Leiden, Leiden, The Netherlands}\\
-% 	\email{bnobakht@liacs.nl}
-% \alignauthor Frank S. de Boer\\
-% 	\affaddr{Centrum Wiskunde and Informatica}\\
-% 	\affaddr{Amsterdam, The Netherlands}\\
-% 	\email{frb@cwi.nl}
-% \and
-% \alignauthor Mohammad Mahdi Jaghoori\\
-% 	\affaddr{Centrum Wiskunde and Informatica}\\
-% 	\affaddr{Amsterdam, The Netherlands}\\
-% 	\email{jaghoori@cwi.nl}
-% \alignauthor Rudolf Schlatte\\
-% 	\affaddr{University of Oslo}\\
-% 	\affaddr{Oslo, Norway}\\
-% 	\email{rudi@ifi.uio.no}
-% }
 
@@
 based on active objects which allows for high-level control of deployment related issues.
+
+\subparagraph*{Conference Publication}
+\emph{Proceedings of the 27th Annual ACM Symposium on Applied Computing -- ACM SAC 2012, Pages 1883--1888, DOI \smalltt{10.1145/2245276.2232086}}
 
@@
 % End:
+
+\lstset{escapeinside={}{}}

### 2015-11-30 p00-ch00-intro/intro.tex
Add publication information to chapter abstracts
Polish intro (spelling and grammar)

@@
 Method invocations are \emph{blocking}; i.e. the caller object waits until it receives the result of the method call from the callee object.
-Surprisingly, such model of interaction was \emph{never} the intention of the creator of the paradigm.
-Originally, object interactions were meant to be \emph{messaging} among objects and objects behaved as autonomous entities possibly on remote locations on a network; Alan Kay clarified later~\cite{alank1,alank2}.
+Surprisingly, this model of interaction was \emph{never} the intention of the creator of the paradigm.
+Originally, object interactions were meant to be \emph{messages} among objects and objects behaved as autonomous entities possibly on remote locations on a network; Alan Kay clarified later~\cite{alank1,alank2}.
 On the contrary, almost all the object-oriented languages at hand have followed the blocking and synchronous model of messaging based on a common misunderstanding. 
-The original definition of object-oriented is close to another model of computation: actor model.
+The original definition of object-oriented programming is close to another model of computation: actor model.
 
@@
 One object, the sender, communicates a message to the other object, the receiver.
-At the receiver side, a message eventually leads to a method invocation (in  object-oriented paradigm) or a function call (in functional paradigm).
-In actor model, the locality of the objects is transparent from the messages.
+At the receiver side, a message eventually leads to a method invocation (in the object-oriented paradigm) or a function call (in the functional paradigm).
+In the actor model, the locality of the objects is not visible in the messages.
 A system is composed of objects with communicating messages.
 
-Considerable amount of research have been carried out to mix object-oriented programming languages with actor model. 
-Language extensions, libraries, and even new programming languages are the naturual outcomes of such research.
+A considerable amount of research has been carried out to mix object-oriented programming languages with the actor model. 
+Language extensions, libraries, and even new programming languages are the natural outcomes of such research.
 The more mainstream the language, the more research and development are available in the intersection of object orientation and actor model.
-For example, \cite{actor_frameworks_jvm:agha} presents a comparative analysis of such research on Java and JVM languages.
+For example, \cite{actor_frameworks_jvm:agha} presents a comparative analysis of such research for Java and JVM languages.
 
-With rapid increase of computation power, a new aspect added to create a triangle: object-oriented, actor model, and \emph{concurrency}.
+The rapid increase of computational power added a new aspect added to create a triangle: object-oriented, actor model, and \emph{concurrency}.
 Concurrency makes it harder to verify programs in terms of correctness of runtime behavior. 
@@
 In addition, in the presence of scheduling and thread interleaving, it is important that objects are able to protect their state from concurrent modifications for correctness properties.
-In this research, we take advantage of a modelling language that aims to mix the three aspects in addition to the capabilities of formal semantics and verification.
+In this thesis, we take advantage of a modelling language that aims to mix the three aspects in addition to the capabilities of formal semantics and verification.
 
 ABS~\cite{johnsen2012abs,hahnlehjlssw11} is a modelling language for concurrent objects and distributed systems.
-ABS provides the semantics of actor model~\cite{actors:agha} as a modelling language.
-ABS language provides a set of core features including algebraic data types (ADT), functional layer, and concurrency layer.
-ABS concurrency layer lays its basis on co-operative scheduling of active objects~\cite{creol:broch:owe}.
+ABS is based on the semantics of the actor model~\cite{actors:agha}.
+ABS language provides a set of core features including algebraic data types (ADT), a functional layer, and a concurrency layer.
+The ABS concurrency layer uses co-operative scheduling of active objects~\cite{creol:broch:owe}.
 ABS semantics is completely expressed in structural operational semantics~\cite{plotkin:sos}.
@@
 % 
-Co-operative scheduling in ABS has been additionally extended for real-time scheduling considering priorities and time constraints~\cite{bjork2013:rtabs,johnsen2012modeling}.
+Co-operative scheduling in ABS has been additionally extended for real-time scheduling taking priorities and time constraintsinto account~\cite{bjork2013:rtabs,johnsen2012modeling}.
 Thus, in general, ABS concurrency layer is a perfect fit if it can be integrated with industry-level real-world programming languages.
@@
 \footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}}, Java~\cite{gosling2000java} is one of the most commonly-used.
-Additionally, since Java 1.5 and the rise of concurrency API in Java~\cite{jsr166}, a remarkable effort by the community focus on how concurrency and distribution can be improved in Java as in provided by ABS semantics and actor model.
-However, not all the development as Java libraries and frameworks form their basis on formal semantics. 
+Since Java 1.5 and the rise of the concurrency API in Java~\cite{jsr166}, substantial effort is spend by the community on how concurrency and distribution can be improved in Java, along the lines of the actor model, as provided by ABS.
+However, not all the development on Java libraries and frameworks are based on a rigorous formal semantics. 
 Therefore, this gives rise to issues and challenges in terms of correctness, semantic preservation, and reasoning.
@@
 % However, 
-Java language specification (JLS)~\cite{gosling2000java} clearly concludes that Java language is not comptabile and ready to be extended as a functional language with the support of algebraic data types.
+The Java language specification (JLS)~\cite{gosling2000java} clearly concludes that Java language is not comptabile and ready to be extended with a functional language that supports algebraic data types.
 This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
-Scala language~\cite{odersky2004scala} is a result of the research that is a dynamic, functional, and object-oriented language on JVM.
+The Scala language~\cite{odersky2004scala} is a result of the research. It is a dynamic, functional, and object-oriented language on JVM.
 Although such extensions over Java towards a functional language exist, they lack formal semantics in the context of concurrency and actor model.
@@
 Our approach enables programmers to model their Java systems in terms of ABS concurrency semantics and then create a correct mapping and implementation from ABS to Java using an API.
-In addition, we aim to facilitate ways to actually program in ABS models as if the programmer is able to take advantage of standard Java libraries.
+In addition, we aim to facilitate ways to actually program ABS models taking advantage of the standard Java libraries.
 
@@
 In this section, we present a high-level overview of design goals that we pursue in this research. 
-Equivalently, the effectivness and success of this research is measured upon the achievement of the discussed goals and objectives.
+%Equivalently, the effectiveness and success of this research is measured upon the achievement of the discussed goals and objectives.
 
@@
 Polyglot programming essentially enables software practice to apply the \emph{right} language in the appropriate layer of the problem statement. 
-In this research, we follow the question in which we try to deliver ABS semantics and features in a polyglot approach. 
-The programmer develops models with ABS that \emph{partically} take advantage of the target language features (e.g. from Java).
+In this research, we aim to deliver ABS semantics and features in a polyglot approach. 
+The programmer develops models with ABS that \emph{partially} take advantage of the target language features (e.g. from Java).
 This approach is also referred to as \emph{Foreign Function Interface} in the context of ABS modelling.
-Listing~\ref{lst:abs:java} shows an ABS code that uses \jtt{java.util.ArrayList} as a data strucuture.
+Listing~\ref{lst:abs:java} shows ABS code that uses \jtt{java.util.ArrayList} as a data strucuture.
 With Polyglot programming in ABS, the programmer does \emph{not} need to define the abstractions necessary for the list construct.
@@
 \item \emph{Compiling ABS to a target programming language}
-One first objective is to compile an ABS model to a target programmig language.
+One first objective is to compile an ABS model to a target programming language.
 The compilation can be at \emph{source-to-source} or \emph{source-to-bytecode}
@@
 \item \emph{Using ABS semantics as an API in a programming language}
-ABS language is precisely expressed with structural operational semantics~\cite{johnsen2012abs} in addition to its syntax definition.
-Therefore, the semantics of ABS can be delivered in a programming language
+The ABS language is precisely expressed with structural operational semantics~\cite{johnsen2012abs} in addition to its syntax definition.
+The semantics of ABS can be delivered in a programming language
 as Application Programming Interface (API) as long as the programming language provides sufficient constructs to respect ABS language semantics.
 In addition, such ABS programming API should be \emph{verifiable}.
-If such ABS mapping to a programming API is provided, a programmer is able to
+If an ABS mapping to a programming API is provided, a programmer is able to
 take advantage of ABS semantics without directly programming in ABS.
-Such capability from ABS enables industry users of mainstraim languages to
+Such capability from ABS enables industry users of mainstream languages to
 model their systems in ABS semantics using their programming languages and
@@
 \item \emph{Modelling in ABS}
-ABS language provides solid structure and semantics to model concurrent
+ABS language provides a rigorous semantics to model concurrent
 and distributed systems.
-The user (that can be a programmer, an analysist, or a researcher) rightfully
-demands to have access to a tool-set and IDE that allows working with ABS 
+For practical purposes, the user (that can be a programmer, an analysist,
+or a researcher) desires a tool-set and IDE that allows working with ABS 
 models in a user-friendly way.
@@
 \paragraph{Extensibility}
-Delivering ABS language features as an API brings about a new requirement: to configure or extend the API. 
 It is a common practice in software engineering that components are developed with a mixture of \emph{configuration}, \emph{extension} and \emph{composition}.
-Therefore, the ABS programmer should be able to extend such API to custom requirements. 
-The extensibility is provided either through configuration of the API or API extension.
+Hence, delivering the ABS language features as an API brings about a new requirement: to configure or extend the API. 
+In particular, the ABS programmer should be able to extend the API to customized requirements. 
+%The extensibility is provided either through configuration of the API or API extension.
 
@@
 We discuss a brief overview of related work in the context of programming languages, actor model, and concurrency.
-We devide the overview in two levels; one is at the level of the programming languages and the other is for the external (third-party) libraries developed for programming languages.
+We divide the overview in two levels; one is at the level of the programming languages and the other is for the external (third-party) libraries developed for programming languages.
 
@@
 
-In this section, we briefly provide an overview of the programming languages that have targetted the similar problem statements.
+In this section, we briefly provide an overview of the programming languages that have targetted similar problem statements.
 It is important to note the timeline and the rapid evolution of programming languages trying to provide an actor-based model of asynchronous message passing.
@@
 \item[First-Class Citizen]
-is the type of programming languages that deliver actor model programming support as a first-class citizen of the language.
-In such languages, actor model is by-design part of the syntax and semantics of the language.
+%is the type of programming languages that deliver actor model programming support as a first-class citizen of the language.
+Languages in which the actor model is by-design part of the syntax and semantics of the language.
 \item[Implicit By Design]
 refers to the type that there is no explicit notion of actor models in the language syntax or semantics.
-However, the programming language provides fundamental constructs for concurrenty and asynchronous message passing.
-Thus, it becomes a trivial task in such type of programming languages to create an abstraction to support the actor model by coding.
+However, the programming language provides fundamental constructs for concurrency and asynchronous message passing.
+Thus, it becomes an easy task in this kind of programming language to create an abstraction to support the actor model by coding.
 \item[External Library]
-refers to the type of programming languages for which actor model support is possible through using an external library developed for the language.
+refers to the type of programming languages for which actor model support is provided by an external library developed for the language.
 \end{description}
@@
 % In the previous section, we presented an overview of the programming languages and their categories that aim to provide actor model~\cite{agha86book}.
-Since many programming languages faced different challenges to provide the necessary syntax and semantics for actor model and concurrency at the level of the language, many libraries and frameworks have been born with a motivation to fill this gap.
+Since many programming languages faced different challenges to provide the necessary syntax and semantics for actor model and concurrency at the level of the language, many libraries and frameworks aim to fill this gap.
 We observe that the more the language itself is close to the actor model semantics, the less external libraries and frameworks target this gap. 
@@
 We formalize a subset of Java language in regards with ABS.
-We provide the operaional semantics of the concurrency layer for scheduling of actors with an extension for deadlines and priorities. 
+We provide the operational semantics of the concurrency layer for scheduling of actors with an extension for deadlines and priorities. 
 We deliver ABS concurrency as a Java API.
@@
 We sketch a general architecture on how ABS language can be modularly translated into different programming languages.
-We present how ABS API library can be used in the translation of ABS models into Java.
+We present how this ABS API library can be used to translate ABS models into Java.
 
@@
 \\ \midrule 
-Monitoring method call sequences using annotations & Application (Part~\ref{p:app}) & Chapter X
+Monitoring method call sequences using annotations & Application (Part~\ref{p:app}) & Chapter~\ref{ch:p03:ch02}
 \\ \midrule

### 2015-12-06 p00-ch00-intro/intro.tex
Review 1 - Notes and TODO's

@@
 Originally, object interactions were meant to be \emph{messages} among objects and objects behaved as autonomous entities possibly on remote locations on a network; Alan Kay clarified later~\cite{alank1,alank2}.
-On the contrary, almost all the object-oriented languages at hand have followed the blocking and synchronous model of messaging based on a common misunderstanding. 
+On the contrary, almost all the object-oriented languages at hand have followed the blocking and synchronous model of messaging. 
 The original definition of object-oriented programming is close to another model of computation: actor model.
@@
 At the receiver side, a message eventually leads to a method invocation (in the object-oriented paradigm) or a function call (in the functional paradigm).
-In the actor model, the locality of the objects is not visible in the messages.
-A system is composed of objects with communicating messages.
+Actor model features location transparency; i.e. the physical location of objects is not visible to other objects.
+A system is composed of objects that communicate through messages.
 
 A considerable amount of research has been carried out to mix object-oriented programming languages with the actor model. 
++\TODO{Find good references.}
 Language extensions, libraries, and even new programming languages are the natural outcomes of such research.
@@
 Concurrency makes it harder to verify programs in terms of correctness of runtime behavior. 
++\TODO{Why concurrency makes it harder?}
 Scheduling of object messages in threads of executions is one of main challenges.
@@
 In this thesis, we take advantage of a modelling language that aims to mix the three aspects in addition to the capabilities of formal semantics and verification.
++\TODO{Explain the gain of utilization from concurrency perspective.}
 
@@
 Co-operative scheduling in ABS has been additionally extended for real-time scheduling taking priorities and time constraintsinto account~\cite{bjork2013:rtabs,johnsen2012modeling}.
-Thus, in general, ABS concurrency layer is a perfect fit if it can be integrated with industry-level real-world programming languages.
-% With distributed computing as one of the recent de-facto requirements in every business use case, an enormous effort has gone to improve the current mainstream programming languages to fit for the new requirements of distributed and concurrent sett1ings.
+All the above characteristics make the ABS language an attractive choice if it
+can be used as a programming language at industrial and business scale.
+However, ABS has so far remained mainly as a modelling language.
+\TODO{Check with Frank.}
 
+One core challenge is to combine ABS semantics (e.g. concurrency layer) in a
+programming language.
+\TODO{Elaborate the core contributions of this thesis}
 Considering the mainstream programming languages
@@
 The Scala language~\cite{odersky2004scala} is a result of the research. It is a dynamic, functional, and object-oriented language on JVM.
-Although such extensions over Java towards a functional language exist, they lack formal semantics in the context of concurrency and actor model.
+However, these extensions and libraries lack formal semantics in the context
+of concurrency and actor model.
 
@@
 \begin{lstlisting}[float=h,language=Java,caption=Using Java in ABS,label=lst:abs:java]
-List<String> params = new ArrayList<>();
-myObj ! doSomething(params);   
+List<String> params = new ArrayList<>();    // Java
+myObj ! doSomething(params);                // ABS
 \end{lstlisting}
@@
 influence from the number of asynchronous messages delivered in the system.
+\TODO{Discuss with Frank and drop this.}
 
 \paragraph{Modularity} 
+One of the main goals of this research is to create a modular and reusable
+software artifact. 
+This can be achieved only by producing an API for ABS semantics that is
+indepedent from any specific ABS backend.
+\TODO{Elaborate on why an API library is modular for ABS}
 The scope of the research spans to a number of layers revolving around ABS
@@
 One first objective is to compile an ABS model to a target programming language.
-The compilation can be at \emph{source-to-source} or \emph{source-to-bytecode}
-level.
+% The compilation can be at \emph{source-to-source} or \emph{source-to-bytecode} level.
+In addition, to be able to integrate generated code from ABS into other legacy systems, it is crucial that generated code is readable and of acceptable quality.\TODO{Use software engineering terms here}
 Target languages potentially include mainstream programming languages such as
@@
 their ABS models to the target languages.
-\item \emph{Using ABS semantics as an API in a programming language}
+\TODO{Why not the previous ABS Java backend suffered from software engineering}
+\item \emph{Using ABS concurrency as an API in an existing programming language}
 The ABS language is precisely expressed with structural operational semantics~\cite{johnsen2012abs} in addition to its syntax definition.
@@
 and distributed systems.
-For practical purposes, the user (that can be a programmer, an analysist,
-or a researcher) desires a tool-set and IDE that allows working with ABS 
+For practical reasons, the user (that can be a programmer, an analysist, or a researcher) 
+desires to have access to a tool-set and IDE that allows working with ABS 
 models in a user-friendly way.
@@
 Hence, delivering the ABS language features as an API brings about a new requirement: to configure or extend the API. 
-In particular, the ABS programmer should be able to extend the API to customized requirements. 
+In particular, the ABS programmer should be able to extend the API to customized requirements. \TODO{Expand?}
 %The extensibility is provided either through configuration of the API or API extension.
@@
 Figure~\ref{fig:arch} presents the general architecture of the implementation.
+
+\TODO{Relate the criteria in this section on how they lead to a natural modular architecture in the figure.}
 
@@
 In this section, we briefly provide an overview of the programming languages that have targetted similar problem statements.
-It is important to note the timeline and the rapid evolution of programming languages trying to provide an actor-based model of asynchronous message passing.
+It is important to note the rapid evolution of programming languages trying to provide an actor-based model of asynchronous message passing.\TODO{Reference}
 In addition, we identify different types of research and work related to actor model and concurrency programming:
@@
 Languages in which the actor model is by-design part of the syntax and semantics of the language.
+\TODO{Cite polylang as a research recent lang.}
+\TODO{Why this category is the best}
 \item[Implicit By Design]
@@
 % In the previous section, we presented an overview of the programming languages and their categories that aim to provide actor model~\cite{agha86book}.
-Since many programming languages faced different challenges to provide the necessary syntax and semantics for actor model and concurrency at the level of the language, many libraries and frameworks aim to fill this gap.
+Since many programming languages faced different challenges to provide the necessary syntax and semantics for actor model and concurrency at the level of the language, many libraries and frameworks aim to fill this gap (c.f.
+Table~\ref{tbl:actor:libs}).
 We observe that the more the language itself is close to the actor model semantics, the less external libraries and frameworks target this gap. 
@@
 
-\section{Oraganization and Contributions}
+\section{Organization and Contributions}
 \label{sec:intro:contribs}
 
+\TODO{Connect back to the problem statement paraph}
+
 In this thesis, we focus on the concurrency layer of ABS.
-We formalize a subset of Java language in regards with ABS.
+We formalize a subset of Java language in regards with ABS\TODO{Check for the way around}.
 We provide the operational semantics of the concurrency layer for scheduling of actors with an extension for deadlines and priorities. 

### 2016-01-12 p00-ch00-intro/intro.tex
Apply review comments to Introduction
Add a paragraph of the core contributions of the thesis

@@
 
-A considerable amount of research has been carried out to mix object-oriented programming languages with the actor model. 
-+\TODO{Find good references.}
+A considerable amount of research has been carried out to mix object-oriented programming languages with the actor model~\cite{philippsen2000survey}. 
 Language extensions, libraries, and even new programming languages are the natural outcomes of such research.
@@
 The rapid increase of computational power added a new aspect added to create a triangle: object-oriented, actor model, and \emph{concurrency}.
-Concurrency makes it harder to verify programs in terms of correctness of runtime behavior. 
-+\TODO{Why concurrency makes it harder?}
+Concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}. 
 Scheduling of object messages in threads of executions is one of main challenges.
@@
 In this thesis, we take advantage of a modelling language that aims to mix the three aspects in addition to the capabilities of formal semantics and verification.
-+\TODO{Explain the gain of utilization from concurrency perspective.}
+
+The core contributions of this thesis target at the intersection of object 
+orientation, actor model, and concurrency.
+We choose Java as the main programming language and as one of the mainstream 
+object-oriented languages. 
+We formalize a subset of Java and its concurrency API~\cite{jsr166} to 
+facilitate formal verification and reasoning about it.
+We create an abstract mapping from a concurrent-object modelling language, 
+ABS~\cite{johnsen2012abs}, to programming semantics of Java concurrency. 
+We provide the formal semantics of the mapping and runtime properties of 
+the concurrency layer including deadlines and scheduling policies.
+We provide an implementation of ABS concurrency layer as a Java API library 
+and framework utilizing the latest langauge additions 
+in Java 8~\cite{jsr335:lambda:translation}.
+We design a large-scale monitoring system as a real-world application; 
+the monitoring system is built with actors and ABS concurrency 
+with formal semantics of ABS composed with 
+schedulability analysis~\cite{fersman2007task}.
+We elaborate on the elements of this research in the following.
 

### 2016-01-12 p01-ch02-2013/paper.tex
Add publication information to chapter abstracts
Remove all `vspace`

@@
 illustrate the use of its constructs with an industrial case study from distributed e-commerce and marketing domain.
+
+\subparagraph*{Conference Publication}
+\emph{Lecture Notes in Computer Science, Volume 7890, Coordination Models and Languages in 15th conference on Distributed Computing Techniques -- COORD 2013, Pages 181--195, DOI \smalltt{10.1007/978-3-642-38493-6\_13}}
 
@@
 \lstset{language=Creol,escapeinside={(*}{*)}} 
-\vspace{-20pt}
+% \vspace{-20pt}
 \begin{lstlisting}[float=h, label=lst:try:catch, caption=Using try-catch for processing future values]
@@
 \end{lstlisting}
-\vspace{-20pt}
+% \vspace{-20pt}
+
 For example, in Listing \ref{lst:try:catch}, if the \gett \space operation raises an exception  control,  is transferred to line (5);
@@
 
-\begin{wrapfigure}{r}{0.4\textwidth}
-\vspace{-40pt}
+\begin{figure}[h]
+% \begin{wrapfigure}{r}{0.4\textwidth}
+% \vspace{-40pt}
 \begin{center}
@@
 \end{center}
-\vspace{-20pt}
+% \vspace{-20pt}
 \caption{Fredhopper's Controller life cycle for remote data processing}
 \label{fig:frh-dm}
-\vspace{-10pt}
-\end{wrapfigure}
+% \vspace{-10pt}
+% \end{wrapfigure}
+\end{figure}
 
@@
 The right code snippet shows the one that uses messages with deadlines.
+
 \lstset{language=Creol,escapeinside={(*}{*)}} 
@@
 \begin{minipage}[t]{0.48\textwidth}
-\vspace{-25pt}
+% \vspace{-25pt}
 \begin{lstlisting}[label=lst:frh:polling, caption=With polling]
@@
 \begin{minipage}[t]{0.48\textwidth}
-\vspace{-25pt}
+% \vspace{-25pt}
 \begin{lstlisting}[label=lst:frh:deadlines, caption=With deadlines]
@@
 \end{minipage}
-\vspace{-10pt}
+% \vspace{-10pt}
 \end{center}
@@
 To model and implement this custom behavior, a custom scheduler is developed for the data processing node.
+
 \lstset{language=Creol,escapeinside={(*}{*)}} 
@@
 \begin{minipage}[t]{0.48\textwidth}
-\vspace{-30pt}
+% \vspace{-30pt}
 \begin{lstlisting}[label=lst:frh:dm, caption=Data Processor class]
@@
 \begin{minipage}[t]{0.48\textwidth}
-\vspace{-30pt}
+% \vspace{-30pt}
 \begin{lstlisting}[label=lst:frh:sched, caption=Custom scheduler]
@@
 \end{minipage}
-\vspace{-15pt}
+% \vspace{-15pt}
 \end{center}
+
 In the above listings, Listing \ref{lst:frh:sched} defines a custom scheduler that determines the priority of two processes with custom logic for specific customer.
@@
 We define \emph{customer satisfaction} as a property that represents the effectiveness of futures with deadline.
-\begin{wraptable}{l}{0.25\textwidth}
-\vspace{-15pt}
+
+\begin{table}[h]
+% \begin{wraptable}{l}{0.25\textwidth}
+% \vspace{-15pt}
 \begin{tabular}{c c}\thickhline
@@
 \label{tbl:eval}
-\vspace{-15pt}
-\end{wraptable}
+% \vspace{-15pt}
+% \end{wraptable}
+\end{table}
+
 For a customer $c$, the satisfaction can be denoted by $s = \frac{r^F_c}{r_c}$; in which $r^F_c$ is the number of finished data processing jobs and $r_c$ is the total number of requested data processing jobs from customer $c$.
@@
 % \bibliography{references}
+
+\lstset{escapeinside={}{}}

### 2016-01-12 p01-ch01-2012/paper.tex
Remove all `vspace`
Fix long captions in figures and tables

@@
 ~\\[-3.5cm]ADD:\\[2.5cm]
-  \caption{Adding the new MethodInvocations are performed on the Client side.}
+\caption[New MethodInvocation]{Adding the new MethodInvocations are performed on the Client side.}
 \label{fig:add}
@@
 ~\\TAKE:\\
-%\begin{center}
+\begin{center}
 %\vspace{-1.3cm}
   \includegraphics[scale=0.3]{figs/seq-diag-2}
-\vspace{-.7cm}
-  \caption{An active object selects a method invocation based on its local scheduling policy. After a method finishes execution, the whole scenario is repeated.}
+% \vspace{-.7cm}
+\caption[Policy-based selection of a MethodInvocation]{An active object selects a method invocation based on its local scheduling policy.
+After a method finishes execution, the whole scenario is repeated.}
+\end{center}
 \label{fig:take}
-%\end{center}
 \end{figure}
@@
 ~\\EXECUTE:\\
-% \begin{center}
+\begin{center}
 % \vspace{-1.3cm}
  \includegraphics[scale=0.35]{figs/seq-diag-3}
-\vspace{-.7cm}
-\caption{A method invocation is executed in an interruptible process. The execution manager thread is blocked while the interruptible process is running.}
+% \vspace{-.7cm}
+\caption[Execution of a MethodInvocation]{A method invocation is executed in an interruptible process. The execution manager thread is blocked while the interruptible process is running.}
+\end{center}
 \label{fig:execute}
-% \end{center}
 \end{figure}

### 2016-01-12 p02-ch01-2014/paper.tex
Add publication information to chapter abstracts
Remove all `vspace`
Fix long captions in figures and tables

@@
 We present briefly the results of a set of experiments which provide evidence of  the potential impact of lambda expressions in Java~8  regarding the adoption of the actor concurrency model in large-scale distributed applications.
+
+%
+
+\subparagraph*{Conference Publication}
+\emph{Lecture Notes in Computer Science, Volume 8803, 6th International Symposium on Leveraging Applications of Formal Methods, Verification and Validation with Specialized Techniques and Applications -- ISoLA 2014, Pages 37--53, DOI
+\smalltt{10.1007/978-3-662-45231-8\_4}}
+
+%
 
@@
 
-\begin{wrapfigure}{r}{0.48\textwidth}
-\vspace{-20pt}
+\begin{figure}[h]
+% \begin{wrapfigure}{r}{0.48\textwidth}
+% \vspace{-20pt}
 \begin{lstlisting}[caption=main in Akka,label=lst:main:akka]
@@
 \end{lstlisting}
-\vspace{-20pt}
-\end{wrapfigure}
+% \vspace{-20pt}
+% \end{wrapfigure}
+\end{figure}
 
@@
 %
+
 \lstset{language=Java}
-\begin{wrapfigure}{r}{0.48\textwidth}
-\vspace{-20pt}
+\begin{figure}[h]
+% \begin{wrapfigure}{r}{0.48\textwidth}
+% \vspace{-20pt}
 \begin{lstlisting}[caption=main in ABS,label=lst:main:ABS]
@@
 \end{lstlisting}
-\vspace{-20pt}
-\end{wrapfigure}
-% 
+% \vspace{-20pt}
+% \end{wrapfigure}
+\end{figure}
+
 Asynchronous method calls use futures as dynamically generated references  to return values.
@@
 \end{center}
-\caption{
+\caption[Benchmarking comparison of ABS API and Akka]{
 Benchmark results of comparing sampling time of message round trips in ABS API and Akka.

### 2016-01-12 p03-ch01-2015-monitoring/sec05-monitoring-model.tex
Remove all `vspace`
Fix long captions in figures and tables

@@
 \begin{figure}[h]
-\vspace{-15pt}
+% \vspace{-15pt}
 \captionsetup[subfigure]{font=footnotesize}
@@
 }
-\vspace{-10pt}
+% \vspace{-10pt}
 % \caption{{\scriptsize Task Automata for $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$}}
-\vspace{-10pt}
+% \vspace{-10pt}
 \end{figure}
@@
 \begin{figure}[h]
-\vspace{-20pt}
+% \vspace{-20pt}
 \captionsetup[subfigure]{font=scriptsize}
@@
 }
-\vspace{-10pt}
+% \vspace{-10pt}
 % \caption{{\scriptsize Task automata to execute $\jtt{verify}_\alpha$ and $\jtt{verify}_\beta$}}
-\vspace{-10pt}
+% \vspace{-10pt}
 \end{figure}
@@
 \begin{figure}[h]
-\vspace{-20pt}
+% \vspace{-20pt}
 \centering
@@
 \end{tikzpicture}
-\vspace{-15pt}
-\caption{{\scriptsize{$M_{\jtt{A}_s}$: Timed Automaton to execute task type \jtt{allocate} in $M_P$}}
-\vspace{-20pt}
-\label{fig:fsm:alloc_s}}
+% \vspace{-15pt}
+\caption{$M_{\jtt{A}_s}$: Timed Automaton to execute task type \jtt{allocate} in $M_P$}
+% \vspace{-20pt}
+\label{fig:fsm:alloc_s}
 \end{figure}
@@
 \begin{figure}[h!]
-\vspace{-10pt}
+% \vspace{-10pt}
 \centering
@@
 \end{tikzpicture}
-\vspace{-15pt}
-\caption{{\scriptsize{$M_{\jtt{D}_s}$: Timed Automaton to execute task type \jtt{deallocate} in $M_P$}}
-\vspace{-20pt}
-\label{fig:fsm:dealloc_s}}
+% \vspace{-15pt}
+\caption{$M_{\jtt{D}_s}$: Timed Automaton to execute task type \jtt{deallocate} in $M_P$}
+% \vspace{-20pt}
+\label{fig:fsm:dealloc_s}
 \end{figure}
@@
 \end{tikzpicture}
-\vspace{-13pt}
-\caption{\scriptsize{An example behavior for $M_E$}}
+% \vspace{-13pt}
+\caption{An example behavior for $M_E$}
 \label{fsm:env}
-\vspace{-20pt}
+% \vspace{-20pt}
 \end{figure}

### 2016-01-12 p03-ch02-2014-jmseq/jmseq-elsa.tex
Remove all `vspace`
Fix long captions in figures and tables

@@
 \begin{figure}[t]
-\vspace{-30pt}
+% \vspace{-30pt}
 \begin{center}
@@
 \end{center}
-\vspace{-20pt}
+% \vspace{-20pt}
 \caption{Fredhopper's Controller life cycle for data processing requests from customers}
 \label{fig:frh-dm}
-\vspace{-5pt}
+% \vspace{-5pt}
 \end{figure}
@@
 \includegraphics[scale=0.72]{images/frh-case-study-summary-rev2}
-\caption{Result of experiment measurements without and with using JMSeq on Fredhopper Client API directed with live environment data from customers.}
+\caption[Comparison of measurements using JMSeq at Fredhopper]{Result of experiment measurements without and with using JMSeq on Fredhopper Client API directed with live environment data from customers.}
 \label{fig:frh:cs}
@@
 \includegraphics[scale=0.55]{images/perf}
-\caption{Comparison of different techniques applied to the same program. 
+\caption[Comparison of different techniques with JMSeq]{Comparison of different techniques applied to the same program. 
 In this figure, ``original'' denotes the original program that is measured for execution time;

### 2016-01-12 build.sh

@@ -9,7 +9,7 @@ CMD_BIB="bibtex"
 function clean {
    if [[ -f ".gitignore" ]]; then
       for f in $(cat .gitignore); do
-         [[ -f $f ]] && rm $f
+         [[ -f $f && $f != ".todo.md" ]] && rm $f
       done 
    fi
 }

### 2016-01-16 p01-ch02-2013/paper.tex
Fix long captions in figures and tables
Change bib style to 'american' from the original template

@@
 \end{align}
-\caption{
+\caption[A kernel version of the real-time programming language]{
 A kernel version of the real-time programming language.
@@
 % \vspace{-20pt}
-\caption{Fredhopper's Controller life cycle for remote data processing}
+\caption[Life cycle for remote data processing]{Fredhopper's Controller life cycle for remote data processing}
 \label{fig:frh-dm}

### 2016-01-18 p00-ch00-intro/intro.tex
Add table of publications

@@
 
-The rapid increase of computational power added a new aspect added to create a triangle: object-oriented, actor model, and \emph{concurrency}.
+The rapid increase of computational power added a new aspect to create a triangle: object-oriented, actor model, and \emph{concurrency}.
 Concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}. 
@@
 In addition, in the presence of scheduling and thread interleaving, it is important that objects are able to protect their state from concurrent modifications for correctness properties.
-In this thesis, we take advantage of a modelling language that aims to mix the three aspects in addition to the capabilities of formal semantics and verification.
 
@@
 programming language.
-\TODO{Elaborate the core contributions of this thesis}
 Considering the mainstream programming languages
@@
 
-\TODO{Connect back to the problem statement paraph}
-
-In this thesis, we focus on the concurrency layer of ABS.
-We formalize a subset of Java language in regards with ABS\TODO{Check for the way around}.
-We provide the operational semantics of the concurrency layer for scheduling of actors with an extension for deadlines and priorities. 
-We deliver ABS concurrency as a Java API.
-We utilize Java~8 and Java Concurrency API to preserve the semantics of ABS concurrency layer.
-We sketch a general architecture on how ABS language can be modularly translated into different programming languages.
-We present how this ABS API library can be used to translate ABS models into Java.
-
-Table~\ref{tbl:thesis} summarizes the structure of the research and contributions in the chapters of this text:
+Based on the problem statement and approach taken in this research, 
+Table~\ref{tbl:thesis} summarizes the structure of this text:
 
@@
 \end{table}
+
+In addition, Table~\ref{tbl:papers} summarizes the conference and journal 
+publications as a result of this research:
+
+\begin{table}[h]
+\centering
+\begin{tabular}{p{7cm}p{5cm}p{1cm}}
+\textsfb{Topic} & \textsfb{Proceedings / Journal} & \textsfb{Year}   
+\\ \toprule
+Programming and Deployment of Active Objects with Application-Level Scheduling & ACM SAC 2012, Pages 1883--1888 & 2012 
+\\ \midrule
+The Future of a Missed Deadline & COORD 2013, Pages 181--195 & 2013 
+\\ \midrule
+Monitoring Method Call Sequences using Annotations & Journal of Science of Computer Programming, 2014, Volume 94, Part 3, Pages 362--378 & 2014 
+\\ \midrule
+Programming with actors in Java~8 & ISoLA 2014, Pages 37--53 & 2014 
+\\ \midrule
+Formal verification of service level agreements through distributed monitoring & ESOCC 2015, Pages 125--140 & 2015
+\\ \bottomrule
+\end{tabular}
+\caption{Actors At Work -- Conference and Journal Publications}
+\label{tbl:papers}
+\end{table}

### 2016-01-23 abstract.tex
Add the first draft of 'Samenvatting' in English

% !TEX root = thesis.tex
%
\pdfbookmark[0]{Abstract}{Abstract}
\chapter*{Abstract}
\label{sec:abstract}
\vspace*{-10mm}

This thesis contributes to the intersection of object orientation, actor 
model, and concurrency.
We choose Java as the main programming language and as one of the mainstream 
object-oriented languages. 
We formalize a subset of Java and its concurrency API to 
facilitate formal verification and reasoning about it.
We create an abstract mapping from a concurrent-object modelling language, 
ABS, to programming semantics of Java concurrency. 
We provide the formal semantics of the mapping and runtime properties of 
the concurrency layer including deadlines and scheduling policies.
We provide an implementation of ABS concurrency layer as a Java API library 
and framework utilizing the latest langauge additions 
in Java~8.
We design and implement a runtime monitoring framework, JMSeq, to verify the
sequenced execution of methods through code annotatinon in JVM. 
In addition, we design a large-scale monitoring system as a real-world 
application; the monitoring system is built with actors and ABS concurrency 
with formal semantics of ABS composed with schedulability 
analysis.

\vspace*{20mm}

{\usekomafont{chapter}Samenvatting}\label{sec:abstract-diff} \\

\TODO{Translate to Dutch. Kindly by Stijn.}

### 2016-01-23 p00-ch00-intro/intro.tex
Add the first draft of 'Samenvatting' in English

@@
 in Java 8~\cite{jsr335:lambda:translation}.
-We design a large-scale monitoring system as a real-world application; 
-the monitoring system is built with actors and ABS concurrency 
-with formal semantics of ABS composed with 
-schedulability analysis~\cite{fersman2007task}.
+We design and implement a runtime monitoring framework, JMSeq, to verify the
+sequenced execution of methods through code annotatinon in JVM. 
+In addition, we design a large-scale monitoring system as a real-world 
+application; the monitoring system is built with actors and ABS concurrency 
+with formal semantics of ABS composed with schedulability 
+analysis~\cite{fersman2007task}.
 We elaborate on the elements of this research in the following.

### 2016-01-23 refs.sh
Fix invalid Figure references

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+#
+# To use this script, make sure the 'build' script does not "clean".
+# The '.aux' file is needed.
+#
+
+bibtool -v -x thesis.aux -s -d -i refs/refs-p0*.bib -o refs/mrefs.bib

### 2016-01-23 p02-ch01-2014/paper.tex
Fix uncited references from imported papers

@@
 a powerful concept for modeling distributed and concurrent
-systems \cite{Agha90,Agha97}. 
+systems \cite{Agha90,agha97}. 
 Different extensions of actors are proposed in several domains and are claimed to be the most suitable model of computation for many  applications \cite{Hewitt07-commitment}. 
@@
 % For instance, an extension may choose to provide remote actor access through HTTP layer; however, another may opt for a network layer implementation.
-We exploit the main properties of actor model~\cite{Agha97,Agha90} to distribute actors based on our implementation.
+We exploit the main properties of actor model~\cite{agha97,Agha90} to distribute actors based on our implementation.
 From a distributed perspective, the following are the main requirements for  distributing actors:
@@
 Further it supports a formal 
-behavioral specification of interfaces \cite{HahnleHJLSSW11}, to be used as contracts.
+behavioral specification of interfaces \cite{hahnlehjlssw11}, to be used as contracts.
 

### 2016-01-23 build.sh
Unify all references and remove unused

@@ -32,6 +32,5 @@ function build {
 
 # Main
 clean
-build "thesis.tex" "refs/mrefs"
-clean
+build "thesis.tex" "refs/refs"
 exit 0

### 2016-01-23 refs.sh

@@ -5,4 +5,4 @@
 # The '.aux' file is needed.
 #
 
-bibtool -v -x thesis.aux -s -d -i refs/refs-p0*.bib -o refs/mrefs.bib
+bibtool -v -x thesis.aux -s -d -i refs/refs-*.bib -o refs/refs.bib

### 2016-02-01 p00-ch00-intro/intro.tex
Add Pony lang in the by-design category

@@
 %is the type of programming languages that deliver actor model programming support as a first-class citizen of the language.
-Languages in which the actor model is by-design part of the syntax and semantics of the language.
-\TODO{Cite polylang as a research recent lang.}
-\TODO{Why this category is the best}
+Languages in which the actor model is by-design part of the syntax and 
+semantics of the language.
+Pony~\cite{ponylang,ClebschD13} targets high-performance 
+computing using actor models and shared memory. 
+When actor model is part of the design of a language, formal verification
+and correctness of actor semantics at runtime yields naturally from the 
+language semantics at runtime.
 \item[Implicit By Design]

### 2016-02-01 p00-ch00-intro/intro.tex
Apply comments on literature review

@@
 In this section, we briefly provide an overview of the programming languages that have targetted similar problem statements.
-It is important to note the rapid evolution of programming languages trying to provide an actor-based model of asynchronous message passing.\TODO{Reference}
-In addition, we identify different types of research and work related to actor model and concurrency programming:
+Various programming languages, in the past decade, have emerged to provide an actor-based model of asynchronous message passing~\cite{philippsen2000survey}.
+We observe different classes of actor model and concurrent model of 
+programming.
+Table~\ref{tbl:actor:pl} presents a summary.
+
 \begin{description}
 \item[First-Class Citizen]
-%is the type of programming languages that deliver actor model programming support as a first-class citizen of the language.
 Languages in which the actor model is by-design part of the syntax and 
@@
 
-Table~\ref{tbl:actor:pl} presents a summary.
-
-\begin{table}[h]
+\begin{table}[t]
 \centering
@@
 
-% In the previous section, we presented an overview of the programming languages and their categories that aim to provide actor model~\cite{agha86book}.
-Since many programming languages faced different challenges to provide the necessary syntax and semantics for actor model and concurrency at the level of the language, many libraries and frameworks aim to fill this gap (c.f.
-Table~\ref{tbl:actor:libs}).
-We observe that the more the language itself is close to the actor model semantics, the less external libraries and frameworks target this gap. 
-In the following, we briefly provide a few frameworks
+Since programming languages faced challenges to provide the necessary syntax 
+and semantics for actor model and concurrency at the level of the language, 
+many libraries and frameworks aim to fill this gap 
+Table~\ref{tbl:actor:libs} presents a summary.
+We observe that the more the language itself is close to the actor model 
+semantics, the less external libraries and frameworks target this gap. 
+In the following, we briefly enumerate frameworks
 \footnote{A more comprehensive list can be obtained at \cite{KarmaniSA09} and  \url{https://en.wikipedia.org/wiki/Actor_model\#Programming_with_Actors}}
-and libraries for Java as one of the main focus areas of this research is to study this gap for a mainstream language such as Java.
+and libraries for Java.
 
-\begin{table}[h]
+\begin{table}[t]
 \centering
@@
 
-One of the main techniques used in libraries to deliver actor programming in Java is byte-code engineering~\cite{dahm1999byte,bruneton2002asm,asm}.
-Byte-code engineering modifies the generated byte-code for compiled classes in Java either during compilation or at runtime.
-Although, this technique is commonly used and argued to provide better performance optimization~\cite{vallee1999soot}, it introduces deep challenges regarding the verification of the running byte-code~\cite{leroy2001java,leroy2003java}.
+One of the main techniques used in libraries to deliver actor programming in 
+Java is byte-code engineering~\cite{dahm1999byte,bruneton2002asm,asm}.
+Byte-code engineering modifies the generated byte-code for compiled classes in 
+Java either during compilation or at runtime.
+Although, this technique is commonly used and argued to provide better 
+performance optimization~\cite{vallee1999soot}, it introduces other challenges 
+regarding the verification of the running 
+byte-code~\cite{leroy2001java,leroy2003java}.
 

### 2016-02-01 p00-ch00-intro/intro.tex
Improve the intro architecture + objectives section

@@
 
-\section{Architecture}
+\section{Objectives and Architecture}
 \label{sec:intro:arch:crit}
@@
 In this section, we present a high-level overview of design goals that we pursue in this research.
-%Equivalently, the effectiveness and success of this research is measured upon the achievement of the discussed goals and objectives.
 
@@
 environment is not predictable at runtime.
-Therefore, distribution challenges performance efficiency or scalability.
 The goal is to ensure the actor system scales in performance with least 
 influence from the number of asynchronous messages delivered in the system.
-\TODO{Discuss with Frank and drop this.}
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.3]{../figs/Arch.eps}
+\caption[General Architecture]{General Architecture of ABS API and Java Language Backend}
+\label{fig:arch}
+\end{figure}
 
 \paragraph{Modularity} 
-One of the main goals of this research is to create a modular and reusable
-software artifact. 
-This can be achieved only by producing an API for ABS semantics that is
-indepedent from any specific ABS backend.
-\TODO{Elaborate on why an API library is modular for ABS}
-The scope of the research spans to a number of layers revolving around ABS
- language:
+We approach ABS modelling and development with a component-based or modular 
+software engineering practices.
+The scope of the research spans to a number of layers around ABS language:
+
 \begin{itemize}
@@
 One first objective is to compile an ABS model to a target programming language.
-% The compilation can be at \emph{source-to-source} or \emph{source-to-bytecode} level.
-In addition, to be able to integrate generated code from ABS into other legacy systems, it is crucial that generated code is readable and of acceptable quality.\TODO{Use software engineering terms here}
 Target languages potentially include mainstream programming languages such as
@@
 their ABS models to the target languages.
-\TODO{Why not the previous ABS Java backend suffered from software engineering}
+We propose a new architecture for ABS tool-set and engineering that enables
+different programming language utilize the same architecture.
 \item \emph{Using ABS concurrency as an API in an existing programming language}
@@
 
-% Based on the scopes and layers of usage of this research, we aim to provide 
-% a highly modular architecture in design and implementation.
-% The goal is verifiable based on produced artifacts of this research in terms
-% of their \emph{usability}, \emph{re-usability}, and \emph{composability} in
-% different contexts.
+This approach yields to a modular architecture presented in 
+Figure~\ref{fig:arch}.
 
-\paragraph{Extensibility}
-It is a common practice in software engineering that components are developed with a mixture of \emph{configuration}, \emph{extension} and \emph{composition}.
-Hence, delivering the ABS language features as an API brings about a new requirement: to configure or extend the API. 
-In particular, the ABS programmer should be able to extend the API to customized requirements. \TODO{Expand?}
-%The extensibility is provided either through configuration of the API or API extension.
-
-\Behrooz{It's worth to include the figure but needs more work to justify its presence and how it relates to this thesis.}
-Figure~\ref{fig:arch} presents the general architecture of the implementation.
-
-\TODO{Relate the criteria in this section on how they lead to a natural modular architecture in the figure.}
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=0.3]{../figs/Arch.eps}
-\caption[General Architecture]{General Architecture of ABS API and Java Language Backend}
-\label{fig:arch}
-\end{figure}
+% \paragraph{Extensibility}
+% It is a common practice in software engineering that components are developed with a mixture of \emph{configuration}, \emph{extension} and \emph{composition}.
+% Hence, delivering the ABS language features as an API brings about a new requirement: to configure or extend the API. 
+% In particular, the ABS programmer should be able to extend the API to customized requirements. \TODO{Expand?}
 

### 2016-02-08 cv.tex
Add single-page CV to the thesis
While there, re-order some of the sections based on Leiden U ordering.

% -*- program: lualatex -*-

\documentclass[a4paper,11pt]{moderncv}
\moderncvtheme[grey]{classic}

% === Packages
\usepackage[scale=0.8]{geometry}
\usepackage[utf8]{inputenc}
% \usepackage{hyperref}              
% \definecolor{linkcolour}{rgb}{0,0.2,0.6}
% \hypersetup{colorlinks,breaklinks,urlcolor=linkcolour, linkcolor=linkcolour}

% === Fonts
\usepackage[T1]{fontenc}
\usepackage{charter}

% === Bullet Style
\def\labelitemi{{\tiny$\bullet$}}

% === CV Information
\title{ }
\firstname{Curriculum Vitae}
\familyname{ }
\address{ }{ }{ }
% \phone[mobile]{+31 611 770 610}
\email{behrooz.nobakht@gmail.com}
% \social[twitter]{behruz}
% \social[linkedin]{behrooznobakht}
\extrainfo{Geboren 29-07-1981, Teheran}
% \homepage{LinkedIn}

\begin{document}
\maketitle

% Education
\section{Opleiding}
%\cventry{year--year}{Degree}{Institution}{City}{\textit{Grade}}{Description}

\cventry{2011--2015}{PhD Informatica}{Universiteit Leiden}{Leiden}{}
{\emph{Actors At Work}\copyright}

\cventry{2009--2011}{MSc cum laude Informatica}{Universiteit Leiden}{Leiden}{}{\emph{Deployment of active objects onto multi-core}}

\cventry{1998--2002}{BSc Informatica}{Technologie Universiteit Sharif}{Teheran}{\textit{7.0}}{\emph{Online Shopping}}

% Work Experience
\section{Werkervaring}
% \cventry{year--year}{Job title}{Employer}{City}{}{General description no%
% longer than 1--2 lines.\newline{}%
% Detailed achievements:%
% }

\cventry{2011--Nu}{Senior Software Ontwikkelaar}{SDL Fredhopper}{Amsterdam}{}{Groep leider in cloud\newline}

\cventry{2003--2009}{Medeoprichter en architect}{ASTA Co.}{Teheran}{}{%
\begin{itemize}
\item Het beheer van het bedrijf
\item Ontwikkelen
\item Project onderhandeling
\end{itemize}
}

\end{document}

### 2016-02-21 p00-ch00-intro/intro.tex
Cite SIMULA 67 in the first intro paragraph

@@
 Method invocations are \emph{blocking}; i.e. the caller object waits until it receives the result of the method call from the callee object.
-Surprisingly, this model of interaction was \emph{never} the intention of the creator of the paradigm.
+This model of interaction was not the intention of the pioneers of the paradigm.
 Originally, object interactions were meant to be \emph{messages} among objects and objects behaved as autonomous entities possibly on remote locations on a network; Alan Kay clarified later~\cite{alank1,alank2}.
 On the contrary, almost all the object-oriented languages at hand have followed the blocking and synchronous model of messaging. 
-The original definition of object-oriented programming is close to another model of computation: actor model.
+The original definition of object-oriented programming in languages such as SIMULA 67~\cite{Dahl:1968:simula} inspired another model of computation: actor model.
 

### 2016-02-21 p00-ch00-intro/intro.tex
Apply discussion comments on intro

@@
 Language extensions, libraries, and even new programming languages are the natural outcomes of such research.
-The more mainstream the language, the more research and development are available in the intersection of object orientation and actor model.
 For example, \cite{actor_frameworks_jvm:agha} presents a comparative analysis of such research for Java and JVM languages.
 
-The rapid increase of computational power added a new aspect to create a triangle: object-oriented, actor model, and \emph{concurrency}.
+The rapid increase of computational power added a new aspect to create a triangle: object orientation, actor model, and \emph{concurrency}.
 Concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}. 
-Scheduling of object messages in threads of executions is one of main challenges.
-In addition, in the presence of scheduling and thread interleaving, it is important that objects are able to protect their state from concurrent modifications for correctness properties.
+The main and foremost challenge is how to combine concurrency with object-oriented programming.
+Concurrency brings about two aspects in combination of object orientation: \emph{communication} and \emph{execution}.
 
-The core contributions of this thesis target at the intersection of object 
-orientation, actor model, and concurrency.
-We choose Java as the main programming language and as one of the mainstream 
-object-oriented languages. 
-We formalize a subset of Java and its concurrency API~\cite{jsr166} to 
-facilitate formal verification and reasoning about it.
-We create an abstract mapping from a concurrent-object modelling language, 
-ABS~\cite{johnsen2012abs}, to programming semantics of Java concurrency. 
-We provide the formal semantics of the mapping and runtime properties of 
-the concurrency layer including deadlines and scheduling policies.
-We provide an implementation of ABS concurrency layer as a Java API library 
-and framework utilizing the latest langauge additions 
-in Java 8~\cite{jsr335:lambda:translation}.
-We design and implement a runtime monitoring framework, JMSeq, to verify the
-sequenced execution of methods through code annotatinon in JVM. 
-In addition, we design a large-scale monitoring system as a real-world 
-application; the monitoring system is built with actors and ABS concurrency 
-with formal semantics of ABS composed with schedulability 
-analysis~\cite{fersman2007task}.
-We elaborate on the elements of this research in the following.
+There are two fundamental approaches towards communication: synchronous and asynchronous. 
+In synchronous model, methods of objects are scheduled to be executed in different threads.
+The method invocation is synchronous, however, its execution is scheduled to occur in potentially a different thread from that of the caller object.
+This model utilizes future values and takes full advantage of object orientation paradigm.
+Asynchronous communication is a core element of actor model.
+In actor model, every communication is asynchronous. 
+A shared or dedicated queue of messages (inbox) is utilized among actors in the system.
+The unit of communication is a \emph{message}.
+The notion of a message is not bound by a specific definition or object interface.
+When an actor receives a message, it may use pattern matching techniques to determine the content of the message.
+When the actor completes the processing of a message, it may decide to reply to the message by sending another message.
+There are two essential approaches towards execution: run-to-completion (RTC) and co-operative scheduling.
+Actor model uses RTC execution style.
+When an actor receives a message, the execution cannot be preempted until the message is processed. 
+In co-operative scheduling, an object may decide to yield computation power at any time to allow another object make progress.
+In contrast with actor model, concurrent active objects~\cite{creol:broch:owe} combine object orientation and actor model. 
+Concurrent active objects use asynchronous communication; however, a message is bound by an object-oriented interface.
+In addition, co-operation scheduling is the main execution style in concurrent active objects.
 
@@
 can be used as a programming language at industrial and business scale.
-However, ABS has so far remained mainly as a modelling language.
-\TODO{Check with Frank.}
+ABS has mainly developed as a modelling language.
 
-One core challenge is to combine ABS semantics (e.g. concurrency layer) in a
+One core challenge is to combine ABS semantics of its concurrency layer with a
 programming language.
@@
 
-% We deliver ABS concurrency layer in form of a Java library API.
-% However, 
 The Java language specification (JLS)~\cite{gosling2000java} clearly concludes that Java language is not comptabile and ready to be extended with a functional language that supports algebraic data types.
 This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
-The Scala language~\cite{odersky2004scala} is a result of the research. It is a dynamic, functional, and object-oriented language on JVM.
+Scala language~\cite{odersky2004scala} is a result of such relevant research. It is a dynamic, functional, and object-oriented language on JVM.
 However, these extensions and libraries lack formal semantics in the context
@@
 
-We primarily focus on ABS concurrency layer and deliver its semantics as a Java API.
+In this thesis, we primarily focus on ABS concurrency layer and deliver its semantics as a Java API.
 We argue that the advantage of a correct semantic translation from ABS concurrency layer to its Java equivalent constructs overcomes the additional support of functional layer with \emph{partial} formal semantics.
@@
 In addition, we aim to facilitate ways to actually program ABS models taking advantage of the standard Java libraries.
+
+The core contributions of this thesis target at the intersection of object 
+orientation, actor model, and concurrency.
+We choose Java as the main programming language and as one of the mainstream 
+object-oriented languages. 
+We formalize a subset of Java and its concurrency API~\cite{jsr166} to 
+facilitate formal verification and reasoning about it.
+We create an abstract mapping from a concurrent-object modelling language, 
+ABS~\cite{johnsen2012abs}, to programming semantics of Java concurrency (cf. Chapter~\ref{ch:p01:ch02}). 
+We provide the formal semantics of the mapping and runtime properties of 
+the concurrency layer including deadlines and scheduling policies (cf. Chapter~\ref{ch:p01:ch01}).
+We provide an implementation of ABS concurrency layer as a Java API library 
+and framework utilizing the latest langauge additions 
+in Java 8~\cite{jsr335:lambda:translation} (cf. Chapter~\ref{ch:p02:ch01}).
+We design and implement a runtime monitoring framework, JMSeq, to verify the
+sequenced execution of methods through code annotatinon in JVM (cf. Chapter~\ref{ch:p03:ch02}). 
+In addition, we design a large-scale monitoring system as a real-world 
+application; the monitoring system is built with actors and ABS concurrency 
+with formal semantics of ABS composed with schedulability 
+analysis~\cite{fersman2007task} (cf. Chapter~\ref{ch:p03:ch01}).
 

### 2016-02-28 p00-ch00-intro/intro.tex
Apply discussion comments to intro chapter

@@
 \label{ch:intro}
-
-\section{From actor \emph{modelling} to actor \emph{programming}}
 \label{sec:intro:abs}
 
+% \section{From actor \emph{modelling} to actor \emph{programming}}
+
 Object-oriented programming~\cite{booch1982object,meyer1988object} has been of one the dominant paradigms for software engineering in the past three decades.
+SIMULA 67~\cite{Dahl:1968:simula} introduced notions of classes, subclasses, and virtual procedures.
+In the main block of a program, objects are created, then their procedures are called.
 One object interacts with another object using the notion of a method.
@@
 This model of interaction was not the intention of the pioneers of the paradigm.
-Originally, object interactions were meant to be \emph{messages} among objects and objects behaved as autonomous entities possibly on remote locations on a network; Alan Kay clarified later~\cite{alank1,alank2}.
+Object interactions were meant to be \emph{messages} among objects and objects behaved as autonomous entities possibly on remote locations on a network; Alan Kay clarified later~\cite{alank1,alank2}. 
 On the contrary, almost all the object-oriented languages at hand have followed the blocking and synchronous model of messaging. 
-The original definition of object-oriented programming in languages such as SIMULA 67~\cite{Dahl:1968:simula} inspired another model of computation: actor model.
+The above definition of object-oriented programming has inspired another model of computation: actor model.
 
@@
 
-There are two fundamental approaches towards communication: synchronous and asynchronous. 
-In synchronous model, methods of objects are scheduled to be executed in different threads.
-The method invocation is synchronous, however, its execution is scheduled to occur in potentially a different thread from that of the caller object.
-This model utilizes future values and takes full advantage of object orientation paradigm.
-Asynchronous communication is a core element of actor model.
+In a concurrent setting, coroutines~\cite{conway1963design,taocp:knuth} enable interactions with collaborative preemption.
+A coroutine allows multiple entry points for suspending and resuming execution at certain locations.
+A coroutine instance holds state and the state might vary among different instances.
+A coroutine may \emph{yield} to another coroutine;
+there's no \emph{return} statement in a coroutine but transfer of control from one to another without returning.
+This relation is symmetric and not similar to that of caller-callee type.
+This fundamental property makes compositional concurrency of coroutines straightforward.
+Coroutines are not originally established in object-oriented principles.
+Object orientation focuses on an object-view with caller-callee interaction  whereas coroutines focus on process-view with transfer of control.
+
+In a concurrent setting with objects, multi-threading has been a dominant approach to bring the two together.
+A thread holds a single stack of elements.
+Execution in a thread is sequential.
+The stack has a starting location (i.e. start of invocation) and a finish location (i.e. the return point).
+In a caller-callee setting, an object might call methods from other objects.
+With objects running in different threads, the execution of methods are interleaved in different threads. 
+Future values are used to hold the eventual return value of a method in the caller site.
+The unit of interleaving in coroutines is coarse whereas multi-threading is based on more fine-grained scheduling of method invocations.
+In addition, interactions in both are blocking and synchronous.
+Actor model relies on asynchronous communication as one of its core elements.
+
 In actor model, every communication is asynchronous. 
-A shared or dedicated queue of messages (inbox) is utilized among actors in the system.
+A queue of messages (inbox) is utilized among actors in the system.
 The unit of communication is a \emph{message}.
@@
 When the actor completes the processing of a message, it may decide to reply to the message by sending another message.
-There are two essential approaches towards execution: run-to-completion (RTC) and co-operative scheduling.
-Actor model uses RTC execution style.
-When an actor receives a message, the execution cannot be preempted until the message is processed. 
-In co-operative scheduling, an object may decide to yield computation power at any time to allow another object make progress.
-In contrast with actor model, concurrent active objects~\cite{creol:broch:owe} combine object orientation and actor model. 
-Concurrent active objects use asynchronous communication; however, a message is bound by an object-oriented interface.
-In addition, co-operation scheduling is the main execution style in concurrent active objects.
+On the other hand, actor model works based on run-to-completion style of execution.
+When a message is being processed, an actor cannot be preempted or intentially yield to allow other actors in the system make progress.
+
+Considering the triangle, it is interesting and challenging to follow the question that how to bring coroutine execution (co-operative scheduling) to an object-oriented setting with asynchronous communication with binding to object interfaces for the messages.
 
 ABS~\cite{johnsen2012abs,hahnlehjlssw11} is a modelling language for concurrent objects and distributed systems.
-ABS is based on the semantics of the actor model~\cite{actors:agha}.
-ABS language provides a set of core features including algebraic data types (ADT), a functional layer, and a concurrency layer.
-The ABS concurrency layer uses co-operative scheduling of active objects~\cite{creol:broch:owe}.
+ABS uses asynchronous communication among objects.
+A message in ABS in eventually bound to a method invocation;
+this defines the interface of the messages sent.
+In addition, ABS supports future values in its asynchronous communication.
+ABS introduces the notion of \bfjtt{release} that is based on co-operative scheduling of objects; i.e. similar to that of \emph{yield} in coroutines~\cite{creol:broch:owe}.
 ABS semantics is completely expressed in structural operational semantics~\cite{plotkin:sos}.
@@
 % 
-Co-operative scheduling in ABS has been additionally extended for real-time scheduling taking priorities and time constraintsinto account~\cite{bjork2013:rtabs,johnsen2012modeling}.
+Co-operative scheduling in ABS has been additionally extended for real-time scheduling with priorities and time constraints~\cite{bjork2013:rtabs,johnsen2012modeling}.
 All the above characteristics make the ABS language an attractive choice if it
@@
 
-One core challenge is to combine ABS semantics of its concurrency layer with a
-programming language.
 Considering the mainstream programming languages
 \footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}}, Java~\cite{gosling2000java} is one of the most commonly-used.
-Since Java 1.5 and the rise of the concurrency API in Java~\cite{jsr166}, substantial effort is spend by the community on how concurrency and distribution can be improved in Java, along the lines of the actor model, as provided by ABS.
+Since Java 1.5 and the rise of the concurrency API in Java JSR-166~\cite{jsr166}, substantial effort focus on how concurrency can be improved in Java.
 However, not all the development on Java libraries and frameworks are based on a rigorous formal semantics. 
 Therefore, this gives rise to issues and challenges in terms of correctness, semantic preservation, and reasoning.
+In addition, the Java language specification (JLS)~\cite{gosling2000java} clearly reveals that Java language is not comptabile and ready to be extended as a functional language that supports algebraic data types.
+This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
+Scala~\cite{odersky2004scala} is a result of such relevant research. It is a dynamic, functional, and object-oriented language on JVM.
 
-The Java language specification (JLS)~\cite{gosling2000java} clearly concludes that Java language is not comptabile and ready to be extended with a functional language that supports algebraic data types.
-This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
-Scala language~\cite{odersky2004scala} is a result of such relevant research. It is a dynamic, functional, and object-oriented language on JVM.
-However, these extensions and libraries lack formal semantics in the context
-of concurrency and actor model.
+One core challenge is how to create a mapping of coroutines to multi-threading in Java.
+Supporting coroutines in Java can be mostly classified in two major categories.
+One category relies a \emph{modified} JVM platform (e.g.
+Da Vinci JVM) in the implementation of thread stack and state such as \cite{Stadler:2009:LCJ:1596655.1596679}, \cite{Stadler:2010:ECJ:1852761.1852765} and \cite{Liu:2006:II:1111320.1111063}.
+Other category involves libraries such as Commons Javaflow\footnote{\url{http://commons.apache.org/sandbox/commons-javaflow/}} or Coroutines\footnote{\url{https://github.com/offbynull/coroutines}} that utilize byte-code modification~\cite{dahm1999byte} at runtime in JVM.
+In this research, we did not intend to use any of the two above approaches.
+We rely on the mainstream JVM platform released by Java team at Oracle; in addition, we do not use byte-code engineering.
 
-In this thesis, we primarily focus on ABS concurrency layer and deliver its semantics as a Java API.
-We argue that the advantage of a correct semantic translation from ABS concurrency layer to its Java equivalent constructs overcomes the additional support of functional layer with \emph{partial} formal semantics.
-Our approach enables programmers to model their Java systems in terms of ABS concurrency semantics and then create a correct mapping and implementation from ABS to Java using an API.
-In addition, we aim to facilitate ways to actually program ABS models taking advantage of the standard Java libraries.
+Since a thread owns a single stack, a \emph{naive} approach is to map every invocation (entry point) in a coroutine to a new thread in Java. 
+This mapping unsurprisingly leads to a poor performance;
+the number of threads is not scalable at runtime based on resource limitations in JVM when the number of object increase (cf. Chapter~\ref{ch:p01:ch01}).
+Therefore, it is reasonable to use a pool of threads (JSR-166~\cite{jsr166}) to direct the execution of all method invocations (messages). 
+However, to deploy a message onto a thread for execution, the message is required to be abstracted in a generic way.
+We utilize Java~8 features (JSR 335~\cite{jsr335} and JSR 292~\cite{jsr292:invokedyn}) to abstract a message as a data structure expressed in a lambda expression (cf. Chapter~\ref{ch:p02:ch01}).
 

### 2016-02-29 p00-ch00-intro/intro.tex
Polish intro

@@
 The unit of interleaving in coroutines is coarse whereas multi-threading is based on more fine-grained scheduling of method invocations.
-In addition, interactions in both are blocking and synchronous.
-Actor model relies on asynchronous communication as one of its core elements.
+In addition, interactions in both coroutines and multi-threading are blocking and synchronous.
+In contrast, actor model relies on asynchronous communication as one of its core elements.
 
 In actor model, every communication is asynchronous.
+The unit of communication is a \emph{message}.
 A queue of messages (inbox) is utilized among actors in the system.
-The unit of communication is a \emph{message}.
 The notion of a message is not bound by a specific definition or object interface.
@@
 On the other hand, actor model works based on run-to-completion style of execution.
-When a message is being processed, an actor cannot be preempted or intentially yield to allow other actors in the system make progress.
+When a message is being processed, an actor cannot be preempted or intentially yield to allow other actors in the system to make progress.
 
-Considering the triangle, it is interesting and challenging to follow the question that how to bring coroutine execution (co-operative scheduling) to an object-oriented setting with asynchronous communication with binding to object interfaces for the messages.
+Considering coroutines, multi-threading, and actor model, it is interesting and challenging to follow the question of how to bring coroutine execution (co-operative scheduling) to an object-oriented setting with asynchronous communication with binding to object interfaces for the messages.
 
@@
 ABS uses asynchronous communication among objects.
-A message in ABS in eventually bound to a method invocation;
-this defines the interface of the messages sent.
+A message in ABS is bound to a method invocation;
+this defines the interface of the sent messages.
 In addition, ABS supports future values in its asynchronous communication.
-ABS introduces the notion of \bfjtt{release} that is based on co-operative scheduling of objects; i.e. similar to that of \emph{yield} in coroutines~\cite{creol:broch:owe}.
+ABS introduces \bfjtt{release} semantics that is based on co-operative scheduling of objects; 
+i.e. similar to that of \emph{yield} in coroutines~\cite{creol:broch:owe}.
 ABS semantics is completely expressed in structural operational semantics~\cite{plotkin:sos}.

### 2016-03-02 p00-ch00-intro/intro.tex
Apply comments to intro based on discussion

@@
 A coroutine allows multiple entry points for suspending and resuming execution at certain locations.
-A coroutine instance holds state and the state might vary among different instances.
+An instance of a coroutine spawns a new process with its own state.
 A coroutine may \emph{yield} to another coroutine;
 there's no \emph{return} statement in a coroutine but transfer of control from one to another without returning.
-This relation is symmetric and not similar to that of caller-callee type.
+The yield relation is symmetric and not similar to that of caller-callee type.
+A coroutine has explicit ability to transfer control to another coroutine.
 This fundamental property makes compositional concurrency of coroutines straightforward.
 Coroutines are not originally established in object-oriented principles.
+Object orientation is based on caller-callee interaction which is asymmetric.
+Caller invokes a method in the callee and waits until the \emph{return} occurs in the method of the callee.
 Object orientation focuses on an object-view with caller-callee interaction  whereas coroutines focus on process-view with transfer of control.
 
-In a concurrent setting with objects, multi-threading has been a dominant approach to bring the two together.
-A thread holds a single stack of elements.
+In a concurrent setting with objects, multi-threading has been a dominant approach to provide an object-oriented concurrency model.
+A thread holds a single stack of synchronous method calls.
 Execution in a thread is sequential.
@@
 In a caller-callee setting, an object might call methods from other objects.
-With objects running in different threads, the execution of methods are interleaved in different threads. 
+With objects running in different threads, the execution of methods are interleaved in uncontrollable threads. 
 Future values are used to hold the eventual return value of a method in the caller site.
-The unit of interleaving in coroutines is coarse whereas multi-threading is based on more fine-grained scheduling of method invocations.
+The unit of interleaving in coroutines is coarse whereas multi-threading is based on implicit scheduling of method invocations.
 In addition, interactions in both coroutines and multi-threading are blocking and synchronous.
@@
 
-Considering coroutines, multi-threading, and actor model, it is interesting and challenging to follow the question of how to bring coroutine execution (co-operative scheduling) to an object-oriented setting with asynchronous communication with binding to object interfaces for the messages.
+A question naturally rises: how to bring coroutine execution (co-operative scheduling) to an object-oriented setting with asynchronous communication with binding to object interfaces for the messages?
+There are a few modelling languages that focus on the question.
+
+Rebeca~\cite{sirjani2002simulation,sirjani2004formal,sirjani2007rebeca} is a modelling language for reactive and concurrent systems. 
+In Rebeca, a number of reactive objects (\emph{rebecs}) interact at runtime.
+Rebecs interactions is based on asynchronous message passing that leads to a method invocation.
+In Rebeca, each rebec has a unique thread of control and an unbounded queue of messages.
+Rebeca uses run-to-completion execution and does not support future values.
 
@@
 In this research, we did not intend to use any of the two above approaches.
+Custom or modified JVM platform implementations are not mainstream and not officially supported by Java team.
+Research and development on a modified JVM requires explicit and periodic maintenance and upgrade effort.
+Moreover, since byte-code modification changes the byte-code of a running program or inserts new byte-code into JVM at runtime, it makes reasoning and correctness analysis/verification more complicated~\cite{leroy2001java,leroy2003java}.
 We rely on the mainstream JVM platform released by Java team at Oracle; in addition, we do not use byte-code engineering.
 
-Since a thread owns a single stack, a \emph{naive} approach is to map every invocation (entry point) in a coroutine to a new thread in Java. 
+Since a thread owns a single stack, a straightforward approach is to map every invocation (entry point) in a coroutine to a new thread in Java~\cite{SchaferP10,schafer2010programming}. 
 This mapping unsurprisingly leads to a poor performance;
@@
 This approach is also referred to as \emph{Foreign Function Interface} in the context of ABS modelling.
-Listing~\ref{lst:abs:java} shows ABS code that uses \jtt{java.util.ArrayList} as a data strucuture.
-With Polyglot programming in ABS, the programmer does \emph{not} need to define the abstractions necessary for the list construct.
 
 \begin{lstlisting}[float=h,language=Java,caption=Using Java in ABS,label=lst:abs:java]
-List<String> params = new ArrayList<>();    // Java
+java.util.List<String> params = new java.util.ArrayList<>();    // Java
 myObj ! doSomething(params);                                    // ABS
 \end{lstlisting}
+
+Listing~\ref{lst:abs:java} shows a snippet of ABS code that uses \jtt{java.util.ArrayList} as a data strucuture.
+Ideally in ABS, the programmer is able to directly use the libraries and API from Java.
+This removes the necessity to redundantly repeat definition of common data structures and API at ABS.
 
@@
 influence from the number of asynchronous messages delivered in the system.
-
-\begin{figure}[t]
-\centering
-\includegraphics[scale=0.3]{../figs/Arch.eps}
-\caption[General Architecture]{General Architecture of ABS API and Java Language Backend}
-\label{fig:arch}
-\end{figure}
 
@@
 
-% \paragraph{Extensibility}
-% It is a common practice in software engineering that components are developed with a mixture of \emph{configuration}, \emph{extension} and \emph{composition}.
-% Hence, delivering the ABS language features as an API brings about a new requirement: to configure or extend the API. 
-% In particular, the ABS programmer should be able to extend the API to customized requirements. \TODO{Expand?}
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.3]{../figs/Arch.eps}
+\caption[General Architecture]{General Architecture of ABS API and Java Language Backend}
+\label{fig:arch}
+\end{figure}
 

### 2016-03-08 abstract.tex
Apply comments from Lucas S
Kindly proof-read by the first external reader for intro chapter.

@@
 In addition, we design a large-scale monitoring system as a real-world 
-application; the monitoring system is built with actors and ABS concurrency 
-with formal semantics of ABS composed with schedulability 
-analysis.
+application; the monitoring system is built with ABS concurrent objects 
+and formal semantics composed with schedulability analysis.
 

### 2016-03-08 p00-ch00-intro/intro.tex
Apply comments from Lucas S
Kindly proof-read by the first external reader for intro chapter.

@@
 
-Object-oriented programming~\cite{booch1982object,meyer1988object} has been of one the dominant paradigms for software engineering in the past three decades.
+Object-oriented programming~\cite{booch1982object,meyer1988object} has been of one the dominant paradigms for software engineering.
+Object orientation provides principles for abstraction and ecapsulation.
 SIMULA 67~\cite{Dahl:1968:simula} introduced notions of classes, subclasses, and virtual procedures.
@@
 On the contrary, almost all the object-oriented languages at hand have followed the blocking and synchronous model of messaging. 
-The above definition of object-oriented programming has inspired another model of computation: actor model.
+Object-oriented programming has inspired another paradigm: actor model.
 
 One of the fundamental elements of actor model~\cite{actors:agha,agha97} is \emph{asynchronous} message passing.
-In this approach, interactions between objects are modelled as \emph{non-blocking} messages.
+In this approach, interactions between objects are modelled as non-blocking messages.
 One object, the sender, communicates a message to the other object, the receiver.
-At the receiver side, a message eventually leads to a method invocation (in the object-oriented paradigm) or a function call (in the functional paradigm).
+In contrast with abstraction in object orientation, a message can be any object and is not bound by any interface.
+At the receiver side, a message is matched with possible patterns and when a match is found, the message is processed by the receiver. 
 Actor model features location transparency; i.e. the physical location of objects is not visible to other objects.
@@
 
-A considerable amount of research has been carried out to mix object-oriented programming languages with the actor model~\cite{philippsen2000survey}. 
-Language extensions, libraries, and even new programming languages are the natural outcomes of such research.
+A considerable amount of research has been carried out to combine object-oriented programming with the actor model~\cite{philippsen2000survey}. 
+Language extensions, libraries, and even new programming languages are the  outcomes of such research.
 For example, \cite{actor_frameworks_jvm:agha} presents a comparative analysis of such research for Java and JVM languages.
 
-The rapid increase of computational power added a new aspect to create a triangle: object orientation, actor model, and \emph{concurrency}.
-Concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}. 
-The main and foremost challenge is how to combine concurrency with object-oriented programming.
+The rapid growth of hardware added a new aspect to create a triangle: object orientation, actor model, and \emph{concurrency}.
+Concurrency motivates utilizing computational power to make programs run faster.
+One goal has been to combine concurrency with object-oriented programming.
+However, concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}. 
 Concurrency brings about two aspects in combination of object orientation: \emph{communication} and \emph{execution}.
@@
 there's no \emph{return} statement in a coroutine but transfer of control from one to another without returning.
+A coroutine has explicit ability to transfer control to another coroutine.
 The yield relation is symmetric and not similar to that of caller-callee type.
-A coroutine has explicit ability to transfer control to another coroutine.
 This fundamental property makes compositional concurrency of coroutines straightforward.
@@
 Object orientation is based on caller-callee interaction which is asymmetric.
-Caller invokes a method in the callee and waits until the \emph{return} occurs in the method of the callee.
-Object orientation focuses on an object-view with caller-callee interaction  whereas coroutines focus on process-view with transfer of control.
+Caller invokes a method in the callee and \emph{blocks} until the \emph{return} occurs in the method of the callee.
+Object orientation focuses on an object-view with caller-callee interaction  whereas coroutines focus on a process-view with transfer of control.
 
-In a concurrent setting with objects, multi-threading has been a dominant approach to provide an object-oriented concurrency model.
+In a concurrent setting with objects, multi-threading has been a common approach to provide an object-oriented concurrency model.
 A thread holds a single stack of synchronous method calls.
@@
 The stack has a starting location (i.e. start of invocation) and a finish location (i.e. the return point).
-In a caller-callee setting, an object might call methods from other objects.
+In a caller-callee setting, an object calls methods from other objects.
 With objects running in different threads, the execution of methods are interleaved in uncontrollable threads. 
@@
 On the other hand, actor model works based on run-to-completion style of execution.
-When a message is being processed, an actor cannot be preempted or intentially yield to allow other actors in the system to make progress.
+When a message is being processed, an actor cannot be preempted or intentionally yield to allow other actors in the system to make progress.
 
-A question naturally rises: how to bring coroutine execution (co-operative scheduling) to an object-oriented setting with asynchronous communication with binding to object interfaces for the messages?
-There are a few modelling languages that focus on the question.
+A question naturally rises: how to bring coroutine execution (co-operative scheduling) to an object-oriented setting with asynchronous communication through messages bound to object interfaces?
+This question is the main problem statement of this thesis; cf. Table~\ref{tbl:problem}.
 
+\begin{table}[h]
+\centering
+\begin{tabular}{llll}
+\emph{Paradigm} & Abstraction & Communication & Execution 
+\\ \toprule
+OOP & Class/Object & Sync. Method Call & Blocking + Multi-threading
+\\ \midrule
+Actor Model & Actor & Async. Messages & Blocking + RTC
+\\ \midrule
+Coroutines & n.a. & \emph{yield} & Co-operative scheduling
+\\ \bottomrule
+\end{tabular}
+\caption[Summarized aspects of problem statement]{
+Considering the different aspects, the problem statement of thesis focuses to combine abstraction from OOP with asynchronous messages from Actor model but in a bounded context.
+In addition, this thesis focuses to bring in co-operative scheduling into the context of object-oriented programming.
+}
+\label{tbl:problem}
+\end{table}
+
+There are modelling languages that focus on the problem statement.
 Rebeca~\cite{sirjani2002simulation,sirjani2004formal,sirjani2007rebeca} is a modelling language for reactive and concurrent systems. 
@@
 Therefore, this gives rise to issues and challenges in terms of correctness, semantic preservation, and reasoning.
-In addition, the Java language specification (JLS)~\cite{gosling2000java} clearly reveals that Java language is not comptabile and ready to be extended as a functional language that supports algebraic data types.
+In addition, the Java language specification (JLS)~\cite{gosling2000java} clearly reveals that Java language is not compatible and ready to be extended as a functional language that supports algebraic data types.
 This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
@@
 Supporting coroutines in Java can be mostly classified in two major categories.
-One category relies a \emph{modified} JVM platform (e.g.
+One category relies on a \emph{modified} JVM platform (e.g.
 Da Vinci JVM) in the implementation of thread stack and state such as \cite{Stadler:2009:LCJ:1596655.1596679}, \cite{Stadler:2010:ECJ:1852761.1852765} and \cite{Liu:2006:II:1111320.1111063}.
-Other category involves libraries such as Commons Javaflow\footnote{\url{http://commons.apache.org/sandbox/commons-javaflow/}} or Coroutines\footnote{\url{https://github.com/offbynull/coroutines}} that utilize byte-code modification~\cite{dahm1999byte} at runtime in JVM.
+The other category involves libraries such as Commons Javaflow\footnote{\url{http://commons.apache.org/sandbox/commons-javaflow/}} or Coroutines\footnote{\url{https://github.com/offbynull/coroutines}} that utilize byte-code modification~\cite{dahm1999byte} at runtime in JVM.
 In this research, we did not intend to use any of the two above approaches.
@@
 This mapping unsurprisingly leads to a poor performance;
-the number of threads is not scalable at runtime based on resource limitations in JVM when the number of object increase (cf. Chapter~\ref{ch:p01:ch01}).
+the number of threads is not scalable at runtime based on resource limitations in JVM when the number of objects increase (cf. Chapter~\ref{ch:p01:ch01}).
 Therefore, it is reasonable to use a pool of threads (JSR-166~\cite{jsr166}) to direct the execution of all method invocations (messages). 
@@
 In addition, we design a large-scale monitoring system as a real-world 
-application; the monitoring system is built with actors and ABS concurrency 
-with formal semantics of ABS composed with schedulability 
+application; the monitoring system is built with ABS concurrent objects 
+and formal semantics composed with schedulability 
 analysis~\cite{fersman2007task} (cf. Chapter~\ref{ch:p03:ch01}).
@@
 In this research, we aim to deliver ABS semantics and features in a polyglot approach. 
-The programmer develops models with ABS that \emph{partially} take advantage of the target language features (e.g. from Java).
+The programmer develops models with ABS that \emph{partially} take advantage of the target language features (e.g. Java).
 This approach is also referred to as \emph{Foreign Function Interface} in the context of ABS modelling.
@@
 \item \emph{Modelling in ABS}
-ABS language provides a rigorous semantics to model concurrent
+ABS language provides rigorous semantics to model concurrent
 and distributed systems.
@@
 & External Library 
+\\ \midrule
+Pony\cite{ponylang,ClebschD13} & actor & First-Class Citizen
 \\ \bottomrule
@@
 
+\clearpage
+
 \section{Organization and Contributions}
@@
 \\ \midrule
-Design and implementation of ABS concurrency layer in Java & Implementation (Part~\ref{p:impl}) & Chapter~\ref{ch:p02:ch01} and \ref{ch:p02:ch02}
+Design and implementation of ABS concurrency layer in Java & Implementation (Part~\ref{p:impl}) & Chapter~\ref{ch:p02:ch01} 
+% and \ref{ch:p02:ch02}
 \\ \midrule 

### 2016-04-11 p00-ch00-intro/intro.tex
Polished the intro

@@
 Object-oriented programming~\cite{booch1982object,meyer1988object} has been of one the dominant paradigms for software engineering.
-Object orientation provides principles for abstraction and ecapsulation.
+Object orientation provides principles for abstraction and encapsulation.
 SIMULA 67~\cite{Dahl:1968:simula} introduced notions of classes, subclasses, and virtual procedures.
@@
 Method invocations are \emph{blocking}; i.e. the caller object waits until it receives the result of the method call from the callee object.
-This model of interaction was not the intention of the pioneers of the paradigm.
-Object interactions were meant to be \emph{messages} among objects and objects behaved as autonomous entities possibly on remote locations on a network; Alan Kay clarified later~\cite{alank1,alank2}. 
+This model of interaction was not the intention of the pioneers of the paradigm:
+object interactions were meant to be \emph{messages} among objects and objects behaved as autonomous entities possibly on remote locations on a network; Alan Kay clarified later~\cite{alank1,alank2}. 
 On the contrary, almost all the object-oriented languages at hand have followed the blocking and synchronous model of messaging. 
-Object-oriented programming has inspired another paradigm: actor model.
+Object-oriented programming has inspired another paradigm: the actor model.
 
-One of the fundamental elements of actor model~\cite{actors:agha,agha97} is \emph{asynchronous} message passing.
-In this approach, interactions between objects are modelled as non-blocking messages.
+One of the fundamental elements of the actor model~\cite{actors:agha,agha97} is \emph{asynchronous} message passing.
+In that approach, interactions between objects are modelled as non-blocking messages.
 One object, the sender, communicates a message to the other object, the receiver.
@@
 However, concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}. 
-Concurrency brings about two aspects in combination of object orientation: \emph{communication} and \emph{execution}.
+Concurrency brings about two aspects in relation to object orientation:
+\emph{communication} and \emph{execution}.
 
-In a concurrent setting, coroutines~\cite{conway1963design,taocp:knuth} enable interactions with collaborative preemption.
+In a concurrent setting, coroutines~\cite{conway1963design,taocp:knuth} enable interactions with collaborative pre-emption.
 A coroutine allows multiple entry points for suspending and resuming execution at certain locations.
 An instance of a coroutine spawns a new process with its own state.
-A coroutine may \emph{yield} to another coroutine;
-there's no \emph{return} statement in a coroutine but transfer of control from one to another without returning.
-A coroutine has explicit ability to transfer control to another coroutine.
-The yield relation is symmetric and not similar to that of caller-callee type.
+A coroutine has no \emph{return} statement, but it may explicitly \emph{yield}
+to another coroutine (transfer control of the execution).
+%A coroutine has explicit ability to transfer control to another coroutine.
+The yield relation is symmetric and not similar to that between caller and callee.
 This fundamental property makes compositional concurrency of coroutines straightforward.
-Coroutines are not originally established in object-oriented principles.
-Object orientation is based on caller-callee interaction which is asymmetric.
-Caller invokes a method in the callee and \emph{blocks} until the \emph{return} occurs in the method of the callee.
+Coroutines are not originally established in the object-oriented paradigm.
+Object orientation is based on caller-callee interaction, which is asymmetric.
+The caller invokes a method in the callee and \emph{blocks} until the
+corresponding \emph{return} occurs in the method of the callee.
 Object orientation focuses on an object-view with caller-callee interaction  whereas coroutines focus on a process-view with transfer of control.
@@
 In a caller-callee setting, an object calls methods from other objects.
-With objects running in different threads, the execution of methods are interleaved in uncontrollable threads. 
-Future values are used to hold the eventual return value of a method in the caller site.
+If multiple threads execute simultaneously in an object, their instructions are
+interleaved in an uncontrolled manner.
+Future values are used to hold the eventual return value of a method at the call site.
 The unit of interleaving in coroutines is coarse whereas multi-threading is based on implicit scheduling of method invocations.
-In addition, interactions in both coroutines and multi-threading are blocking and synchronous.
-In contrast, actor model relies on asynchronous communication as one of its core elements.
+Furthermore, interactions in both coroutines and multi-threading are blocking and synchronous.
+In contrast, the actor model relies on asynchronous communication (which is non-blocking)
+as one of its core elements.
 
-In actor model, every communication is asynchronous.
+In the actor model, all communication is asynchronous.
 The unit of communication is a \emph{message}.
-A queue of messages (inbox) is utilized among actors in the system.
-The notion of a message is not bound by a specific definition or object interface.
-When an actor receives a message, it may use pattern matching techniques to determine the content of the message.
+A queue of messages (the inbox) is utilized among actors in the system.
+The notion of a message is not bound to a specific definition or object interface.
+When an actor receives a message, it may use pattern matching techniques to extract the content of the message.
 When the actor completes the processing of a message, it may decide to reply to the message by sending another message.
-On the other hand, actor model works based on run-to-completion style of execution.
-When a message is being processed, an actor cannot be preempted or intentionally yield to allow other actors in the system to make progress.
+On the other hand, the actor model works based on run-to-completion style of execution.
+While a message is processed, an actor cannot be pre-empted or intentionally yield to allow other actors in the system to make progress.
 
-A question naturally rises: how to bring coroutine execution (co-operative scheduling) to an object-oriented setting with asynchronous communication through messages bound to object interfaces?
+A question naturally rises: how to bring coroutine-style execution (co-operative scheduling) to an object-oriented setting with asynchronous communication through messages bound to object interfaces?
 This question is the main problem statement of this thesis; cf. Table~\ref{tbl:problem}.
@@
 
-There are modelling languages that focus on the problem statement.
+There already exist modelling languages that focus on the problem statement.
 Rebeca~\cite{sirjani2002simulation,sirjani2004formal,sirjani2007rebeca} is a modelling language for reactive and concurrent systems. 
@@
 Rebecs interactions is based on asynchronous message passing that leads to a method invocation.
-In Rebeca, each rebec has a unique thread of control and an unbounded queue of messages.
+In Rebeca, each rebec has its own unique thread of control and an unbounded queue of messages.
 Rebeca uses run-to-completion execution and does not support future values.
@@
 i.e. similar to that of \emph{yield} in coroutines~\cite{creol:broch:owe}.
-ABS semantics is completely expressed in structural operational semantics~\cite{plotkin:sos}.
-This allows ABS models to take advantage of various verification methods and static analysis techniques.
+The ABS semantics is completely formalized by a structural operational semantics~\cite{plotkin:sos}.
+This allows ABS models to take advantage of a wide range of static and dynamic analysis techniques.
 % 
@@
 \footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}}, Java~\cite{gosling2000java} is one of the most commonly-used.
-Since Java 1.5 and the rise of the concurrency API in Java JSR-166~\cite{jsr166}, substantial effort focus on how concurrency can be improved in Java.
-However, not all the development on Java libraries and frameworks are based on a rigorous formal semantics. 
+Since Java 1.5 and the rise of the concurrency API in Java JSR-166~\cite{jsr166}, substantial effort focuses on how concurrency can be improved in Java.
+However, not all development of Java libraries and frameworks proceeds based on a rigorous formal semantics. 
 Therefore, this gives rise to issues and challenges in terms of correctness, semantic preservation, and reasoning.
-In addition, the Java language specification (JLS)~\cite{gosling2000java} clearly reveals that Java language is not compatible and ready to be extended as a functional language that supports algebraic data types.
-This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
-Scala~\cite{odersky2004scala} is a result of such relevant research. It is a dynamic, functional, and object-oriented language on JVM.
+In addition, the Java language specification (JLS)~\cite{gosling2000java} clearly reveals that the Java language is not compatible and ready to be extended as a functional language that supports algebraic data types.
+This has driven research and development that focuses on extending Java with support for functional programming~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
+Scala~\cite{odersky2004scala} is a result of such research. It is a dynamic, functional, and object-oriented language on JVM.
 
@@
 In this research, we did not intend to use any of the two above approaches.
-Custom or modified JVM platform implementations are not mainstream and not officially supported by Java team.
-Research and development on a modified JVM requires explicit and periodic maintenance and upgrade effort.
-Moreover, since byte-code modification changes the byte-code of a running program or inserts new byte-code into JVM at runtime, it makes reasoning and correctness analysis/verification more complicated~\cite{leroy2001java,leroy2003java}.
+Custom or modified JVM platform implementations are not mainstream and not officially supported by the Java team.
+Research and development of a modified JVM requires explicit and periodic maintenance and upgrade effort.
+Moreover, as byte-code modification changes the byte-code of a running program or inserts new byte-code into JVM at runtime, this complicates reasoning and correctness analysis/verification~\cite{leroy2001java,leroy2003java}.
 We rely on the mainstream JVM platform released by Java team at Oracle; in addition, we do not use byte-code engineering.
 
-Since a thread owns a single stack, a straightforward approach is to map every invocation (entry point) in a coroutine to a new thread in Java~\cite{SchaferP10,schafer2010programming}. 
-This mapping unsurprisingly leads to a poor performance;
+Another, straightforward way to support coroutines in Java multi-threading is that since a thread owns a single stack, we can translate every invocation (entry point) in a coroutine to a new thread in Java~\cite{SchaferP10,schafer2010programming}.
+This naive approach unsurprisingly leads to a poor performance;
 the number of threads is not scalable at runtime based on resource limitations in JVM when the number of objects increase (cf. Chapter~\ref{ch:p01:ch01}).
@@
 
-The core contributions of this thesis target at the intersection of object 
+The core contributions of this thesis target the intersection of object 
 orientation, actor model, and concurrency.
@@
 We create an abstract mapping from a concurrent-object modelling language, 
-ABS~\cite{johnsen2012abs}, to programming semantics of Java concurrency (cf. Chapter~\ref{ch:p01:ch02}). 
+ABS~\cite{johnsen2012abs}, to the programming semantics of concurrent Java (cf. Chapter~\ref{ch:p01:ch02}). 
 We provide the formal semantics of the mapping and runtime properties of 
@@
 We provide an implementation of ABS concurrency layer as a Java API library 
-and framework utilizing the latest langauge additions 
+and framework utilizing the latest language additions 
 in Java 8~\cite{jsr335:lambda:translation} (cf. Chapter~\ref{ch:p02:ch01}).
 We design and implement a runtime monitoring framework, JMSeq, to verify the
-sequenced execution of methods through code annotatinon in JVM (cf. Chapter~\ref{ch:p03:ch02}). 
+the correct ordering of execution of methods through code annotations in JVM (cf. Chapter~\ref{ch:p03:ch02}). 
 In addition, we design a large-scale monitoring system as a real-world 
 application; the monitoring system is built with ABS concurrent objects 
-and formal semantics composed with schedulability 
-analysis~\cite{fersman2007task} (cf. Chapter~\ref{ch:p03:ch01}).
+and formal semantics that leverages schedulability 
+analysis to verify correctness of the monitors~\cite{fersman2007task} (cf. Chapter~\ref{ch:p03:ch01}).
 
@@
 software engineering practices.
-The scope of the research spans to a number of layers around ABS language:
+The scope of the research spans a number of layers around ABS language:
 
@@
 Java, Erlang, Haskell, and Scala.
-Thus, the ABS programmer should be able to use an ABS compiler to compile
-their ABS models to the target languages.
-We propose a new architecture for ABS tool-set and engineering that enables
-different programming language utilize the same architecture.
+%Thus, the ABS programmer should be able to use an ABS compiler to compile
+%their ABS models to the target languages.
+We propose a new architecture for the ABS tool-set and engineering that enables
+different programming languages to utilize the same architecture.
 \item \emph{Using ABS concurrency as an API in an existing programming language}
-The ABS language is precisely expressed with structural operational semantics~\cite{johnsen2012abs} in addition to its syntax definition.
+The ABS language syntax and semantics are formalized precisely and rigorously by a structural operational semantics~\cite{johnsen2012abs}.
 The semantics of ABS can be delivered in a programming language
-as Application Programming Interface (API) as long as the programming language provides sufficient constructs to respect ABS language semantics.
-In addition, such ABS programming API should be \emph{verifiable}.
-If an ABS mapping to a programming API is provided, a programmer is able to
+as Application Programming Interface (API) as long as the programming language provides sufficient constructs to respect the ABS language semantics.
+In addition, such an ABS programming API should be \emph{verifiable}.
+If a mapping from ABS to a programming API is provided, a programmer is able to
 take advantage of ABS semantics without directly programming in ABS.
-Such capability from ABS enables industry users of mainstream languages to
-model their systems in ABS semantics using their programming languages and
-platforms.
+%Such capability from ABS
+This enables industry users of mainstream languages to
+model their systems in ABS semantics using the programming languages and
+platforms they are already familiar with.
 \item \emph{Modelling in ABS}
@@
 and distributed systems.
-For practical reasons, the user (that can be a programmer, an analysist, or a researcher) 
-desires to have access to a tool-set and IDE that allows working with ABS 
+For practical reasons, it is important that the user (that can be a programmer, an analyst, or a researcher) 
+has access to a tool-set and IDE that allows working with ABS 
 models in a user-friendly way.
@@
 
-This approach yields to a modular architecture presented in 
+The above objectives and design principles are realized by the modular architecture presented in 
 Figure~\ref{fig:arch}.
@@
 
-We discuss a brief overview of related work in the context of programming languages, actor model, and concurrency.
-We divide the overview in two levels; one is at the level of the programming languages and the other is for the external (third-party) libraries developed for programming languages.
+We briefly discuss related work in the context of programming languages, actor model, and concurrency.
+In the overview, we distinguish two levels; one is at the level of the programming languages and the other is for the external (third-party) libraries developed for programming languages.
 
@@
 
-In this section, we briefly provide an overview of the programming languages that have targetted similar problem statements.
+In this section, we briefly provide an overview of the programming languages that have targeted similar problem statements.
 Various programming languages, in the past decade, have emerged to provide an actor-based model of asynchronous message passing~\cite{philippsen2000survey}.
@@
 \item[Implicit By Design]
-refers to the type that there is no explicit notion of actor models in the language syntax or semantics.
-However, the programming language provides fundamental constructs for concurrency and asynchronous message passing.
+Refers to languages that have no explicit notion of actor models in their syntax or semantics, but do provide fundamental constructs for concurrency and asynchronous message passing.
 Thus, it becomes an easy task in this kind of programming language to create an abstraction to support the actor model by coding.
 \item[External Library]
-refers to the type of programming languages for which actor model support is provided by an external library developed for the language.
+Refers to the type of programming languages for which actor model support is provided by an external library.
 \end{description}
@@
 semantics, the less external libraries and frameworks target this gap. 
-In the following, we briefly enumerate frameworks
-\footnote{A more comprehensive list can be obtained at \cite{KarmaniSA09} and  \url{https://en.wikipedia.org/wiki/Actor_model\#Programming_with_Actors}}
-and libraries for Java.
+In the following, we briefly enumerate frameworks and libraries for Java
+\footnote{A more comprehensive list can be obtained at \cite{KarmaniSA09} and  \url{https://en.wikipedia.org/wiki/Actor_model\#Programming_with_Actors}}.
 
@@
 Although, this technique is commonly used and argued to provide better 
-performance optimization~\cite{vallee1999soot}, it introduces other challenges 
+performance optimization~\cite{vallee1999soot}, it introduces challenges 
 regarding the verification of the running 

### 2016-04-17 p00-ch00-intro/intro.tex
Apply comments from meeting with Stijn

@@
 Object orientation provides principles for abstraction and encapsulation.
+Abstraction is accomplished by high-level concepts of interfaces, classes, and objects (class instances).
+Encapsulation provides means to hide implementation details of objects such as their internal state.
 SIMULA 67~\cite{Dahl:1968:simula} introduced notions of classes, subclasses, and virtual procedures.
@@
 One object, the sender, communicates a message to the other object, the receiver.
-In contrast with abstraction in object orientation, a message can be any object and is not bound by any interface.
+In contrast with abstractions provided by object orientation, a message is not bound to any interface in the actor model.
 At the receiver side, a message is matched with possible patterns and when a match is found, the message is processed by the receiver. 
@@
 
-The rapid growth of hardware added a new aspect to create a triangle: object orientation, actor model, and \emph{concurrency}.
+Multicore and distributed computing raised a new challenge: combining object orientation, the actor model, and \emph{concurrency}.
 Concurrency motivates utilizing computational power to make programs run faster.
 One goal has been to combine concurrency with object-oriented programming.
-However, concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}. 
-Concurrency brings about two aspects in relation to object orientation:
-\emph{communication} and \emph{execution}.
+However, due to an exponential number of interleavings, concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}.
+% Concurrency brings about two aspects in relation to object orientation:
+% \emph{communication} and \emph{execution}.
+Concurrency has different forms in different paradigms.
 
-In a concurrent setting, coroutines~\cite{conway1963design,taocp:knuth} enable interactions with collaborative pre-emption.
+\TODO{Put a titled paraph here? e.g. ``Concurrency Approaches''}
+
+In a concurrent setting, coroutines~\cite{conway1963design,taocp:knuth} enable interactions with collaborative pre-emption:
+a coroutine has no \emph{return} statement, but it may explicitly \emph{yield}
+to another coroutine (transfer control of the execution).
+The yield relation is symmetric. 
+Creating an instance of a coroutine spawns a new process with its own state.
 A coroutine allows multiple entry points for suspending and resuming execution at certain locations.
-An instance of a coroutine spawns a new process with its own state.
-A coroutine has no \emph{return} statement, but it may explicitly \emph{yield}
-to another coroutine (transfer control of the execution).
-%A coroutine has explicit ability to transfer control to another coroutine.
-The yield relation is symmetric and not similar to that between caller and callee.
-This fundamental property makes compositional concurrency of coroutines straightforward.
+\TODO{and this property makes compositional concurrency of coroutines straightforward}.
 Coroutines are not originally established in the object-oriented paradigm.
-Object orientation is based on caller-callee interaction, which is asymmetric.
-The caller invokes a method in the callee and \emph{blocks} until the
+Object orientation is based on caller-callee interaction, which is asymmetric:
+the caller invokes a method in the callee and \emph{blocks} until the
 corresponding \emph{return} occurs in the method of the callee.
-Object orientation focuses on an object-view with caller-callee interaction  whereas coroutines focus on a process-view with transfer of control.
+Object orientation focuses on an object-view with caller-callee interaction and stack-based execution whereas coroutines focus on a process-view with flexible transfer of control.
 
@@
 The stack has a starting location (i.e. start of invocation) and a finish location (i.e. the return point).
+Future values can be used to hold the eventual return value of a method at the call site. \TODO{Do we need a citation for futures?}
 In a caller-callee setting, an object calls methods from other objects.
@@
 interleaved in an uncontrolled manner.
-Future values are used to hold the eventual return value of a method at the call site.
-The unit of interleaving in coroutines is coarse whereas multi-threading is based on implicit scheduling of method invocations.
+In coroutines, the points of interleaving are explicit (i.e. \emph{yield}) 
+whereas multi-threading is based on implicit scheduling of method invocations.
 Furthermore, interactions in both coroutines and multi-threading are blocking and synchronous.
@@
 When the actor completes the processing of a message, it may decide to reply to the message by sending another message.
-On the other hand, the actor model works based on run-to-completion style of execution.
+The actor model works based on run-to-completion style of execution.
 While a message is processed, an actor cannot be pre-empted or intentionally yield to allow other actors in the system to make progress.
 
-A question naturally rises: how to bring coroutine-style execution (co-operative scheduling) to an object-oriented setting with asynchronous communication through messages bound to object interfaces?
+\paragraph*{Problem Statement and Approach}
+A question naturally rises: how to bring coroutine-style execution (co-operative scheduling) to an object-oriented setting with asynchronous communication through messages bound to object interfaces in a programming language?
 This question is the main problem statement of this thesis; cf. Table~\ref{tbl:problem}.
+
+\TODO{Discuss the existence of this table. Where's concurrency? Remove ``Coroutine'' row. Add a column ``Concurrency''. Fix ``Execution'' column.}
 
@@
 \centering
-\begin{tabular}{llll}
-\emph{Paradigm} & Abstraction & Communication & Execution 
+\begin{tabular}{lll}
+ & OOP & Actor Model 
 \\ \toprule 
-OOP & Class/Object & Sync. Method Call & Blocking + Multi-threading
+\emph{Abstraction} & Class/Object & Actor 
 \\ \midrule
-Actor Model & Actor & Async. Messages & Blocking + RTC
+\emph{Communication} & Sync. Method Call & Async. Message 
 \\ \midrule
-Coroutines & n.a. & \emph{yield} & Co-operative scheduling
+\emph{Execution} & Blocking + Stack-based & Blocking 
+\\ \midrule
+\emph{Concurrency} & Mutli-threading & Multi-threading + RTC
 \\ \bottomrule
@@
 In Rebeca, a number of reactive objects (\emph{rebecs}) interact at runtime.
-Rebecs interactions is based on asynchronous message passing that leads to a method invocation.
-In Rebeca, each rebec has its own unique thread of control and an unbounded queue of messages.
+Each rebec has its own unique thread of control and an unbounded queue of messages.
+Rebecs interactions are based on asynchronous message passing.
+Each message is put in the unbounded queue of the
+receiver rebec and specifies a unique method to be invoked when the message
+is processed.
 Rebeca uses run-to-completion execution and does not support future values.
@@
 i.e. similar to that of \emph{yield} in coroutines~\cite{creol:broch:owe}.
-The ABS semantics is completely formalized by a structural operational semantics~\cite{plotkin:sos}.
-This allows ABS models to take advantage of a wide range of static and dynamic analysis techniques.
+The ABS semantics is completely formalized by a structural operational semantics~\cite{plotkin:sos,abs:2012}.
+This allows ABS models to take advantage of a wide range of static and dynamic analysis techniques~\cite{DBLP:journals/sttt/WongBBGGHMS15,DBLP:conf/cade/DinBH15,DBLP:conf/ifm/GiachinoGLLW13,DBLP:conf/tacas/AlbertAFGGMPR14}.
 % 
@@
 can be used as a programming language at industrial and business scale.
-ABS has mainly developed as a modelling language.
+ABS has mainly developed as a modelling language;
+for example, ABS does not support common I/O API for files or network,
+and it does not provide a standard library of data structures.
 
@@
 Therefore, this gives rise to issues and challenges in terms of correctness, semantic preservation, and reasoning.
-In addition, the Java language specification (JLS)~\cite{gosling2000java} clearly reveals that the Java language is not compatible and ready to be extended as a functional language that supports algebraic data types.
+In addition, the Java language specification (JLS)~\cite{gosling2000java,jls8} clearly reveals that the Java language is not yet ready to support algebraic data types.
 This has driven research and development that focuses on extending Java with support for functional programming~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
@@
 In this research, we did not intend to use any of the two above approaches.
-Custom or modified JVM platform implementations are not mainstream and not officially supported by the Java team.
-Research and development of a modified JVM requires explicit and periodic maintenance and upgrade effort.
+Custom or modified JVM platform implementations are not mainstream and not officially supported by the Java team which jeopardizes portability.
+To keep up with new versions of the Java language, research and development of a modified JVM requires explicit and periodic maintenance and upgrade effort.
 Moreover, as byte-code modification changes the byte-code of a running program or inserts new byte-code into JVM at runtime, this complicates reasoning and correctness analysis/verification~\cite{leroy2001java,leroy2003java}.
-We rely on the mainstream JVM platform released by Java team at Oracle; in addition, we do not use byte-code engineering.
+We aim to rely only on the mainstream JVM platform released by the Java team at Oracle; in addition, we do not intend to use byte-code engineering.
 
-Another, straightforward way to support coroutines in Java multi-threading is that since a thread owns a single stack, we can translate every invocation (entry point) in a coroutine to a new thread in Java~\cite{SchaferP10,schafer2010programming}.
-This naive approach unsurprisingly leads to a poor performance;
-the number of threads is not scalable at runtime based on resource limitations in JVM when the number of objects increase (cf. Chapter~\ref{ch:p01:ch01}).
+Another straightforward way to support coroutines in Java multi-threading is that since a thread owns a single stack, we can translate every invocation (entry point) in a coroutine to a new thread in Java~\cite{SchaferP10,schafer2010programming}.
+This naive approach unsurprisingly leads to a poor performance since threads are resource-intensive in Java and the number of threads is not scalable at runtime due to resource limitations in JVM when the number of objects increase (cf.
+Chapter~\ref{ch:p01:ch01}).
 Therefore, it is reasonable to use a pool of threads (JSR-166~\cite{jsr166}) to direct the execution of all method invocations (messages). 
-However, to deploy a message onto a thread for execution, the message is required to be abstracted in a generic way.
-We utilize Java~8 features (JSR 335~\cite{jsr335} and JSR 292~\cite{jsr292:invokedyn}) to abstract a message as a data structure expressed in a lambda expression (cf. Chapter~\ref{ch:p02:ch01}).
+% To deploy a message onto a thread for execution, the message can be abstracted in a generic way.
+We utilize Java~8 features (JSR 335~\cite{jsr335} and JSR 292~\cite{jsr292:invokedyn}) to model a message as a data structure expressed in a lambda expression (cf. Chapter~\ref{ch:p02:ch01}).
+
+\TODO{Where does the following paragraph have to move?}
 
@@
 We formalize a subset of Java and its concurrency API~\cite{jsr166} to 
-facilitate formal verification and reasoning about it.
+facilitate formal verification and reasoning about it (cf. Chapter~\ref{ch:p01:ch02}).
 We create an abstract mapping from a concurrent-object modelling language, 
@@
 
-In this section, we present a high-level overview of design goals that we pursue in this research.
+In this section, we present a high-level overview of design goals of our framework and its implementation.
 
@@
 In this approach, different languages with different focus and abstractions contribute to the same problem statement in different layers. 
-Polyglot programming essentially enables software practice to apply the \emph{right} language in the appropriate layer of the problem statement. 
-In this research, we aim to deliver ABS semantics and features in a polyglot approach. 
-The programmer develops models with ABS that \emph{partially} take advantage of the target language features (e.g. Java).
-This approach is also referred to as \emph{Foreign Function Interface} in the context of ABS modelling.
+Polyglot programming essentially enables software practice to apply the \emph{right} language in the appropriate layer of a system.
+Layers of a system require points of integration.
+ABS is an attractive choice for the concurrency and distributed layers.
+Therefore, ABS should be able to provide integration points to other layers.
+The programmer develops models with ABS that \emph{partially} take advantage of features of another language e.g. Java.
+This approach is also referred to as \emph{Foreign Function Interface}.
 
@@
 This removes the necessity to redundantly repeat definition of common data structures and API at ABS.
+Furthermore, this allows to take advantage of the already rich and mature library API of Java.
 
 \paragraph{Scalability} 
-Asynchronous message passing in ABS is a fit for distributed systems.
-In such systems, the number of messsages delivered among actors in the 
+In distributed systems, the number of messsages delivered among objects in the 
 environment is not predictable at runtime.
@@
 influence from the number of asynchronous messages delivered in the system.
+Due to support of co-operative scheduling for asynchronous messages, ABS is a fit for distributed systems.
 
-\paragraph{Modularity} 
+\paragraph{Separation of Concerns} 
 We approach ABS modelling and development with a component-based or modular 
@@
 The ABS language syntax and semantics are formalized precisely and rigorously by a structural operational semantics~\cite{johnsen2012abs}.
-The semantics of ABS can be delivered in a programming language
-as Application Programming Interface (API) as long as the programming language provides sufficient constructs to respect the ABS language semantics.
-In addition, such an ABS programming API should be \emph{verifiable}.
 If a mapping from ABS to a programming API is provided, a programmer is able to
 take advantage of ABS semantics without directly programming in ABS.
-%Such capability from ABS
 This enables industry users of mainstream languages to
@@
 platforms they are already familiar with.
-\item \emph{Modelling in ABS}
+\item \emph{Modular Architecture of ABS Tools}
 ABS language provides rigorous semantics to model concurrent
@@
 models in a user-friendly way.
-The ABS IDE and tool-set developers should be able to easily reuse and compose
-over existing modules and components.
+The ABS IDE and tool-set should be easy to reuse and extend.
 \end{itemize}
@@
 Various programming languages, in the past decade, have emerged to provide an actor-based model of asynchronous message passing~\cite{philippsen2000survey}.
-We observe different classes of actor model and concurrent model of 
-programming.
-Table~\ref{tbl:actor:pl} presents a summary.
+Table~\ref{tbl:actor:pl} presents different classes of actor model and concurrent model of programming.
 
@@
 computing using actor models and shared memory. 
-When actor model is part of the design of a language, formal verification
-and correctness of actor semantics at runtime yields naturally from the 
-language semantics at runtime.
+Having the actor model as part of a language design simplifies formal verification.
 \item[Implicit By Design]
-Refers to languages that have no explicit notion of actor models in their syntax or semantics, but do provide fundamental constructs for concurrency and asynchronous message passing.
+Refers to languages that have no explicit notion of actors in their syntax or semantics, but do provide fundamental constructs for concurrency and asynchronous message passing.
 Thus, it becomes an easy task in this kind of programming language to create an abstraction to support the actor model by coding.
-\item[External Library]
-Refers to the type of programming languages for which actor model support is provided by an external library.
+% \item[External Library]
+% Refers to the type of programming languages for which actor model support is provided by an external library.
 \end{description}
@@
 semantics, the less external libraries and frameworks target this gap. 
-In the following, we briefly enumerate frameworks and libraries for Java
+In the following, we briefly enumerate frameworks and libraries for JVM
 \footnote{A more comprehensive list can be obtained at \cite{KarmaniSA09} and  \url{https://en.wikipedia.org/wiki/Actor_model\#Programming_with_Actors}}.
@@
 One of the main techniques used in libraries to deliver actor programming in 
-Java is byte-code engineering~\cite{dahm1999byte,bruneton2002asm,asm}.
+JVM is byte-code engineering~\cite{dahm1999byte,bruneton2002asm,asm}.
 Byte-code engineering modifies the generated byte-code for compiled classes in 
@@
 \end{tabular}
-\caption{Thesis Contributions Summary}
+\caption{Thesis Organization}
 \label{tbl:thesis}
@@
 \centering
-\begin{tabular}{p{7cm}p{5cm}p{1cm}}
+\begin{tabular}{p{2cm}p{7cm}p{1cm}}
 \textsfb{Topic} & \textsfb{Proceedings / Journal} & \textsfb{Year}   
 \\ \toprule
-Programming and Deployment of Active Objects with Application-Level Scheduling & ACM SAC 2012, Pages 1883--1888 & 2012 
+Chapter~\ref{ch:p01:ch01} & ACM SAC 2012, Pages 1883--1888 & 2012 
 \\ \midrule
-The Future of a Missed Deadline & COORD 2013, Pages 181--195 & 2013 
+Chapter~\ref{ch:p01:ch02} & COORD 2013, Pages 181--195 & 2013 
 \\ \midrule
-Monitoring Method Call Sequences using Annotations & Journal of Science of Computer Programming, 2014, Volume 94, Part 3, Pages 362--378 & 2014 
+Chapter~\ref{ch:p02:ch01} & ISoLA 2014, Pages 37--53 & 2014 
 \\ \midrule
-Programming with actors in Java~8 & ISoLA 2014, Pages 37--53 & 2014 
+Chapter~\ref{ch:p03:ch02} & Journal of Science of Computer Programming, 2014, Volume 94, Part 3, Pages 362--378 & 2014 
 \\ \midrule
-Formal verification of service level agreements through distributed monitoring & ESOCC 2015, Pages 125--140 & 2015
+Chapter~\ref{ch:p03:ch01} & ESOCC 2015, Pages 125--140 & 2015
 \\ \bottomrule
@@
 \end{table}
+
+All implementations of this thesis can be found at
+
+\begin{center}
+\url{https://github.com/CrispOSS/abs-api-parent}
+\end{center}
+
+and the source of thesis can be found at
+
+\begin{center}
+\url{https://github.com/nobeh/thesis}
+\end{center}

### 2016-04-22 p00-ch00-intro/intro.tex
Apply comments based on review meeting

@@
 One of the fundamental elements of the actor model~\cite{actors:agha,agha97} is \emph{asynchronous} message passing.
-In that approach, interactions between objects are modelled as non-blocking messages.
+In that approach, interactions between objects are modeled as non-blocking messages.
 One object, the sender, communicates a message to the other object, the receiver.
@@
 
-\TODO{Put a titled paraph here? e.g. ``Concurrency Approaches''}
-
+\paragraph*{Existing Concurrency Models}
 In a concurrent setting, coroutines~\cite{conway1963design,taocp:knuth} enable interactions with collaborative pre-emption:
@@
 Creating an instance of a coroutine spawns a new process with its own state.
-A coroutine allows multiple entry points for suspending and resuming execution at certain locations.
-\TODO{and this property makes compositional concurrency of coroutines straightforward}.
-Coroutines are not originally established in the object-oriented paradigm.
+A coroutine allows multiple entry points for suspending and resuming execution at explicitly specified locations.
+The explicit interleaving points supports a compositional understanding of the code~\cite{DBLP:journals/cacm/Hoare74}.
+coroutines are not originally established in the object-oriented paradigm.
 Object orientation is based on caller-callee interaction, which is asymmetric:
@@
 The stack has a starting location (i.e. start of invocation) and a finish location (i.e. the return point).
-Future values can be used to hold the eventual return value of a method at the call site. \TODO{Do we need a citation for futures?}
+Future values can be used to hold the eventual return value of a method at the call site~\cite{DBLP:books/daglib/0092230,de2007complete}.
 In a caller-callee setting, an object calls methods from other objects.
@@
 While a message is processed, an actor cannot be pre-empted or intentionally yield to allow other actors in the system to make progress.
+Integration of actor model and object orientation leads to the use of asynchronous method calls.
 
 \paragraph*{Problem Statement and Approach}
-A question naturally rises: how to bring coroutine-style execution (co-operative scheduling) to an object-oriented setting with asynchronous communication through messages bound to object interfaces in a programming language?
-This question is the main problem statement of this thesis; cf. Table~\ref{tbl:problem}.
+The main challenge is to generate production code from an actor-based language which supports asynchronous method calls and coroutine-style execution.
+We take Java~\cite{gosling2000java} as the target programming language because it is one of the mainstream languages~\cite{tiobe} and it offers a wide range of mature and production-ready libraries and frameworks to use.
 
-\TODO{Discuss the existence of this table. Where's concurrency? Remove ``Coroutine'' row. Add a column ``Concurrency''. Fix ``Execution'' column.}
+% A question naturally rises: how to bring coroutine-style execution (co-operative scheduling) to an object-oriented setting with asynchronous communication through messages bound to object interfaces in a programming language?
+% This question is the main problem statement of this thesis; cf. Table~\ref{tbl:problem}.
 
-\begin{table}[h]
-\centering
-\begin{tabular}{lll}
- & OOP & Actor Model 
-\\ \toprule 
-\emph{Abstraction} & Class/Object & Actor 
-\\ \midrule
-\emph{Communication} & Sync. Method Call & Async. Message 
-\\ \midrule
-\emph{Execution} & Blocking + Stack-based & Blocking 
-\\ \midrule
-\emph{Concurrency} & Mutli-threading & Multi-threading + RTC
-\\ \bottomrule
-\end{tabular}
-\caption[Summarized aspects of problem statement]{
-Considering the different aspects, the problem statement of thesis focuses to combine abstraction from OOP with asynchronous messages from Actor model but in a bounded context.
-In addition, this thesis focuses to bring in co-operative scheduling into the context of object-oriented programming.
-}
-\label{tbl:problem}
-\end{table}
+% \begin{table}[h]
+% \centering
+% \begin{tabular}{lll}
+%  & OOP & Actor Model 
+% \\ \toprule 
+% \emph{Abstraction} & Class/Object & Actor 
+% \\ \midrule
+% \emph{Communication} & Sync. Method Call & Async. Message 
+% \\ \midrule
+% \emph{Execution} & Blocking + Stack-based & Blocking 
+% \\ \midrule
+% \emph{Concurrency} & Multi-threading & Multi-threading + RTC
+% \\ \bottomrule
+% \end{tabular}
+% \caption[Summarized aspects of problem statement]{
+% Considering the different aspects, the problem statement of thesis focuses to combine abstraction from OOP with asynchronous messages from Actor model but in a bounded context.
+% In addition, this thesis focuses to bring in co-operative scheduling into the context of object-oriented programming.
+% }
+% \label{tbl:problem}
+% \end{table}
 
-There already exist modelling languages that focus on the problem statement.
-Rebeca~\cite{sirjani2002simulation,sirjani2004formal,sirjani2007rebeca} is a modelling language for reactive and concurrent systems. 
+There exist actor-based executable modeling languages that support asynchronous method calls and coroutine style execution; e.g. Rebeca and ABS.
+Modeling languages are used to \emph{model} software systems for understanding, analysis, and verification. 
+Execution of models is restricted to simulation.
+As such, they do not generate production-ready code intended to run industrial environments.
+
+Rebeca~\cite{sirjani2002simulation,sirjani2004formal,sirjani2007rebeca} is a modeling language for reactive and concurrent systems. 
 In Rebeca, a number of reactive objects (\emph{rebecs}) interact at runtime.
@@
 
-ABS~\cite{johnsen2012abs,hahnlehjlssw11} is a modelling language for concurrent objects and distributed systems.
+ABS~\cite{johnsen2012abs,hahnlehjlssw11} is a modeling language for concurrent objects and distributed systems.
 ABS uses asynchronous communication among objects.
@@
 can be used as a programming language at industrial and business scale.
-ABS has mainly developed as a modelling language;
+ABS has mainly developed as a modeling language;
 for example, ABS does not support common I/O API for files or network,
@@
 
-Considering the mainstream programming languages
-\footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}}, Java~\cite{gosling2000java} is one of the most commonly-used.
-Since Java 1.5 and the rise of the concurrency API in Java JSR-166~\cite{jsr166}, substantial effort focuses on how concurrency can be improved in Java.
-However, not all development of Java libraries and frameworks proceeds based on a rigorous formal semantics. 
-Therefore, this gives rise to issues and challenges in terms of correctness, semantic preservation, and reasoning.
-In addition, the Java language specification (JLS)~\cite{gosling2000java,jls8} clearly reveals that the Java language is not yet ready to support algebraic data types.
-This has driven research and development that focuses on extending Java with support for functional programming~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
-Scala~\cite{odersky2004scala} is a result of such research. It is a dynamic, functional, and object-oriented language on JVM.
+JSR 166~\cite{jsr166} introduced a new concurrency model~\cite{DBLP:conf/java/Lea00} in Java 1.5.
+The concurrency model lays the grounds to support actor model with asynchronous method calls.
+This gave rise to various research to support actor model on Java~\cite{actor_frameworks_jvm:agha} and further extend Java as a functional programming language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
+From an industrial perspective, the development of Java language from version 6 to version 8 experienced a slow pace that influenced the growth of businesses and justified alternatives.
+Scala~\cite{odersky2004scala}, a dynamic, functional, and object-oriented language on JVM, rose to the challenge to fill the gap during the period between Java~7 and Java~8.
+Java~8~\cite{jls8} released fundamental features such as lambda expressions that are essential for concurrency and functional programming.
 
@@
 Moreover, as byte-code modification changes the byte-code of a running program or inserts new byte-code into JVM at runtime, this complicates reasoning and correctness analysis/verification~\cite{leroy2001java,leroy2003java}.
-We aim to rely only on the mainstream JVM platform released by the Java team at Oracle; in addition, we do not intend to use byte-code engineering.
+We aim to rely only on the mainstream JVM platform released by the Java team at Oracle.
+Moreover, we do not intend to use byte-code engineering used for instance in Encore~\cite{fernandez2015parallel}.
 
@@
 We utilize Java~8 features (JSR 335~\cite{jsr335} and JSR 292~\cite{jsr292:invokedyn}) to model a message as a data structure expressed in a lambda expression (cf. Chapter~\ref{ch:p02:ch01}).
-
-\TODO{Where does the following paragraph have to move?}
-
-The core contributions of this thesis target the intersection of object 
-orientation, actor model, and concurrency.
-We choose Java as the main programming language and as one of the mainstream 
-object-oriented languages. 
-We formalize a subset of Java and its concurrency API~\cite{jsr166} to 
-facilitate formal verification and reasoning about it (cf. Chapter~\ref{ch:p01:ch02}).
-We create an abstract mapping from a concurrent-object modelling language, 
-ABS~\cite{johnsen2012abs}, to the programming semantics of concurrent Java (cf. Chapter~\ref{ch:p01:ch02}). 
-We provide the formal semantics of the mapping and runtime properties of 
-the concurrency layer including deadlines and scheduling policies (cf. Chapter~\ref{ch:p01:ch01}).
-We provide an implementation of ABS concurrency layer as a Java API library 
-and framework utilizing the latest language additions 
-in Java 8~\cite{jsr335:lambda:translation} (cf. Chapter~\ref{ch:p02:ch01}).
-We design and implement a runtime monitoring framework, JMSeq, to verify the
-the correct ordering of execution of methods through code annotations in JVM (cf. Chapter~\ref{ch:p03:ch02}). 
-In addition, we design a large-scale monitoring system as a real-world 
-application; the monitoring system is built with ABS concurrent objects 
-and formal semantics that leverages schedulability 
-analysis to verify correctness of the monitors~\cite{fersman2007task} (cf. Chapter~\ref{ch:p03:ch01}).
 
@@
 
-Listing~\ref{lst:abs:java} shows a snippet of ABS code that uses \jtt{java.util.ArrayList} as a data strucuture.
+Listing~\ref{lst:abs:java} shows a snippet of ABS code that uses \jtt{java.util.ArrayList} as a data structure.
 Ideally in ABS, the programmer is able to directly use the libraries and API from Java.
@@
 \paragraph{Scalability} 
-In distributed systems, the number of messsages delivered among objects in the 
+In distributed systems, the number of massages delivered among objects in the 
 environment is not predictable at runtime.
@@
 \paragraph{Separation of Concerns} 
-We approach ABS modelling and development with a component-based or modular 
+We approach ABS modeling and development with a component-based or modular 
 software engineering practices.
@@
 
-\clearpage
-
-\section{Organization and Contributions}
+\section{Outline and Contributions}
 \label{sec:intro:contribs}
 
-Based on the problem statement and approach taken in this research, 
-Table~\ref{tbl:thesis} summarizes the structure of this text:
+The core contributions of this thesis target the intersection of object 
+orientation, actor model, and concurrency.
+We choose Java as the main programming language and as one of the mainstream 
+object-oriented languages. 
+We formalize a subset of Java and its concurrency API~\cite{jsr166} to 
+facilitate formal verification and reasoning about it (cf. Chapter~\ref{ch:p01:ch02}).
+We create an abstract mapping from a concurrent-object modeling language, 
+ABS~\cite{johnsen2012abs}, to the programming semantics of concurrent Java (cf. Chapter~\ref{ch:p01:ch02}). 
+We provide the formal semantics of the mapping and runtime properties of 
+the concurrency layer including deadlines and scheduling policies (cf. Chapter~\ref{ch:p01:ch01}).
+We provide an implementation of ABS concurrency layer as a Java API library 
+and framework utilizing the latest language additions 
+in Java 8~\cite{jsr335:lambda:translation} (cf. Chapter~\ref{ch:p02:ch01}).
+We design and implement a runtime monitoring framework, JMSeq, to verify the
+the correct ordering of execution of methods through code annotations in JVM (cf. Chapter~\ref{ch:p03:ch02}). 
+In addition, we design a large-scale monitoring system as a real-world 
+application; the monitoring system is built with ABS concurrent objects 
+and formal semantics that leverages schedulability 
+analysis to verify correctness of the monitors~\cite{fersman2007task} (cf. Chapter~\ref{ch:p03:ch01}).
+Table~\ref{tbl:thesis} summarizes the structure of this text.
 
-\begin{table}[h]
+\begin{table}[t]
 \centering
@@
 \end{table}
+
+\clearpage
 

### 2016-04-25 abstract.tex
Update summary

@@
 
-This thesis contributes to the intersection of object orientation, actor 
-model, and concurrency.
+This thesis contributes to the intersection of object orientation, actor model, and concurrency.
 We choose Java as the main programming language and as one of the mainstream 
@@
 facilitate formal verification and reasoning about it.
-We create an abstract mapping from a concurrent-object modelling language, 
-ABS, to programming semantics of Java concurrency. 
+We create an abstract mapping from a concurrent-object modeling language, 
+ABS, to the programming semantics of concurrent Java. 
 We provide the formal semantics of the mapping and runtime properties of 
@@
 We provide an implementation of ABS concurrency layer as a Java API library 
-and framework utilizing the latest langauge additions 
+and framework utilizing the latest language additions 
 in Java~8.
 We design and implement a runtime monitoring framework, JMSeq, to verify the
-sequenced execution of methods through code annotatinon in JVM. 
+the correct ordering of execution of methods through code annotations in JVM. 
 In addition, we design a large-scale monitoring system as a real-world 
 application; the monitoring system is built with ABS concurrent objects 
-and formal semantics composed with schedulability analysis.
+and formal semantics that leverages schedulability 
+analysis to verify correctness of the monitors.
 

### 2016-04-26 p00-ch00-intro/intro.tex
polish

@@
 The explicit interleaving points supports a compositional understanding of the code~\cite{DBLP:journals/cacm/Hoare74}.
-coroutines are not originally established in the object-oriented paradigm.
+Coroutines are not originally established in the object-oriented paradigm.
 Object orientation is based on caller-callee interaction, which is asymmetric:
@@
 ABS has mainly developed as a modeling language;
-for example, ABS does not support common I/O API for files or network,
+for example, ABS does not support a common I/O API for files or networking,
 and it does not provide a standard library of data structures.
@@
 JSR 166~\cite{jsr166} introduced a new concurrency model~\cite{DBLP:conf/java/Lea00} in Java 1.5.
-The concurrency model lays the grounds to support actor model with asynchronous method calls.
+The concurrency model lays the grounds to support for the actor model with asynchronous method calls.
 This gave rise to various research to support actor model on Java~\cite{actor_frameworks_jvm:agha} and further extend Java as a functional programming language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
-From an industrial perspective, the development of Java language from version 6 to version 8 experienced a slow pace that influenced the growth of businesses and justified alternatives.
+From an industrial perspective, the development of the Java language from version 6 to version 8 experienced a slow pace that influenced the growth of businesses and justified alternatives.
 Scala~\cite{odersky2004scala}, a dynamic, functional, and object-oriented language on JVM, rose to the challenge to fill the gap during the period between Java~7 and Java~8.
@@
 \paragraph{Scalability} 
-In distributed systems, the number of massages delivered among objects in the 
+In distributed systems, the number of messages delivered among objects in the 
 environment is not predictable at runtime.
@@
 \paragraph{Separation of Concerns} 
-We approach ABS modeling and development with a component-based or modular 
+We approach ABS modeling and development with a component-based and modular 
 software engineering practices.
@@
 the concurrency layer including deadlines and scheduling policies (cf. Chapter~\ref{ch:p01:ch01}).
-We provide an implementation of ABS concurrency layer as a Java API library 
+We provide an implementation of the ABS concurrency layer as a Java API library 
 and framework utilizing the latest language additions 

### 2016-04-26 p00-ch00-intro/intro.tex
Update based on comments from Stijn

@@
 ABS uses asynchronous communication among objects.
-A message in ABS is bound to a method invocation;
+A message in ABS is generated from a method enclosed by an interface.
 this defines the interface of the sent messages.
@@
 Scala~\cite{odersky2004scala}, a dynamic, functional, and object-oriented language on JVM, rose to the challenge to fill the gap during the period between Java~7 and Java~8.
-Java~8~\cite{jls8} released fundamental features such as lambda expressions that are essential for concurrency and functional programming.
+Java~8~\cite{jls8} alleviated this gap by releasing fundamental features such as lambda expressions that are essential for concurrency and functional programming.
 
@@
 \end{tabular}
-\caption{Thesis Organization}
+\caption{\thesisTitle -- Thesis Organization}
 \label{tbl:thesis}
@@
 \\ \toprule
-Chapter~\ref{ch:p01:ch01} & ACM SAC 2012, Pages 1883--1888 & 2012 
+Chapter~\ref{ch:p01:ch01} & ACM SAC 2012, pp. 1883--1888 & 2012 
 \\ \midrule
-Chapter~\ref{ch:p01:ch02} & COORD 2013, Pages 181--195 & 2013 
+Chapter~\ref{ch:p01:ch02} & COORD 2013, pp. 181--195 & 2013 
 \\ \midrule
-Chapter~\ref{ch:p02:ch01} & ISoLA 2014, Pages 37--53 & 2014 
+Chapter~\ref{ch:p02:ch01} & ISoLA 2014, pp. 37--53 & 2014 
 \\ \midrule
-Chapter~\ref{ch:p03:ch02} & Journal of Science of Computer Programming, 2014, Volume 94, Part 3, Pages 362--378 & 2014 
+Chapter~\ref{ch:p03:ch02} & 
+FACS 2010, pp. 53--70 \emph{and} 
+\newline
+Journal of Science of Computer Programming, vol. 94, part 3, pp. 362--378 & 
+2010 \emph{and}
+\newline
+2014 
 \\ \midrule
-Chapter~\ref{ch:p03:ch01} & ESOCC 2015, Pages 125--140 & 2015
+Chapter~\ref{ch:p03:ch01} & ESOCC 2015, pp. 125--140 & 2015
 \\ \bottomrule
 \end{tabular}
-\caption{Actors At Work -- Conference and Journal Publications}
+\caption{\thesisTitle -- Conference and Journal Publications}
 \label{tbl:papers}

### 2016-05-04 p00-ch00-intro/intro.tex
intro: removed duplicate word

--- 2016-04-26 p00-ch00-intro/intro.tex
@@
 We design and implement a runtime monitoring framework, JMSeq, to verify the
-the correct ordering of execution of methods through code annotations in JVM (cf. Chapter~\ref{ch:p03:ch02}). 
+correct ordering of execution of methods through code annotations in JVM (cf. Chapter~\ref{ch:p03:ch02}). 
 In addition, we design a large-scale monitoring system as a real-world 

### 2016-05-04 abstract.tex
polish
abstract: removed duplicate word

@@
 the concurrency layer including deadlines and scheduling policies.
-We provide an implementation of ABS concurrency layer as a Java API library 
+We provide an implementation of the ABS concurrency layer as a Java API library 
 and framework utilizing the latest language additions 
@@
 We design and implement a runtime monitoring framework, JMSeq, to verify the
-the correct ordering of execution of methods through code annotations in JVM. 
+correct ordering of execution of methods through code annotations in JVM. 
 In addition, we design a large-scale monitoring system as a real-world 

### 2016-05-04 p01-ch01-2012/paper.tex
Fix invalid Figure references
Fix ref/label for Chapter 01

@@
 
-\section{Introduction} \label{sec:introduction}
+\section{Introduction} \label{ch01:sec:introduction}
 One of the major challenges in the design of programming
@@
 
-Section \ref{sec:creol} first provides an overview of the Creol language with application-level scheduling. 
-In Section \ref{sec:compiler}, we elaborate on the design  of the tool-set and the
-prototype. The use of the tool-set is exemplified by a case study in Section \ref{sec:caseStudy}.
-Section \ref{sec:relwork} summarizes the related work. 
-Finally, we conclude in Section \ref{sec:conclusion}.
+Section \ref{ch01:sec:creol} first provides an overview of the Creol language with application-level scheduling. 
+In Section \ref{ch01:sec:compiler}, we elaborate on the design  of the tool-set and the
+prototype. The use of the tool-set is exemplified by a case study in Section \ref{ch01:sec:caseStudy}.
+Section \ref{ch01:sec:relwork} summarizes the related work. 
+Finally, we conclude in Section \ref{ch01:sec:conclusion}.
 
-%\section{Preliminaries: Creol} \label{sec:creol}
+%\section{Preliminaries: Creol} \label{ch01:sec:creol}
 %\input{Creol}
@@
 \section{Application-Level Scheduling}
-\label{sec:creol}
+\label{ch01:sec:creol}
 
@@
 interface Resource begin
-	(* \label{lst:priority_range} *)(*
-	%with User  \label{lst:with} 
+	(* \label{ch01:lst:priority_range} *)(*
+	%with User  \label{ch01:lst:with} 
 	*)
@@
 	var taken := false;
-	(* \label{lst:priority_class} *)
+	(* \label{ch01:lst:priority_class} *)
 	op request () == 
-		await (*$\sim$*)taken; 			(* \label{lst:await_taken} *)
+		await (*$\sim$*)taken; 			(* \label{ch01:lst:await_taken} *)
 		taken := true;
@@
 range} that is available to all clients. For instance, in Line
-\ref{lst:priority_range} of {\footnotesize\texttt{Resource}} in Listing
-\ref{lst:exres}, we can define a priority range:
+\ref{ch01:lst:priority_range} of {\footnotesize\texttt{Resource}} in Listing
+\ref{ch01:lst:exres}, we can define a priority range:
 
@@
 This involves a declaration of a priority range generally depending on the structure of the class. 
-In our example, assuming a range 0..2 added in Line \ref{lst:priority_class}, this requires changing the method signatures in the {\ftt{Exclusive\-Resource}} class:
+In our example, assuming a range 0..2 added in Line \ref{ch01:lst:priority_class}, this requires changing the method signatures in the {\ftt{Exclusive\-Resource}} class:
 
@@
 being executed. 
-For example, a process waiting at Line \ref{lst:await_taken} could be given a higher priority over new requests by assigning a smaller value:
+For example, a process waiting at Line \ref{ch01:lst:await_taken} could be given a higher priority over new requests by assigning a smaller value:
 
@@
 % Method definitions without an explicit priority are given a default priority value. 
-% %As discussed in Section \ref{sec:conclusion}, 
+% %As discussed in Section \ref{ch01:sec:conclusion}, 
 % Other possible levels of priority include method invocation level, specified by the caller, or interface level priorities. 
@@
 We may declare a scheduling policy for the \ftt{ExclusiveResource} class by adding at Line
-\ref{lst:priority_class} in Listing \ref{lst:exres}: 
+\ref{ch01:lst:priority_class} in Listing \ref{ch01:lst:exres}: 
 %
@@
 
-\section{Tool Architecture} \label{sec:compiler}
+\section{Tool Architecture} \label{ch01:sec:compiler}
 
@@
   \caption{\Crisp Architecture: Structural Overview}
-  \label{fig:api}
+  \label{ch01:fig:api}
 \end{center}
@@
 \Crisp provides a one-to-one mapping from Creol classes and methods to their equivalent Java constructs. 
-In order to implement active objects in Creol, we use the \jtt{java.util.concurrent} package (see Figure \ref{fig:api}).
+In order to implement active objects in Creol, we use the \jtt{java.util.concurrent} package (see Figure \ref{ch01:fig:api}).
 Each active object consists of an instance of a process store  and an execution manager to store and execute the method invocations.
@@
 \subsection{A New Method Invocation}
-\label{sec:add}
+\label{ch01:sec:add}
 
@@
 \caption[New MethodInvocation]{Adding the new MethodInvocations are performed on the Client side.}
-\label{fig:add}
+\label{ch01:fig:add}
 %\end{center}
@@
 client. 
-We adopt the second approach as depicted in Figure \ref{fig:add}.
+We adopt the second approach as depicted in Figure \ref{ch01:fig:add}.
 At any time, there can be concurrent clients that are
@@
 %Storing a method invocation in the process store of the server takes place in
-%the client side thread of execution. Listing \ref{lst:jfib_method_fibonacci}
+%the client side thread of execution. Listing \ref{ch01:lst:jfib_method_fibonacci}
 %presents a sample generated Java code for an asynchronous method call in Listing
-%\ref{lst:cfib}.
+%\ref{ch01:lst:cfib}.
 
@@
 repeatedly fetches an instance of method invocation from its process
-store for execution (cf. Fig \ref{fig:take}).
+store for execution (cf. Fig \ref{ch01:fig:take:next}).
 The process store uses its instance of
@@
 and customized based on the requirements by the user. 
-%Listing \ref{lst:aoloop} presents the main loop for the active object.
+%Listing \ref{ch01:lst:aoloop} presents the main loop for the active object.
 
@@
 % \vspace{-.7cm}
+\end{center}
 \caption[Policy-based selection of a MethodInvocation]{An active object selects a method invocation based on its local scheduling policy.
 After a method finishes execution, the whole scenario is repeated.}
-\end{center}
-\label{fig:take}
+\label{ch01:fig:take:next}
 \end{figure}
@@
 % storage. 
-%An example of a user-defined priority manager is given in Section \ref{sec:caseStudy}.
+%An example of a user-defined priority manager is given in Section \ref{ch01:sec:caseStudy}.
 Postponing the act of resolving priorities to this point rather than when inserting new processes to the store enables us to handle dynamic priorities.
@@
 
-As shown in Figure \ref{fig:execute}, the thread factory creates threads of type  {\ftt{InterruptibleProcess}}.
+As shown in Figure \ref{ch01:fig:execute}, the thread factory creates threads of type  {\ftt{InterruptibleProcess}}.
 The execution manager thread blocks as soon as it starts the interruptible process which executes the associated  method invocation. 
-If the method releases the processor before completion, it will be added back to the process store as explained in Section \ref{sec:add}.
+If the method releases the processor before completion, it will be added back to the process store as explained in Section \ref{ch01:sec:add}.
 When a suspended method invocation is resumed, the execution manager skips the creation of a new thread and  reuses the one that was assigned to the method invocation before.
@@
 % \vspace{-.7cm}
+\end{center}
 \caption[Execution of a MethodInvocation]{A method invocation is executed in an interruptible process. The execution manager thread is blocked while the interruptible process is running.}
-\end{center}
-\label{fig:execute}
+\label{ch01:fig:execute}
 \end{figure}
@@
 
-\section{Case Study} \label{sec:caseStudy}
+\section{Case Study} \label{ch01:sec:caseStudy}
 
@@
 %
-To avoid erroneous behavior, we use the actual parameter $n$ in \ftt{divide} method to define its priority level, too (see line \ref{line:priority}). 
+To avoid erroneous behavior, we use the actual parameter $n$ in \ftt{divide} method to define its priority level, too (see line \ref{ch01:line:priority}). 
 As a result, every invocation of this method generates a process with a priority equal to its parameter.
@@
 
-class Prime(p: Int) implements IPrime begin (*\label{lst:priority}*)
+class Prime(p: Int) implements IPrime begin (*\label{ch01:lst:priority}*)
 	var next: IPrime
-	op divide(n: Int) priority (n) == (*\label{line:priority}*)
+	op divide(n: Int) priority (n) == (*\label{ch01:line:priority}*)
 		if (n % p) /= 0 then 
@@
 The first interesting observation was that \Crisp prime sieve utilizes all the CPU's on the underlying hardware as much as possible during  execution. 
-This can be seen in Figure \ref{fig:multi-cpu-1} which shows the CPU usage. 
+This can be seen in Figure \ref{ch01:fig:multi-cpu-1} which shows the CPU usage. 
 Both CPUs are fairly in use while running this program.
-Figure \ref{fig:prime_sieve_cacoj} depicts the results of the monitoring of the parallel prime sieve in \Crisp using Visual VM tool. 
+Figure \ref{ch01:fig:prime_sieve_cacoj} depicts the results of the monitoring of the parallel prime sieve in \Crisp using Visual VM tool. 
 It depicts the number of threads generated for the program.
@@
 In JVM, the size of thread stack can be configured using \ftt{-Xss} option for every run. 
-To demonstrate this feature of \Crisp, we collected the minimum stack size needed for every program run in Table \ref{tbl:cmp-thread-stack}. 
+To demonstrate this feature of \Crisp, we collected the minimum stack size needed for every program run in Table \ref{ch01:tbl:cmp-thread-stack}. 
 All \Crisp runs use the minimum thread stack size of \ftt{64k} that is possible for the JVM.
@@
 % The results depict that the sequential version, through time, increasingly required memory allocation for the thread stack, however, \Crisp performs on a constant size of thread stack. 
-% Table \ref{tbl:cmp-thread-stack} summarizes the thread stack allocation on different runs.
+% Table \ref{ch01:tbl:cmp-thread-stack} summarizes the thread stack allocation on different runs.
 
@@
 \caption{Thread stack allocated for different executions}
-\label{tbl:cmp-thread-stack}
+\label{ch01:tbl:cmp-thread-stack}
 \end{center}
@@
 It can handle huge number of thread generation if required.
-Table \ref{tbl:cmp-thread-num-gen} summarizes the thread generation data for parallel prime sieve in \Crisp.
+Table \ref{ch01:tbl:cmp-thread-num-gen} summarizes the thread generation data for parallel prime sieve in \Crisp.
 It shows scalability of \Crisp as $p$ rises for parallel prime sieve.
@@
 \caption{Number of live threads and total threads created for different runs of parallel prime sieve}
-\label{tbl:cmp-thread-num-gen}
+\label{ch01:tbl:cmp-thread-num-gen}
 \end{center}
@@
 application level. We further discuss this in Section
-\ref{sec:conclusion}.
+\ref{ch01:sec:conclusion}.
 
@@
 \caption{Increasing parallelism in \Crisp for Prime Sieve}
-\label{fig:prime_sieve_cacoj}
+\label{ch01:fig:prime_sieve_cacoj}
 % \end{center}
@@
   \caption{Utilizing both CPUs with Prime Sieve in \Crisp}
-  \label{fig:multi-cpu-1}
+  \label{ch01:fig:multi-cpu-1}
 \end{center}
@@
 
-\section{Related Work} \label{sec:relwork}
+\section{Related Work} \label{ch01:sec:relwork}
 \input{related2}
 
-\section{Conclusion} \label{sec:conclusion}
+\section{Conclusion} \label{ch01:sec:conclusion}
 In this paper, we proposed \Crisp as an implementation scheme for
@@
 complete control over scheduling of messages and threads in
-\jtt{Executor\-Service}'s internal queue. Table \ref{tbl:eval}
+\jtt{Executor\-Service}'s internal queue. Table \ref{ch01:tbl:eval}
 summarizes this discussion.
@@
 \caption{Overview of evaluation of challenges}
-\label{tbl:eval}
+\label{ch01:tbl:eval}
 \end{center}

### 2016-05-04 p01-ch02-2013/paper.tex
Fix ref/label for Chapter 02

@@
 
-\section{Introduction} \label{sec:introduction}
+\section{Introduction} \label{ch02:sec:introduction}
 
@@
 
-The paper continues as follows: Section \ref{sec:deadlines} introduces the language constructs
-and provides informal semantics of the language with a case study in Section \ref{sec:frh}. 
-Section \ref{sec:sos} presents the operational semantics of Crisp.
-Section \ref{sec:impl} follows to provide a detailed discussion on the implementation.
+The paper continues as follows: Section \ref{ch02:sec:deadlines} introduces the language constructs
+and provides informal semantics of the language with a case study in Section \ref{ch02:sec:frh}. 
+Section \ref{ch02:sec:sos} presents the operational semantics of Crisp.
+Section \ref{ch02:sec:impl} follows to provide a detailed discussion on the implementation.
 The case study continues in this section with further details and code examples.
-Section \ref{sec:relwork} discusses related work of research and 
-finally Section \ref{sec:conclusion} concludes the paper and proposes future line of research.
+Section \ref{ch02:sec:relwork} discusses related work of research and 
+finally Section \ref{ch02:sec:conclusion} concludes the paper and proposes future line of research.
 
-\section{Programming with deadlines} \label{sec:deadlines}
+\section{Programming with deadlines} \label{ch02:sec:deadlines}
 In this section, we introduce the basic concepts underlying  the notion of ``deadlines'' for asynchronous messages between actors. 
 The main new constructs specify how a message can be sent with a deadline, how the message response can be processed, and what happens when a deadline is missed.
-We discuss the informal semantics of these concepts  and illustrate them using a case study in Section \ref{sec:frh}.
+We discuss the informal semantics of these concepts  and illustrate them using a case study in Section \ref{ch02:sec:frh}.
 
-Listing \ref{fig:grammar} introduces a minimal version of the  real-time
+Listing \ref{ch02:fig:grammar} introduces a minimal version of the  real-time
 actor-based language Crisp.
@@
 }
-\label{fig:grammar}
+\label{ch02:fig:grammar}
 \end{figure}
@@
 to  \emph{run to completion}.
-In  Section \ref{sec:impl} we discuss the implementation details of this design choice.
+In  Section \ref{ch02:sec:impl} we discuss the implementation details of this design choice.
 
@@
 
-For example, in Listing \ref{lst:try:catch}, if the \gett \space operation raises an exception  control,  is transferred to line (5);
+For example, in Listing \ref{ch02:lst:try:catch}, if the \gett \space operation raises an exception  control,  is transferred to line (5);
 otherwise, the execution continues in line (3). 
@@
 
-% Since, we provide a reference implementation in Java (c.f. Section \ref{sec:impl}), we take advantage of exception ``stack traces''; 
+% Since, we provide a reference implementation in Java (c.f. Section \ref{ch02:sec:impl}), we take advantage of exception ``stack traces''; 
 % i.e. an instance of exception can produce a chain of ``causes'' that has led to the occurrence of the exception.
@@
 % \emph{Discussion?} Compare the approaches: event-based with no get() and exceptions as messages vs. messages with exceptions
-\subsection{Case Study: Fredhopper Distributed Data Processing} \label{sec:frh}
+\subsection{Case Study: Fredhopper Distributed Data Processing} \label{ch02:sec:frh}
 
@@
 \caption[Life cycle for remote data processing]{Fredhopper's Controller life cycle for remote data processing}
-\label{fig:frh-dm}
+\label{ch02:fig:frh-dm}
 % \vspace{-10pt}
@@
 When the data processing is complete, the result is again published to customer environment and additionally becomes available through FAS services.
-Figure \ref{fig:frh-dm} illustrates an example scenario that is described above.
+Figure \ref{ch02:fig:frh-dm} illustrates an example scenario that is described above.
 
@@
 
-To illustrate the contribution of this paper, we extract a closed-world simplified version of the scenario in Figure \ref{fig:frh-dm} from Controller.
-In Section \ref{sec:impl}, we provide an implementation-level usage of our work applied to this case study.
+To illustrate the contribution of this paper, we extract a closed-world simplified version of the scenario in Figure \ref{ch02:fig:frh-dm} from Controller.
+In Section \ref{ch02:sec:impl}, we provide an implementation-level usage of our work applied to this case study.
 
-\section{Operational Semantics} \label{sec:sos}
+\section{Operational Semantics} \label{ch02:sec:sos}
 We describe the semantics of the language by means of a two-tiered labeled transition system:
@@
 where   $\mathsf{insert}(q, m(\tau))$  defines the result of inserting the process  $(S,\tau)$, where $S$ denotes the body of method $m$,
-in $q$, according to some application-specific policy (described below in Section \ref{sec:impl}).
+in $q$, according to some application-specific policy (described below in Section \ref{ch02:sec:impl}).
 
@@
 % in which the operation \textsf{new()} creates a new instance of a process for the concurrent object.
-% Additionally, as we discssued in Equation \ref{eq:orderedq}, each concurrent object owns an ordered queue for the processes.
+% Additionally, as we discssued in Equation \ref{ch02:eq:orderedq}, each concurrent object owns an ordered queue for the processes.
 % The process queue is ordered based on the result of $\sigma$ function:
@@
 
-\section{Implementation} \label{sec:impl}
+\section{Implementation} \label{ch02:sec:impl}
 We base our implementation on Java's concurrent package: \jtt{java.util.concurrent}. 
@@
 \parbf{Fredhopper case study.}
-As introduced in Section \ref{sec:frh}, we extract a closed-world simplified version from Fredhopper Controller.
+As introduced in Section \ref{ch02:sec:frh}, we extract a closed-world simplified version from Fredhopper Controller.
 We apply the approach discussed in this paper to use deadlines for asynchronous messages.
-% Controller operates in a high-demand distributed setting as in Figure \ref{fig:ctl:deploy}.
+% Controller operates in a high-demand distributed setting as in Figure \ref{ch02:fig:ctl:deploy}.
 
-Listing \ref{lst:frh:polling} and \ref{lst:frh:deadlines} present the difference in the previous Controller and the approach in Crisp.
+Listing \ref{ch02:lst:frh:polling} and \ref{ch02:lst:frh:deadlines} present the difference in the previous Controller and the approach in Crisp.
 The left code snippet shows the Controller that uses polling to retrieve data processing results.
@@
 % \caption{Fredhopper's Controller distribution model}
-% \label{fig:ctl:deploy}
+% \label{ch02:fig:ctl:deploy}
 % \vspace{-15pt}
@@
 However, it does not provide language constructs to specify a deadline for a message that is sent to data processing service.
-A deadline may be simulated using a combination of timeout and periodic polling approaches (Listing \ref{lst:frh:polling}).
-Though, this approach cannot guarantee eventual termination in all cases; as discussed before that Step 4 in Figure \ref{fig:frh-dm} may never complete.
+A deadline may be simulated using a combination of timeout and periodic polling approaches (Listing \ref{ch02:lst:frh:polling}).
+Though, this approach cannot guarantee eventual termination in all cases; as discussed before that Step 4 in Figure \ref{ch02:fig:frh-dm} may never complete.
 Controller is required to meet certain customer expectations based on an SLA.
@@
 
-In the above listings, Listing \ref{lst:frh:sched} defines a custom scheduler that determines the priority of two processes with custom logic for specific customer.
-To use the custom scheduler, the only requirement is that the class \jtt{DataProcessor} defines a specific class variable called \jtt{scheduler} in Listing \ref{lst:frh:dm}.
+In the above listings, Listing \ref{ch02:lst:frh:sched} defines a custom scheduler that determines the priority of two processes with custom logic for specific customer.
+To use the custom scheduler, the only requirement is that the class \jtt{DataProcessor} defines a specific class variable called \jtt{scheduler} in Listing \ref{ch02:lst:frh:dm}.
 The \textsl{custom scheduler} is picked up by Crisp core architecture and is used to schedule the queued processes.
@@
 \caption{Evaluation Results}
-\label{tbl:eval}
+\label{ch02:tbl:eval}
 % \vspace{-15pt}
@@
 In this evaluation, a total number of about 25000 data processing requests were included.
-The results show $6\%$ improvement in Table \ref{tbl:eval} (that amounts to around 1600 better data processing requests). 
+The results show $6\%$ improvement in Table \ref{ch02:tbl:eval} (that amounts to around 1600 better data processing requests). 
 Because of data issues or wrong parameters in the data processing requests, there are requests that still fail or never end and should be handled by a human resource.
@@
 
-\section{Related Work} \label{sec:relwork}
+\section{Related Work} \label{ch02:sec:relwork}
 The programming language presented in this paper is a real-time  extension of the  language introduced in  \cite{crisp-sac}. This new extension features
@@
 
-\section{Conclusion and future work} \label{sec:conclusion}
+\section{Conclusion and future work} \label{ch02:sec:conclusion}
 In this paper, we presented both a formal semantics and  an implementation of a real-time actor-based programming language.

### 2016-05-04 p02-ch01-2014/paper.tex
Fix ref/label for Chapter 03

@@
 \section{Introduction}
-\label{sec:introduction}
+\label{ch03:sec:introduction}
 Java is beyond doubt one of the mainstream object oriented programming languages 
@@
 The paper continues as follows: 
-in Section \ref{sec:relatedwork}, we briefly discuss a set of related works on actors and concurrent models especially on JVM platform. 
-Section \ref{sec:example} presents an example that we use throughout the paper, we start to model the example using a library. 
-Section \ref{sec:actor:programming} briefly introduces a concurrent modeling language and implements the example.
-Section \ref{sec:j8:features} briefly discusses Java 8 features that this works uses for implementation.
-Section \ref{sec:j8:actors:realization} presents how an actor model maps into programming in Java~8.
-Section \ref{sec:impl:arch} discusses in detail the implementation architecture of the actor API.
-Section \ref{sec:experiments} discusses how a number of benchmarks were performed for the implementation of the API and how they compare with current related works.
-Section \ref{sec:conclusion} concludes the paper and discusses the future work.
+in Section \ref{ch03:sec:relatedwork}, we briefly discuss a set of related works on actors and concurrent models especially on JVM platform. 
+Section \ref{ch03:sec:example} presents an example that we use throughout the paper, we start to model the example using a library. 
+Section \ref{ch03:sec:actor:programming} briefly introduces a concurrent modeling language and implements the example.
+Section \ref{ch03:sec:j8:features} briefly discusses Java 8 features that this works uses for implementation.
+Section \ref{ch03:sec:j8:actors:realization} presents how an actor model maps into programming in Java~8.
+Section \ref{ch03:sec:impl:arch} discusses in detail the implementation architecture of the actor API.
+Section \ref{ch03:sec:experiments} discusses how a number of benchmarks were performed for the implementation of the API and how they compare with current related works.
+Section \ref{ch03:sec:conclusion} concludes the paper and discusses the future work.
 
 \section{Related Work}
-\label{sec:relatedwork}
+\label{ch03:sec:relatedwork}
 
@@
 \section{State of the Art: An example}
-\label{sec:example}
+\label{ch03:sec:example}
 In the following, we illustrate the state of the art in actor programming by means of a simple example using
@@
 We want to model in Akka an ``asynchronous ping-pong match'' between two actors represented by the 
-two interfaces \jtt{IPing} and \jtt{IPong} which are depicted in Listings \ref{lst:ping:interface} and \ref{lst:pong:interface}.
+two interfaces \jtt{IPing} and \jtt{IPong} which are depicted in Listings \ref{ch03:lst:ping:interface} and \ref{ch03:lst:pong:interface}.
 An asynchronous call by the actor implementing the \jtt{IPong} interface of the \jtt{ping} method of the actor
@@
 a given pre-defined  class \jtt{UntypedActor}  and \emph{implement} the interface \jtt{IPing}, 
-as depicted in Listings \ref{lst:ping:akka} and \ref{lst:pong:akka}.
+as depicted in Listings \ref{ch03:lst:ping:akka} and \ref{ch03:lst:pong:akka}.
 The class \jtt{UntypedActor} provides two Akka framework methods \jtt{tell} and \jtt{onReceive} 
@@
 Access to the sender of the message in Akka is provided by \jtt{sender()}.
-In the main method as described in Listing \ref{lst:main:akka} we show how the initialize and start the ping/pong match.
+In the main method as described in Listing \ref{ch03:lst:main:akka} we show how the initialize and start the ping/pong match.
 Note that a reference to the ``pong'' actor is passed to the ``ping'' actor.
@@
 \section{Actor Programming in Java}
-\label{sec:actor:programming}
+\label{ch03:sec:actor:programming}
 
@@
 Release statements can be conditional (e.g., checking a future for the return value).
-Listings \ref{lst:ping:abs}, \ref{lst:pong:abs} and \ref{lst:main:ABS} present an implementation of ping-pong example in ABS.
-By means of the statement on line 6 of Listing \ref{lst:ping:abs}
+Listings \ref{ch03:lst:ping:abs}, \ref{ch03:lst:pong:abs} and \ref{ch03:lst:main:ABS} present an implementation of ping-pong example in ABS.
+By means of the statement on line 6 of Listing \ref{ch03:lst:ping:abs}
 a ``ping'' object  directly calls asynchronously the \jtt{pong} method  of its ``pong'' object, and vice versa.
@@
 \section{Java 8 Features}
-\label{sec:j8:features}
+\label{ch03:sec:j8:features}
 
@@
 \section{Modeling actors in Java~8}
-\label{sec:j8:actors:realization}
+\label{ch03:sec:j8:actors:realization}
 
@@
 \paragraph{The \jtt{Actor} Interface}
-We introduce an interface to model actors  using Java~8 features discussed in Section~\ref{sec:j8:features}.
+We introduce an interface to model actors  using Java~8 features discussed in Section~\ref{ch03:sec:j8:features}.
 Implementing an interface in Java means that the object exposes public APIs specified by the interface that is considered the behavior of the object.
@@
 
-The \jtt{Actor} interface (Listings \ref{lst:j8:run} and \ref{lst:j8:enq})
+The \jtt{Actor} interface (Listings \ref{ch03:lst:j8:run} and \ref{ch03:lst:j8:enq})
 provides a set of \bfjtt{default} methods, namely the \jtt{run} and \jtt{send} methods,  which the implementing classes do not need to re-implement.
@@
 \paragraph{Modeling Asynchronous Messages} 
-\label{par:asynchronous_messages}
+\label{ch03:par:asynchronous_messages}
 We model an asynchronous call 
@@
 $$
-to a method in ABS by the Java~8 code snippet of Listing \ref{lst:j8:call0}.
+to a method in ABS by the Java~8 code snippet of Listing \ref{ch03:lst:j8:call0}.
 The final local variables $\jtt{u}_1$, $\ldots$, $\jtt{u}_n$ (of the caller) are used to store the values of
@@
 Note that the generated object which  represents the  lambda expression will contain the local context of the caller of the method ``$m$'' (including the local variables storing the values of the expressions $e_1,\ldots,e_n$), which will be restored upon execution of the lambda expression.
-Listing \ref{lst:j8:call1} models an asynchronous call to a method without a return value.
+Listing \ref{ch03:lst:j8:call1} models an asynchronous call to a method without a return value.
 
-As an example, Listings \ref{lst:ping} and \ref{lst:pong} present the running ping/pong example, using the above API.
-The main program to use ping and pong implementation is presented in Listing~\ref{lst:main:api}.
+As an example, Listings \ref{ch03:lst:ping} and \ref{ch03:lst:pong} present the running ping/pong example, using the above API.
+The main program to use ping and pong implementation is presented in Listing~\ref{ch03:lst:main:api}.
 
@@
 
-% In Listing \ref{lst:ping}, the implementation uses a reference to another actor; namely \jtt{pong}.
+% In Listing \ref{ch03:lst:ping}, the implementation uses a reference to another actor; namely \jtt{pong}.
 % It sends a message to the actor, captures the result into a future value and waits for the result.
 % The processing of the message in \jtt{pong} leads to the invocation of the method ``pong'' 
-% using the provided argument and returns the result (Listing \ref{lst:pong}).
+% using the provided argument and returns the result (Listing \ref{ch03:lst:pong}).
 % 
 As demonstrated in the above examples, the ``ping'' and ``pong'' objects \emph{preserve} 
-their own \emph{interfaces}  contrary to the example depicted in Section \ref{sec:example} 
+their own \emph{interfaces}  contrary to the example depicted in Section \ref{ch03:sec:example} 
 in which the objects \emph{extend} a specific ``universal actor abstraction'' to inherit methods and behaviors to become an actor.
-Further,  messages are processed \emph{generically} by the \jtt{run} method described in Listing \ref{lst:j8:run}.
+Further,  messages are processed \emph{generically} by the \jtt{run} method described in Listing \ref{ch03:lst:j8:run}.
 Although, in the first place, sending an asynchronous may look like to be able to change the recipient actor's state, this is not correct.
@@
 % We describe the  translation of await statements in the context of a general translation of statements in the ABS language.
-We model cooperative scheduling by means of a call to the \jtt{await} method in Listing~\ref{lst:j8:await}.
+We model cooperative scheduling by means of a call to the \jtt{await} method in Listing~\ref{ch03:lst:j8:await}.
 Note that the preempted process is thus passed as an additional parameter and as such queued in case the guard is false,
@@
 % \textsl{Actor API in Java~8} 
-% \label{par:actor_api_in_java_8}
+% \label{ch03:par:actor_api_in_java_8}
 
 \section{Implementation Architecture}
-\label{sec:impl:arch}
+\label{ch03:sec:impl:arch}
 
-Figure \ref{fig:actors:arch} presents the general layered architecture of the actor API in Java 8.
+Figure \ref{ch03:fig:actors:arch} presents the general layered architecture of the actor API in Java 8.
 It consists of  three layers: the  routing layer  which
@@
 \caption{Architecture of Actor API in Java~8}
-\label{fig:actors:arch}
+\label{ch03:fig:actors:arch}
 \end{figure}
@@
 % \section{Distribution of Actors}
-% \label{sec:distributtion}
+% \label{ch03:sec:distributtion}
 
 We discuss next the distribution of actors in this architecture.
-In the architecture presented  in Figure~\ref{fig:actors:arch}, each layer can be \emph{distributed} independently of another layer in a transparent way.
+In the architecture presented  in Figure~\ref{ch03:fig:actors:arch}, each layer can be \emph{distributed} independently of another layer in a transparent way.
 Not only the routing layer can provide distribution, the queue layer of the architecture may also be remote to take advantage of cluster storage for actor messages. 
@@
 A complete example of using the remote API has been developed\footnote{An example of ABS Remote API is available at \surl{https://github.com/CrispOSS/abs-api-remote-sample}.}. 
-Expanding our ping-pong example in this paper, Listing~\ref{lst:ping:remote} and \ref{lst:pong:remote} present how a remote server of actors is created for the ping and pong actors.
+Expanding our ping-pong example in this paper, Listing~\ref{ch03:lst:ping:remote} and \ref{ch03:lst:pong:remote} present how a remote server of actors is created for the ping and pong actors.
 In the following listings, \jtt{java.util.Properties} is used provide input parameters of the actor server;
@@
 
-In Listing~\ref{lst:ping:remote}, a remote reference to a pong actor is created that exposes the \jtt{IPong} interface.
+In Listing~\ref{ch03:lst:ping:remote}, a remote reference to a pong actor is created that exposes the \jtt{IPong} interface.
 This interface is proxied
@@
 When an actor is initialized in a distributed setting it transparently identifies its actor server and registers with it.
-The above two listings are aligned with the similar main program presented in Listing~\ref{lst:main:api} that presents the same in a local setting.
+The above two listings are aligned with the similar main program presented in Listing~\ref{ch03:lst:main:api} that presents the same in a local setting.
 The above two listings run in separate JVM instances and therefore do not share any objects.
@@
 \section{Experiments}
-\label{sec:experiments}
+\label{ch03:sec:experiments}
 
@@
 }
-\label{fig:benchmarks}
+\label{ch03:fig:benchmarks}
 \end{figure}
@@
 
-The results are presented in Figure \ref{fig:benchmarks}.
+The results are presented in Figure \ref{ch03:fig:benchmarks}.
 The performance difference observed in the measurements can be explained as follows.
-An actor in Akka is expected to expose a certain behavior as discussed in Section \ref{sec:example} (i.e. \jtt{onReceive}).
+An actor in Akka is expected to expose a certain behavior as discussed in Section \ref{ch03:sec:example} (i.e. \jtt{onReceive}).
 This means that every message leads to an eventual invocation of this method inside actor.
@@
 \section{Conclusion}
-\label{sec:conclusion}
+\label{ch03:sec:conclusion}
 

### 2016-05-04 p03-ch01-2015-monitoring/paper-esocc.tex
Add publication information to chapter abstracts
Fix ref/label for Chapter 04

@@
 \section*{Abstract}
-\label{abstract}
+\label{ch04:abstract}
 In this paper, we introduce a  formal model of the  availability, budget compliance
@@
 
+%
+
+\subparagraph*{Conference Publication}
+\emph{Lecture Notes in Computer Science, Volume 9306, Service Oriented Cloud Computing, 4th European Conference -- ESOCC 2015, Pages 125--140, DOI 
+\smalltt{10.1007/978-3-319-24072-5\_9}}
+
+%
+
 \input{sec01-intro}

### 2016-05-04 p03-ch01-2015-monitoring/sec02-related.tex
Fix ref/label for Chapter 04

@@
 \section{Related Work}
-\label{sec:relatedwork}
+\label{ch04:sec:relatedwork}
 
@@
 They show how two examples of real-world SLAs would lead one service provider to report 0\% availability while another would report 100\% for the same system state history but using a different period of time.
-The transparency that the authors attempt to reach is addressed in our work by the concept of monitoring window and expectation tolerance in Section \ref{sec:modelling}.
+The transparency that the authors attempt to reach is addressed in our work by the concept of monitoring window and expectation tolerance in Section \ref{ch04:sec:modelling}.
 Additionally, the authors take a continuous time approach contrasted with ours that uses discrete time advancements.

### 2016-05-04 p03-ch01-2015-monitoring/sec045-defs.tex
Add publication information to chapter abstracts
Remove all `vspace`
Replace `wrapfigure` with `figure`
Fix ref/label for Chapter 04

@@
 \section{Distributed Monitoring Model}
-\label{sec:modelling}
+\label{ch04:sec:modelling}
 
@@
 %\caption{{\footnotesize High-level deployment architecture}}
-%\label{fig:deployment}
+%\label{ch04:fig:deployment}
 %\vspace{-10pt}
 %\end{wrapfigure}
-%Figure \ref{fig:deployment} presents a high-level overview of services deployed in a distributed environment (e.g. ``the cloud'').
+%Figure \ref{ch04:fig:deployment} presents a high-level overview of services deployed in a distributed environment (e.g. ``the cloud'').
 In the deployment environment (e.g., ``the cloud''), every  server from the IaaS provider is  used for a \emph{single} service of a customer, such as
-the Query Service API for a customer of SDL Fredhopper (c.f. Section \ref{sec:fredhopper_example}).
+the Query Service API for a customer of SDL Fredhopper (c.f. Section \ref{ch04:sec:fredhopper_example}).
 Typically, multiple servers are allocated to a single customer.
@@
 characteristics in a deployment environment.
- % given in Figure \ref{fig:deployment}.
+ % given in Figure \ref{ch04:fig:deployment}.
 The monitoring platform works in two cyclic phases: \emph{observation} and \emph{reaction}.
@@
 %\begin{assm}[Unlimited Capacity]
-%\label{def:platform:cap}
+%\label{ch04:def:platform:cap}
 %We assume that the external infrastructure provider is capable to provision an \emph{unlimited} number of resources.
@@
 %\begin{assm}[Single Resource Type]
-%\label{def:single:resource:type}
+%\label{ch04:def:single:resource:type}
 %To simplify reasoning, we assume that all resources are of the \emph{same type}; i.e. they have the same computing power, memory, and IO capacity.
@@
 %\begin{assm}[Resource Initialization Time $t_i$]
-%\label{def:resource:init:time}
+%\label{ch04:def:resource:init:time}
 %We assume that every resource $r$ that is initialized is ready for use in at most $t_i$ amount of time.
@@
 %Initialization time can vary between different resource types.
-The above Assumption \ref{def:single:resource:type} simplifies this: we can just choose the initialization time (a fixed constant) of the single resource type that was chosen.
+The above Assumption \ref{ch04:def:single:resource:type} simplifies this: we can just choose the initialization time (a fixed constant) of the single resource type that was chosen.
 %The initialization time is also part of the contract with the IaaS provider.
-%As discussed above in Assumption \ref{def:single:resource:type}, when a resource is launched and initialized, it might take a different amount of time to be in a state that is \emph{operational} and ready to be used.
+%As discussed above in Assumption \ref{ch04:def:single:resource:type}, when a resource is launched and initialized, it might take a different amount of time to be in a state that is \emph{operational} and ready to be used.
 %Thus,
@@
 A service exposes an API that is accessible through a delivery layer, such as HTTP.
-In our example, a service is the Query API (c.f. Section~\ref{sec:fredhopper_example}) that is accessible through a single HTTP endpoint.
+In our example, a service is the Query API (c.f. Section~\ref{ch04:sec:fredhopper_example}) that is accessible through a single HTTP endpoint.
 
@@
 \emph{monitoring platform}  $P$ is responsible for (de-)allocation of resources for computation or storage.
-We abstract from a specific implementation of the monitoring platform $P$ through an API in Listing~\ref{lst:platform}. 
+We abstract from a specific implementation of the monitoring platform $P$ through an API in Listing~\ref{ch04:lst:platform}. 
 %
+
 \lstset{language=java,aboveskip=-20pt,belowskip=-20pt}
-\begin{wrapfigure}{r}{0.5\textwidth}
-\vspace{-15pt}
+\begin{figure}[h]
+% \begin{wrapfigure}{r}{0.5\textwidth}
+% \vspace{-15pt}
 \begin{center}
@@
 \end{center}   
-\vspace{-15pt}
-\end{wrapfigure}
+% \vspace{-15pt}
+% \end{wrapfigure}
+\end{figure}
+\lstset{mathescape=false}
 %
+
 There is only \emph{one} instance of $P$ available.
@@
 \begin{defn}[Service Availability $\alpha(s,\tau,t_c)$]
-\label{def:service:availability}
+\label{ch04:def:service:availability}
 % 
@@
 We define the availability of a service $\alpha(s,\tau,t_c)$ in every monitoring window $\tau$ as:
-\vspace{-10pt}
+% \vspace{-10pt}
 \[
@@
 \begin{defn}[Budget Compliance $\beta(s,\tau)$]
-\label{def:budget:compliance}
+\label{ch04:def:budget:compliance}
 % 

### 2016-05-04 p03-ch01-2015-monitoring/sec05-monitoring-model.tex
Fix ref/label for Chapter 04

@@
 \section{Service Characteristics Verification}
-\label{sec:timed:fsm}
+\label{ch04:sec:timed:fsm}
 
@@
 is within the thresholds, taking tolerance into account
-(Definition~\ref{def:service:availability}).
+(Definition~\ref{ch04:def:service:availability}).
 If the condition fails, $M_{\alpha_s}$ generates a task for monitoring platform $P$ to allocate a new resource to service $s$ with a deadline of $\tau$.
@@
 We use the semantics of a task automata in \cite{jaghoori2010time} p. 92 in the transitions of the task automata.
-Figure~\ref{fig:fsm:task:alpha} and \ref{fig:fsm:task:beta} present $M_{\alpha_s}$ and $M_{\beta_s}$.
+Figure~\ref{ch04:fig:fsm:task:alpha} and \ref{ch04:fig:fsm:task:beta} present $M_{\alpha_s}$ and $M_{\beta_s}$.
 Both $M_{\alpha_s}$ and $M_{\beta_s}$ share state with the monitoring platform $P$.
@@
 \centering
-\subcaptionbox{\scriptsize{$M_{\alpha_s}$ task automata for $\alpha(s,\tau,t_c)$}\label{fig:fsm:task:alpha}}[0.45\textwidth]{
+\subcaptionbox{\scriptsize{$M_{\alpha_s}$ task automata for $\alpha(s,\tau,t_c)$}\label{ch04:fig:fsm:task:alpha}}[0.45\textwidth]{
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=7cm,thick,scale=0.6,every node/.style={scale=0.6},level distance=20mm]
@@
 }%
-\subcaptionbox{\scriptsize{$M_{\beta_s}$ task automata for $\beta(s,\tau)$}\label{fig:fsm:task:beta}}[0.45\textwidth]{
+\subcaptionbox{\scriptsize{$M_{\beta_s}$ task automata for $\beta(s,\tau)$}\label{ch04:fig:fsm:task:beta}}[0.45\textwidth]{
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=7cm,thick,scale=0.6,every node/.style={scale=0.6},level distance=20mm]
@@
 We use a separate task automaton for each service characteristic to verify the SLA of the service after $t_G$ time.
-Respectively, $M^{\alpha}_{\jtt{V}}$ and $M^{\beta}_{\jtt{V}}$ execute tasks $\jtt{verify}_\alpha$ and $\jtt{verify}_\beta$ (Figures~\ref{fig:fsm:task:verify:alpha} and \ref{fig:fsm:task:verify:beta}).
+Respectively, $M^{\alpha}_{\jtt{V}}$ and $M^{\beta}_{\jtt{V}}$ execute tasks $\jtt{verify}_\alpha$ and $\jtt{verify}_\beta$ (Figures~\ref{ch04:fig:fsm:task:verify:alpha} and \ref{ch04:fig:fsm:task:verify:beta}).
 $M^{\alpha}_{\jtt{V}}$ uses \bfjtt{await} to ensure the condition of the SLA.
@@
 \centering
-\subcaptionbox{\scriptsize{$M^{\alpha}_{\jtt{V}}$ to execute $\jtt{verify}_\alpha$}\label{fig:fsm:task:verify:alpha}}[0.45\textwidth]{
+\subcaptionbox{\scriptsize{$M^{\alpha}_{\jtt{V}}$ to execute $\jtt{verify}_\alpha$}\label{ch04:fig:fsm:task:verify:alpha}}[0.45\textwidth]{
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=6.5cm,thick,scale=0.6,every node/.style={scale=0.6},level distance=20mm]
@@
 }%
-\subcaptionbox{\scriptsize{$M^{\beta}_{\jtt{V}}$ to execute $\jtt{verify}_\beta$}\label{fig:fsm:task:verify:beta}}[0.45\textwidth]{
+\subcaptionbox{\scriptsize{$M^{\beta}_{\jtt{V}}$ to execute $\jtt{verify}_\beta$}\label{ch04:fig:fsm:task:verify:beta}}[0.45\textwidth]{
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=6cm,thick,scale=0.6,every node/.style={scale=0.6},level distance=20mm]
@@
 We define $M_{\jtt{A}_s}$ as the timed automata to execute the tasks of type \jtt{allocate} in $M_P$.
-We use the model  semantics presented in \cite{jaghoori2010time} p. 92 to design $M_{\jtt{A}_s}$. The resulting automata is presented in Figure~\ref{fig:fsm:alloc_s}.
+We use the model  semantics presented in \cite{jaghoori2010time} p. 92 to design $M_{\jtt{A}_s}$. The resulting automata is presented in Figure~\ref{ch04:fig:fsm:alloc_s}.
 
@@
 % \vspace{-20pt}
-\label{fig:fsm:alloc_s}
+\label{ch04:fig:fsm:alloc_s}
 \end{figure}
@@
 % 
-Similarly, we design $M_{\jtt{D}_s}$ to execute task type \jtt{deallocate} in Figure~\ref{fig:fsm:dealloc_s}.
+Similarly, we design $M_{\jtt{D}_s}$ to execute task type \jtt{deallocate} in Figure~\ref{ch04:fig:fsm:dealloc_s}.
 Therefore, we also have $M_{\jtt{D}}$ in $M_P$ as: $M_{\jtt{D}} = \; \parallel_s M_{\jtt{D}_s}$.
@@
 % \vspace{-20pt}
-\label{fig:fsm:dealloc_s}
+\label{ch04:fig:fsm:dealloc_s}
 \end{figure}
@@
 In $M_E$, in every monitoring window, there is a probability that some resources fail.
-For example, we present a particular instance of $M_E$ in Figure~\ref{fsm:env}.
+For example, we present a particular instance of $M_E$ in Figure~\ref{ch04:fsm:env}.
 In this environment, in every monitoring, an unspecified constant ($c$) number of resources fail.
@@
 \caption{An example behavior for $M_E$}
-\label{fsm:env}
+\label{ch04:fsm:env}
 % \vspace{-20pt}
@@
 \begin{thm}
-\label{thm:alpha}
+\label{ch04:thm:alpha}
 If the SLA for service $s$ on $\alpha(s,\tau,t_c)$ is violated, either:
@@
 \begin{thm}
-\label{thm:beta}
+\label{ch04:thm:beta}
 If the SLA for service $s$ on $\beta(s,\tau)$ is violated, either:
@@
 \begin{proof}
-Similar to the proof of Theorem~\ref{thm:alpha}.
+Similar to the proof of Theorem~\ref{ch04:thm:alpha}.
 \end{proof}
@@
 \begin{thm}
-\label{thm:gamma}
+\label{ch04:thm:gamma}
 If $\mathcal{S}_\gamma$ is \emph{schedulable} given input parameters $(\tau, t_i, t_G)$, then the SLA for both service characteristics $\alpha(s,\tau,t_c)$ and $\beta(s,\tau)$ is satisfied within $t_G$ time after a violation.
@@
 \begin{proof}
-When a violation of the SLA occurs in $\mathcal{S}_\gamma$, either $\mathcal{S}_\alpha$ or $\mathcal{S}_\beta$ (or both) start to evolve the service based on Theorems~\ref{thm:alpha} and \ref{thm:beta}.
+When a violation of the SLA occurs in $\mathcal{S}_\gamma$, either $\mathcal{S}_\alpha$ or $\mathcal{S}_\beta$ (or both) start to evolve the service based on Theorems~\ref{ch04:thm:alpha} and \ref{ch04:thm:beta}.
 Therefore, there exists at least one task of $\jtt{verify}_\alpha$ or $\jtt{verify}_\beta$ with a deadline $t_G$.

### 2016-05-04 p03-ch01-2015-monitoring/sec06-impl.tex
Remove all `vspace`
Replace `wrapfigure` with `figure`
Fix long captions in figures and tables
Fix ref/label for Chapter 04

@@
 \section{Evaluation of the monitoring model} % (fold)
-\label{sec:implementation}
+\label{ch04:sec:implementation}
 
@@
 
-Figure~\ref{fig:alpha} plots the service availability $\alpha(s,\tau,t_c)$ over time with the different monitoring windows.
+Figure~\ref{ch04:fig:alpha} plots the service availability $\alpha(s,\tau,t_c)$ over time with the different monitoring windows.
 The following summarizes the behavior:
@@
 % 
-\begin{wrapfigure}{r}{0.6\textwidth}
-\vspace{-40pt}
+\begin{figure}[h]
+% \begin{wrapfigure}{r}{0.6\textwidth}
+% \vspace{-40pt}
 \begin{center}
@@
 \end{center}
-\vspace{-20pt}
-\caption{{\scriptsize Evolving $\alpha(s,\tau,t_c)$ with different $\tau$}}
-\label{fig:alpha}
-\vspace{-20pt}
-\end{wrapfigure}
+% \vspace{-20pt}
+\caption{Evolving $\alpha(s,\tau,t_c)$ with different $\tau$}
+\label{ch04:fig:alpha}
+% \vspace{-20pt}
+% \end{wrapfigure}
+\end{figure}
 %
+
 Every monitoring measurement is performed in a monitoring window $\tau$.
@@
 
-If $\tau < t_i$, Theorem \ref{thm:alpha} does not hold because every task \jtt{allocate} in $M_{\jtt{A}}$ misses its deadline.
+If $\tau < t_i$, Theorem \ref{ch04:thm:alpha} does not hold because every task \jtt{allocate} in $M_{\jtt{A}}$ misses its deadline.
 Thus, it is essential that $\tau \geq t_i$.
@@
 % 
-% Assumption \ref{def:single:resource:type} says all the resources are of the same type and as such $t_i$ is the same for all resources.
+% Assumption \ref{ch04:def:single:resource:type} says all the resources are of the same type and as such $t_i$ is the same for all resources.
 In a real setting, different services may use different types of resources.

### 2016-05-04 p03-ch01-2015-monitoring/sec55-app.tex
Fix ref/label for Chapter 04

@@
 \section{Applications and Future Work} % (fold)
-\label{sec:application}
+\label{ch04:sec:application}
 
@@
 The focus of this research is not to enter this domain and yet build on top of it a model to be able to reason about ``distributed service properties'' such as service availability or budget compliance.
-In the following, we present a set of pseudo algorithms for the model in Section \ref{sec:modelling} that present how the monitoring can be generated based on the agreements that are specified
+In the following, we present a set of pseudo algorithms for the model in Section \ref{ch04:sec:modelling} that present how the monitoring can be generated based on the agreements that are specified
 \footnote{This is a work in progress in ENVISAGE project WP2}.
 
-In Algorithm \ref{alg:gen:monitor}, the monitoring actor method \jtt{check} is presented that collects the measurements and publishes them to the platform.
+In Algorithm \ref{ch04:alg:gen:monitor}, the monitoring actor method \jtt{check} is presented that collects the measurements and publishes them to the platform.
 
@@
 \caption{Monitor for service $s$}
-\label{alg:gen:monitor}
+\label{ch04:alg:gen:monitor}
 \begin{algorithmic}
@@
 When platform receives a measurement, it updates its state, computes the service properties and initiates an asynchronous message to process and react to them.
-This is presented in Algorithm \ref{alg:publish}.
+This is presented in Algorithm \ref{ch04:alg:publish}.
 
@@
 \caption{Process measurements}
-\label{alg:publish}
+\label{ch04:alg:publish}
 \begin{algorithmic}
@@
 As a consequence of publishing measurements, the platform react to them in case they do not match the expectations.
-The behavior is discussed in Theorem \ref{thm:sustain:conv}.
-Algorithm \ref{alg:process} presents the pseudo code for the behavior that makes changes to $M_\alpha$ and $M_\beta$.
+The behavior is discussed in Theorem \ref{ch04:thm:sustain:conv}.
+Algorithm \ref{ch04:alg:process} presents the pseudo code for the behavior that makes changes to $M_\alpha$ and $M_\beta$.
 
@@
 \caption{$M_\alpha$ and $M_\beta$}
-\label{alg:process}
+\label{ch04:alg:process}
 \begin{algorithmic}
@@
 \item We defined $\hat{t}$ to be the time a resources needs to initialized in the platform.
-Under Assumption \ref{def:resource:init:time}, all resources require same amount of time to initialize.
-If $\tau < \hat{t}$, Theorem \ref{def:service:availability} invalidates because the measurements are done in a time window that is not enough to be able to take the newly allocated resource into account for a service.
+Under Assumption \ref{ch04:def:resource:init:time}, all resources require same amount of time to initialize.
+If $\tau < \hat{t}$, Theorem \ref{ch04:def:service:availability} invalidates because the measurements are done in a time window that is not enough to be able to take the newly allocated resource into account for a service.
 Thus, it is essential that $\tau \geq \hat{t}$.
@@
 Widening the monitoring window causes to lose fine granularity of monitoring measurements to be able to react to them in proper time.
-\item For simplicity of reasoning, we considered Assumption \ref{def:single:resource:type};
+\item For simplicity of reasoning, we considered Assumption \ref{ch04:def:single:resource:type};
 i.e. all the resources provisioned by the platform are of the same type and as such $\hat{t}$ is the same.
 However, different services may fit into different types or resources and using different types of resource may fit better for budget compliance as they cost differently.
-In such setting, the monitoring window should be at least the maximum $\hat{t}$ of any resource type that is available in the platform: $\tau \geq \mathsf{max}(\hat{t}) \; \forall r \in p$; otherwise, Theorem \ref{thm:service:budget:conv} violates.
+In such setting, the monitoring window should be at least the maximum $\hat{t}$ of any resource type that is available in the platform: $\tau \geq \mathsf{max}(\hat{t}) \; \forall r \in p$; otherwise, Theorem \ref{ch04:thm:service:budget:conv} violates.
 \end{itemize}

### 2016-05-04 p03-ch01-2015-monitoring/sec99-appendix.tex
Fix ref/label for Chapter 04

@@
 \appendix
-\label{appendix}
+\label{ch04:appendix}
 
@@
 \subsection{Operational Semantics} % (fold)
-\label{sub:operational_semantics}
+\label{ch04:sub:operational_semantics}
 We use operational semantics to present the semantics of the runtime monitoring. 
@@
 \subsection{A Timed Automata for Budget Compliance}
-\label{sec:prop:budget}
+\label{ch04:sec:prop:budget}
 
-Similar to Section \ref{sec:service:avail:fsm}, we use a timed automata (see Definition \ref{def:timed:automata}) to model the behavior of service budget compliance $\beta(s,\tau)$ (see Definition \ref{def:budget:compliance}) represented in Figure \ref{fig:fsm:budgetcomp}.
+Similar to Section \ref{ch04:sec:service:avail:fsm}, we use a timed automata (see Definition \ref{ch04:def:timed:automata}) to model the behavior of service budget compliance $\beta(s,\tau)$ (see Definition \ref{ch04:def:budget:compliance}) represented in Figure \ref{ch04:fig:fsm:budgetcomp}.
 % 
@@
 \caption{A timed automata for Budget Compliance $\beta(s,\tau)$}
-\label{fig:fsm:budgetcomp}
+\label{ch04:fig:fsm:budgetcomp}
 \end{figure}
@@
 \section{Distributed Service Sustainability} % (fold)
-\label{sec:distributed_service_sustainability}
+\label{ch04:sec:distributed_service_sustainability}
 \begin{defn}[Service Sustainability $\gamma(s,\tau)$]
-\label{def:service:sustain}
+\label{ch04:def:service:sustain}
 We define the service sustainability as:
@@
 \begin{theorem}[Budget Compliance Convergence]
-\label{thm:budget:comp:conv}
-The automata $M_\beta$ on budget compliance $\beta(s,\tau)$ makes a transition to state $q_B$ in a finite amount of time under Assumptions \ref{def:single:resource:type} and \ref{def:resource:cost:constant} given that $\tau \geq \hat{t}$ and agreement $a$ for service $s$ is fixed and not changing through time.
+\label{ch04:thm:budget:comp:conv}
+The automata $M_\beta$ on budget compliance $\beta(s,\tau)$ makes a transition to state $q_B$ in a finite amount of time under Assumptions \ref{ch04:def:single:resource:type} and \ref{ch04:def:resource:cost:constant} given that $\tau \geq \hat{t}$ and agreement $a$ for service $s$ is fixed and not changing through time.
 \end{theorem}
@@
 \begin{proof}[Budget Compliance Convergence]
-Proof can be completed using same technique in Theorem \ref{thm:service:avail:conv}.
+Proof can be completed using same technique in Theorem \ref{ch04:thm:service:avail:conv}.
 \newline$\Box$
@@
 \subsection{A Timed Automata for Service Sustainability}
-\label{sec:sustain}
+\label{ch04:sec:sustain}
 
@@
 \begin{theorem}[Sustainable Service Convergence]
-\label{thm:sustain:conv}
+\label{ch04:thm:sustain:conv}
 The sustainability of service $s$ converges to:
@@
 At time $T$, if $g_{\overline{A}}$ holds, $M_\alpha$ makes a transition to $q_{\overline{A}}$ and starts to allocate a new resource $r$ for $s$.
-At a later time $T' \geq T + \tau$, $M_\alpha$ makes a transition to $q_A$ based Theorem \ref{thm:service:avail:conv}.
+At a later time $T' \geq T + \tau$, $M_\alpha$ makes a transition to $q_A$ based Theorem \ref{ch04:thm:service:avail:conv}.
 However, if $g_{\overline{B}}$ holds at time $T'$, $M_\beta$ makes a transition to $q_{\overline{B}}$ because the newly allocated resource $r$ makes budget compliance $\beta(s,\tau)$ violate.
-Based on Theorem \ref{thm:budget:comp:conv}, $M_\beta$ makes a transition to $q_B$ after deallocating at least one resource from $s$ at a later time $T'' \geq T' + \tau$.
+Based on Theorem \ref{ch04:thm:budget:comp:conv}, $M_\beta$ makes a transition to $q_B$ after deallocating at least one resource from $s$ at a later time $T'' \geq T' + \tau$.
 At time $T''$, the system recurs the behavior at time $T$ and thus $\gamma(s,\tau)$ for every subsequent $\tau$ cannot converge to $(1-\varepsilon)\times(1-\varepsilon_{\euro})$.
@@
 \section{Application of Actor-based Monitoring Model}
-\label{sec:app}
+\label{ch04:sec:app}
 
@@
 
-Theorem \ref{thm:sustain:conv} proved that $\gamma(s,\tau)$ converges to $(E(s,\tau)-\varepsilon)\times(E_B(s,\tau)-\varepsilon_{\euro})$ given certain conditions.
+Theorem \ref{ch04:thm:sustain:conv} proved that $\gamma(s,\tau)$ converges to $(E(s,\tau)-\varepsilon)\times(E_B(s,\tau)-\varepsilon_{\euro})$ given certain conditions.
 The values of $E(s,\tau)$ and $E_B(s,\tau)$ are bound to $1$.
@@
 On its own, this function is interesting.
-Figure \ref{fig:tolerance:func} presents the landscape of this function with $\varepsilon$ and $\varepsilon_{\euro}$ as its variables.
+Figure \ref{ch04:fig:tolerance:func} presents the landscape of this function with $\varepsilon$ and $\varepsilon_{\euro}$ as its variables.
 %
@@
 Additionally, choosing such model allows to design a business in which exclusively either service availability or budget compliance is crucial but not both.
-\item Apart from the corner values, the value of the function is \emph{confined} by the arcs that are projected on the ceiling of the box in Figure \ref{fig:tolerance:func}.
+\item Apart from the corner values, the value of the function is \emph{confined} by the arcs that are projected on the ceiling of the box in Figure \ref{ch04:fig:tolerance:func}.
 This means that the function values changes co-dependently based on the changes of $\varepsilon$ and $\varepsilon_{\euro}$; i.e.
@@
 The focus of this research is not enter this domain and yet build on top it a model to be able to reason about ``distributed service properties'' such as service availability or budget compliance. 
-In the following, we present a set of pseudo algorithms for the model in Section \ref{sec:modelling} that present how the monitoring can be generated based on the agreements that are specified. 
-In Algorithm \ref{alg:gen:monitor}, the monitoring actor method \jtt{check} is presented that collects the measurements and publishes them to the platform.
+In the following, we present a set of pseudo algorithms for the model in Section \ref{ch04:sec:modelling} that present how the monitoring can be generated based on the agreements that are specified. 
+In Algorithm \ref{ch04:alg:gen:monitor}, the monitoring actor method \jtt{check} is presented that collects the measurements and publishes them to the platform.
 
@@
 } 
-\label{fig:sustain:factor}
+\label{ch04:fig:sustain:factor}
 \end{figure}
@@
 \begin{itemize}
-  \item $\forall \mathcal{M}_k \in \mathcal{M}^+$ Thoerem \ref{thm:service:avail:conv} holds for $\mathcal{M}_k$.
-  \item $\forall \mathcal{M}_k \in \mathcal{M}^-$ Theorem \ref{thm:service:budget:conv} holds for $\mathcal{M}_k$.
-  \item $\forall \mathcal{M}_k \in \mathcal{M}^+ \; , \mathcal{M}_l \in \mathcal{M}^- \;, \mathcal{M}_m = \mathcal{M}_k \times \mathcal{M}_l$ Theorem \ref{thm:service:sust} holds for $\mathcal{M}_m$. 
+  \item $\forall \mathcal{M}_k \in \mathcal{M}^+$ Thoerem \ref{ch04:thm:service:avail:conv} holds for $\mathcal{M}_k$.
+  \item $\forall \mathcal{M}_k \in \mathcal{M}^-$ Theorem \ref{ch04:thm:service:budget:conv} holds for $\mathcal{M}_k$.
+  \item $\forall \mathcal{M}_k \in \mathcal{M}^+ \; , \mathcal{M}_l \in \mathcal{M}^- \;, \mathcal{M}_m = \mathcal{M}_k \times \mathcal{M}_l$ Theorem \ref{ch04:thm:service:sust} holds for $\mathcal{M}_m$.
 \end{itemize}
@@
 \subsection{A Timed Automata for Platform} 
-\label{sec:automata:platform}
+\label{ch04:sec:automata:platform}
 
@@
 In practice, a deadline for a message is part of the implementation details of any platform API for an IaaS provider.
-Figure \ref{fig:fsm:platform} presents a timed automata for platform $P$.
+Figure \ref{ch04:fig:fsm:platform} presents a timed automata for platform $P$.
 
@@
 Every reactive transition has a fixed deadline equal to $\tau$.
-A reactive transition is presented with a ``dotted'' line in Figure~\ref{fig:fsm} whereas a periodic transition is presented with a ``solid'' line.
+A reactive transition is presented with a ``dotted'' line in Figure~\ref{ch04:fig:fsm} whereas a periodic transition is presented with a ``solid'' line.
 Every transition is presented as $\frac{\textsf{condition}}{\textsf{action}}$.
@@
 % M_P
-\subcaptionbox{{\small{$M_P$: Timed Automaton for Platform $P$}}\label{fig:fsm:platform}}
+\subcaptionbox{{\small{$M_P$: Timed Automaton for Platform $P$}}\label{ch04:fig:fsm:platform}}
 {
@@
 % M_alpha
-\subcaptionbox{{\small{$M_\alpha$: Timed Automaton $\alpha(s,\tau,t_c)$}}\label{fig:fsm:serviceavail}}
+\subcaptionbox{{\small{$M_\alpha$: Timed Automaton $\alpha(s,\tau,t_c)$}}\label{ch04:fig:fsm:serviceavail}}
 {
@@
 % M_beta
-\subcaptionbox{{\small{$M_\beta$: Timed Automaton for $\beta(s,\tau)$}}\label{fig:fsm:budgetcompliance}}
+\subcaptionbox{{\small{$M_\beta$: Timed Automaton for $\beta(s,\tau)$}}\label{ch04:fig:fsm:budgetcompliance}}
 {
@@
 % M_gamma
-\subcaptionbox{{\small{$M_\gamma$: Timed Automaton for $\gamma(s,\tau)$}}\label{fig:fsm:sustain}}
+\subcaptionbox{{\small{$M_\gamma$: Timed Automaton for $\gamma(s,\tau)$}}\label{ch04:fig:fsm:sustain}}
 {
@@
 \caption{Timed automata for elements of the monitoring platform}
-\label{fig:fsm}
+\label{ch04:fig:fsm}
 \end{figure}
@@
 \subsection{A Timed Automata for Service Availability}
-\label{sec:service:avail:fsm}
+\label{ch04:sec:service:avail:fsm}
 
-We use timed automata~\cite{alur:1994:timedautomata} $M_\alpha$ to manage service availability (see Definition \ref{def:service:availability}).
+We use timed automata~\cite{alur:1994:timedautomata} $M_\alpha$ to manage service availability (see Definition \ref{ch04:def:service:availability}).
 $M_\alpha$ is designed to specifically address an \emph{under-capacity} service:
@@
 \begin{defn}[Service Availability Convergence]
-\label{thm:service:avail:conv}
+\label{ch04:thm:service:avail:conv}
 Given that: 
 \begin{itemize}
-  \item platform $P$ provides resources with no limits (Assumption~\ref{def:platform:cap}), and,
-  \item monitoring window $\tau$ is chosen such that it's larger that any resource initialization time ($\tau > \hat{t}$) (Assumption~\ref{def:resource:init:time}), 
+  \item platform $P$ provides resources with no limits (Assumption~\ref{ch04:def:platform:cap}), and,
+  \item monitoring window $\tau$ is chosen such that it's larger that any resource initialization time ($\tau > \hat{t}$) (Assumption~\ref{ch04:def:resource:init:time}), 
   \item the number of failed resources
@@
 \subsection{A Timed Automata for Service Budget Compliance}
-\label{sec:automata:budget}
+\label{ch04:sec:automata:budget}
 
-Analogously, we define a timed automata $M_\beta$ for service budget compliance (Definition~\ref{def:budget:compliance}).
-The automaton is shown in Figure~\ref{fig:fsm:budgetcompliance}.
+Analogously, we define a timed automata $M_\beta$ for service budget compliance (Definition~\ref{ch04:def:budget:compliance}).
+The automaton is shown in Figure~\ref{ch04:fig:fsm:budgetcompliance}.
 $M_\beta$ is specifically designed to address a service that is \emph{over-budget}.
@@
 \begin{defn}[Service Budget Compliance Convergence]
-\label{thm:service:budget:conv}
+\label{ch04:thm:service:budget:conv}
 The timed automata $M_\beta$ for service budget compliance $\beta(s,\tau)$ ensures that service $s$ is budget compliant in a finite amount of time.
@@
 \begin{proof}%[Service Budget Compliance Convergence]
-Similar to that of Theorem~\ref{thm:service:avail:conv}.
+Similar to that of Theorem~\ref{ch04:thm:service:avail:conv}.
 \end{proof}
@@
 \subsection{Service Sustainability}
-\label{sec:service:sustain}
+\label{ch04:sec:service:sustain}
 
 In the previous two sections, we modelled service availability $\alpha(s,\tau,t_c)$ and service budget compliance $\beta(s,\tau)$ using two timed automata.
-In isolation, the automata guarantee two important theorems: Theorem \ref{thm:service:avail:conv} and \ref{thm:service:budget:conv}.
+In isolation, the automata guarantee two important theorems: Theorem \ref{ch04:thm:service:avail:conv} and \ref{ch04:thm:service:budget:conv}.
 %  The automata two theorems on the two behave in the absence of the other limitation.

### 2016-05-04 p03-ch02-2014-jmseq/jmseq-elsa.tex
Change bib style to 'american' from the original template
Fix ref/label for Chapter 05

@@
 
- \section{Introduction} \label{sec:intro}
+ \section{Introduction} \label{ch05:sec:intro}
 Monitoring refers to the activity of tracking observable information from an
@@
 \caption{JMSeq Execution Architecture}
-\label{fig:arch-overview}
+\label{ch05:fig:arch-overview}
 \end{center}
@@
 %The program is verified against the specifications during runtime and fails as soon as a violation is detected. 
-%The verification is performed in a separate virtual machine to protect the original application from any modification (Figure \ref{fig:arch-overview}). 
-%In the following, we provide a  discussion of what distinguishes JMSeq from other related work (c.f. Section \ref{sec:relworks}) in verification:
+%The verification is performed in a separate virtual machine to protect the original application from any modification (Figure \ref{ch05:fig:arch-overview}). 
+%In the following, we provide a  discussion of what distinguishes JMSeq from other related work (c.f. Section \ref{ch05:sec:relworks}) in verification:
 The main distinguishing overall characteristic of JMSeq is
@@
    JMSeq uses Java Platform Debugger API (JPDA) \cite{JPDA_Home} to initiate a separate JVM and verify the target application using events from the target JVM.
-  Figure \ref{fig:arch-overview} displays the high-level execution architecture of JMSeq.
+  Figure \ref{ch05:fig:arch-overview} displays the high-level execution architecture of JMSeq.
    \item[Fail-fast approach.] JMSeq terminates execution as soon as a violation of a specification occurs.  This approach prevents unsafe behavior,
@@
 By annotating the  interfaces of a system any service object that implements the interface behavior inherits the annotations and can be verified by JMSeq.
-In this  respect, JMSeq also introduces integration with frameworks such as JUnit (cf. Section \ref{sec:junit}).
+In this  respect, JMSeq also introduces integration with frameworks such as JUnit (cf. Section \ref{ch05:sec:junit}).
 
@@
 The first three sections describe the usage of JMSeq.
-Section \ref{sec:spec} presents the language for specifying sequences of method calls, 
-and Section \ref{sec:impl_annot} discusses the types of annotations available in JMSeq.
-These two sections are connected via examples in Section \ref{sec:samples}.
-Afterwards, implementation of the JMSeq framework is explained in detail in Section \ref{sec:implementation}.
-The practical applicability of JMSeq is demonstrated by means of a case study from industry in Section \ref{sec:frh} and performance evaluation in Section \ref{sec:results}. 
-Finally, related work is discussed in Section \ref{sec:relworks}.
-Section \ref{sec:conclusion} concludes the paper. %discuss possible future work.
-
-\section{Method Call Sequence Specification} \label{sec:spec}
+Section \ref{ch05:sec:spec} presents the language for specifying sequences of method calls, 
+and Section \ref{ch05:sec:impl_annot} discusses the types of annotations available in JMSeq.
+These two sections are connected via examples in Section \ref{ch05:sec:samples}.
+Afterwards, implementation of the JMSeq framework is explained in detail in Section \ref{ch05:sec:implementation}.
+The practical applicability of JMSeq is demonstrated by means of a case study from industry in Section \ref{ch05:sec:frh} and performance evaluation in Section \ref{ch05:sec:results}. 
+Finally, related work is discussed in Section \ref{ch05:sec:relworks}.
+Section \ref{ch05:sec:conclusion} concludes the paper. %discuss possible future work.
+
+\section{Method Call Sequence Specification} \label{ch05:sec:spec}
 
@@
 sequences of messages among a small, fixed number of objects. In
-Figure \ref{fig:seq-spec}, we see two UML message sequence diagrams each
+Figure \ref{ch05:fig:seq-spec}, we see two UML message sequence diagrams each
 describing how, and in which order, the four objects interact with each other. In this
@@
 needs to distinguish between the following two cases (depicted in
-Figure \ref{fig:seq-spec}).
+Figure \ref{ch05:fig:seq-spec}).
 \begin{description}
@@
   \caption{Examples of Method Call Sequence Specification}
-  \label{fig:seq-spec}
+  \label{ch05:fig:seq-spec}
 \end{center}
@@
 sequences of method calls can be specified by means
-of the grammar in  Figure \ref{fig:spec_grammar}.
+of the grammar in  Figure \ref{ch05:fig:spec_grammar}.
 
@@
 \caption{Method Sequence Specification Grammar}
-\label{fig:spec_grammar}
+\label{ch05:fig:spec_grammar}
 \end{center}
@@
 For instance, the specification of the sequence in case $1$ of
-Figure \ref{fig:seq-spec} is given by:
+Figure \ref{ch05:fig:seq-spec} is given by:
 %\begin{flushleft}
@@
 % It is \emph{notable} that the meta-grammar provided in Figure
-% \ref{fig:spec_grammar} is a context-free grammar; 
+% \ref{ch05:fig:spec_grammar} is a context-free grammar; 
 It is notable that the actual specifications are regular expressions;
@@
 
-\section{Annotations for method call sequences} \label{sec:impl_annot}
+\section{Annotations for method call sequences} \label{ch05:sec:impl_annot}
 Java enables developers to attach extra metadata to their programs using annotations.
@@
 These metadata include the specification of the sequences of method calls
-based on the grammar discussed in Section \ref{sec:spec}. The metadata
+based on the grammar discussed in Section \ref{ch05:sec:spec}. The metadata
 is placed on classes or methods in a Java program. 
@@
 for such classes and notifies the annotation metadata loader (cf.
-Section \ref{sec:annotation_repo}) that the objects instantiated from the annotated
+Section \ref{ch05:sec:annotation_repo}) that the objects instantiated from the annotated
 class contain methods that specify a sequence of method executions. 
-The code is given in Listing \ref{lst:seq_obj}.
+The code is given in Listing \ref{ch05:lst:seq_obj}.
 
@@
 
-In Listing~\ref{lst:seq_obj}, {\small\texttt{@Retention(RetentionPolicy.RUNTIME)}} declares that this
+In Listing~\ref{ch05:lst:seq_obj}, {\small\texttt{@Retention(RetentionPolicy.RUNTIME)}} declares that this
 annotation is only applicable during runtime and may  not be used in other scenarios.
@@
 property declaring the sequence specification discussed in Section
-\ref{sec:spec}. Listing \ref{lst:seq_method} presents the declaration of
+\ref{ch05:sec:spec}. Listing \ref{ch05:lst:seq_method} presents the declaration of
 this annotation type.
@@
 {\small\texttt{value()}} holds the sequence specification (cf.
-Section \ref{sec:spec}). 
+Section \ref{ch05:sec:spec}). 
 Additionally, a sequenced method annotation allows verification of
@@
 receives information about the execution through a parameter of
-type \smalltt{Execution} (cf. Section \ref{sec:implementation}).
+type \smalltt{Execution} (cf. Section \ref{ch05:sec:implementation}).
 \smalltt{Execution} is the central data model that carries information
@@
 
-Section \ref{sec:samples} provides more discussion along with different
+Section \ref{ch05:sec:samples} provides more discussion along with different
 examples on using JMSeq.
 
-\section{JMSeq by example} \label{sec:samples}
+\section{JMSeq by example} \label{ch05:sec:samples}
 In this section, we go through different usage scenarios of JMSeq 
@@
 \subsection{Sequenced Execution Specification}
-In  Listing \ref{lst:annotated_spec}, we again go through the example
-introduced in Section \ref{sec:spec}.
+In  Listing \ref{ch05:lst:annotated_spec}, we again go through the example
+introduced in Section \ref{ch05:sec:spec}.
 % give two examples of
 % annotations of the class {\small\texttt{Main}} in Figure
-% \ref{fig:seq-spec}. 
+% \ref{ch05:fig:seq-spec}. 
 In this example, we annotate the method \texttt{main()} with the sequences of method calls describing the two scenarios
-given in Figure \ref{fig:seq-spec}. 
-The JMSeq specifications used here are discussed in Section \ref{sec:spec}.
+given in Figure \ref{ch05:fig:seq-spec}. 
+The JMSeq specifications used here are discussed in Section \ref{ch05:sec:spec}.
 The listing shows how to use them in a Java program.
@@
 execution of an application using the example given in Figure
-\ref{fig:seq-spec} Case 1.
+\ref{ch05:fig:seq-spec} Case 1.
 
@@
 totally ignored. We use the following annotation in Listing
-\ref{lst:case-1-exceptions} to express this situation.
+\ref{ch05:lst:case-1-exceptions} to express this situation.
 
@@
 JMSeq additionally supports integration with testing frameworks such as JUnit. 
-An example of such integration is discussed in Section \ref{sec:junit}.
-
-\section{The JMSeq framework} \label{sec:implementation}
+An example of such integration is discussed in Section \ref{ch05:sec:junit}.
+
+\section{The JMSeq framework} \label{ch05:sec:implementation}
 
@@
 Only the annotated methods will be monitored by JMSeq.
-The overall verification process in JMSeq is depicted in Figure \ref{fig:alg-verify}.
+The overall verification process in JMSeq is depicted in Figure \ref{ch05:fig:alg-verify}.
 
@@
 \caption{Overview of JMSeq's process to verify a program}
-\label{fig:alg-verify}
+\label{ch05:fig:alg-verify}
 \end{figure*}
@@
 Before verification starts, JMSeq inspects the classes in the execution class path searching for methods that are annotated for sequenced verification; during this process, it collects all of the available method call sequence metadata.
-This step (denoted as Load Annotated Metadata in Figure \ref{fig:alg-verify}) does not need the source code, but rather uses the annotations metadata available in the compiled code.
+This step (denoted as Load Annotated Metadata in Figure \ref{ch05:fig:alg-verify}) does not need the source code, but rather uses the annotations metadata available in the compiled code.
 
@@
   Then, at verification time, the instance of \smalltt{Execution} is transformed to an instance of \smalltt{CallExpression} to hold method signature details used by the state machine to make a transition.
-(Section \ref{sec:event_handling})
+(Section \ref{ch05:sec:event_handling})
   \item Using the metadata available from the annotations of the methods, then, the   next ``possible call expressions'' of the current state, corresponding to the possible next transitions of the state machine,  are built.
-(Section \ref{sec:annotation_repo}, \ref{sec:exec_verify})
+(Section \ref{ch05:sec:annotation_repo}, \ref{ch05:sec:exec_verify})
   \item A match making is done between the possible call expressions (state machine transitions) and the   current call expression as the candidate.  
@@
   When a failure occurs, JMSeq will execute the custom verification failure handler that is implemented either as part of
-  the component code by the programmer, or externally by the system tester. (Section \ref{sec:exec_verify})
+  the component code by the programmer, or externally by the system tester. (Section \ref{ch05:sec:exec_verify})
 \end{enumerate}
@@
 
-The overall architecture of JMSeq is given in Figure \ref{fig:arch-diag}. It basically
+The overall architecture of JMSeq is given in Figure \ref{ch05:fig:arch-diag}. It basically
 consists of three main modules: one for handling the events raised by the JVM executing the target program, another module for storing the annotation information, and a third
@@
   \caption{Software Architecture for Method Sequence Specification}
-  \label{fig:arch-diag}
+  \label{ch05:fig:arch-diag}
 \end{center}
@@
 % The current design of JMSeq is completely \textsl{general} and \textsl{modular};  as it
-% allows for replacing the grammar in Figure \ref{fig:spec_grammar} with other
+% allows for replacing the grammar in Figure \ref{ch05:fig:spec_grammar} with other
 % specification modules, based, for example, on temporal logics or extended
@@
 
-\subsubsection{Event Handling Module} \label{sec:event_handling}
+\subsubsection{Event Handling Module} \label{ch05:sec:event_handling}
 Java Debugger Interface (JDI) is the interface
@@
 
-\subsubsection{Annotation Metadata Module} \label{sec:annotation_repo}
+\subsubsection{Annotation Metadata Module} \label{ch05:sec:annotation_repo}
 
@@
 
-\subsubsection{Execution Verification Module} \label{sec:exec_verify}
+\subsubsection{Execution Verification Module} \label{ch05:sec:exec_verify}
 
@@
 At this point, JMSeq checks if an implementation of the
-\smalltt{Verification\-Failure\-Handler} interface is present (cf. Section \ref{sec:impl_annot}).
+\smalltt{Verification\-Failure\-Handler} interface is present (cf. Section \ref{ch05:sec:impl_annot}).
 If present, JMSeq tries to instantiate an instance of the verification failure handler and hands over the execution metadata to this handler. 
@@
 An event may also be sent due to occurrence of an exception. 
-In this case, the execution is verified based on the exception specifications expressed using the mechanism described in Section \ref{sec:impl_annot}. 
+In this case, the execution is verified based on the exception specifications expressed using the mechanism described in Section \ref{ch05:sec:impl_annot}. 
 If the exception is allowed and verified the execution continues; otherwise, it fails and control is delegated to verification failure handler as explained above.
@@
     \item A string which is in the format of the JMSeq specification grammar as
-    in Figure \ref{fig:spec_grammar}. This service is used the first time a
+    in Figure \ref{ch05:fig:spec_grammar}. This service is used the first time a
     sequential execution is detected to build the root of the future possible
@@
   in the annotation. As described by the grammar in Figure
-  \ref{fig:spec_grammar}, for each current call expression there can
+  \ref{ch05:fig:spec_grammar}, for each current call expression there can
   be several possible next call expressions that may appear as the next event, but for
@@
 
-\subsection{JUnit Support} \label{sec:junit}
+\subsection{JUnit Support} \label{ch05:sec:junit}
 To ease the process of verification, JMSeq provides a seamless integration to JUnit \cite{JUnit}. 
@@
 \smalltt{get()} in \smalltt{Basket}. We specify this scenario in Listing
-\ref{lst:apples}.
+\ref{ch05:lst:apples}.
 
@@
 the \smalltt{JUnit4Support} that is provided in the library. The code is
-presented in Listing \ref{lst:apples-junit}.
+presented in Listing \ref{ch05:lst:apples-junit}.
 
@@
 
-\section{The Fredhopper Access Server: A Case Study} \label{sec:frh}
+\section{The Fredhopper Access Server: A Case Study} \label{ch05:sec:frh}
 
@@
 % \vspace{-20pt}
-\caption{Fredhopper's Controller life cycle for data processing requests from customers}
-\label{fig:frh-dm}
+\caption[Life cycle for data processing requests from customers]{Fredhopper's Controller life cycle for data processing requests from customers}
+\label{ch05:fig:frh-dm}
 % \vspace{-5pt}
@@
 When the data processing is complete, the result is published to the customer environment and becomes also available through FAS services.
-Figure \ref{fig:frh-dm} illustrates this example scenario.
+Figure \ref{ch05:fig:frh-dm} illustrates this example scenario.
 
@@
 
-Figure \ref{fig:frh:cs} presents the measurements for each setting without using JMSeq and with using JMSeq for both rebuild and update triggers.
+Figure \ref{ch05:fig:frh:cs} presents the measurements for each setting without using JMSeq and with using JMSeq for both rebuild and update triggers.
 The measurements were based on an estimated number of 30300 triggers in the same period.
@@
 \caption[Comparison of measurements using JMSeq at Fredhopper]{Result of experiment measurements without and with using JMSeq on Fredhopper Client API directed with live environment data from customers.}
-\label{fig:frh:cs}
+\label{ch05:fig:frh:cs}
 \end{center}
@@
 
-The relevant part of the API of the Fredhopper Controller is presented in Listing \ref{lst:frh:datadispatcher}.
+The relevant part of the API of the Fredhopper Controller is presented in Listing \ref{ch05:lst:frh:datadispatcher}.
 Customer applications use the interface to create a data object reference (using the \smalltt{create(Object)} method) and then use the created data reference to start a data job (using the overloaded \smalltt{dispatch} method, which has an optional \smalltt{Data} parameter).
-Note that the dispatch method allows for two different uses by the customer, for instance in Listing~\ref{lst:frh:c1} the data job to be executed performs only a rebuild of the customer's product catalog and as such does not need to have a data reference.
+Note that the dispatch method allows for two different uses by the customer, for instance in Listing~\ref{ch05:lst:frh:c1} the data job to be executed performs only a rebuild of the customer's product catalog and as such does not need to have a data reference.
 \lstset{language=java}
@@
 \end{lstlisting}
-If the customer needs to update their catalog with new products, then first a data reference is created. This reference is used to initiate a data job, as described in Listing \ref{lst:frh:c2}.
+If the customer needs to update their catalog with new products, then first a data reference is created. This reference is used to initiate a data job, as described in Listing \ref{ch05:lst:frh:c2}.
 \lstset{language=java}
@@
 with JMSeq we can create a \smalltt{CustomerAppDriver} class which calls the \smalltt{CustomerApp} methods and monitors the subsequent
-method calls.  The expected usage of the API of the two scenarios above is described by the SequencedMethod annotations in Listing~\ref{lst:frh:c3}.
+method calls.  The expected usage of the API of the two scenarios above is described by the SequencedMethod annotations in Listing~\ref{ch05:lst:frh:c3}.
 If the \smalltt{CustomerApp} does not behave as expected, the verifier detects an error.
@@
 
-%Figure \ref{fig:frh:cs} shows how using JMSeq in an evaluation environment for the customers detects violations and
+%Figure \ref{ch05:fig:frh:cs} shows how using JMSeq in an evaluation environment for the customers detects violations and
 %improves the usage of Controller API and eventually affects the success rate of triggers for customers.
@@
 % \newpage
-\subsection{Discussion} \label{sec:discuss-jmop}
+\subsection{Discussion} \label{ch05:sec:discuss-jmop}
 
 We use JavaMOP \cite{chen_rosu_jmop,chen_rosu_mop,MOP} for the same case study from Fredhopper. 
-Listing \ref{lst:frh:jmop} shows a JavaMOP script to describe the same verification goal.
+Listing \ref{ch05:lst:frh:jmop} shows a JavaMOP script to describe the same verification goal.
 
@@
 
-\section{Performance Results} \label{sec:results}
+\section{Performance Results} \label{ch05:sec:results}
 
@@
 We use JMSeq to verify an example code for this scenario. 
-We use the code snippet displayed in Listing \ref{lst:jmseq-iter} that uses JMSeq annotations to express the correct usage. 
+We use the code snippet displayed in Listing \ref{ch05:lst:jmseq-iter} that uses JMSeq annotations to express the correct usage. 
 In the code, we use a variable \smalltt{count} to control the number of times \smalltt{hasNext()} is called before there is a call to \smalltt{next()}. 
@@
 
-Figure \ref{fig:jmseq-performance} displays how different cases compare to each other regarding the execution time. 
+Figure \ref{ch05:fig:jmseq-performance} displays how different cases compare to each other regarding the execution time. 
 As expected, the original program execution has the lowest execution time.
@@
  and ``original{\textbackslash}JMSeq'' denotes the measurement of the execution time for the program apart from times for JMSeq.}
-\label{fig:jmseq-performance}
+\label{ch05:fig:jmseq-performance}
 \end{center}
@@
 % \caption{Execution data model performance results with JMSeq}
-% \label{tab:results-handleevent}
+% \label{ch05:tab:results-handleevent}
 % \end{table}
@@
 % activity, and a computed time for a single JMSeq activity. Table
-% \ref{tab:results-handleevent} summarizes the performance results for
-% execution data model and Table \ref{tab:results-verify} summarizes the
+% \ref{ch05:tab:results-handleevent} summarizes the performance results for
+% execution data model and Table \ref{ch05:tab:results-verify} summarizes the
 % same facts for verification.
@@
 % \caption{Verification performance results with JMSeq}
-% \label{tab:results-verify}
+% \label{ch05:tab:results-verify}
 % \end{table}
@@
 % number of invocations that it needs to verify (Figure
-% \ref{fig:jmseq-performance}). In this example, we set
+% \ref{ch05:fig:jmseq-performance}). In this example, we set
 % up \smalltt{count} to increase in an ``exponential'' way, thus, JMSeq
@@
 
-\section{Related Work} \label{sec:relworks}
+\section{Related Work} \label{ch05:sec:relworks}
 
@@
 logic when there is a match or fail based on the event expression logic.
-In Section \ref{sec:discuss-jmop}, we discussed how JMSeq contrasts with JavaMOP and similar approaches.
-
-% In Listing \ref{lst:sample_mop} we show an example of JavaMOP code with ERE
-% logic for the two specifications given in Figure \ref{fig:seq-spec}.
+In Section \ref{ch05:sec:discuss-jmop}, we discussed how JMSeq contrasts with JavaMOP and similar approaches.
+
+% In Listing \ref{ch05:lst:sample_mop} we show an example of JavaMOP code with ERE
+% logic for the two specifications given in Figure \ref{ch05:fig:seq-spec}.
 
@@
 
-% The code snippet presented in Listing \ref{lst:sample_mop} is one of
+% The code snippet presented in Listing \ref{ch05:lst:sample_mop} is one of
 % several ways to express the problem described in Figure
-% \ref{fig:seq-spec}. JavaMOP allows the use of AspectJ language to
+% \ref{ch05:fig:seq-spec}. JavaMOP allows the use of AspectJ language to
 % define the scope of the program points to be verified. This may look
@@
 % the operational behavior using regular expressions. For instance,
-% in Section \ref{sec:spec}, we used
+% in Section \ref{ch05:sec:spec}, we used
 % %\begin{flushleft}
@@
 % %\end{flushleft}
-% to specify the same program in Figure \ref{fig:seq-spec}.
+% to specify the same program in Figure \ref{ch05:fig:seq-spec}.
 % Another major difference between JMSeq and JavaMOP is the use of
 % annotations versus instrumentation as already discussed in Section
-% \ref{sec:intro}.
+% \ref{ch05:sec:intro}.
 
@@
 
-\section{Conclusion and future work} \label{sec:conclusion}
+\section{Conclusion and future work} \label{ch05:sec:conclusion}
 

### 2016-05-04 abstract.tex
Add the problem statement to summary

@@
 \vspace*{-10mm}
+
+Object orientation provides principles for abstraction and encapsulation.
+Abstraction is accomplished by high-level concepts of interfaces, classes, and objects (class instances).
+Many object-oriented languages have followed the blocking and synchronous model of messaging. 
+One of the fundamental elements of the actor model is \emph{asynchronous} message passing.
+However, a message is not bound to any interface in the actor model.
+It has been an ongoing research to combine object-oriented programming with the actor model.
+Multicore and distributed computing raised a new challenge: combining object orientation, the actor model, and \emph{concurrency}.
+In a concurrent setting, coroutines enable interactions with collaborative pre-emption.
+Coroutines are not originally established in the object-oriented paradigm.
+In a concurrent setting with objects, multi-threading has been a common approach to provide an object-oriented concurrency model.
+Furthermore, interactions in both coroutines and multi-threading are blocking and synchronous.
+In contrast, the actor model relies on asynchronous communication.
+While a message is processed, an actor cannot be pre-empted or intentionally yield to allow other actors in the system to make progress.
+Integration of actor model and object orientation leads to the use of asynchronous method calls.
 

### 2016-05-04 p01-ch02-2013/paper.tex
Fix a cascaded paraph spacing issue

@@
 
-\parbf{How to send a message with a deadline?} 
+\paragraph*{How to send a message with a deadline?} 
 % Let us consider that object $o$ sends message $m$ to object $o'$ with deadline $d$:
@@
 
-\parbf{What happens when a deadline is missed?}
+\paragraph*{What happens when a deadline is missed?}
 Messages received by an actor generate processes.
@@
 
-\parbf{How to process the response of a message with a deadline?}
+\paragraph*{How to process the response of a message with a deadline?}
 In the above example of an asynchronous message, the \emph{future} result of processing the  message   
@@
 Given a future variable $f$, the programmer can query  the  availability of the  result by the construct
-
 $$
@@
 Here $(S,\sigma,\phi)\uparrow v$ indicates that $S$  raises an exception $v$:
-\begin{itemize}\itemsep5pt
+\begin{itemize}
  \item $ (x=f.\gett(),\sigma,\phi) \uparrow \sigma(f)$ where $\phi(\sigma(f)).\mathsf{aborted}=\jtt{true}$,
@@
 
-\parbf{Assignment statement} is used to assign a value to a variable:
+\paragraph*{Assignment statement} is used to assign a value to a variable:
 
@@
 
-\parbf{Returning a result} consists of setting the field 
+\paragraph*{Returning a result} consists of setting the field 
 \textsf{val} of the future of the process:
@@
 
-\parbf{Initialization of \jtt{timeout} in \gett  \space operation} 
+\paragraph*{Initialization of \jtt{timeout} in \gett  \space operation} 
 assigns to a  distinguished (local) variable \jtt{timeout} its initial \emph{absolute} value:
@@
 
-\parbf{The \gett \space operation} is used to assign the value of a future to a variable:
+\paragraph*{The \gett \space operation} is used to assign the value of a future to a variable:
 $$ \langle  (x=f.\gett(); S, \sigma), q, \phi\rangle \rightarrow \langle (S, \sigma [x \mapsto \phi(\sigma(f)).\mathsf{val}]), q, \phi\rangle $$
@@
 
-\parbf{Timeout} is operationally presented by the following transition:
+\paragraph*{Timeout} is operationally presented by the following transition:
 $$ \langle  (x=f.\gett(); S, \sigma), q, \phi\rangle \rightarrow \langle (S, \sigma), q, \phi\rangle $$
@@
 
-\parbf{The \textnormal{\textbf{\jtt{try-catch}}} block} semantics is presented by:
+\paragraph*{The \textnormal{\textbf{\jtt{try-catch}}} block} semantics is presented by:
 
@@
 
-\parbf{Exception handling.} 
+\paragraph*{Exception handling.} 
 We provide the operational semantics of exception handling in a general way in the following:
@@
 
-\parbf{Abnormal termination} of the active process is generated by an uncaught exception:
+\paragraph*{Abnormal termination} of the active process is generated by an uncaught exception:
 $$
@@
 
-\parbf{Normal termination} is presented by:
+\paragraph*{Normal termination} is presented by:
 $$
@@
 
-\parbf{Deadline missed.}
+\paragraph*{Deadline missed.}
 Let  $(S', \tau)$ be some  pending process  in $q$ such that $\tau(\textbf{\jtt{deadline}}) < \sigma(\jtt{time})$.
@@
 
-\parbf{Incoming message.}
+\paragraph*{Incoming message.}
 Let the active process $p$ belong to the actor $\tau(\jtt{this})$ (i.e., 
@@
 
-\parbf{Outgoing message.} 
+\paragraph*{Outgoing message.} 
 We model an outgoing message by:
@@
 where 
-\begin{itemize}\itemsep5pt
+\begin{itemize}
 \item $\phi'$ results from $\phi$ by extending its domain  with a new future object $o$
@@
 
-\parbf{Local computation step.} 
+\paragraph*{Local computation step.} 
 The interleaving of local computation steps of the individual actors is
@@
 
-\parbf{Communication.} 
+\paragraph*{Communication.} 
 Matching a  message sent by one actor with its  reception by  the specified callee
@@
 
-\parbf{Progress of Time.}
+\paragraph*{Progress of Time.}
 The following transition uniformly updates the local clocks (represented by the instance variable \jtt{time})
@@
 
-\parbf{Deploying actors onto JVM threads.}
+\paragraph*{Deploying actors onto JVM threads.}
 %As a major improvement to our previous work \cite{crisp-sac}, we changed the model of deployment of actors onto Java threads for simplicity, performance, and better reasoning.
@@
 
-%\parbf{Implementation of $(S, \tau)$ and $(S, \sigma)$.} 
+%\paragraph*{Implementation of $(S, \tau)$ and $(S, \sigma)$.} 
 %As discussed, when an actor receives an incoming message, a process is created for the messages.
@@
 
-\parbf{Implementation of processes with deadlines.}
+\paragraph*{Implementation of processes with deadlines.}
 A process itself is represented in the implementation by a data structure
@@
 
-\parbf{Fredhopper case study.}
+\paragraph*{Fredhopper case study.}
 As introduced in Section \ref{ch02:sec:frh}, we extract a closed-world simplified version from Fredhopper Controller.

### 2016-05-17 abstract.tex
Update abstract based on comments from Stijn

--- 2016-05-04 abstract.tex
@@
 
+Object orientation provides principles for abstraction and encapsulation.
+Abstraction is accomplished by high-level concepts of interfaces, classes, and objects.
+Many object-oriented languages have followed the blocking and synchronous model of messaging. 
+One of the fundamental elements of the actor model is \emph{asynchronous} message passing.
+Unlike object orientation, a message is not bound to any interface in the actor model.
+Ongoing research focus to combine object-oriented programming with the actor model and concurrency.
+In a concurrent setting with objects, multi-threading is a common approach to provide an object-oriented concurrency model.
+In contrast, a message in actor model cannot be pre-empted while being processed.
+Integration of actor model and object orientation leads to the use of asynchronous method calls.
+% Multicore and distributed computing raised a new challenge: combining object orientation, the actor model, and \emph{concurrency}.
+Coroutines can be used for collaborative pre-emption.
+Coroutines are not originally established in the object-oriented paradigm.
+Furthermore, interactions in both coroutines and multi-threading are blocking and synchronous.
+The main challenge is to generate production code from an actor-based language which supports asynchronous method calls and coroutine-style execution.
+
 This thesis contributes to the intersection of object orientation, actor model, and concurrency.
@@
 object-oriented languages. 
-We formalize a subset of Java and its concurrency API to 
-facilitate formal verification and reasoning about it.
-We create an abstract mapping from a concurrent-object modeling language, 
-ABS, to the programming semantics of concurrent Java. 
+We formalize a subset of Java and its concurrency API.
+We create an abstract mapping from a actor modeling language, ABS, to the programming semantics of concurrent Java. 
+The mapping ensures a correct translation from a model to its equivalent production-ready code.
 We provide the formal semantics of the mapping and runtime properties of 
 the concurrency layer including deadlines and scheduling policies.
-We provide an implementation of the ABS concurrency layer as a Java API library 
-and framework utilizing the latest language additions 
-in Java~8.
+We provide an implementation of the ABS concurrency layer as a Java API library and framework using Java~8 features.
+The implementation is generic and extensible as it can be used either as a standalone library or as part of a pluggable code generator.
 We design and implement a runtime monitoring framework, JMSeq, to verify the
-correct ordering of execution of methods through code annotations in JVM. 
+correct ordering of execution of methods in possibly interleaved threads through code annotations in JVM. 
 In addition, we design a large-scale monitoring system as a real-world 
 application; the monitoring system is built with ABS concurrent objects 
-and formal semantics that leverages schedulability 
-analysis to verify correctness of the monitors.
+and formal semantics that leverages schedulability analysis to verify correctness of the monitors.
 

### 2016-05-23 abstract.tex
polished abstract

@@
 Many object-oriented languages have followed the blocking and synchronous model of messaging. 
-One of the fundamental elements of the actor model is \emph{asynchronous} message passing.
-Unlike object orientation, a message is not bound to any interface in the actor model.
-Ongoing research focus to combine object-oriented programming with the actor model and concurrency.
+\emph{Asynchronous} message passing is one of the fundamental elements of the actor model.
+Unlike in object orientation, a message is not bound to any interface in the actor model.
+Ongoing research focuses on combining object-oriented programming with the actor model and concurrency.
 In a concurrent setting with objects, multi-threading is a common approach to provide an object-oriented concurrency model.
@@
 
-This thesis contributes to the intersection of object orientation, actor model, and concurrency.
-We choose Java as the main programming language and as one of the mainstream 
+This thesis contributes to the intersection of object orientation, the actor model, and concurrency.
+We choose Java as the main target programming language and as one of the mainstream 
 object-oriented languages. 
 We formalize a subset of Java and its concurrency API.
-We create an abstract mapping from a actor modeling language, ABS, to the programming semantics of concurrent Java. 
+We create an abstract mapping from an actor-based modeling language, ABS, to the programming semantics of concurrent Java. 
 The mapping ensures a correct translation from a model to its equivalent production-ready code.

### 2016-05-25 cv.tex
Updates for the review-ready version

@@
 
-\cventry{2011--Nu}{Senior Software Ontwikkelaar}{SDL Fredhopper}{Amsterdam}{}{Groep leider in cloud\newline}
+\cventry{mei 2016}{Lead Software Ontwikkelaar}{HERE}{Eindhoven}{}{CTO Office Applied Technologies\newline}
+
+\cventry{2011--apr. 2016}{Senior Software Ontwikkelaar}{SDL Fredhopper}{Amsterdam}{}{Groep leider in cloud\newline}
 

### 2016-05-25 p00-ch00-intro/intro.tex
Update abstract based on comments from Stijn
Updates for the review-ready version

@@
 
-In a concurrent setting with objects, multi-threading has been a common approach to provide an object-oriented concurrency model.
+In a concurrent setting with objects, multi-threading is a common approach to provide an object-oriented concurrency model.
 A thread holds a single stack of synchronous method calls.
@@
 orientation, actor model, and concurrency.
-We choose Java as the main programming language and as one of the mainstream 
+We choose Java as the main target programming language and as one of the mainstream 
 object-oriented languages. 

### 2016-06-07 abstract.tex
Updates for the review-ready version
Initial commit of dutch summary

@@
 {\usekomafont{chapter}Samenvatting}\label{sec:abstract-diff} \\
+%Object orientation provides principles for abstraction and encapsulation.
+Objectgeoriënteerd programmeren biedt ondersteuning voor abstractie en encapsulatie.
+%Abstraction is accomplished by high-level concepts of interfaces, classes, and objects.
+Abstractie wordt bereikt door de introductie van high-level concepten als interfaces,
+klassen en objecten.
+%Many object-oriented languages have followed the blocking and synchronous model of messaging. 
+Veel objectgeoriënteerde talen gebruiken blokkerende, synchrone communicatie voor het zenden van berichten (messages) tussen objecten.
+%\emph{Asynchronous} message passing is one of the fundamental elements of the actor model.
+\emph{Asynchrone} communicatie (message passing) vormt een fundamenteel element van het zogenaamde actor model.
+%Unlike in object orientation, a message is not bound to any interface in the actor model.
+Anders dan in objectoriëntatie is een message niet gebonden aan of beperkt tot een specifieke interface.
 
-\TODO{Translate to Dutch. Kindly by Stijn.}
+%Ongoing research focuses on combining object-oriented programming with the actor model and concurrency.
+Huidig onderzoek focust op het combineren van objectoriëntatie met het actor model en concurrency (meerdere taken kunnen tegelijk actief zijn).
+%In a concurrent setting with objects, multi-threading is a common approach to provide an object-oriented concurrency model.
+Voor het uitbreiden van objectoriëntatie naar een concurrent omgeving is multi-threading een veelgebruikte aanpak.
+%In contrast, a message in actor model cannot be pre-empted while being processed.
+??? TODO
+%Integration of actor model and object orientation leads to the use of asynchronous method calls.
+Integratie van objectoriëntatie met het actor model leidt tot het gebruik van asynchrone methode aanroepen.
+%Coroutines can be used for collaborative pre-emption.
+TODO
+%Coroutines are not originally established in the object-oriented paradigm.
+TODO
+%Furthermore, interactions in both coroutines and multi-threading are blocking and synchronous.
+TODO
+%The main challenge is to generate production code from an actor-based language which supports asynchronous method calls and coroutine-style execution.
+De belangrijkste uitdaging is de generatie van productie code uit een actor gebaseerde taal die ondersteuning biedt voor asynchrone methode aanroepen en een flexibele co-routine stijl van executie.
+
+%This thesis contributes to the intersection of object orientation, the actor model, and concurrency.
+Dit proefschrift draagt bij aan de kennis over de combinatie van objectoriëntatie, het actor model en concurrency.
+%We choose Java as the main target programming language and as one of the mainstream 
+%object-oriented languages. 
+We kiezen Java, als een volwassen en een van de meest gebruikte programmeertalen in productie omgevingen, als de target language voor code generatie vanuit een actor gebaseerde taal.
+%We formalize a subset of Java and its concurrency API.
+We formaliseren een deel van Java en zijn concurrency library (API).
+%We create an abstract mapping from an actor-based modeling language, ABS, to the programming semantics of concurrent Java. 
+Vervolgens creëren we een abstracte relatie (een functie) van de actor gebaseerde taal ABS naar concurrent Java.
+%The mapping ensures a correct translation from a model to its equivalent production-ready code.
+Deze abstracte functie garandeert een correcte vertaling van een abstract ABS model naar executeerbare productie code.
+%We provide the formal semantics of the mapping and runtime properties of 
+%the concurrency layer including deadlines and scheduling policies.
+We geven de formele semantiek van de functie en de eigenschappen van de concurrency
+laag tijdens executie, inclusief deadlines en scheduling policies.
+%We provide an implementation of the ABS concurrency layer as a Java API library and framework using Java~8 features.
+Daarna implementeren we de concurrency laag van ABS als een Java API library, met behulp van Java~8.
+%The implementation is generic and extensible as it can be used either as a standalone library or as part of a pluggable code generator.
+Deze implementatie is generiek en uitbreidbaar, en kan zowel gebruikt worden als
+standalone library, of als een inplugbare code generator.
+%We design and implement a runtime monitoring framework, JMSeq, to verify the
+%correct ordering of execution of methods in possibly interleaved threads through code annotations in JVM. 
+We ontwerpen en implementeren een monitoring framework genaamd JMSeq waarmee
+de correcte ordening van methode aanroepen, gespecificeerd met code annotaties, geverifieerd kan worden in een multi-threaded omgeving.
+%In addition, we design a large-scale monitoring system as a real-world 
+%application; the monitoring system is built with ABS concurrent objects 
+%and formal semantics that leverages schedulability analysis to verify correctness of the monitors.
+Tevens ontwerpen we een grootschalig real-world monitoring systeem met
+ABS concurrent objecten en een formele semantiek. We gebruiken state-of-the-art technologie voor schedulability analyse om de correctheid van de monitors te verifiëren.

### 2016-06-13 abstract.tex
abstract: polished 2nd paragraph on coroutines and translated it

@@
 Voor het uitbreiden van objectoriëntatie naar een concurrent omgeving is multi-threading een veelgebruikte aanpak.
-%In contrast, a message in actor model cannot be pre-empted while being processed.
-??? TODO
 %Integration of actor model and object orientation leads to the use of asynchronous method calls.
 Integratie van objectoriëntatie met het actor model leidt tot het gebruik van asynchrone methode aanroepen.
-%Coroutines can be used for collaborative pre-emption.
-TODO
+%However, neither of object-orientation and actor model support explicit control of pre-emption for messages or asynchronous method calls.
+Echter, nog objectoriëntatie, nog het actor model biedt ondersteuning voor expliciete controle voor het tijdelijk onderbreken (pre-emption) van de verwerking van berichten of asynchrone methode aanroepen.
+%In contrast, a message in actor model cannot be pre-empted while being processed.
+%REMOVED THIS SENTENCE
+%In comparison to this rigid ``run to completion''-execution style, coroutines are more flexible: they allow multiple entry points and can be used for collaborative pre-emption between tasks by giving up control with explicit yield or suspend instructions.
+In vergelijking met deze starre ``run to completion''-executie stijl van programma's zijn coroutines flexibeler: er zijn meerdere ``entry points'' voor de executie, en taken coöpereren voor controle over de processor, door tijdelijk controle op te geven met expliciete yield of suspend instructies.
 %Coroutines are not originally established in the object-oriented paradigm.
-TODO
+Coroutines zijn echter niet standaard in objectoriëntatie,
 %Furthermore, interactions in both coroutines and multi-threading are blocking and synchronous.
-TODO
+en interacties met zowel coroutines als multi-threading zijn traditioneel synchroon en blokkerend.
 %The main challenge is to generate production code from an actor-based language which supports asynchronous method calls and coroutine-style execution.

### 2016-06-13 abstract.tex
abstract: applied polishments to actual (non-commented) english version

@@
 Unlike in object orientation, a message is not bound to any interface in the actor model.
+
 Ongoing research focuses on combining object-oriented programming with the actor model and concurrency.
 In a concurrent setting with objects, multi-threading is a common approach to provide an object-oriented concurrency model.
-In contrast, a message in actor model cannot be pre-empted while being processed.
 Integration of actor model and object orientation leads to the use of asynchronous method calls.
-% Multicore and distributed computing raised a new challenge: combining object orientation, the actor model, and \emph{concurrency}.
-Coroutines can be used for collaborative pre-emption.
+However, neither of object-orientation and actor model support explicit control of pre-emption for messages or asynchronous method calls.
+%In contrast, a message in actor model cannot be pre-empted while being processed.
+In comparison to this rigid ``run to completion''-execution style, coroutines are more flexible: they allow multiple entry points and can be used for collaborative pre-emption between tasks by giving up control with explicit yield or suspend instructions.
 Coroutines are not originally established in the object-oriented paradigm.

### 2016-06-14 abstract.tex
Use LaTeX accent characters for Dutch abstract

@@
 {\usekomafont{chapter}Samenvatting}\label{sec:abstract-diff} \\
+
 %Object orientation provides principles for abstraction and encapsulation.
-Objectgeoriënteerd programmeren biedt ondersteuning voor abstractie en encapsulatie.
+Objectgeori\"{e}nteerd programmeren biedt ondersteuning voor abstractie en encapsulatie.
 %Abstraction is accomplished by high-level concepts of interfaces, classes, and objects.
-Abstractie wordt bereikt door de introductie van high-level concepten als interfaces,
-klassen en objecten.
+Abstractie wordt bereikt door de introductie van high-level concepten als interfaces, klassen en objecten.
 %Many object-oriented languages have followed the blocking and synchronous model of messaging. 
-Veel objectgeoriënteerde talen gebruiken blokkerende, synchrone communicatie voor het zenden van berichten (messages) tussen objecten.
+Veel objectgeori\"{e}nteerde talen gebruiken blokkerende, synchrone communicatie voor het zenden van berichten (messages) tussen objecten.
 %\emph{Asynchronous} message passing is one of the fundamental elements of the actor model.
@@
 %Unlike in object orientation, a message is not bound to any interface in the actor model.
-Anders dan in objectoriëntatie is een message niet gebonden aan of beperkt tot een specifieke interface.
+Anders dan in objectori\"{e}ntatie is een message niet gebonden aan of beperkt tot een specifieke interface.
 
 %Ongoing research focuses on combining object-oriented programming with the actor model and concurrency.
-Huidig onderzoek focust op het combineren van objectoriëntatie met het actor model en concurrency (meerdere taken kunnen tegelijk actief zijn).
+Huidig onderzoek focust op het combineren van objectori\"{e}ntatie met het actor model en concurrency (meerdere taken kunnen tegelijk actief zijn).
 %In a concurrent setting with objects, multi-threading is a common approach to provide an object-oriented concurrency model.
-Voor het uitbreiden van objectoriëntatie naar een concurrent omgeving is multi-threading een veelgebruikte aanpak.
+Voor het uitbreiden van objectori\"{e}ntatie naar een concurrent omgeving is multi-threading een veelgebruikte aanpak.
 %Integration of actor model and object orientation leads to the use of asynchronous method calls.
-Integratie van objectoriëntatie met het actor model leidt tot het gebruik van asynchrone methode aanroepen.
+Integratie van objectori\"{e}ntatie met het actor model leidt tot het gebruik van asynchrone methode aanroepen.
 %However, neither of object-orientation and actor model support explicit control of pre-emption for messages or asynchronous method calls.
-Echter, nog objectoriëntatie, nog het actor model biedt ondersteuning voor expliciete controle voor het tijdelijk onderbreken (pre-emption) van de verwerking van berichten of asynchrone methode aanroepen.
+Echter, nog objectori\"{e}ntatie, nog het actor model biedt ondersteuning voor expliciete controle voor het tijdelijk onderbreken (pre-emption) van de verwerking van berichten of asynchrone methode aanroepen.
 %In contrast, a message in actor model cannot be pre-empted while being processed.
@@
 %In comparison to this rigid ``run to completion''-execution style, coroutines are more flexible: they allow multiple entry points and can be used for collaborative pre-emption between tasks by giving up control with explicit yield or suspend instructions.
-In vergelijking met deze starre ``run to completion''-executie stijl van programma's zijn coroutines flexibeler: er zijn meerdere ``entry points'' voor de executie, en taken coöpereren voor controle over de processor, door tijdelijk controle op te geven met expliciete yield of suspend instructies.
+In vergelijking met deze starre ``run to completion''-executie stijl van programma's zijn coroutines flexibeler: er zijn meerdere ``entry points'' voor de executie, en taken co\"{e}pereren voor controle over de processor, door tijdelijk controle op te geven met expliciete yield of suspend instructies.
 %Coroutines are not originally established in the object-oriented paradigm.
-Coroutines zijn echter niet standaard in objectoriëntatie,
+Coroutines zijn echter niet standaard in objectori\"{e}ntatie,
 %Furthermore, interactions in both coroutines and multi-threading are blocking and synchronous.
@@
 %The main challenge is to generate production code from an actor-based language which supports asynchronous method calls and coroutine-style execution.
-De belangrijkste uitdaging is de generatie van productie code uit een actor gebaseerde taal die ondersteuning biedt voor asynchrone methode aanroepen en een flexibele co-routine stijl van executie.
+De belangrijkste uitdaging is de generatie van productie code uit een actor gebaseerde taal die ondersteuning biedt voor asynchrone methode aanroepen en een flexibele coroutine stijl van executie.
 
@@
 Tevens ontwerpen we een grootschalig real-world monitoring systeem met
-ABS concurrent objecten en een formele semantiek. We gebruiken state-of-the-art technologie voor schedulability analyse om de correctheid van de monitors te verifiëren.
+ABS concurrent objecten en een formele semantiek. We gebruiken state-of-the-art technologie voor schedulability analyse om de correctheid van de monitors te verifi\"{e}ren.

### 2016-08-18 p01-ch01-2012/paper.tex
Apply formatting and typography fixes
- Fix the mis-referenced listings and figures
- Fix typos

Signed-off-by: Behrooz Nobakht <nobeh5@gmail.com>

@@
 % 
-\chapterauthor{Behrooz Nobakht, Frank S. de Boer, Mohammad Mahdi Jaghoori,  Rudolph Schlatte}
+\chapterauthor{Behrooz Nobakht, Frank S. de Boer, Mohammad Mahdi Jaghoori, Rudolf Schlatte}
 
@@
 
-\begin{lstlisting}[float=t, label=lst:exres, caption=Exclusive Resource in Creol %, multicols=2
-]
+\begin{lstlisting}[float=t, label=ch01:lst:exres, caption=Exclusive Resource in Creol]
 interface Resource begin

### 2016-08-18 p01-ch02-2013/paper.tex
Apply formatting and typography fixes
- Fix the mis-referenced listings and figures
- Fix typos

Signed-off-by: Behrooz Nobakht <nobeh5@gmail.com>

@@
 % \vspace{-20pt}
-\begin{lstlisting}[float=h, label=lst:try:catch, caption=Using try-catch for processing future values]
+\begin{lstlisting}[float=h, label=ch02:lst:try:catch, caption=Using try-catch for processing future values]
 try {
@@
 % \vspace{-25pt}
-\begin{lstlisting}[label=lst:frh:polling, caption=With polling]
+\begin{lstlisting}[label=ch02:lst:frh:polling, caption=With polling]
 class DataProcessor begin
@@
 % \vspace{-25pt}
-\begin{lstlisting}[label=lst:frh:deadlines, caption=With deadlines]
+\begin{lstlisting}[label=ch02:lst:frh:deadlines, caption=With deadlines]
 class DataProcessor begin
@@
 % \vspace{-30pt}
-\begin{lstlisting}[label=lst:frh:dm, caption=Data Processor class]
+\begin{lstlisting}[label=ch02:lst:frh:dm, caption=Data Processor class]
 class DataProcessor begin
@@
 % \vspace{-30pt}
-\begin{lstlisting}[label=lst:frh:sched, caption=Custom scheduler]
+\begin{lstlisting}[label=ch02:lst:frh:sched, caption=Custom scheduler]
 class DataScheduler extends DefaultSchedulingManager {
@@
 
-\begin{table}[h]
+\begin{table}[t]
+\centering
 % \begin{wraptable}{l}{0.25\textwidth}
 % \vspace{-15pt}
-\begin{tabular}{c c}\thickhline
-$s_1$ & $s_2$ \\ \thickhline
-$88.71\%$ & $94.57\%$ \\ \hline
+\begin{tabular}{c c}
+\toprule
+$s_1$ & $s_2$ \\ 
+\toprule
+$88.71\%$ & $94.57\%$ \\ 
+\bottomrule
 \end{tabular}

### 2016-08-18 p02-ch01-2014/paper.tex
Apply formatting and typography fixes
- Fix the mis-referenced listings and figures
- Fix typos

Signed-off-by: Behrooz Nobakht <nobeh5@gmail.com>

@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Ping as an interface,label=lst:ping:interface]
+\begin{lstlisting}[caption=Ping as an interface,label=ch03:lst:ping:interface]
 public interface IPing {
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Pong as an interface,label=lst:pong:interface]
+\begin{lstlisting}[caption=Pong as an interface,label=ch03:lst:pong:interface]
 public interface IPong {
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Ping actor in Akka,label=lst:ping:akka]
+\begin{lstlisting}[caption=Ping actor in Akka,label=ch03:lst:ping:akka]
 public class Ping(ActorRef pong)
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Pong class in Akka,label=lst:pong:akka]
+\begin{lstlisting}[caption=Pong class in Akka,label=ch03:lst:pong:akka]
 public class Pong 
@@
 % \vspace{-20pt}
-\begin{lstlisting}[caption=main in Akka,label=lst:main:akka]
+\begin{lstlisting}[caption=main in Akka,label=ch03:lst:main:akka]
 ActorSystem s = ActorSystem.create();
@@
 % \vspace{-20pt}
-\begin{lstlisting}[caption=main in ABS,label=lst:main:ABS]
+\begin{lstlisting}[caption=main in ABS,label=ch03:lst:main:ABS]
 ABSIPong pong;
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Ping in ABS,label=lst:ping:abs]
+\begin{lstlisting}[caption=Ping in ABS,label=ch03:lst:ping:abs]
 interface ABSIPing {
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Pong in ABS,label=lst:pong:abs]
+\begin{lstlisting}[caption=Pong in ABS,label=ch03:lst:pong:abs]
 interface ABSIPong {
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Actor interface (1) ,label=lst:j8:run]
+\begin{lstlisting}[caption=Actor interface (1) ,label=ch03:lst:j8:run]
 public interface Actor {
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Actor interface (2),label=lst:j8:enq]
+\begin{lstlisting}[caption=Actor interface (2),label=ch03:lst:j8:enq]
 
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[mathescape,caption=Async messages with futures,label=lst:j8:call0]
+\begin{lstlisting}[mathescape,caption=Async messages with futures,label=ch03:lst:j8:call0]
 final T$_1$ u$_1$ = e$_1$;
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[mathescape,caption=Async messages w/o futures,label=lst:j8:call1]
+\begin{lstlisting}[mathescape,caption=Async messages w/o futures,label=ch03:lst:j8:call1]
 final T$_1$ u$_1$ = e$_1$;
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[mathescape, caption=Ping as an Actor,label=lst:ping]
+\begin{lstlisting}[mathescape, caption=Ping as an Actor,label=ch03:lst:ping]
 public class Ping(IPong pong) implements IPing, Actor {
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[mathescape, caption=Pong as an Actor,label=lst:pong]
+\begin{lstlisting}[mathescape, caption=Pong as an Actor,label=ch03:lst:pong]
 public class Pong implements IPong, Actor {
@@
 % \end{lstlisting}
-\begin{lstlisting}[mathescape,caption=main in ABS API,label=lst:main:api]
+\begin{lstlisting}[mathescape,caption=main in ABS API,label=ch03:lst:main:api]
 IPong pong = new Pong();
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[mathescape,caption=Java~8 await implementation,label=lst:j8:await]
+\begin{lstlisting}[mathescape,caption=Java~8 await implementation,label=ch03:lst:j8:await]
 void await(final Boolean guard, 
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Remote ping actor main,label=lst:ping:remote]
+\begin{lstlisting}[caption=Remote ping actor main,label=ch03:lst:ping:remote]
 Properties p =  new Properties();
@@
 \begin{minipage}[t]{0.48\textwidth}
-\begin{lstlisting}[caption=Remote pong actor main,label=lst:pong:remote]
+\begin{lstlisting}[caption=Remote pong actor main,label=ch03:lst:pong:remote]
 Properties p = new Properties();

### 2016-08-18 p03-ch01-2015-monitoring/sec045-defs.tex
Apply formatting and typography fixes
- Fix the mis-referenced listings and figures
- Fix typos

Signed-off-by: Behrooz Nobakht <nobeh5@gmail.com>

@@
 
-\lstset{language=java,aboveskip=-20pt,belowskip=-20pt}
+\lstset{language=java}
 \begin{figure}[h]
@@
 \begin{center}
-\begin{lstlisting}[mathescape,caption=Platform API,label=lst:platform]
+\begin{lstlisting}[mathescape,caption=Platform API,label=ch04:lst:platform]
 interface Platform {

### 2016-08-18 p03-ch02-2014-jmseq/jmseq-elsa.tex
Apply formatting and typography fixes
- Fix the mis-referenced listings and figures
- Fix typos

Signed-off-by: Behrooz Nobakht <nobeh5@gmail.com>

@@
 \lstset{language=Java}
-\begin{lstlisting}[label=lst:seq_obj, caption=SequencedObject Annotation Declaration]
+\begin{lstlisting}[label=ch05:lst:seq_obj, caption=SequencedObject Annotation Declaration]
 @Retention(RetentionPolicy.RUNTIME)
@@
 \lstset{language=Java}
-\begin{lstlisting}[label=lst:seq_method, caption=SequencedMethod Annotation Declaration]
+\begin{lstlisting}[label=ch05:lst:seq_method, caption=SequencedMethod Annotation Declaration]
 @Retention(RetentionPolicy.RUNTIME)
@@
 \lstset{language=Java}
-\begin{lstlisting}[label=lst:annotated_spec, caption=Sample annotated specification]
+\begin{lstlisting}[label=ch05:lst:annotated_spec, caption=Sample annotated specification]
 
@@
 \lstset{language=Java}
-\begin{lstlisting}[label=lst:case-1-exceptions, caption=Exception
+\begin{lstlisting}[label=ch05:lst:case-1-exceptions, caption=Exception
 Specification for Case 1]
@@
 \lstset{language=Java}
-\begin{lstlisting}[label=lst:apples, caption=Apples JMSeq Specification]
+\begin{lstlisting}[label=ch05:lst:apples, caption=Apples JMSeq Specification]
 @SequencedObject
@@
 \lstset{language=Java}
-\begin{lstlisting}[label=lst:apples-junit, caption=Apples JMSeq JUnit
+\begin{lstlisting}[label=ch05:lst:apples-junit, caption=Apples JMSeq JUnit
 Test Unit]
@@
 \lstset{language=java}
-\begin{lstlisting}[label=lst:frh:datadispatcher, caption=Controller Data Dispatcher Interface API]
+\begin{lstlisting}[label=ch05:lst:frh:datadispatcher, caption=Controller Data Dispatcher Interface API]
 interface DataDispatcher {
@@
 \lstset{language=java}
-\begin{lstlisting}[label=lst:frh:c1, caption=A use case with no data requirement]
+\begin{lstlisting}[label=ch05:lst:frh:c1, caption=A use case with no data requirement]
 class CustomerApp {
@@
 \lstset{language=java}
-\begin{lstlisting}[label=lst:frh:c2, caption=A use case with data requirement]
+\begin{lstlisting}[label=ch05:lst:frh:c2, caption=A use case with data requirement]
 class CustomerApp {
@@
 \lstset{language=java}
-\begin{lstlisting}[float=t,label=lst:frh:c3, caption=How to verify Fredhopper API usage]
+\begin{lstlisting}[float=t,label=ch05:lst:frh:c3, caption=How to verify Fredhopper API usage]
 @SequencedObject
@@
 \lstset{language=[AspectJ]Java, morekeywords={suffix, event,ere, CFG, @fail, @match, __RESET}} 
-\begin{lstlisting}[label=lst:frh:jmop, caption=Fredhopper case study with JavaMOP]
+\begin{lstlisting}[label=ch05:lst:frh:jmop, caption=Fredhopper case study with JavaMOP]
 UpdateCatalog(Catalog catalog) {
@@
 \lstset{language=Java}
-\begin{lstlisting}[label=lst:jmseq-iter, caption=Iterator usage with
+\begin{lstlisting}[label=ch05:lst:jmseq-iter, caption=Iterator usage with
 JMSeq Annotations]
@@
 
-% \lstset{language=[AspectJ]Java, morekeywords={suffix, event,ere, CFG, @fail, @match, __RESET}} \begin{lstlisting}[label=lst:sample_mop, caption=Sample JavaMOP Specification using CFG logic] // Case 1
+% \lstset{language=[AspectJ]Java, morekeywords={suffix, event,ere, CFG, @fail, @match, __RESET}} \begin{lstlisting}[label=ch05:lst:sample_mop, caption=Sample JavaMOP Specification using CFG logic] // Case 1
 % SampleCase1(Main m) {

### 2016-08-18 abstract.tex
Switch to 'biber' backend for biblio
Fix multiple issues in the bibliography file
and special characters for the author names
from different locales.

Biber supports using 'IDS' to alias multiple
keys for the same citation.

Signed-off-by: Behrooz Nobakht <nobeh5@gmail.com>

@@
 %This thesis contributes to the intersection of object orientation, the actor model, and concurrency.
-Dit proefschrift draagt bij aan de kennis over de combinatie van objectoriëntatie, het actor model en concurrency.
+Dit proefschrift draagt bij aan de kennis over de combinatie van objectori\"{e}ntatie, het actor model en concurrency.
 %We choose Java as the main target programming language and as one of the mainstream 
@@
 %We create an abstract mapping from an actor-based modeling language, ABS, to the programming semantics of concurrent Java. 
-Vervolgens creëren we een abstracte relatie (een functie) van de actor gebaseerde taal ABS naar concurrent Java.
+Vervolgens cre\"{e}ren we een abstracte relatie (een functie) van de actor gebaseerde taal ABS naar concurrent Java.
 %The mapping ensures a correct translation from a model to its equivalent production-ready code.

### 2016-08-18 acknowledgement.tex
Switch to 'biber' backend for biblio
Fix multiple issues in the bibliography file
and special characters for the author names
from different locales.

Biber supports using 'IDS' to alias multiple
keys for the same citation.

Signed-off-by: Behrooz Nobakht <nobeh5@gmail.com>

%
\pdfbookmark[0]{Acknowledgement}{Acknowledgement}
\chapter*{Acknowledgement}
\label{sec:acknowledgement}
\vspace*{-10mm}

There will be acknowledgments.

### 2016-08-18 build.sh

@@ -3,13 +3,13 @@
 # Commands
 CMD_LATEX="latex -interaction=batchmode -halt-on-error -shell-escape -enable-write18"
 CMD_PDFLATEX="pdflatex -interaction=batchmode -halt-on-error -shell-escape -enable-write18"
-CMD_BIB="bibtex"
+CMD_BIB="biber"
 
 # Cleans LaTeX files
 function clean {
    if [[ -f ".gitignore" ]]; then
       for f in $(cat .gitignore); do
-         [[ -f $f && $f != ".todo.md" ]] && rm $f
+         [[ -f $f && $f != ".todo.md" ]] && rm -rf "$f" && echo "Removed $f"
       done 
    fi
 }

### 2016-08-19 build.sh
Add pdfTex info into the document
Signed-off-by: Behrooz Nobakht <nobeh5@gmail.com>

@@ -7,11 +7,15 @@ CMD_BIB="biber"
 
 # Cleans LaTeX files
 function clean {
+   echo "Cleaning ... "
+   echo -n "      "
    if [[ -f ".gitignore" ]]; then
       for f in $(cat .gitignore); do
-         [[ -f $f && $f != ".todo.md" ]] && rm -rf "$f" && echo "Removed $f"
+         [[ -f $f && $f != ".todo.md" ]] && rm -rf "$f" && echo -n "$f "
       done 
    fi
+   echo ""
+   echo ""
 }
 
 # Build the LaTeX documents

### 2016-10-04 acknowledgement.tex
Update non-scientific parts
Trying to get ready for publication

@@
 
-There will be acknowledgments.
+To mom, who sacrificed all she could for us and left us very early.
+To dad, who never loses hope in us.
+To Shima, whose love and patience are my only sources of strength.
+To Maryam and Ali, together with who, we never stop making our parents proud.
+
+This thesis like many others has been a journey.
+A journey that passed through the lives of family, friends, and colleagues.
+This is to extend a warm thankful message to all; especially
+Hossein Rahmani, Ramin Etemadi, Mahdi Jaghoori, Mohsen Davarinejad,
+Vlad Serbanescu, Nikolaos Bezirg, Keyvan Azadbakht.
+And, thank you Mohsen Davarinejad and Hossein Rezazadeh Mehrizi to honor 
+me for the defense session.
+It was a pleasure for me to meet and work with great minds in the context of
+ENVISAGE and HATS projects.
+
+I would like to thank Dr. Peter Wong who accepted to honor me for the opposition
+committee of the defense session.

### 2016-10-13 acknowledgement.tex
Update Acknowledgements

--- 2016-08-18 acknowledgement.tex
@@
 %
-\pdfbookmark[0]{Acknowledgement}{Acknowledgement}
-\chapter*{Acknowledgement}
+\pdfbookmark[0]{Acknowledgements}{Acknowledgements}
+\chapter*{Acknowledgements}
 \label{sec:acknowledgement}
@@
 
-There will be acknowledgments.
+To mom, who sacrificed all she could for us and left us very early.
+To dad, who never loses hope in us.
+To Shima, whose love and patience are my only sources of strength.
+To Maryam and Ali, who selflessly made room for me to get to where I am.
+
+This thesis like many others has been a journey.
+A journey that passed through the lives of family, friends, and colleagues.
+This is to extend a warm thankful message to all; especially
+Hossein Rahmani, Ramin Etemadi, Mahdi Jaghoori, Mohsen Davarinejad,
+Vlad Serbanescu, Nikolaos Bezirgiannis, Keyvan Azadbakht.
+It was a pleasure for me to meet and work with great minds during
+ENVISAGE and HATS projects together with Fredhopper team.
+I would like to thank Amir Hajari who has been like a wiser brother 
+for me during the years that I have lived in The Netherlands.
+
+I would like to thank Dr. Peter Wong who accepted to honor me for the opposition
+committee of the defense session.

### 2016-10-13 cv.tex
Switch to 'Source Code Pro' for monospace font
Signed-off-by: Behrooz Nobakht <nobeh5@gmail.com>
Update CV section

--- 2016-02-08 cv.tex
@@
 % \social[linkedin]{behrooznobakht}
-\extrainfo{Geboren 29-07-1981, Teheran}
+\extrainfo{Born 29-07-1981, Tehran, Iran}
 % \homepage{LinkedIn}
@@
 % Education
-\section{Opleiding}
+\section{Education}
 %\cventry{year--year}{Degree}{Institution}{City}{\textit{Grade}}{Description}
 
-\cventry{2011--2015}{PhD Informatica}{Universiteit Leiden}{Leiden}{}
-{\emph{Actors At Work}\copyright}
+\cventry{2011--2015}{PhD Computer Science}{Universiteit Leiden}{Leiden}{}
+{\emph{Actors At Work}\textcopyright}
 
-\cventry{2009--2011}{MSc cum laude Informatica}{Universiteit Leiden}{Leiden}{}{\emph{Deployment of active objects onto multi-core}}
+\cventry{2009--2011}{MSc cum laude Computer Science}{Universiteit Leiden}{Leiden}{}{\emph{Deployment of active objects onto multi-core}}
 
-\cventry{1998--2002}{BSc Informatica}{Technologie Universiteit Sharif}{Teheran}{\textit{7.0}}{\emph{Online Shopping}}
+\cventry{1998--2002}{BSc Software Engineering}{Sharif University of Technology}{Tehran}{\textit{7.0}}{\emph{Online Shopping}}
 
 % Work Experience
-\section{Werkervaring}
+\section{Epxerience}
 % \cventry{year--year}{Job title}{Employer}{City}{}{General description no%
@@
 
-\cventry{2011--Nu}{Senior Software Ontwikkelaar}{SDL Fredhopper}{Amsterdam}{}{Groep leider in cloud\newline}
+\cventry{May 2016}{Lead Software Engineer}{HERE}{Eindhoven}{}{CTO Office Applied Technologies\newline}
 
-\cventry{2003--2009}{Medeoprichter en architect}{ASTA Co.}{Teheran}{}{%
+\cventry{2011--Apr. 2016}{Senior Software Engineer}{SDL Fredhopper}{Amsterdam}{}{Team Lead -- Cloud Platform\newline}
+
+\cventry{2003--2009}{Co-Founder and Architect}{ASTA Co.}{Tehran}{}{%
 \begin{itemize}
-\item Het beheer van het bedrijf
-\item Ontwikkelen
-\item Project onderhandeling
+\item Software Development and Architecture
+\item Project Management
 \end{itemize}

### 2016-10-13 propositions.tex
Add thesis propositions

% -*- program: pdflatex -*-

\documentclass[a5paper,10pt]{article}
\usepackage[scale=0.9,centering]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{charter}

\pagenumbering{gobble}

\begin{document}

\begin{center}
{\Large \uppercase{Stellingen}}\\
Propositions belonging to the Ph.D. dissertation:\\
{\large Actors at Work}\\
by \textsl{Behrooz Nobakht}\\[1.5em]
\end{center}

\begin{enumerate}
\item Application-level priorities is one of the cross-functional requirements
of concurrent systems as important as functional requirements. (Chapter 2)

\item Co-operative scheduling of application-level priorities 
is resourceful and costly in thread-based programming languages. (Chapter 2)

\item Real-time programming with deadlines and timeouts requires 
fine granularity and flexibility of the programming language. (Chapter 3)

\item Separation of invocation from execution of asynchronous messages
drastically minimizes resourcefulness of co-operative scheduling
in a multi-threaded runtime. (Chapter 4)

\item Java 8 features enable to create a more efficient abstraction for 
asynchronous messages. (Chapter 4)

\item Runtime verification of multi-thread Java applications through
non-intrusive code annotations processing improves correctness and de-coupling in
comparison to bytecode instrumentation. (Chapter 5)

\item Actors are a natural fit for a distributed monitoring model
based on observation and reaction to guarantee composable 
multi-objective service characteristics. (Chapter 6)

\item Bridging modelling and programming for the purpose of verification 
and reasoning is a necessity for concurrent and distributed programming.

\item Explicit transfer of control in co-operative scheduling in ABS makes
programs easier to understand and reason about compared to the implicit
behavior in multi-threaded models.

\item Supporting hybrid thread mapping model in object-oriented languages
such as Java is a desirable but currently unrealized feature because of implementation
complexity.

\item The orthogonal properties of object orientation, co-operative scheduling,
and actor model has given rise to numerous challenges to devise a new 
programming model at their intersection.

\item The increasing importance of concurrent and distributed programming,
similar to the shift of structured programming to object orientation,
requires more of mentality change rather than languages and tools.

\item Concurrency could be hard even for a computer despite the presence of an 
experienced programmer.
\end{enumerate}

\end{document}

### 2016-10-16 propositions.tex
Update propositions

@@
 
-\item Supporting hybrid thread mapping model in object-oriented languages
+\item Supporting a hybrid thread mapping model in object-oriented languages
 such as Java is a desirable but currently unrealized feature because of implementation
@@
 similar to the shift of structured programming to object orientation,
-requires more of mentality change rather than languages and tools.
+requires more of a mentality change rather than languages and tools.
 

### 2016-10-16 p01-ch01-2012/paper.tex
Fix spacing issues

@@
 
-%\Mahdi{Do we want to talk about with clauses?}
-
 To explain extending Creol with priority specification and scheduling, we take a {\em client/server} perspective.
@@
 \begin{lstlisting}[float=t, label=ch01:lst:exres, caption=Exclusive Resource in Creol]
-interface Resource begin
-	(* \label{ch01:lst:priority_range} *)(*
-	%with User  \label{ch01:lst:with} 
-	*)
+interface Resource begin (* \label{ch01:lst:priority_range} *)
 		op request()
@@
 
-\begin{lstlisting}[frame=none, numbers=none]
-priority range 0..9
-\end{lstlisting}
+\begin{quote}
+\lstinline{priority range 0..9}
+\end{quote}
 
@@
 
-\begin{lstlisting}[frame=none, numbers=none]
-mutex ! request() priority(7);
-\end{lstlisting}
+\begin{quote}
+\lstinline{mutex ! request() priority(7);}
+\end{quote}
 
@@
 
-\begin{lstlisting}[frame=none, numbers=none]
-op request() priority(2) == ...
-op release() priority(0) == ...
-\end{lstlisting}
+\begin{quote}
+\lstinline{op request() priority(2) == ...}\\
+\lstinline{op release() priority(0) == ...}
+\end{quote}
 
@@
 
-\begin{lstlisting}[frame=none, numbers=none]
-await (*$\sim$*)taken priority(1);
-\end{lstlisting}
+\begin{quote}
+\lstinline{await ~taken priority(1);}
+\end{quote}
 
@@
 So, we define $\delta: P_1 \times P_2 \rightarrow P$ as:
-%
+
 \[

### 2016-10-31 acknowledgement.tex
Apply fixes based on regulations
Update Acknowledgements

@@
 To Maryam and Ali, who selflessly made room for me to get to where I am.
+To my daughter, whose birth has given us passion to go through more than 
+we have ever thought we can.
 
@@
 A journey that passed through the lives of family, friends, and colleagues.
-This is to extend a warm thankful message to all; especially
-Hossein Rahmani, Ramin Etemadi, Mahdi Jaghoori, Mohsen Davarinejad,
-Vlad Serbanescu, Nikolaos Bezirgiannis, Keyvan Azadbakht.
+This is to extend a warm thankful message to my friends; especially
+Hossein Rahmani, Ramin Etemadi, Mahdi Jaghoori, Lucas Sakizloglou,
+Mohammad Rezazade Mehrizi, and Mohsen Davarinejad.
+I would like to thank Amir Hajari who has been like a wiser brother 
+for me during the years that I have lived in The Netherlands.
 It was a pleasure for me to meet and work with great minds during
 ENVISAGE and HATS projects together with Fredhopper team.
-I would like to thank Amir Hajari who has been like a wiser brother 
-for me during the years that I have lived in The Netherlands.
-
-I would like to thank Dr. Peter Wong who accepted to honor me for the opposition
-committee of the defense session.

### 2016-11-01 acknowledgement.tex
Update Acknowledgements

@@
 To Maryam and Ali, who selflessly made room for me to get to where I am.
-To my daughter, whose birth has given us passion to go through more than 
-we have ever thought we can.
+To Fatemeh, whose birth has given us passion for life more than we have ever thought.
 
@@
 This is to extend a warm thankful message to my friends; especially
-Hossein Rahmani, Ramin Etemadi, Mahdi Jaghoori, Lucas Sakizloglou,
-Mohammad Rezazade Mehrizi, and Mohsen Davarinejad.
-I would like to thank Amir Hajari who has been like a wiser brother 
+Hossein Rahmani, Ramin Etemadi, Mohammad Rezazade Mehrizi, Lucas Sakizloglou,
+and Mohammad Mahdi Jaghoori.
+I would like to thank Amir Hajari who is like a wiser brother 
 for me during the years that I have lived in The Netherlands.

### 2016-11-02 booklet.tex
Add invitation booklet

% -*- program: pdflatex -*-

\documentclass[10pt]{article}
\usepackage[centering,paperheight=240mm,paperwidth=60mm,margin=10mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{charter}
\usepackage{graphicx}

\pagenumbering{gobble}

\begin{document}

\begin{center}

{\large Universiteit Leiden} \\[1em]
\includegraphics[width=3cm]{figs/ullogo} \\[3em]

{\Large \uppercase{\textbf{Actors\\ at\\ Work}}}\\[3em]
{\large \textsc{Proefschrift}\\[2em]}
{ter verkrijging van\\
de graad van doctor aan de Universiteit Leiden\\
op gezag van de Rector Magnificus prof. dr. C. J. J. M. Stolker,\\
volgens besluit van het College voor Promoties\\
te verdedigen       
op donderdag 15 december 2016\\
klokke 11.15 uur \\[1em]}
{door\\[2em]}
{\large Behrooz Nobakht\\[2em]}    
{geboren te Tehran, Iran, in 1981}
\end{center}

\end{document}
